### 2018-09-13 manuscript/manuscript.Rmd
add manuscript back to repo...  prior to turning it private

---
title: "25 Years of Water Quality Change in Rhode Island Lakes and Ponds"
output: 
  word_document:
    reference_docx: manuscript_template.docx   
author:
- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
- "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
- "^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston, RI 02881"
- "^*3*^ ORISE, Narragansett, RI 02882"
- "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
bibliography: crossref.bib
csl: plos.csl
abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on dozens of Rhode Island lakes and ponds for over 25 years, allowing us to explore long-term trends in common water quality parameters.  Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g. in urbanized northern Rhode Island).  Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis. This provides a robust and commonly scaled measurement to explore this data for long-term trends.  State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.  While yearly trends are useful, they do mask month-to-month variability differences across sites.  Additionally, while most sites track the yearly trend in decreasing water quality, there are bright spots with some sites improving over the 25 years. Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.  Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like Watershed Watch, for identifying trends in environmental condition. 
---

```{r setup, include=FALSE}
library(knitr)
library(captioner)
opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
tab_num <- captioner(prefix = "Table")
fig_num <- captioner(prefix = "Figure")

```

# Introduction

General stuff about long-term trends, monitoring programs and watersehd watch details

Given what is known about long term trends in water quality within the broader New England region [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.  Something about spatial and temporal variability of water quality

The goals of this study were to look at 25 years (still 25? need to check) of lakes and ponds data in Rhode Island.  The particular questions we were hoping to answer are: 1.) What are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and temperature; 2.) are water quality trends in Rhode Island similar to regional trends in the northeastern United states; 3.) Given trends at the state level, are there individual lakes and ponds that have water quality that deviate from the state-wide trends.  In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/green_murky_warm>.

# Methods

# Results

# Discussion

# Conclusions

# Bibliography

### 2018-09-13 manuscript/manuscript.Rmd
day one of dediated writing time

@@
 
-Given what is known about long term trends in water quality within the broader New England region [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
+Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
 Something about spatial and temporal variability of water quality
@@
 
+## Study Area and Data
+
+## Trend Analysis
+
+### Z-score
+
+### Kendall's Tau
+
+## Individual Lakes
+
 # Results
 
-# Discussion
+## State-wide trends in water quality
 
-# Conclusions
+## Individual lake trends in water quality
+
+# Discussion and conclusions
 

### 2018-10-18 manuscript/manuscript.Rmd
few sentences

@@
 
-General stuff about long-term trends, monitoring programs and watersehd watch details
+General stuff about long-term trends, monitoring programs and watershed watch details
 
@@
 
+For this study we combined a long-term dataset on water quality of lakes in Rhode Island, with a trend analysis similar to what is used to assess long term temperature trends [Need Citation].
+Details of both are outlined below.
+
 ## Study Area and Data
+
+The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (RIWW) program (Figure X).
+MORE DETAILS ON RIWW.
+
+In particular we selected data from the RIWW data that ADD WHAT WE FILTERED.
+The final dataset used in our analysis included XXX lakes and ponds which had approximately XXX samples for temperature, XXX samples for chlorophyll *a*, XXX samples for Total Nitrogen (TN), and XXX samples for Total Phosphorus (TP)
 

### 2018-11-01 manuscript/manuscript.Rmd
updates to manuscript

@@
 In particular we selected data from the RIWW data that ADD WHAT WE FILTERED.
-The final dataset used in our analysis included XXX lakes and ponds which had approximately XXX samples for temperature, XXX samples for chlorophyll *a*, XXX samples for Total Nitrogen (TN), and XXX samples for Total Phosphorus (TP)
+The final dataset used in our analysis included XXX lakes and ponds which had approximately XXX samples for temperature, XXX samples for chlorophyll *a*, XXX samples for Total Nitrogen (TN), and XXX samples for Total Phosphorus (TP).
+
+Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included XX, XX, XX, and XX states [@Oliver_2017] and found little change in water quality trends across the region.
+We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@stachelek_lagosne_2017; @soranno_lagosne_2017].
+Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a*.
 
 ## Trend Analysis
+
+There are many different methods for analyzing time series data for trends.
+One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.
+For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
+Thus, it is necessary to account for this type of site-specific variation.
+This is very similar to how the long-term temperature trends have been calculated by the XXX [FILL IN GROUP DOING THIS]
 
@@
 
-## Individual Lakes
+### Comparison of Rhode Island to the Region
 
@@
 
-## Individual lake trends in water quality
+[TN, TP, CHLA, TEMP Figures]
+
+## Regional trends in water quality
 
@@
 
+## Trends
+
+### Local scale vs Regional/Global scale
+
+## Caveats
+
 # Bibliography

### 2018-11-02 manuscript/manuscript.Rmd
Refined URIWW acronym

@@
 
-The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (RIWW) program (Figure X).
+The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (Figure X).
 MORE DETAILS ON RIWW.

### 2018-11-05 manuscript/manuscript.Rmd
drafted some z-score and tred text

@@
 Thus, it is necessary to account for this type of site-specific variation.
-This is very similar to how the long-term temperature trends have been calculated by the XXX [FILL IN GROUP DOING THIS]
+This is very similar to how the long-term temperature trends have been calculated by the XXX [FILL IN GROUP DOING THIS].
+The general approach is to calculate site-specific deviations from a pre-determined reference period.
+This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the referenc period.
 
 ### Z-score
+
+The anomaly approach is very useful and is the approach we chose to follow.
+However, using anomalies works best with a single measure (e.g temperature) or with mulitple measurements that are on the same scale.
+The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
+For instance, temperature anomalies will like range +/- 2 to 3 degrees, whereas phosphorus anomalies, might range +/- XX to XX ug/l.
+To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.
+This value is commonly referred to as a Z-score.
+We use these z-scores to examine each water quality variable for a trend over the time period of 1990 to 2016.
 

### 2018-11-07 manuscript/manuscript.Rmd
updated figures and some more text.  Actually calcing kendalls now!

@@
 - "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
-bibliography: crossref.bib
+bibliography: manuscript.bib
 csl: plos.csl
@@
 
-The goals of this study were to look at 25 years (still 25?
-need to check) of lakes and ponds data in Rhode Island.
-The particular questions we were hoping to answer are: 1.) What are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and temperature; 2.) are water quality trends in Rhode Island similar to regional trends in the northeastern United states; 3.) Given trends at the state level, are there individual lakes and ponds that have water quality that deviate from the state-wide trends.
-In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/green_murky_warm>.
+The goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island.
+We will answer two particular questions.
+First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
+In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends>.
 
@@
 
-For this study we combined a long-term dataset on water quality of lakes in Rhode Island, with a trend analysis similar to what is used to assess long term temperature trends [Need Citation].
+For this study we combined a long-term dataset on water quality of lakes in Rhode Island, with a trend analysis based on centered and scaled water quality values (i.e.
+z-scores) to find increasing or decreasing annual water quality trends.
 Details of both are outlined below.
@@
 Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included XX, XX, XX, and XX states [@Oliver_2017] and found little change in water quality trends across the region.
-We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@stachelek_lagosne_2017; @soranno_lagosne_2017].
+We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@stacheleklagosne2017; @sorannolagosne2017].
 Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a*.
@@
 One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.
-For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
+For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds wer sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
 Thus, it is necessary to account for this type of site-specific variation.
-This is very similar to how the long-term temperature trends have been calculated by the XXX [FILL IN GROUP DOING THIS].
-The general approach is to calculate site-specific deviations from a pre-determined reference period.
-This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the referenc period.
+This is similar to how the long-term temperature trends are analyzed using temperature anomolies [FILL IN GROUP DOING THIS].
+The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.
+This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the reference period.
 
@@
 
+Furthermore, since we are interest in water quality trends over time, we wanted to explore how each site was responding.
+Thus, our z-scores were calculate over the reference period, 1990-2016, for each site.
+In other words the long-term mean and standard deviation were calculated on a site-by-site basis.
+
 ### Kendall's Tau
 
+Kendall's Tau and the Mann-Kendall test are commonly used to test for monotonic trends water quality data [@mann1945nonparametric; @helsel2002statistical; @hipel1994time].
+We followed these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
+
 ### Comparison of Rhode Island to the Region
+
+Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
+The data we used in our analysis, the University of Rhode Island's Watershed Watch data, was inlcuded in that, but we were curious if regional trends were masking local scale trends in Rhode Island.
+In addition to the analysis condcuted by Oliver et al.
+[-@Oliver_2017], we analyzed the same dataset but using the approach outlined here.
 

### 2018-11-07 R/functions.R

@@ -8,6 +8,7 @@ library(tidyr)
 library(here)
 library(ggplot2)
 library(hrbrthemes)
+library(Kendall)
 if(!require(LAGOSNE)){devtools::install_github("cont-limno/LAGOSNE")}
 library(LAGOSNE)
 
@@ -48,15 +49,14 @@ wq_trend_gg <- function(df, wqparam,
     write_csv(df2, write, append = FALSE)
   }
   
-  kt <- with(df2, cor(year, mn_value,use = "pairwise.complete.obs"))
-  ktp <- with(df2, cor.test(year, mn_value)$p.value)
+  kt <- with(df2, Kendall(year, mn_value))
 
   gg <- ggplot(df2,aes(x = year, y = mn_value)) + 
     geom_point(aes(color = col_group), size=2.5) +
     geom_smooth(method = "lm", se=FALSE, color = "black") +
     theme_ipsum() +
-    labs(..., title = paste0("Kendall's Tau: ", round(kt,3),
-                             " p-value: ", round(ktp, 3))) +
+    labs(..., title = paste0("Kendall's Tau: ", round(kt$tau,3),
+                             " p-value: ", round(kt$sl, 4))) +
     scale_color_manual(values = c("red3","darkblue")) + 
     theme(legend.position="none", plot.title = element_text(size=10, face="plain")) + 
     #geom_label(aes(x = 2006, y = min(mn_value)*0.88), 

### 2018-11-30 manuscript/manuscript.Rmd
some progress on ms!

@@
 Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.
-Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like Watershed Watch, for identifying trends in environmental condition.
+Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition.
+editor_options: 
+  chunk_output_type: console
 ---
@@
 library(captioner)
+source(here::here("R/functions.R"))
 opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
@@
 fig_num <- captioner(prefix = "Figure")
+```
 
+```{r add_analysis, include=FALSE}
+#got to figure out path stuff before render.
+source("../R/data_available.R")
+source("../R/summary_stats.R")  
+source("../R/figures.R")
 ```
@@
 z-scores) to find increasing or decreasing annual water quality trends.
-Details of both are outlined below.
+Details are outlined below.
 
@@
 
-In particular we selected data from the RIWW data that ADD WHAT WE FILTERED.
-The final dataset used in our analysis included XXX lakes and ponds which had approximately XXX samples for temperature, XXX samples for chlorophyll *a*, XXX samples for Total Nitrogen (TN), and XXX samples for Total Phosphorus (TP).
+In particular we selected data from the RIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data.
+The final dataset used in our analysis included `r nrows(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
 
-Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included XX, XX, XX, and XX states [@Oliver_2017] and found little change in water quality trends across the region.
-We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@stacheleklagosne2017; @sorannolagosne2017].
-Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a*.
+Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+The authors found little change in water quality trends across this region [@Oliver_2017] .
+We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
+Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a* from the LAGOSNE database.
 
@@
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature anomalies will like range +/- 2 to 3 degrees, whereas phosphorus anomalies, might range +/- XX to XX ug/l.
+For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees celsius to a high of 30 degrees celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
 To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.
@@
 Furthermore, since we are interest in water quality trends over time, we wanted to explore how each site was responding.
-Thus, our z-scores were calculate over the reference period, 1990-2016, for each site.
+Thus, our z-scores were calculated over the reference period, 1990-2016, for each site.
 In other words the long-term mean and standard deviation were calculated on a site-by-site basis.

### 2018-11-30 R/data_available.R

@@ -49,5 +49,3 @@ ww_avail_data_locations <- ww_data %>%
   right_join(data_avail) %>%
   select(station_name, lon_dd, lat_dd, params_avail) %>%
   write_csv("data/ww_avail_data_locations.csv")
-
-table(ww_avail_data_locations$params_avail)

### 2018-11-30 R/data_clean.R

@@ -98,6 +98,24 @@ ww_lake_trend_data <- ww_all %>%
   mutate(measurement_scale = scale(mn_measurement),
          measurement_anmly = mn_measurement - mean(mn_measurement),
          lt_mean = mean(mn_measurement))
+
+# Few known issues to remove
+# Site ww 536, not really a pond.  sampled only at shallow outflow
+ww_lake_trend_data <- ww_lake_trend_data %>% 
+  filter(station_name != "WW536")
+
+# Almy Pond sample from 2016-07-23 likely had sediment contamination
+# Keeps just temp from this sample.
+# Per discussion with Elizabeth Herron
+idx <- !with(ww_lake_trend_data, station_name == "WW199" &
+              year == 2016 &
+              month == 7 &
+              day == 23 & 
+              param != "temp" &
+              param != "secchi" &
+              param != "ph" &
+              param != "enterococci")
+ww_lake_trend_data <- ww_lake_trend_data[idx,]
   
 
 write_csv(ww_lake_trend_data, here("data/ww_lake_trend_data.csv"))

### 2018-11-30 R/summary_stats.R

@@ -1,0 +1,29 @@
+source(here::here("R/functions.R"))
+ww_all<-read_csv(here("data/ww_lake_trend_data.csv"))
+param_ranges <- ww_all %>%
+  filter(param == "chla" |
+           param == "temp" |
+           param == "total_n" |
+           param == "total_p") %>%
+  group_by(param) %>%
+  summarize(min = round(min(mn_measurement, na.rm = T), 2),
+            percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(mn_measurement, na.rm = T), 2),
+            median = round(median(mn_measurement, na.rm = T), 2),
+            percentile_75 = round(quantile(mn_measurement, probs = 0.75, na.rm = T), 2),
+            max = round(max(mn_measurement, na.rm = T), 2),
+            sd = round(sd(mn_measurement, na.rm = T), 2)) 
+
+param_anomaly_ranges <- ww_all %>%
+  filter(param == "chla" |
+           param == "temp" |
+           param == "total_n" |
+           param == "total_p") %>%
+  group_by(param) %>%
+  summarize(min = round(min(measurement_anmly, na.rm = T), 2),
+            percentile_25 = round(quantile(measurement_anmly, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(measurement_anmly, na.rm = T), 2),
+            median = round(median(measurement_anmly, na.rm = T), 2),
+            percentile_75 = round(quantile(measurement_anmly, probs = 0.75, na.rm = T), 2),
+            max = round(max(measurement_anmly, na.rm = T), 2),
+            sd = round(sd(measurement_anmly, na.rm = T), 2))

### 2018-12-03 manuscript/manuscript.Rmd
render with some numbers

@@
 
-```{r add_analysis, include=FALSE}
+```{r analysis, include=FALSE}
 #got to figure out path stuff before render.
-source("../R/data_available.R")
-source("../R/summary_stats.R")  
-source("../R/figures.R")
+source(here::here("R/data_available.R"))
+source(here::here("R/summary_stats.R"))  
+source(here::here("R/figures.R"))
 ```
@@
 For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data.
-The final dataset used in our analysis included `r nrows(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
 

### 2018-12-03 R/data_available.R

@@ -47,5 +47,5 @@ ww_avail_data_locations <- ww_data %>%
   select(station_name, lon_dd, lat_dd) %>%
   unique() %>%
   right_join(data_avail) %>%
-  select(station_name, lon_dd, lat_dd, params_avail) %>%
-  write_csv("data/ww_avail_data_locations.csv")
+  select(station_name, lon_dd, lat_dd, params_avail) #%>%
+  #write_csv("data/ww_avail_data_locations.csv")

### 2018-12-03 R/figures.R

@@ -39,8 +39,7 @@ ggsave(here("figures/ww_temp_trends.jpg"), temp_gg, width = 5, height = 3.75,
 lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
 lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", 
-                             y = "Average Yearly Scaled Chlorophyll", x = "Year",
-                             write = "temporary.csv")
+                             y = "Average Yearly Scaled Chlorophyll", x = "Year")
 ggsave(here("figures/lagos_chla_trends.jpg"), lagos_chla_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 

### 2018-12-19 manuscript/manuscript.Rmd
figures added to paper

@@
 
-The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (Figure X).
+The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
 MORE DETAILS ON RIWW.
 
+![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`](../figures/ww_map.jpg)
+
 In particular we selected data from the RIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data.
+For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
+Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(avail_data$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.
 
@@
 
-[TN, TP, CHLA, TEMP Figures]
+![`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.")`](../figures/ww_temp_trends.jpg)
+
+![`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.")`](../figures/ww_tn_trends.jpg)
+
+![`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.")`](../figures/ww_tp_trends.jpg)
+
+![`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll *a* (average z-score) in Rhode Island lakes and ponds.")`](../figures/ww_chla_trends.jpg)
 
 ## Regional trends in water quality
+
+![`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.")`](../figures/lagos_tn_trends.jpg)
+
+![`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.")`](../figures/lagos_tp_trends.jpg)
+
+![`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll *a* (average z-score) in the LAGOSNE states.")`](../figures/lagos_chla_trends.jpg)
 

### 2018-12-19 R/data_available.R

@@ -47,5 +47,5 @@ ww_avail_data_locations <- ww_data %>%
   select(station_name, lon_dd, lat_dd) %>%
   unique() %>%
   right_join(data_avail) %>%
-  select(station_name, lon_dd, lat_dd, params_avail) #%>%
-  #write_csv("data/ww_avail_data_locations.csv")
+  select(station_name, params_avail) %>%
+  write_csv(here("data/ww_avail_data_stations.csv"))

### 2018-12-19 R/figures.R

@@ -9,7 +9,7 @@ chla_gg <- ww_data %>%
   wq_trend_gg( "chla", yvar = "measurement_scale", 
                        y = "Average Yearly Scaled Chlorophyll", x = "Year",
                write = here("data/chla_data.csv"))
-ggsave(here("figures/ww_chla_trends.jpg"), chla_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_chla_trends.jpg"), chla_gg, width = 8, height = 6, 
        units = "in", dpi = 600)
 
 tn_gg <-  ww_data %>%
@@ -17,7 +17,7 @@ tn_gg <-  ww_data %>%
   wq_trend_gg("total_n", yvar = "measurement_scale", 
                       y = "Average Yearly Scaled Total Nitrogen", x = "Year",
               write = here("data/total_n_data.csv"))
-ggsave(here("figures/ww_tn_trends.jpg"), tn_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_tn_trends.jpg"), tn_gg, width = 8, height = 6, 
        units = "in", dpi = 600)
 
 tp_gg <-  ww_data %>%
@@ -25,7 +25,7 @@ tp_gg <-  ww_data %>%
   wq_trend_gg("total_p", yvar = "measurement_scale", 
                       y = "Average Yearly Scaled Total Phosphorus", x = "Year",
                       write = here("data/total_p_data.csv"))
-ggsave(here("figures/ww_tp_trends.jpg"), tp_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_tp_trends.jpg"), tp_gg, width = 8, height = 6, 
        units = "in", dpi = 600)
 
 temp_gg <- ww_data %>%
@@ -33,23 +33,52 @@ temp_gg <- ww_data %>%
   wq_trend_gg("temp", yvar = "measurement_scale", 
                        y = "Average Yearly Scaled Temperature", x = "Year",
               write = here("data/temp_data.csv"))
-ggsave(here("figures/ww_temp_trends.jpg"), temp_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_temp_trends.jpg"), temp_gg, width = 8, height = 6, 
        units = "in", dpi = 600)
 
 lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
 lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", 
[…]

### 2018-12-19 R/functions.R

@@ -11,6 +11,9 @@ library(hrbrthemes)
 library(Kendall)
 if(!require(LAGOSNE)){devtools::install_github("cont-limno/LAGOSNE")}
 library(LAGOSNE)
+library(sf)
+library(USAboundaries)
+library(USAboundariesData)
 
 
 #' Function to filter on number of years sampled per site
@@ -65,7 +68,8 @@ wq_trend_gg <- function(df, wqparam,
     #         hjust = "left", label.size = NA, size = 2.5) +
     scale_x_continuous(labels = c(1990,1995,2000,2005,2010,2015),
                        breaks = c(1990,1995,2000,2005,2010,2015),
-                       minor_breaks = NULL)
+                       minor_breaks = NULL) +
+    scale_y_continuous(limits = c(-0.5, 0.75))
   
   gg
 

### 2018-12-19 manuscript/manuscript.Rmd
some progress on the paper.

@@
 
+During the selected sampling period of 1990 to 2016 average lake temperature was `r ww_param_ranges%>%filter(param == "temp")%>%pull(mean)`, average total nitrogen was `r ww_param_ranges%>%filter(param == "total_n")%>%pull(mean)`, average total phosphorus was `r ww_param_ranges%>%filter(param == "total_p")%>%pull(mean)`, and average chlorophyll was `r ww_param_ranges%>%filter(param == "chla")%>%pull(mean)` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`)
+
+```{r ww_params, echo=FALSE, results="asis"}
+knitr::kable(ww_param_ranges, format = "markdown")
+tab_num("ww_params")
+```
+
 ## State-wide trends in water quality

### 2018-12-19 R/summary_stats.R

@@ -1,10 +1,11 @@
 source(here::here("R/functions.R"))
 ww_all<-read_csv(here("data/ww_lake_trend_data.csv"))
-param_ranges <- ww_all %>%
+ww_param_ranges <- ww_all %>%
   filter(param == "chla" |
            param == "temp" |
            param == "total_n" |
            param == "total_p") %>%
+  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
   group_by(param) %>%
   summarize(min = round(min(mn_measurement, na.rm = T), 2),
             percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
@@ -14,11 +15,42 @@ param_ranges <- ww_all %>%
             max = round(max(mn_measurement, na.rm = T), 2),
             sd = round(sd(mn_measurement, na.rm = T), 2)) 
 
-param_anomaly_ranges <- ww_all %>%
+ww_param_anomaly_ranges <- ww_all %>%
   filter(param == "chla" |
            param == "temp" |
            param == "total_n" |
            param == "total_p") %>%
+  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
+  group_by(param) %>%
+  summarize(min = round(min(measurement_anmly, na.rm = T), 2),
+            percentile_25 = round(quantile(measurement_anmly, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(measurement_anmly, na.rm = T), 2),
+            median = round(median(measurement_anmly, na.rm = T), 2),
+            percentile_75 = round(quantile(measurement_anmly, probs = 0.75, na.rm = T), 2),
+            max = round(max(measurement_anmly, na.rm = T), 2),
+            sd = round(sd(measurement_anmly, na.rm = T), 2))
+
+lagos_all <- read_csv(here("data/lagos_lake_trend_data.csv"))
+
+lagos_param_ranges <- lagos_all %>%
+  filter(param == "chla" |
+           param == "total_n" |
+           param == "total_p") %>%
+  mutate(param = factor(param,levels = c("total_n", "total_p", "chla"))) %>%
+  group_by(param) %>%
+  summarize(min = round(min(measurement, na.rm = T), 2),
+            percentile_25 = round(quantile(measurement, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(measurement, na.rm = T), 2),
[…]

### 2018-12-20 manuscript/manuscript.Rmd
close to figuring out the weird stuff...

--- 2018-12-03 manuscript/manuscript.Rmd
@@
 author:
-- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
+- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
 - "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
@@
 
+```{r ww_params2, echo=FALSE}
+source(here::here("R/functions.R"))
+ww_all<-read_csv(here("data/ww_lake_trend_data.csv"))
+ww_param_ranges <- ww_all %>%
+  filter(param == "chla" |
+           param == "temp" |
+           param == "total_n" |
+           param == "total_p") %>%
+  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
+  group_by(param) %>%
+  summarize(min = round(min(mn_measurement, na.rm = T), 2),
+            percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(mn_measurement, na.rm = T), 2),
+            median = round(median(mn_measurement, na.rm = T), 2),
+            percentile_75 = round(quantile(mn_measurement, probs = 0.75, na.rm = T), 2),
+            max = round(max(mn_measurement, na.rm = T), 2),
+            sd = round(sd(mn_measurement, na.rm = T), 2)) %>%
+  ungroup() %>%
+  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
+  select(Parameter = param, Units = units, 
+         "25th Percentile" = percentile_25,
+         Mean = mean, Median = median, 
+         "75th Percentile" = percentile_75,
+         Max = max, "Std. Dev" = sd )
+knitr::kable(ww_param_ranges)
+```
+
 ```{r setup, include=FALSE}
@@
 ```
+
+*`r tab_num("ww_params")`*
 
@@
 
-The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (Figure X).
+The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
 MORE DETAILS ON RIWW.
 
+*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`*
+![](../figures/ww_map.jpg)
+
 In particular we selected data from the RIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data.
+For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
+Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.
 
@@
 
+During the selected sampling period of 1990 to 2016 average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+
+*`r tab_num("ww_params")`*
+```{r ww_params, echo=FALSE, results="asis"}
+knitr::kable(ww_param_ranges, format = "markdown")
+```
+
 ## State-wide trends in water quality
 
-[TN, TP, CHLA, TEMP Figures]
+*`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.")`*
+![](../figures/ww_temp_trends.jpg)
+
+*`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.")`*
+![](../figures/ww_tn_trends.jpg)
+
+*`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.")`*
+![](../figures/ww_tp_trends.jpg)
+
+*`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll *a* (average z-score) in Rhode Island lakes and ponds.")`*
+![](../figures/ww_chla_trends.jpg)
 
 ## Regional trends in water quality
+
+*`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.")`*
+![](../figures/lagos_tn_trends.jpg)
+
+*`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.")`*
+![](../figures/lagos_tp_trends.jpg)
+
+*`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll *a* (average z-score) in the LAGOSNE states.")`*
+![](../figures/lagos_chla_trends.jpg)
 
@@
 
+### RI is downstream - Bryan will write
+
 ## Caveats

### 2018-12-20 test.Rmd
close to figuring out the weird stuff...

---
title: "Untitled"
output:
  word_document: 
    reference_docx: "manuscript/manuscript_template.docx"
---

# Plain Data Frame

```{r plain}
x<-data.frame(unit = "µg/l", stringsAsFactors = FALSE)
knitr::kable(x)
```

# Tibble
```{r tibble}
x<-dplyr::tbl_df(data.frame(unit = "µg/l", stringsAsFactors = FALSE))
knitr::kable(x, format = "markdown")
```

# My stuff
```{r culprit}
source(here::here("R/functions.R"))
ww_all<-read_csv(here("data/ww_lake_trend_data.csv"))
ww_param_ranges <- ww_all %>%
  filter(param == "chla" |
           param == "temp" |
           param == "total_n" |
           param == "total_p") %>%
  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
  group_by(param) %>%
  summarize(min = round(min(mn_measurement, na.rm = T), 2),
            percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
            mean = round(mean(mn_measurement, na.rm = T), 2),
            median = round(median(mn_measurement, na.rm = T), 2),
            percentile_75 = round(quantile(mn_measurement, probs = 0.75, na.rm = T), 2),
            max = round(max(mn_measurement, na.rm = T), 2),
            sd = round(sd(mn_measurement, na.rm = T), 2)) %>%
  ungroup() %>%
  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
  select(Parameter = param, Units = units, 
         "25th Percentile" = percentile_25,
         Mean = mean, Median = median, 
         "75th Percentile" = percentile_75,
         Max = max, "Std. Dev" = sd )
knitr::kable(ww_param_ranges, format = "markdown")
```

### 2018-12-20 R/data_clean.R

@@ -98,24 +98,6 @@ ww_lake_trend_data <- ww_all %>%
   mutate(measurement_scale = scale(mn_measurement),
          measurement_anmly = mn_measurement - mean(mn_measurement),
          lt_mean = mean(mn_measurement))
-
-# Few known issues to remove
-# Site ww 536, not really a pond.  sampled only at shallow outflow
-ww_lake_trend_data <- ww_lake_trend_data %>% 
-  filter(station_name != "WW536")
-
-# Almy Pond sample from 2016-07-23 likely had sediment contamination
-# Keeps just temp from this sample.
-# Per discussion with Elizabeth Herron
-idx <- !with(ww_lake_trend_data, station_name == "WW199" &
-              year == 2016 &
-              month == 7 &
-              day == 23 & 
-              param != "temp" &
-              param != "secchi" &
-              param != "ph" &
-              param != "enterococci")
-ww_lake_trend_data <- ww_lake_trend_data[idx,]
   
 
 write_csv(ww_lake_trend_data, here("data/ww_lake_trend_data.csv"))

### 2018-12-20 R/figures.R

@@ -9,7 +9,7 @@ chla_gg <- ww_data %>%
   wq_trend_gg( "chla", yvar = "measurement_scale", 
                        y = "Average Yearly Scaled Chlorophyll", x = "Year",
                write = here("data/chla_data.csv"))
-ggsave(here("figures/ww_chla_trends.jpg"), chla_gg, width = 8, height = 6, 
+ggsave(here("figures/ww_chla_trends.jpg"), chla_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 
 tn_gg <-  ww_data %>%
@@ -17,7 +17,7 @@ tn_gg <-  ww_data %>%
   wq_trend_gg("total_n", yvar = "measurement_scale", 
                       y = "Average Yearly Scaled Total Nitrogen", x = "Year",
               write = here("data/total_n_data.csv"))
-ggsave(here("figures/ww_tn_trends.jpg"), tn_gg, width = 8, height = 6, 
+ggsave(here("figures/ww_tn_trends.jpg"), tn_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 
 tp_gg <-  ww_data %>%
@@ -25,7 +25,7 @@ tp_gg <-  ww_data %>%
   wq_trend_gg("total_p", yvar = "measurement_scale", 
                       y = "Average Yearly Scaled Total Phosphorus", x = "Year",
                       write = here("data/total_p_data.csv"))
-ggsave(here("figures/ww_tp_trends.jpg"), tp_gg, width = 8, height = 6, 
+ggsave(here("figures/ww_tp_trends.jpg"), tp_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 
 temp_gg <- ww_data %>%
@@ -33,52 +33,24 @@ temp_gg <- ww_data %>%
   wq_trend_gg("temp", yvar = "measurement_scale", 
                        y = "Average Yearly Scaled Temperature", x = "Year",
               write = here("data/temp_data.csv"))
-ggsave(here("figures/ww_temp_trends.jpg"), temp_gg, width = 8, height = 6, 
+ggsave(here("figures/ww_temp_trends.jpg"), temp_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 
 lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
 lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", 
[…]

### 2018-12-20 R/functions.R

@@ -14,7 +14,9 @@ library(LAGOSNE)
 library(sf)
 library(USAboundaries)
 library(USAboundariesData)
-
+#if(!require(sass)){devtools::install_github("rstudio/sass")}
+#if(!require(gt)){devtools::install_github("rstudio/gt")}
+#library(gt)
 
 #' Function to filter on number of years sampled per site
 #' @param df data frame to filter

### 2018-12-20 R/summary_stats.R

@@ -13,7 +13,14 @@ ww_param_ranges <- ww_all %>%
             median = round(median(mn_measurement, na.rm = T), 2),
             percentile_75 = round(quantile(mn_measurement, probs = 0.75, na.rm = T), 2),
             max = round(max(mn_measurement, na.rm = T), 2),
-            sd = round(sd(mn_measurement, na.rm = T), 2)) 
+            sd = round(sd(mn_measurement, na.rm = T), 2)) %>%
+  ungroup() %>%
+  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
+  select(Parameter = param, Units = units, 
+         "25th Percentile" = percentile_25,
+         Mean = mean, Median = median, 
+         "75th Percentile" = percentile_75,
+         Max = max, "Std. Dev" = sd )
 
 ww_param_anomaly_ranges <- ww_all %>%
   filter(param == "chla" |
@@ -28,7 +35,14 @@ ww_param_anomaly_ranges <- ww_all %>%
             median = round(median(measurement_anmly, na.rm = T), 2),
             percentile_75 = round(quantile(measurement_anmly, probs = 0.75, na.rm = T), 2),
             max = round(max(measurement_anmly, na.rm = T), 2),
-            sd = round(sd(measurement_anmly, na.rm = T), 2))
+            sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
+  ungroup() %>%
+  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
+  select(Parameter = param, Units = units, 
+         "25th Percentile" = percentile_25,
+         Mean = mean, Median = median, 
+         "75th Percentile" = percentile_75,
+         Max = max, "Std. Dev" = sd )
 
 lagos_all <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
@@ -44,7 +58,14 @@ lagos_param_ranges <- lagos_all %>%
             median = round(median(measurement, na.rm = T), 2),
             percentile_75 = round(quantile(measurement, probs = 0.75, na.rm = T), 2),
             max = round(max(measurement, na.rm = T), 2),
-            sd = round(sd(measurement, na.rm = T), 2)) 
+            sd = round(sd(measurement, na.rm = T), 2)) %>%
+  ungroup() %>%
+  mutate(units = c("µg/l", "µg/l","µg/l")) %>%
+  select(Parameter = param, Units = units, 
[…]

### 2018-12-31 manuscript/manuscript.Rmd
last push before furlough

@@
 
-```{r ww_params2, echo=FALSE}
-source(here::here("R/functions.R"))
-ww_all<-read_csv(here("data/ww_lake_trend_data.csv"))
-ww_param_ranges <- ww_all %>%
-  filter(param == "chla" |
-           param == "temp" |
-           param == "total_n" |
-           param == "total_p") %>%
-  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
-  group_by(param) %>%
-  summarize(min = round(min(mn_measurement, na.rm = T), 2),
-            percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
-            mean = round(mean(mn_measurement, na.rm = T), 2),
-            median = round(median(mn_measurement, na.rm = T), 2),
-            percentile_75 = round(quantile(mn_measurement, probs = 0.75, na.rm = T), 2),
-            max = round(max(mn_measurement, na.rm = T), 2),
-            sd = round(sd(mn_measurement, na.rm = T), 2)) %>%
-  ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
-  select(Parameter = param, Units = units, 
-         "25th Percentile" = percentile_25,
-         Mean = mean, Median = median, 
-         "75th Percentile" = percentile_75,
-         Max = max, "Std. Dev" = sd )
-knitr::kable(ww_param_ranges)
-```
-
 ```{r setup, include=FALSE}
@@
 library(captioner)
-source(here::here("R/functions.R"))
+source(here::here("R/functions.R"), encoding = "UTF-8")
 opts_chunk$set(echo = TRUE, fig.width=6, fig.height=5)
@@
 #got to figure out path stuff before render.
-source(here::here("R/data_available.R")) 
-source(here::here("R/summary_stats.R"))  
-source(here::here("R/figures.R"))
+source(here::here("R/data_available.R"), encoding = "UTF-8") 
+source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
+source(here::here("R/figures.R"), encoding = "UTF-8")
 ```
-
-*`r tab_num("ww_params")`*
 
@@
 *`r tab_num("ww_params")`*
+
 ```{r ww_params, echo=FALSE, results="asis"}
@@
 *`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.")`*
+
 ![](../figures/ww_temp_trends.jpg)
@@
 *`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.")`*
+
 ![](../figures/ww_tn_trends.jpg)
@@
 *`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.")`*
+
 ![](../figures/ww_tp_trends.jpg)
@@
 *`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll *a* (average z-score) in Rhode Island lakes and ponds.")`*
+
 ![](../figures/ww_chla_trends.jpg)
@@
 *`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.")`*
+
 ![](../figures/lagos_tn_trends.jpg)
@@
 *`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.")`*
+
 ![](../figures/lagos_tp_trends.jpg)
@@
 *`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll *a* (average z-score) in the LAGOSNE states.")`*
+
 ![](../figures/lagos_chla_trends.jpg)

### 2018-12-31 R/figures.R

@@ -40,17 +40,19 @@ lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
 lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", 
                              y = "Average Yearly Scaled Chlorophyll", x = "Year",
-                             write = "temporary.csv")
+                             write = here("data/chla_data_lagos.csv"))
 ggsave(here("figures/lagos_chla_trends.jpg"), lagos_chla_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 
 lagos_tn_gg <- wq_trend_gg(lagos_data, "total_n", yvar = "measurement_scale", 
-                             y = "Average Yearly Scaled Total Nitrogen", x = "Year")
+                             y = "Average Yearly Scaled Total Nitrogen", x = "Year",
+                           write = here("data/total_n_data_lagos.csv"))
 ggsave(here("figures/lagos_tn_trends.jpg"), lagos_tn_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 
 lagos_tp_gg <- wq_trend_gg(lagos_data, "total_p", yvar = "measurement_scale", 
-                             y = "Average Yearly Scaled Total Phosphorus", x = "Year")
+                             y = "Average Yearly Scaled Total Phosphorus", x = "Year",
+                           write = here("data/total_n_data_lagos.csv"))
 ggsave(here("figures/lagos_tp_trends.jpg"), lagos_tp_gg, width = 5, height = 3.75, 
        units = "in", dpi = 600)
 

### 2018-12-31 R/summary_stats.R

@@ -1,5 +1,6 @@
 source(here::here("R/functions.R"))
-ww_all<-read_csv(here("data/ww_lake_trend_data.csv"))
+
+ww_all<-read.csv(here("data/ww_lake_trend_data.csv"), stringsAsFactors = FALSE)
 ww_param_ranges <- ww_all %>%
   filter(param == "chla" |
            param == "temp" |

### 2019-01-07 manuscript/manuscript.Rmd
Add paragraph describing URI WW
Focuses on lakes & ponds

@@
 General stuff about long-term trends, monitoring programs and watershed watch details
+
+The University of Rhode Island's Watershed Watch (URI WW) Volunteer Monitoring program has been collecting water quality data from Rhode Island lakes and ponds for over 25 years.
+The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
+URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
+Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen.
+Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.
 

### 2019-01-29 manuscript/manuscript.Rmd
added some lagos results

@@
 ---
-title: "25 Years of Water Quality Change in Rhode Island Lakes and Ponds"
+title: "26 Years of Water Quality Change in Rhode Island Lakes and Ponds"
 output: 
@@
 csl: plos.csl
-abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on dozens of Rhode Island lakes and ponds for over 25 years, allowing us to explore long-term trends in common water quality parameters.
+abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.
 Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g.
@@
 
-During the selected sampling period of 1990 to 2016 average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagosne_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagosne_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagosne_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagosne_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagosne_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagosne_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagosne_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 knitr::kable(ww_param_ranges, format = "markdown")
+```
+
+*`r tab_num("lagosne_params")`*
+
+```{r ww_params, echo=FALSE, results="asis"}
+knitr::kable(lagosne_param_ranges, format = "markdown")
 ```

### 2019-01-29 manuscript/manuscript.Rmd
fresh render from new additions

@@
 During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
-For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagosne_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagosne_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagosne_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagosne_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagosne_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagosne_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagosne_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 
-*`r tab_num("lagosne_params")`*
+*`r tab_num("lagos_params")`*
 
-```{r ww_params, echo=FALSE, results="asis"}
-knitr::kable(lagosne_param_ranges, format = "markdown")
+```{r lagos_params, echo=FALSE, results="asis"}
+knitr::kable(lagos_param_ranges, format = "markdown")
 ```

### 2019-01-29 manuscript/manuscript.Rmd
started adding in citation fig numbers to trend figures.

@@
 
-*`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.")`*
+(`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+
+*`r fig_num(name = "temp_ri")`*
 
@@
 
-*`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.")`*
+(`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+
+*`r fig_num(name = "tn_ri")`*
 

### 2019-01-30 manuscript/manuscript.Rmd
few updates

@@
 During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
-For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 ```
+
+For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 
-*`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.")`*
+(`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+
+*`r fig_num(name = "tp_ri")`*
 
@@
 
-*`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll *a* (average z-score) in Rhode Island lakes and ponds.")`*
+(`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+
+*`r fig_num(name = "chla_ri")`*
 
@@
 
-*`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.")`*
+(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
+
+*`r fig_num(name = "tn_lagos")`*
 
@@
 
-*`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.")`*
+(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).
+
+*`r fig_num(name = "tp_lagos")`*
 
@@
 
-*`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll *a* (average z-score) in the LAGOSNE states.")`*
+(`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
+
+*`r fig_num(name = "chla_lagos")`*
 

### 2019-02-01 manuscript/manuscript.Rmd
first line of results text for state trends!

@@
 
-(`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Across the years of this study, average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r` , p-value: `r`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
 

### 2019-02-06 manuscript/manuscript.Rmd
new render, with tau and pvalues available now.

--- 2019-01-30 manuscript/manuscript.Rmd
@@
 
-(`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Across the years of this study, average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
 

### 2019-02-06 R/figures.R

@@ -9,50 +9,62 @@ chla_gg <- ww_data %>%
   wq_trend_gg( "chla", yvar = "measurement_scale", 
                        y = "Average Yearly Scaled Chlorophyll", x = "Year",
                write = here("data/chla_data.csv"))
-ggsave(here("figures/ww_chla_trends.jpg"), chla_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_chla_trends.jpg"), chla_gg[[1]], width = 5, height = 3.75, 
        units = "in", dpi = 600)
+
+ww_chla_kt <- chla_gg[[2]]
 
 tn_gg <-  ww_data %>%
   filter(state == "RI") %>%
   wq_trend_gg("total_n", yvar = "measurement_scale", 
                       y = "Average Yearly Scaled Total Nitrogen", x = "Year",
               write = here("data/total_n_data.csv"))
-ggsave(here("figures/ww_tn_trends.jpg"), tn_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_tn_trends.jpg"), tn_gg[[1]], width = 5, height = 3.75, 
        units = "in", dpi = 600)
+
+ww_tn_kt <- tn_gg[[2]]
 
 tp_gg <-  ww_data %>%
   filter(state == "RI") %>%
   wq_trend_gg("total_p", yvar = "measurement_scale", 
                       y = "Average Yearly Scaled Total Phosphorus", x = "Year",
                       write = here("data/total_p_data.csv"))
-ggsave(here("figures/ww_tp_trends.jpg"), tp_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_tp_trends.jpg"), tp_gg[[1]], width = 5, height = 3.75, 
        units = "in", dpi = 600)
+
+ww_tp_kt <- tp_gg[[2]]
 
 temp_gg <- ww_data %>%
   filter(state == "RI") %>%
   wq_trend_gg("temp", yvar = "measurement_scale", 
                        y = "Average Yearly Scaled Temperature", x = "Year",
               write = here("data/temp_data.csv"))
-ggsave(here("figures/ww_temp_trends.jpg"), temp_gg, width = 5, height = 3.75, 
+ggsave(here("figures/ww_temp_trends.jpg"), temp_gg[[1]], width = 5, height = 3.75, 
        units = "in", dpi = 600)
+
+ww_temp_kt <- temp_gg[[2]]
 
[…]

### 2019-02-06 R/functions.R

@@ -73,7 +73,6 @@ wq_trend_gg <- function(df, wqparam,
                        minor_breaks = NULL) +
     scale_y_continuous(limits = c(-0.5, 0.75))
   
-  gg
-
+  list(gg, kt)
 }
  

### 2019-02-11 manuscript/manuscript.Rmd
draft results for RI done, fixed map, moved WW to methods. closes #11, closes #10, closes #3.

--- 2019-01-30 manuscript/manuscript.Rmd
@@
 ---
-title: "26 Years of Water Quality Change in Rhode Island Lakes and Ponds"
+title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
 output: 
@@
 
-The University of Rhode Island's Watershed Watch (URI WW) Volunteer Monitoring program has been collecting water quality data from Rhode Island lakes and ponds for over 25 years.
-The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
-Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
-For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen.
-Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.
-
 Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
@@
 
-The goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island.
-We will answer two particular questions.
+The goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
 First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
@@
 The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
-MORE DETAILS ON RIWW.
+The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 25 years.
+The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
+URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
+Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen.
+Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.
 
@@
 
-In particular we selected data from the RIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+In particular we selected data from the URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
 For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
@@
 
-(`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 
-(`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+*`r fig_num(name = "chla_ri")`*
+
+![](../figures/ww_chla_trends.jpg)
+
+```{r no_early_tn, echo=FALSE}
+ww_tn_kt_no9394 <- Kendall(ww_tn_df[c(-1,-2),]$year, ww_tn_df[c(-1,-2),]$mn_value)
+```
+
+Average yearly trends for nutrients showed weaker or no trends over time.
+Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).
+Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 
-(`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-
 *`r fig_num(name = "tp_ri")`*
@@
 ![](../figures/ww_tp_trends.jpg)
-
-(`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-
-*`r fig_num(name = "chla_ri")`*
-
-![](../figures/ww_chla_trends.jpg)
 

### 2019-02-11 manuscript/manuscript.Rmd

@@
 
-General stuff about long-term trends, monitoring programs and watershed watch details
+1. Water quality in last several decades (Why Water Quality Paragraph(s))
+  - Why of interest
+    - cyanos stuff
+    - CWA stuff
+  - Understand long-term trends from local, to regional, to national
+  - What variables are of interest and why
+    - productivity
+    - climate change
+    - Nutrient enrichment stuff
 
-Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
-Something about spatial and temporal variability of water quality
+2. Temporal and Spatial scaling of water quality trends Paragraph
 
-The goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
+3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
+  - Shown to be stable in LAGOS (Maybe other studies in NE)
+  - "Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds."
+
+4. Rhode Island Paragraph (May be combined with #3)
+
+4.
+Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
 First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.

### 2019-02-12 manuscript/manuscript.Rmd
converted mostly to slope.  need to update methods.

@@
 
-Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 ```{r no_early_tn, echo=FALSE}
-ww_tn_kt_no9394 <- Kendall(ww_tn_df[c(-1,-2),]$year, ww_tn_df[c(-1,-2),]$mn_value)
+ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
+  tidy() %>%
+  slice(2) %>%
+  select(slope = estimate, p.value)
 ```
@@
 Average yearly trends for nutrients showed weaker or no trends over time.
-Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).
-Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Total nitrogen did have a significant trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Since 1995, the yearly trend is not significant (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 4)`).
+Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 
-(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
+In general, broad regional trends were less obvious and in the case of chlorophyll *a* and total phosphorus, the trends were not significant.
+In particular, chlorophyll *a* showed a slight, yet non-significant negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 4)`, `r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
 
-*`r fig_num(name = "tn_lagos")`*
+*`r fig_num(name = "chla_lagos")`*
 
-![](../figures/lagos_tn_trends.jpg)
+![](../figures/lagos_chla_trends.jpg)
 
-(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).
+Total phosphorus shows a slight and non-significant increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`) and total nitrogen shows a slight negative non-significant trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 4)`, `r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
 
@@
 
-(`r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
+*`r fig_num(name = "tn_lagos")`*
 
-*`r fig_num(name = "chla_lagos")`*
-
-![](../figures/lagos_chla_trends.jpg)
+![](../figures/lagos_tn_trends.jpg)
 

### 2019-02-12 R/figures.R

@@ -14,6 +14,7 @@ ggsave(here("figures/ww_chla_trends.jpg"), chla_gg[[1]], width = 7.5, height = 5
 
 ww_chla_kt <- chla_gg[[2]]
 ww_chla_df <- chla_gg[[3]]
+ww_chla_rl <- chla_gg[[4]]
 
 tn_gg <-  ww_data %>%
   filter(state == "RI") %>%
@@ -25,6 +26,7 @@ ggsave(here("figures/ww_tn_trends.jpg"), tn_gg[[1]], width = 7.5, height = 5.625
 
 ww_tn_kt <- tn_gg[[2]]
 ww_tn_df <- tn_gg[[3]]
+ww_tn_rl <- tn_gg[[4]]
 
 tp_gg <-  ww_data %>%
   filter(state == "RI") %>%
@@ -36,6 +38,7 @@ ggsave(here("figures/ww_tp_trends.jpg"), tp_gg[[1]], width = 7.5, height = 5.625
 
 ww_tp_kt <- tp_gg[[2]]
 ww_tp_df <- tp_gg[[3]]
+ww_tp_rl <- tp_gg[[4]]
 
 temp_gg <- ww_data %>%
   filter(state == "RI") %>%
@@ -47,6 +50,7 @@ ggsave(here("figures/ww_temp_trends.jpg"), temp_gg[[1]], width = 7.5, height = 5
 
 ww_temp_kt <- temp_gg[[2]]
 ww_temp_df <- temp_gg[[3]]
+ww_temp_rl <- temp_gg[[4]]
 
 lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
@@ -57,6 +61,7 @@ ggsave(here("figures/lagos_chla_trends.jpg"), lagos_chla_gg[[1]], width = 7.5, h
        units = "in", dpi = 600)
 lagos_chla_kt <- lagos_chla_gg[[2]]
 lagos_chla_df <- lagos_chla_gg[[3]]
+lagos_chla_rl <- lagos_chla_gg[[4]]
 
 lagos_tn_gg <- wq_trend_gg(lagos_data, "total_n", yvar = "measurement_scale", 
                              y = "Average Yearly Scaled Total Nitrogen", x = "Year",
@@ -65,6 +70,7 @@ ggsave(here("figures/lagos_tn_trends.jpg"), lagos_tn_gg[[1]], width = 7.5, heigh
[…]

### 2019-02-12 manuscript/manuscript.Rmd
changed method from kendalls tau to regression.  closes #12

@@
 
-### Kendall's Tau
+### Linear regression
 
-Kendall's Tau and the Mann-Kendall test are commonly used to test for monotonic trends water quality data [@mann1945nonparametric; @helsel2002statistical; @hipel1994time].
-We followed these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
+Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
+We used these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
+For each parameter we fit a regression line to the z scores as a function of year and tested the null hypothesis that no trend existed (e.g.
+$\beta$~1~ = 0).
 

### 2019-02-20 manuscript/manuscript.Rmd
added section 1 of 3 to introduction along with associated literature to .bib

--- 2019-02-11 manuscript/manuscript.Rmd
@@
   - Why of interest
-    - cyanos stuff
+    - cyanos stuff (toxins, increasing problems)
     - CWA stuff
@@
     - Nutrient enrichment stuff
+
+  Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global].
+Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes].
+Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
+Eutrophication has caused the frequency and severity of algal blooms to increase, and these blooms are predicted to worsen under a warming climate [@michalak2013record; @kosten2012warmer; @taranu2015acceleration].
+Increasing HABs will cause further declines in water quality, greater risks for human and animal health, and will be economically costly [@brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+Management strategies have historically focused on reducing P inputs to lakes, but research also suggests that concurrently reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
+These contrasting studies indicate that spatial differences of HABs exist and that long-term studies are needed to identify trends at local, regional, and national scales.
 

### 2019-03-05 manuscript/manuscript.Rmd
added another paragraph and related literature

@@
 title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
+author:
+- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
+  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A
+  ^*2*^
+- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
+  Ecology Division, Narragansett, RI 02882
+- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
+  RI 02881
+- ^*3*^ ORISE, Narragansett, RI 02882
+- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
+bibliography: manuscript.bib
 output:
+  html_document:
+    df_print: paged
   word_document:
     reference_docx: manuscript_template.docx
-author:
-- "Hollister. J. W. ^\\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A ^*2*^"
-- "^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic Ecology Division, Narragansett, RI 02882"
-- "^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston, RI 02881"
-- "^*3*^ ORISE, Narragansett, RI 02882"
-- "^\\*^  *corresponding author: hollister.jeff@epa.gov*"
-bibliography: manuscript.bib
-csl: plos.csl
-abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.
-Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g.
-in urbanized northern Rhode Island).
-Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis.
-This provides a robust and commonly scaled measurement to explore this data for long-term trends.
-State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen.
-Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.
-While yearly trends are useful, they do mask month-to-month variability differences across sites.
-Additionally, while most sites track the yearly trend in decreasing water quality, there are bright spots with some sites improving over the 25 years.
-Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.
-Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition.
 editor_options:
   chunk_output_type: console
+csl: plos.csl
+abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring
+  Program has been collecting water quality data on Rhode Island lakes and ponds for
+  close to three decades, allowing us to explore long-term trends in common water
+  quality parameters.  Not all lakes and ponds in the study area were sampled across
+  the full time period and lakes were often added in geographic clusters (e.g. in
+  urbanized northern Rhode Island).  Not unlike how long-term temperature records
+  are analyzed, we centered and scaled (i.e., the z-score) water quality measurements
+  on a per-station basis. This provides a robust and commonly scaled measurement to
+  explore this data for long-term trends.  State-wide aggregation shows increasing
+  temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus
+  is showing a decline, perhaps reflecting the management focus on phosphorus reductions.  While
+  yearly trends are useful, they do mask month-to-month variability differences across
+  sites.  Additionally, while most sites track the yearly trend in decreasing water
+  quality, there are bright spots with some sites improving over the 25 years. Contrary
+  to previously reported analyses that show relatively stable water quality at the
+  regional scale, our analysis shows that long-term water quality trends within Rhode
+  Island show some parameters improving while others are in decline.  Importantly,
+  this analysis also points out the incredible value and importance of data from long-term
+  monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for
+  identifying trends in environmental condition.
 ---
@@
 Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
-Eutrophication has caused the frequency and severity of algal blooms to increase, and these blooms are predicted to worsen under a warming climate [@michalak2013record; @kosten2012warmer; @taranu2015acceleration].
+Eutrophication has caused the frequency and severity of harmful algal blooms (HABs) to increase, and these blooms are predicted to worsen under a warming climate [@michalak2013record; @kosten2012warmer; @taranu2015acceleration].
 Increasing HABs will cause further declines in water quality, greater risks for human and animal health, and will be economically costly [@brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
 Management strategies have historically focused on reducing P inputs to lakes, but research also suggests that concurrently reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These contrasting studies indicate that spatial differences of HABs exist and that long-term studies are needed to identify trends at local, regional, and national scales.
+These contrasting studies indicate that spatial differences and relationships between N, P, chlorophyll *a*, and HABs exist and that long-term studies are needed to identify trends at local, regional, and national scales.
 
@@
 
+Stoddard 2016
+
+ Despite these challenges, it is important to study at multiple spatial scales because emergent trends on a regional scales may not be evident when studying individual lakes @cheruvelil2013multi; @lottig2014long].
+
 3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
-  - Shown to be stable in LAGOS (Maybe other studies in NE)
-  - "Given what is known about long term trends in water quality within the broader region of the northeastern United States [@Oliver_2017], we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds."
 
-4. Rhode Island Paragraph (May be combined with #3)
-
-4.
+   Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Macroscale (i.e.
+subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
+This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
+Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
+Given what is known about long term trends in water quality within the broader region of the northeastern United States , we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
  Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.

### 2019-03-07 manuscript/manuscript.Rmd
fresh render with stephen's draft intro

@@
 output:
-  html_document:
-    df_print: paged
   word_document:
@@
 editor_options:
-  chunk_output_type: console
+  chunk_output_type: inline
 csl: plos.csl
-abstract: The University of Rhode Island’s Watershed Watch Volunteer Monitoring
+abstract: The University of Rhode Island's Watershed Watch Volunteer Monitoring
   Program has been collecting water quality data on Rhode Island lakes and ponds for
@@
 fig_num <- captioner(prefix = "Figure")
+options(scipen = 2, digits = 5)
 ```
@@
 
-1. Water quality in last several decades (Why Water Quality Paragraph(s))
-  - Why of interest
-    - cyanos stuff (toxins, increasing problems)
-    - CWA stuff
-  - Understand long-term trends from local, to regional, to national
-  - What variables are of interest and why
-    - productivity
-    - climate change
-    - Nutrient enrichment stuff
-
 Aquatic ecosystems have been altered as the result of human activities modifying nutrient cycling on a global scale [@vitousek1997human; @finlay2013human; @filippelli2008global].
@@
 
-2. Temporal and Spatial scaling of water quality trends Paragraph
-
-Stoddard 2016
-
- Despite these challenges, it is important to study at multiple spatial scales because emergent trends on a regional scales may not be evident when studying individual lakes @cheruvelil2013multi; @lottig2014long].
-
-3. Work has been done in the larger Northeastern Region of US (Northeastern Paragraph)
+Programs such as USEPA's National Lakes Assessment (NLA) provide data that allows for continental scale water quality analysis.
+These analyses can be used for managing water resources by developing water quality criteria for N, P, chlorophyll *a*, and microcystin [@herlihy2013using; @yuan2014managing].
+Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
+Large-scale data can also be used for water quality modeling across broad spatial scales including predicting lake trophic state, which can predict ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+These trophic state models indicate that landscape variables (i.e.
+ecoregion, elevation, and latitude) are important and that regional trends exist.
+Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
+Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
 
@@
 Given what is known about long term trends in water quality within the broader region of the northeastern United States , we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
+
 Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
@@
 
-Average yearly scaled temperature in lakes and ponds appear to be increasing (Kendall's Tau: `r round(ww_temp_kt$tau,3)` , p-value: `r round(ww_temp_kt$sl, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (Kendall's Tau: `r round(ww_chla_kt$tau,3)` , p-value: `r round(ww_chla_kt$sl, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 Average yearly trends for nutrients showed weaker or no trends over time.
-Total nitrogen did have a significant trend (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-Since 1995, the yearly trend is not significant (Kendall's Tau: `r round(ww_tn_kt_no9394$tau,3)` , p-value: `r round(ww_tn_kt_no9394$sl, 4)`).
-Total phosphorus does not show a trend over time in the yearly scaled values (Kendall's Tau: `r round(ww_tn_kt$tau,3)` , p-value: `r round(ww_tn_kt$sl, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Total nitrogen did have a significant trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Since 1995, the yearly trend is not significant (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
+Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
 
@@
 
-(`r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
+In general, broad regional trends were less obvious and in the case of chlorophyll *a* and total phosphorus, the trends were not significant.
+In particular, chlorophyll *a* showed a slight, yet non-significant negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
 
-*`r fig_num(name = "tn_lagos")`*
+*`r fig_num(name = "chla_lagos")`*
 
-![](../figures/lagos_tn_trends.jpg)
+![](../figures/lagos_chla_trends.jpg)
 
-(`r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`).
+Total phosphorus shows a slight and non-significant increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`) and total nitrogen shows a slight negative non-significant trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
 

### 2019-03-07 manuscript/manuscript.Rmd
few small edits to intro

@@
 Increasing HABs will cause further declines in water quality, greater risks for human and animal health, and will be economically costly [@brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
-Management strategies have historically focused on reducing P inputs to lakes, but research also suggests that concurrently reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
+To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that concurrently reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
 These contrasting studies indicate that spatial differences and relationships between N, P, chlorophyll *a*, and HABs exist and that long-term studies are needed to identify trends at local, regional, and national scales.
@@
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long term trends in water quality within the broader region of the northeastern United States , we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
+Given what is known about long term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
 
@@
 First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends>.
+In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOS project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [NEED CITATIONS].
 

### 2019-03-07 manuscript/manuscript.Rmd
outline of description and possible journals

@@
   temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus
-  is showing a decline, perhaps reflecting the management focus on phosphorus reductions.  While
-  yearly trends are useful, they do mask month-to-month variability differences across
-  sites.  Additionally, while most sites track the yearly trend in decreasing water
+  is showing a decline, perhaps reflecting the management focus on phosphorus reductions.   While yearly trends are useful, they do mask month-to-month variability differences 
+  across sites.  Additionally, while most sites track the yearly trend in decreasing water
   quality, there are bright spots with some sites improving over the 25 years. Contrary
@@
 Given what is known about long term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
+Paragraph about awesomeness of WW and volunteer monitoring groups - Betty
 
@@
 
-For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
 
@@
 
+## Volunteer Monitoring Data
+
+### Why this is possible - Fantastic WW data
+
 ## Trends
+
+### Parameter by parameter discussion of trends
+
+### How do trends break down by trophic state?
 
@@
 
+## Management/Implications
+
+### Season Length Changes
+
 ## Caveats

### 2019-03-19 manuscript/manuscript.Rmd
Added paragraph to into and some citations

@@
 Paragraph about awesomeness of WW and volunteer monitoring groups - Betty
+
+Examining long term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
+URI’s Watershed Watch is a XX year-old citizen science project that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
+Over the past XX years, volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
+These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
+Often due to lack of resources, these efforts would not happen otherwise.
+Here specifically, URIWW data contributed to not only the larger subcontinent extent study, but also allowed us to examine the long-term trends of Rhode Island.
 

### 2019-03-20 R/functions.R
added package installations

@@ -1,3 +1,16 @@
+# Installations
+pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
+          "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
+          "USAboundariesData")
+
+for(i in pkgs){
+  if(!require(i)){
+    install.packages(i)
+  }
+}
+
+if(!require(LAGOSNE)){devtools::install_github("cont-limno/LAGOSNE")}
+
 # Library
 library(lubridate)
 library(broom)
@@ -10,7 +23,6 @@ library(here)
 library(ggplot2)
 library(hrbrthemes)
 library(Kendall)
-if(!require(LAGOSNE)){devtools::install_github("cont-limno/LAGOSNE")}
 library(LAGOSNE)
 library(sf)
 library(USAboundaries)

### 2019-03-20 manuscript/manuscript.Rmd
fresh render prior to next week meeting.

@@
 ```{r setup, include=FALSE}
-library(knitr)
-library(captioner)
 source(here::here("R/functions.R"), encoding = "UTF-8")

### 2019-03-20 R/functions.R

@@ -1,11 +1,13 @@
 # Installations
 pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
           "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
-          "USAboundariesData")
+          "USAboundariesData","knitr","captioner")
 
 for(i in pkgs){
-  if(!require(i)){
-    install.packages(i)
+  if(!i %in% installed.packages()){
+    browser()
+    install.packages(i, repos = c("http://packages.ropensci.org", 
+                                  "https://cran.rstudio.com"))
   }
 }
 
@@ -27,6 +29,8 @@ library(LAGOSNE)
 library(sf)
 library(USAboundaries)
 library(USAboundariesData)
+library(knitr)
+library(captioner)
 
 #' Function to filter on number of years sampled per site
 #' @param df data frame to filter

### 2019-04-08 manuscript/manuscript.Rmd
still need to force length of x-axis to 1993 to 2016...

@@
 ---
-title: "Long-term Water Quality Change in Rhode Island Lakes and Ponds"
+title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
 author:
@@
 csl: plos.csl
-abstract: The University of Rhode Island's Watershed Watch Volunteer Monitoring
-  Program has been collecting water quality data on Rhode Island lakes and ponds for
-  close to three decades, allowing us to explore long-term trends in common water
-  quality parameters.  Not all lakes and ponds in the study area were sampled across
-  the full time period and lakes were often added in geographic clusters (e.g. in
-  urbanized northern Rhode Island).  Not unlike how long-term temperature records
-  are analyzed, we centered and scaled (i.e., the z-score) water quality measurements
-  on a per-station basis. This provides a robust and commonly scaled measurement to
-  explore this data for long-term trends.  State-wide aggregation shows increasing
-  temperature, chlorophyll *a*, and total nitrogen. Interestingly, total phosphorus
-  is showing a decline, perhaps reflecting the management focus on phosphorus reductions.   While yearly trends are useful, they do mask month-to-month variability differences 
-  across sites.  Additionally, while most sites track the yearly trend in decreasing water
-  quality, there are bright spots with some sites improving over the 25 years. Contrary
-  to previously reported analyses that show relatively stable water quality at the
-  regional scale, our analysis shows that long-term water quality trends within Rhode
-  Island show some parameters improving while others are in decline.  Importantly,
-  this analysis also points out the incredible value and importance of data from long-term
-  monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for
-  identifying trends in environmental condition.
+abstract: "The University of Rhode Island's Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.
+Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g.
+in urbanized northern Rhode Island).
+Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis.
+This provides a robust and commonly scaled measurement to explore this data for long-term trends.
+State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen.
+Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.
+Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.
+Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition."
 ---

### 2019-04-08 R/data_clean.R

@@ -67,7 +67,7 @@ ww_lake_trend_data <- ww_all %>%
                                     TRUE ~ `Station Name`)) %>% #Fixes 4 digit ids
   #filter(`Station Name` %in% filter_year(., 25)) %>% # Get stations with 20+ years - moved to plots...
   filter(year >= 1993) %>% #only get data for 1993 and beyond as that is earliest TN data for WW.  Make sure all ranges match
-  filter(year <= 2013) %>% #only get data from 2013 and prior to match Oliver
+  #filter(year <= 2013) %>% #only get data from 2013 and prior to match Oliver
   filter(Parameter %in% params) %>% #Filter out subset of parameters
   filter(Concentration != `Detection Limit`) %>% #filters out measurements at detect limit (temporary)
   mutate(Parameter = case_when(Parameter == params[1] ~ "temp", #NAMES!!!!!
@@ -132,7 +132,7 @@ lagos_data <- lagosne_select(table = "epi_nutr",
          day = as(day(ymd(sampledate)), "integer"),
          `Station Name` = lagoslakeid, np_ratio = tn/tp) %>%
   filter(year >= 1993) %>% #only get data for 1993 and beyond as that is earliest TN data for WW.  Make sure all ranges match
-  filter(year <= 2013) %>% #only get data from 2013 and prior to match Oliver
+  #filter(year <= 2013) %>% #only get data from 2013 and prior to match Oliver
   select(`Station Name`, programname, year, month, day, chla, total_p = tp, total_n = tn, np_ratio) %>%
   gather(param, measurement, chla:np_ratio) %>%
   filter(!is.na(measurement)) %>%

### 2019-04-08 R/functions.R

@@ -33,6 +33,7 @@ library(USAboundaries)
 library(USAboundariesData)
 library(knitr)
 library(captioner)
+library(cowplot)
 
 #' Function to filter on number of years sampled per site
 #' @param df data frame to filter

### 2019-04-09 manuscript/manuscript.Rmd
got new figures with improved sd error bars closes #40

@@
 
-## Trend Analysis
+## Water Quality Trend Analysis
 
@@
 For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees celsius to a high of 30 degrees celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
-To standardize these values we used the common approach of dividing each anomlaly by the standard deviation for the reference period.
+To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period.
 This value is commonly referred to as a Z-score.
@@
 
-Furthermore, since we are interest in water quality trends over time, we wanted to explore how each site was responding.
-Thus, our z-scores were calculated over the reference period, 1990-2016, for each site.
+Furthermore, since we are interested in water quality trends over time, we wanted to explore how each site was responding.
+Thus, our z-scores were calculated over the reference period, 1993-2016, for each site.
 In other words the long-term mean and standard deviation were calculated on a site-by-site basis.
 
-### Linear regression
+### Summarizing z-scores
+
+Add stuff here. Include an example figure with columns as years, rows as lakes showing how to calc z and summarize.
+
+### Linear regression on yearly means
 
@@
 
-Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.
+Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
+Panel A.
+shows yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.
+Panel B.
+shows yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
 
@@
 
-*`r fig_num(name = "chla_ri")`*
+*`r fig_num(name = "chla_trend")`*
 
-![](../figures/ww_chla_trends.jpg)
+![](../figures/paired_chla_trends.jpg)
 
@@
 Average yearly trends for nutrients showed weaker or no trends over time.
-Total nitrogen did have a significant trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
-Since 1995, the yearly trend is not significant (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
-Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).
+Total nitrogen did suggest a positive trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of total nitrogen (average z-score).
+Panel A.
+shows yearly averaged total nitrogen z-scores from the URI Watershed Watch data.
+Panel B.
+shows yearly averaged total nitrogen z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
+Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of total phosphorus (average z-score).
+Panel A.
+shows yearly averaged total phosphorus z-scores from the URI Watershed Watch data.
+Panel B.
+shows yearly averaged total phosphorus z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+The pattern is the same for the total nitrogren to total phosphorus ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of total nitrogen relative to the concentration of total phosphorus (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the total nitrogen:total phosphorus ratio (average z-score).
+Panel A.
+shows yearly averaged total nitrogen:total phosphorus ratio z-scores from the URI Watershed Watch data.
+Panel B.
+shows yearly averaged total nitrogen:total phosphorus ratio z-scores from the LAGOSNE dataset.
+Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A)
 
-*`r fig_num(name = "tn_ri")`*
+*`r fig_num(name = "tn_trend")`*
 
-![](../figures/ww_tn_trends.jpg)
+![](../figures/paired_tn_trends.jpg)
 
-*`r fig_num(name = "tp_ri")`*
+*`r fig_num(name = "tp_trend")`*
 
-![](../figures/ww_tp_trends.jpg)
+![](../figures/paired_tp_trends.jpg)
+
+*`r fig_num(name = "np_trend")`*
+
+![](../figures/paired_np_trends.jpg)
 
@@
 
-In general, broad regional trends were less obvious and in the case of chlorophyll *a* and total phosphorus, the trends were not significant.
-In particular, chlorophyll *a* showed a slight, yet non-significant negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).
-
-*`r fig_num(name = "chla_lagos")`*
-
-![](../figures/lagos_chla_trends.jpg)
-
-Total phosphorus shows a slight and non-significant increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`) and total nitrogen shows a slight negative non-significant trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
-
-*`r fig_num(name = "tp_lagos")`*
-
-![](../figures/lagos_tp_trends.jpg)
-
-*`r fig_num(name = "tn_lagos")`*
-
-![](../figures/lagos_tn_trends.jpg)
+In general, there was little to suggest broad regional trends.
+Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), total phosphorus shows a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), total nitrogen shows a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the total nitrogen:total phosphorus ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 

### 2019-04-09 R/data_clean.R

@@ -117,7 +117,9 @@ ww_lake_trend_data <- ww_lake_trend_data %>%
   group_by(station_name, param) %>%
   mutate(measurement_scale = scale(mn_measurement),
          measurement_anmly = mn_measurement - mean(mn_measurement),
-         lt_mean = mean(mn_measurement))
+         lt_mean = mean(mn_measurement),
+         lt_sd = sd(mn_measurement),
+         lt_n = n())
   
 
 write_csv(ww_lake_trend_data, here("data/ww_lake_trend_data.csv"))
@@ -144,7 +146,9 @@ lagos_data <- lagosne_select(table = "epi_nutr",
   group_by(station_name, param) %>%
   mutate(measurement_scale = scale(measurement),
          measurement_anmly = measurement - mean(measurement),
-         lt_mean = mean(measurement))
+         lt_mean = mean(measurement),
+         lt_sd = sd(measurement),
+         lt_n = n())
 
 write_csv(lagos_data, here("data/lagos_lake_trend_data.csv"))
   

### 2019-04-09 R/figures.R

@@ -7,8 +7,9 @@ ww_data <- read_csv(here("data/ww_lake_trend_data.csv"))
 chla_gg <- ww_data %>%
   filter(state == "RI") %>%
   wq_trend_gg( "chla", yvar = "measurement_scale", 
-                       y = "Average Yearly Scaled Chlorophyll", x = "Year",
-               write = here("data/chla_data.csv"))
+               y = "Average Yearly Scaled Chlorophyll", x = "Year",
+               write = here("data/chla_data.csv"), error_bar = "sd",
+               title = "A. URI Watershed Watch Chlorophyll")
 ggsave(here("figures/ww_chla_trends.jpg"), chla_gg[[1]], width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
 
@@ -19,8 +20,9 @@ ww_chla_rl <- chla_gg[[4]]
 tn_gg <-  ww_data %>%
   filter(state == "RI") %>%
   wq_trend_gg("total_n", yvar = "measurement_scale", 
-                      y = "Average Yearly Scaled Total Nitrogen", x = "Year",
-              write = here("data/total_n_data.csv"))
+              y = "Average Yearly Scaled Total Nitrogen", x = "Year",
+              write = here("data/total_n_data.csv"), error_bar = "sd",
+              title = "A. URI Watershed Watch Total Nitrogen")
 ggsave(here("figures/ww_tn_trends.jpg"), tn_gg[[1]], width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
 
@@ -31,8 +33,9 @@ ww_tn_rl <- tn_gg[[4]]
 tp_gg <-  ww_data %>%
   filter(state == "RI") %>%
   wq_trend_gg("total_p", yvar = "measurement_scale", 
-                      y = "Average Yearly Scaled Total Phosphorus", x = "Year",
-                      write = here("data/total_p_data.csv"))
+              y = "Average Yearly Scaled Total Phosphorus", x = "Year",
+              write = here("data/total_p_data.csv"), error_bar = "sd",
+              title = "A. URI Watershed Watch Total Phosphorus")
 ggsave(here("figures/ww_tp_trends.jpg"), tp_gg[[1]], width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
 
@@ -43,8 +46,9 @@ ww_tp_rl <- tp_gg[[4]]
 temp_gg <- ww_data %>%
[…]

### 2019-04-09 R/functions.R

@@ -59,30 +59,31 @@ filter_year <- function(df, num_yrs){
 wq_trend_gg <- function(df, wqparam, 
                         yvar = c("measurement_anmly","measurement_scale"), 
                         num_yrs = 10, write = NULL, title = "",
-                        start_yr = 1893, ...){
+                        start_yr = 1993, error_bar = c("se", "sd"), ...){
   yvar <- rlang::sym(match.arg(yvar))
+  error_bar <- rlang::sym(match.arg(error_bar))
   df1 <- df %>%
     filter(param == wqparam) %>%
     filter(year >= start_yr) %>%
     filter(station_name %in% filter_year(., num_yrs))
   
+  
   df2 <- df1 %>%
-    group_by(year) %>%
-    summarize(mn_value = mean(!!yvar),
-              n = n(),
-              se = sd(!!yvar)/sqrt(n())) %>%
+    #This should take care of pseudoreplication by using the per site/year means
+    #results in n for years being equal to number of sites per year
+    group_by(station_name,year) %>%
+    summarize(mn_value_station = mean(!!yvar)) %>%
     ungroup() %>%
-    mutate(col_group = case_when(mn_value <= 0 ~ 
+    group_by(year) %>%
+    summarize(mn_value = mean(mn_value_station),
+              sd = sd(mn_value_station),
+              n = n(),
+              se = sd/sqrt(n())) %>%
+    mutate(col_group = case_when(mn_value < 0 ~ 
                                    "Less than long-term site average",
                                  mn_value > 0 ~ 
-                                   "Greater than long-term site average"))
-  
-   num_lakes <- df1 %>%
-     group_by(year) %>%
-     summarize(n_lakes = length(unique(station_name)))
-  
-   df2 <- df2 %>%
-     left_join(num_lakes)
+                                   "Greater than long-term site average",
+                                 TRUE ~ "Equal to long-term site average"))
   
   if(!is.null(write)){
     write_csv(df2, write, append = FALSE)
@@ -95,7 +96,7 @@ wq_trend_gg <- function(df, wqparam,
     select(slope = estimate, p.value) 
   
   gg <- ggplot(df2,aes(x = year, y = mn_value)) + 
-    geom_pointrange(aes(ymin=mn_value-se, ymax=mn_value+se, 
+    geom_pointrange(aes(ymin=mn_value-!!error_bar, ymax=mn_value+!!error_bar, 
[…]

### 2019-04-30 manuscript/manuscript.Rmd
re-order of discussion topics
cleaned up a bunch of issues, started on z-score schematic

@@
 editor_options:
-  chunk_output_type: inline
+  chunk_output_type: console
 csl: plos.csl
-abstract: "The University of Rhode Island's Watershed Watch Volunteer Monitoring Program has been collecting water quality data on Rhode Island lakes and ponds for close to three decades, allowing us to explore long-term trends in common water quality parameters.
-Not all lakes and ponds in the study area were sampled across the full time period and lakes were often added in geographic clusters (e.g.
-in urbanized northern Rhode Island).
-Not unlike how long-term temperature records are analyzed, we centered and scaled (i.e., the z-score) water quality measurements on a per-station basis.
-This provides a robust and commonly scaled measurement to explore this data for long-term trends.
-State-wide aggregation shows increasing temperature, chlorophyll *a*, and total nitrogen.
-Interestingly, total phosphorus is showing a decline, perhaps reflecting the management focus on phosphorus reductions.
-Contrary to previously reported analyses that show relatively stable water quality at the regional scale, our analysis shows that long-term water quality trends within Rhode Island show some parameters improving while others are in decline.
-Importantly, this analysis also points out the incredible value and importance of data from long-term monitoring programs, like the Univeristy of Rhode Island's Watershed Watch, for identifying trends in environmental condition."
+abstract: ""
 ---
@@
 Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
-Eutrophication has caused the frequency and severity of harmful algal blooms (HABs) to increase, and these blooms are predicted to worsen under a warming climate [@michalak2013record; @kosten2012warmer; @taranu2015acceleration].
-Increasing HABs will cause further declines in water quality, greater risks for human and animal health, and will be economically costly [@brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
-To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that concurrently reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These contrasting studies indicate that spatial differences and relationships between N, P, chlorophyll *a*, and HABs exist and that long-term studies are needed to identify trends at local, regional, and national scales.
+This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutorphic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
+These contrasting studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales.
 
-Programs such as USEPA's National Lakes Assessment (NLA) provide data that allows for continental scale water quality analysis.
-These analyses can be used for managing water resources by developing water quality criteria for N, P, chlorophyll *a*, and microcystin [@herlihy2013using; @yuan2014managing].
+Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental scale water quality analysis.
+These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
 Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
-Large-scale data can also be used for water quality modeling across broad spatial scales including predicting lake trophic state, which can predict ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
 These trophic state models indicate that landscape variables (i.e.
@@
 Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
-Macroscale (i.e.
+Macro-scale (i.e.
 subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
@@
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and ponds.
-Paragraph about awesomeness of WW and volunteer monitoring groups - Betty
+Given what is known about long term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and reservoirs.
 
 Examining long term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
-URI’s Watershed Watch is a XX year-old citizen science project that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
-Over the past XX years, volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
+URI’s Watershed Watch is a citizen science project founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
+Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
+These efforts have been ongoing in some waterbodies since 1988.
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
-Often due to lack of resources, these efforts would not happen otherwise.
-Here specifically, URIWW data contributed to not only the larger subcontinent extent study, but also allowed us to examine the long-term trends of Rhode Island.
+Additionally, the fact that volunteer monitors are used has allowed for this program to persist much longer than agency run programs that are often prone to uncertain funding.
+Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
+[@Oliver_2017], but also allowed us to examine the long-term trends in Rhode Island.
 
-Thus, the goals of this study were to look at approximately 25 years of lakes and ponds data in Rhode Island and answer two particular questions.
-First, what are the state-wide trends in total nitrogen, total phosphorus, chlorophyll *a*, and lake temperature and second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-In addition to these goals, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOS project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [NEED CITATIONS].
+Thus, the goals of this study were to look at approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
+First, what are the state-wide trends in total nitrogen, total phosphorus, nitrogen to phosphorus ratio (N:P), chlorophyll *a*, and lake temperature.
+Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
+Another focus of this paper was to apply existing methodologies for examining long term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long term trends.
+Lastly, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOS project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
 
@@
 
-The study area for this analyis includes data in lakes and ponds in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and pond sampling sites")`).
-The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and ponds for over 25 years.
+The study area for this analysis includes data from lakes and reservoirs in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
+The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
 The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
@@
 Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
-For freshwater lakes and ponds, weekly secchi depth and water temperature are recorded, along with bi-weekly Chl-a and dissolved oxygen.
-Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory for nutrients, alkalinity, pH, and bacteria.
+For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
+Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory.
 
-*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and pond sampling sites")`*
+*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`*
 ![](../figures/ww_map.jpg)
 
-In particular we selected data from the URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, total nitorgen, total phosphorus, and chlorophyll *a* and for each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and ponds which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), and `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
-Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, temp")` had data for all 4 parameters.
+In particular we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+For this analysis we were interested in trends in lake temperature, total nitrogen, total phosphorus, N:P, and chlorophyll *a*.
+For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
+Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
 
@@
 The authors found little change in water quality trends across this region [@Oliver_2017] .
-We repeated our analysis (see **Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
-Temperature data were not availble, thus we only examine trends, using our analytical methods, for Total Nitrogren, Total Phosphorus, and Chlorophyll *a* from the LAGOSNE database.
+We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
+Temperature data were not available, thus we only examine trends, using our analytical methods, for total nitrogen, total phosphorus, N:P, and chlorophyll *a* from the LAGOSNE database.
 
@@
 There are many different methods for analyzing time series data for trends.
-One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been addedd to a dataset.
-For instance, if you have long-term data on water quality for several ponds, yet the cleaner ponds wer sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
+One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been added to a dataset.
+For instance, if you have long-term data on water quality for several waterbodies, yet the cleaner waterbodies were sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
 Thus, it is necessary to account for this type of site-specific variation.
-This is similar to how the long-term temperature trends are analyzed using temperature anomolies [FILL IN GROUP DOING THIS].
+This is similar to how the long-term temperature trends are analyzed using temperature anomalies (e.g., [@jones1996calculating]).
 The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.
@@
 
-The anomaly approach is very useful and is the approach we chose to follow.
-However, using anomalies works best with a single measure (e.g temperature) or with mulitple measurements that are on the same scale.
+The anomaly approach is very useful and informed the approach we chose to follow.
+However, using anomalies works best with a single measure (e.g temperature) or with multiple measurements that are on the same scale.
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees celsius to a high of 30 degrees celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
-To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period.
-This value is commonly referred to as a Z-score.
-We use these z-scores to examine each water quality variable for a trend over the time period of 1990 to 2016.
-
+For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
+To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
+The resultant value is commonly referred to as a z-score.
+We use these z-scores to examine each water quality variable for a trend over the time period of 1993 to 2016.
 Furthermore, since we are interested in water quality trends over time, we wanted to explore how each site was responding.
 Thus, our z-scores were calculated over the reference period, 1993-2016, for each site.
-In other words the long-term mean and standard deviation were calculated on a site-by-site basis.
 
@@
 
-Add stuff here. Include an example figure with columns as years, rows as lakes showing how to calc z and summarize.
+Details for our approach of calculating the yearly z-scores are as follows and are presented graphically in NEED FIGURE FOR THIS.
+The steps below are repeated for each of the water quality parameters.
+
+1. For each site, calculate the reference mean and reference standard deviation for all values.
+2. Calculate the z-score for each measurement by subtracting the reference mean and dividing by the reference standard deviation.
+3.
+Summarize the z-scores by calculating an average z-score for each site in each year.
+This step prevent bias from pseudoreplication of multiple measurements of the same site in a given year ([@hurlbert1984pseudoreplication]).
+The same site averages across years are assumed to be independent.
+4. Summarize the yearly z-scores by calculating the average, standard deviation, and sample size of the per site z-scores.
+
+The resultant values are analyzed for a trend over time.
 
@@
 Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
-We used these standard procedures to test for postive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
+We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
 For each parameter we fit a regression line to the z scores as a function of year and tested the null hypothesis that no trend existed (e.g.
@@
 Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-The data we used in our analysis, the University of Rhode Island's Watershed Watch data, was inlcuded in that, but we were curious if regional trends were masking local scale trends in Rhode Island.
-In addition to the analysis condcuted by Oliver et al.
-[-@Oliver_2017], we analyzed the same dataset but using the approach outlined here.
+While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local scale trends in Rhode Island and thus wanted to compare the trends at the regional scale to the trends at the state scale.
+The analysis conducted by Oliver et al.
+[-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al.
+[-@Oliver_2017] but using the trend analysis approach outlined above.
 
@@
 
-During the sampling period of 1990 to 2016 Rhode Island lakes and ponds average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the sampling period of 1990 to 2016 Rhode Island lakes and reservoirs average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, average N:P ratio was `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 
-For lakes in ponds in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, average N:P ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
 
@@
 
-Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.
+Average yearly scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
 Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`).
@@
 Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
-The pattern is the same for the total nitrogren to total phosphorus ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of total nitrogen relative to the concentration of total phosphorus (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the total nitrogen:total phosphorus ratio (average z-score).
+The pattern is the same for the N:P ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of total nitrogen relative to the concentration of total phosphorus (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the total nitrogen:total phosphorus ratio (average z-score).
 Panel A.
@@
 shows yearly averaged total nitrogen:total phosphorus ratio z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A)
+Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
 
@@
 
-## Volunteer Monitoring Data
-
-### Why this is possible - Fantastic WW data
-
 ## Trends
@@
 
+## Volunteer Monitoring Data
+
+### Why this is possible - Fantastic WW data
+
 ## Caveats

### 2019-04-30 R/data_available.R

@@ -19,19 +19,29 @@ total_p_10yr_sites <- ww_data %>%
          param == "total_p") %>%
   filter_year(10)
 
+np_10yr_sites <- ww_data %>%
+  filter(state == "RI",
+         param == "np_ratio") %>%
+  filter_year(10)
+  
 temp_10yr_sites <- ww_data %>%
   filter(state == "RI",
          param == "temp") %>%
   filter_year(10)
 
 possible_sites <- unique(c(chla_10yr_sites,total_p_10yr_sites, 
-                           total_n_10yr_sites, temp_10yr_sites))
+                           total_n_10yr_sites, temp_10yr_sites, np_10yr_sites))
 data_avail <- data_frame(station_name = possible_sites, 
                          chla_avail = possible_sites %in% chla_10yr_sites,
                          total_n_avail = possible_sites %in% total_n_10yr_sites, 
                          total_p_avail = possible_sites %in% total_p_10yr_sites,
+                         np_avail = possible_sites %in% np_10yr_sites,
                          temp_avail = possible_sites %in% temp_10yr_sites) %>%
   mutate(params_avail = case_when(chla_avail == TRUE & total_n_avail == TRUE &
+                                    total_p_avail == TRUE & np_avail == TRUE &
+                                    temp_avail == TRUE ~ 
+                                    "chla, total_p, total_n, np, temp",
+                                  chla_avail == TRUE & total_n_avail == TRUE &
                                     total_p_avail == TRUE & temp_avail == TRUE ~ 
                                     "chla, total_p, total_n, temp",
                                   chla_avail == TRUE & total_n_avail == FALSE & 

### 2019-04-30 R/functions.R

@@ -5,7 +5,6 @@ pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr",
 
 for(i in pkgs){
   if(!i %in% installed.packages()){
-    browser()
     install.packages(i, repos = c("http://packages.ropensci.org", 
                                   "https://cran.rstudio.com"))
   }

### 2019-04-30 R/summary_stats.R

@@ -5,8 +5,9 @@ ww_param_ranges <- ww_all %>%
   filter(param == "chla" |
            param == "temp" |
            param == "total_n" |
-           param == "total_p") %>%
-  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
+           param == "total_p" |
+           param == "np_ratio") %>%
+  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "np_ratio","chla"))) %>%
   group_by(param) %>%
   summarize(min = round(min(mn_measurement, na.rm = T), 2),
             percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
@@ -16,7 +17,7 @@ ww_param_ranges <- ww_all %>%
             max = round(max(mn_measurement, na.rm = T), 2),
             sd = round(sd(mn_measurement, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
+  mutate(units = c("celsius", "µg/l", "µg/l", "", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 
@@ -27,8 +28,9 @@ ww_param_anomaly_ranges <- ww_all %>%
   filter(param == "chla" |
            param == "temp" |
            param == "total_n" |
-           param == "total_p") %>%
-  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "chla"))) %>%
+           param == "total_p"|
+           param == "np_ratio") %>%
+  mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "np_ratio", "chla"))) %>%
   group_by(param) %>%
   summarize(min = round(min(measurement_anmly, na.rm = T), 2),
             percentile_25 = round(quantile(measurement_anmly, probs = 0.25, na.rm = T), 2),
@@ -38,7 +40,7 @@ ww_param_anomaly_ranges <- ww_all %>%
             max = round(max(measurement_anmly, na.rm = T), 2),
             sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l","µg/l")) %>%
+  mutate(units = c("celsius", "µg/l", "µg/l", "", "µg/l")) %>%
   select(Parameter = param, Units = units, 
[…]

### 2019-05-06 R/data_clean.R
finished diagram of z-score method with example script and data

@@ -105,21 +105,25 @@ ww_np <- ww_lake_trend_data %>%
 
 ww_lake_trend_data <- ww_lake_trend_data %>%
   rbind(ww_np)
+
   
 ww_lake_trend_data <- ww_lake_trend_data %>%
-  left_join(ww_sites) %>%
-  filter(WB_Type == "Lake or Pond" |
-           WB_Type == "Reservoir") %>%
-  select(station_name:location, site_descr = Site_DESCR, mn_measurement, town = Town, county = COUNTY, state = State, 
-         lon_dd = LON_DD, lat_dd = LAT_DD, huc_12 = HUC_12, 
-         huc_10_name = HUC_10_NAME, huc_12_name = HUC_12_NAME) %>%
   filter(month >= 5 & month <= 10) %>%
+  group_by(station_name, year, param) %>%
+  summarize(station_year_mean = mean(mn_measurement)) %>%
+  ungroup() %>%
   group_by(station_name, param) %>%
-  mutate(measurement_scale = scale(mn_measurement),
-         measurement_anmly = mn_measurement - mean(mn_measurement),
-         lt_mean = mean(mn_measurement),
-         lt_sd = sd(mn_measurement),
-         lt_n = n())
+  mutate(measurement_scale = scale(station_year_mean),
+         measurement_anmly = station_year_mean - mean(station_year_mean),
+         lt_mean = mean(station_year_mean),
+         lt_sd = sd(station_year_mean),
+         lt_n = n()) %>%
+  left_join(ww_sites) %>%
+  filter(WB_Type == "Lake or Pond" | WB_Type == "Reservoir") %>%
+  select(station_name:lt_n, site_descr = Site_DESCR, 
+         town = Town, county = COUNTY, state = State, lon_dd = LON_DD, 
+         lat_dd = LAT_DD, huc_12 = HUC_12, huc_10_name = HUC_10_NAME, 
+         huc_12_name = HUC_12_NAME)
   
 
 write_csv(ww_lake_trend_data, here("data/ww_lake_trend_data.csv"))
@@ -143,11 +147,16 @@ lagos_data <- lagosne_select(table = "epi_nutr",
   #filter(`Station Name` %in% filter_year(., 20)) %>% #moved to plots and 10 years is min...
   filter(month >= 5 & month <= 10) %>%
   select(station_name = `Station Name`,year, month, day, param, measurement) %>%
+  #This should take care of pseudoreplication by using the per site/year means
+  #results in n for years being equal to number of sites per year
[…]

### 2019-05-06 R/figures.R

@@ -5,7 +5,7 @@ source(here::here("R/functions.R"))
 ww_data <- read_csv(here("data/ww_lake_trend_data.csv"))
 
 chla_gg <- ww_data %>%
-  filter(state == "RI") %>%
+  #filter(state == "RI") %>%
   wq_trend_gg( "chla", yvar = "measurement_scale", 
                y = "Average Yearly Scaled Chlorophyll", x = "Year",
                write = here("data/chla_data.csv"), error_bar = "sd",
@@ -18,7 +18,7 @@ ww_chla_df <- chla_gg[[3]]
 ww_chla_rl <- chla_gg[[4]]
 
 tn_gg <-  ww_data %>%
-  filter(state == "RI") %>%
+  #filter(state == "RI") %>%
   wq_trend_gg("total_n", yvar = "measurement_scale", 
               y = "Average Yearly Scaled Total Nitrogen", x = "Year",
               write = here("data/total_n_data.csv"), error_bar = "sd",
@@ -31,7 +31,7 @@ ww_tn_df <- tn_gg[[3]]
 ww_tn_rl <- tn_gg[[4]]
 
 tp_gg <-  ww_data %>%
-  filter(state == "RI") %>%
+  #filter(state == "RI") %>%
   wq_trend_gg("total_p", yvar = "measurement_scale", 
               y = "Average Yearly Scaled Total Phosphorus", x = "Year",
               write = here("data/total_p_data.csv"), error_bar = "sd",
@@ -44,7 +44,7 @@ ww_tp_df <- tp_gg[[3]]
 ww_tp_rl <- tp_gg[[4]]
 
 temp_gg <- ww_data %>%
-  filter(state == "RI") %>%
+  #filter(state == "RI") %>%
   wq_trend_gg("temp", yvar = "measurement_scale", 
               y = "Average Yearly Scaled Temperature", x = "Year",
               write = here("data/temp_data.csv"), error_bar = "sd",
@@ -57,7 +57,7 @@ ww_temp_df <- temp_gg[[3]]
 ww_temp_rl <- temp_gg[[4]]
 
 np_gg <- ww_data %>%
-  filter(state == "RI") %>%
+  #filter(state == "RI") %>%
   wq_trend_gg("np_ratio", yvar = "measurement_scale", 
               y = "Average Yearly Scaled N:P", x = "Year",
               write = here("data/np_data.csv"), error_bar = "sd",
[…]

### 2019-05-06 R/functions.R

@@ -61,6 +61,7 @@ wq_trend_gg <- function(df, wqparam,
                         start_yr = 1993, error_bar = c("se", "sd"), ...){
   yvar <- rlang::sym(match.arg(yvar))
   error_bar <- rlang::sym(match.arg(error_bar))
+  
   df1 <- df %>%
     filter(param == wqparam) %>%
     filter(year >= start_yr) %>%
@@ -68,14 +69,9 @@ wq_trend_gg <- function(df, wqparam,
   
   
   df2 <- df1 %>%
-    #This should take care of pseudoreplication by using the per site/year means
-    #results in n for years being equal to number of sites per year
-    group_by(station_name,year) %>%
-    summarize(mn_value_station = mean(!!yvar)) %>%
-    ungroup() %>%
     group_by(year) %>%
-    summarize(mn_value = mean(mn_value_station),
-              sd = sd(mn_value_station),
+    summarize(mn_value = mean(!!yvar),
+              sd = sd(!!yvar),
               n = n(),
               se = sd/sqrt(n())) %>%
     mutate(col_group = case_when(mn_value < 0 ~ 
@@ -112,7 +108,7 @@ wq_trend_gg <- function(df, wqparam,
                        limits = c(1993,2016)) +
     scale_y_continuous(labels = c(-2, -1, 0, 1, 2),
                        breaks = c(-2, -1, 0, 1, 2),
-                       limits = c(-2.25, 2.25))
+                       limits = c(-2.75, 2.75))
   
   list(gg, kt, df2, regress)
 }

### 2019-05-06 R/schematic.R

@@ -1,0 +1,21 @@
+library(readr)
+library(dplyr)
+schematic <- read_csv(here::here("data/schematic.csv"))
+
+station_year_mean <- schematic %>% 
+  group_by(station, year) %>%
+  summarize(station_year_mean = mean(meas)) %>%
+  ungroup()
+
+station_year_z <- station_year_mean %>%
+  group_by(station) %>%
+  mutate(station_year_z = scale(station_year_mean),
+         long_term_mean = mean(station_year_mean),
+         long_term_sd = sd(station_year_mean)) %>%
+  ungroup()
+  
+mean_z <- station_year_z %>%
+  group_by(year) %>%
+  summarize(mean_z = mean(station_year_z),
+            sd_z = sd(station_year_z),
+            n = n())

### 2019-05-06 manuscript/manuscript.Rmd
abstract drafted for NALMS.  Needs word culling!

@@
 - Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
-  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A
+  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
   ^*2*^
@@
 Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
-This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutorphic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].

### 2019-05-08 manuscript/manuscript.Rmd
abstract done! closes #42

@@
 csl: plos.csl
-abstract: ""
+abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
+Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine its effectiveness requires long-term datasets.
+A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States, however, trends at smaller scales may be masked.
+We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
+These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents.
+We assess for trends with z-scores (i.e.
+scaled anomalies) calculated on a per-station basis and examine yearly averages.
+Temperature and chlorophyll *a* are increasing.
+Total nitrogen shows a weak increasing trend driven by low years in the early 1990s.
+Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
+Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.
+In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
+Although not causal, this analysis suggests an association between lake temperature and productivity.
+Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.
+'
 ---
@@
 Thus, our z-scores were calculated over the reference period, 1993-2016, for each site.
+We refer to this approach as the site-specific z-scores
 
-### Summarizing z-scores
+### Summarizing site-specific z-scores
 
-Details for our approach of calculating the yearly z-scores are as follows and are presented graphically in NEED FIGURE FOR THIS.
-The steps below are repeated for each of the water quality parameters.
+Details for our approach of calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
+Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate these calculations is available from <https://github.com/jhollist/ri_wq_trends>.
+The general steps, listed below, are repeated for each of the water quality parameters.
 
+WRONG ORDER JEFF!
 1. For each site, calculate the reference mean and reference standard deviation for all values.

### 2019-05-08 R/data_available.R

@@ -31,7 +31,7 @@ temp_10yr_sites <- ww_data %>%
 
 possible_sites <- unique(c(chla_10yr_sites,total_p_10yr_sites, 
                            total_n_10yr_sites, temp_10yr_sites, np_10yr_sites))
-data_avail <- data_frame(station_name = possible_sites, 
+data_avail <- tibble(station_name = possible_sites, 
                          chla_avail = possible_sites %in% chla_10yr_sites,
                          total_n_avail = possible_sites %in% total_n_10yr_sites, 
                          total_p_avail = possible_sites %in% total_p_10yr_sites,
@@ -52,7 +52,8 @@ data_avail <- data_frame(station_name = possible_sites,
                                     "total_p, total_n",
                                   chla_avail == TRUE & total_n_avail == FALSE &
                                     total_p_avail == TRUE & temp_avail == TRUE ~
-                                    "chla, total_p, temp"))
+                                    "chla, total_p, temp",
+                                  TRUE ~ "other"))
 ww_avail_data_locations <- ww_data %>%
   select(station_name, lon_dd, lat_dd) %>%
   unique() %>%

### 2019-05-08 R/summary_stats.R

@@ -9,13 +9,13 @@ ww_param_ranges <- ww_all %>%
            param == "np_ratio") %>%
   mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "np_ratio","chla"))) %>%
   group_by(param) %>%
-  summarize(min = round(min(mn_measurement, na.rm = T), 2),
-            percentile_25 = round(quantile(mn_measurement, probs = 0.25, na.rm = T), 2),
-            mean = round(mean(mn_measurement, na.rm = T), 2),
-            median = round(median(mn_measurement, na.rm = T), 2),
-            percentile_75 = round(quantile(mn_measurement, probs = 0.75, na.rm = T), 2),
-            max = round(max(mn_measurement, na.rm = T), 2),
-            sd = round(sd(mn_measurement, na.rm = T), 2)) %>%
+  summarize(min = round(min(station_year_mean, na.rm = T), 2),
+            percentile_25 = round(quantile(station_year_mean, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(station_year_mean, na.rm = T), 2),
+            median = round(median(station_year_mean, na.rm = T), 2),
+            percentile_75 = round(quantile(station_year_mean, probs = 0.75, na.rm = T), 2),
+            max = round(max(station_year_mean, na.rm = T), 2),
+            sd = round(sd(station_year_mean, na.rm = T), 2)) %>%
   ungroup() %>%
   mutate(units = c("celsius", "µg/l", "µg/l", "", "µg/l")) %>%
   select(Parameter = param, Units = units, 
@@ -56,13 +56,13 @@ lagos_param_ranges <- lagos_all %>%
            param == "np_ratio") %>%
   mutate(param = factor(param,levels = c("total_n", "total_p", "np_ratio","chla"))) %>%
   group_by(param) %>%
-  summarize(min = round(min(measurement, na.rm = T), 2),
-            percentile_25 = round(quantile(measurement, probs = 0.25, na.rm = T), 2),
-            mean = round(mean(measurement, na.rm = T), 2),
-            median = round(median(measurement, na.rm = T), 2),
-            percentile_75 = round(quantile(measurement, probs = 0.75, na.rm = T), 2),
-            max = round(max(measurement, na.rm = T), 2),
-            sd = round(sd(measurement, na.rm = T), 2)) %>%
+  summarize(min = round(min(station_year_mean, na.rm = T), 2),
[…]

### 2019-05-10 manuscript/manuscript.Rmd
small change on captions, added schematic

@@
   chunk_output_type: console
-csl: plos.csl
+csl: ecology.csl
 abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
@@
 
-*`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`*
-![](../figures/ww_map.jpg)
+![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg)
 
@@
 
-The anomaly approach is very useful and informed the approach we chose to follow.
+Anomalies are very useful and informed the approach we chose to follow.
 However, using anomalies works best with a single measure (e.g temperature) or with multiple measurements that are on the same scale.
@@
 Details for our approach of calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
-Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate these calculations is available from <https://github.com/jhollist/ri_wq_trends>.
-The general steps, listed below, are repeated for each of the water quality parameters.
+Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <https://github.com/jhollist/ri_wq_trends>.
 
-WRONG ORDER JEFF!
-1. For each site, calculate the reference mean and reference standard deviation for all values.
-2. Calculate the z-score for each measurement by subtracting the reference mean and dividing by the reference standard deviation.
-3.
-Summarize the z-scores by calculating an average z-score for each site in each year.
-This step prevent bias from pseudoreplication of multiple measurements of the same site in a given year ([@hurlbert1984pseudoreplication]).
-The same site averages across years are assumed to be independent.
-4. Summarize the yearly z-scores by calculating the average, standard deviation, and sample size of the per site z-scores.
+![`r fig_num("z_schematic")`](../figures/z-score_schematic.jpg)
 
-The resultant values are analyzed for a trend over time.
+The general steps, outlined in `r fig_num("z_schematic", display = "cite")` and listed below, are repeated for each of the water quality parameters.
+
+1.
+For each site, calculate the per site and per year mean values which results in a single mean value for each site and year.
+This step prevent bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].
+The per site means across years are assumed to be independent.
+2. Calculate the long-term reference mean and reference standard deviation for all per site and per year means.  This results in a single long-term mean and standard deviation for each of the sites.
+3. Calculate the z-score for each per site and per year mean by subtracting the reference mean and dividing by the reference standard deviation.
+4. Summarize the per site and per year z-scores by calculating the mean z-score per year.  The resultant values are analyzed for a trend over time.
 
@@
 Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local scale trends in Rhode Island and thus wanted to compare the trends at the regional scale to the trends at the state scale.
+While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local scale trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
 The analysis conducted by Oliver et al.
@@
 
-*`r tab_num("ww_params")`*
-
 ```{r ww_params, echo=FALSE, results="asis"}
@@
 ```
+*`r tab_num("ww_params")`*
 
 For lakes and reservoirs in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, average N:P ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
-
-*`r tab_num("lagos_params")`*
 
@@
 ```
+*`r tab_num("lagos_params")`*
 
@@
 
-*`r fig_num(name = "temp_ri")`*
+![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
 
-![](../figures/ww_temp_trends.jpg)
-
-*`r fig_num(name = "chla_trend")`*
-
-![](../figures/paired_chla_trends.jpg)
+![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends.jpg)
 
@@
 
-*`r fig_num(name = "tn_trend")`*
+![`r fig_num(name = "tn_trend")`](../figures/paired_tn_trends.jpg)
 
-![](../figures/paired_tn_trends.jpg)
+![`r fig_num(name = "tp_trend")`](../figures/paired_tp_trends.jpg)
 
-*`r fig_num(name = "tp_trend")`*
-
-![](../figures/paired_tp_trends.jpg)
-
-*`r fig_num(name = "np_trend")`*
-
-![](../figures/paired_np_trends.jpg)
+![`r fig_num(name = "np_trend")`](../figures/paired_np_trends.jpg)
 

### 2019-05-13 manuscript/manuscript.Rmd
closes #8 and adds some climate anomaly refs

@@
 Details for our approach of calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
-Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <https://github.com/jhollist/ri_wq_trends>.
+Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <fill in with zenodo link> [@zenodo_citation].
 
@@
 Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].
 
@@
 
+Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
+neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
+One possible explanation for this is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
+Second,  scale does indeed matter when trying to identify long-term water quality trends as our analysis shows little increasing trend in broad scale (e.g.
+Northeastern and Mid-western United States) patterns of productivity, but do show clear increasing trends for just the state of Rhode Island.
+In the discussion that follows, we more closely address these trends, discuss broader implications of these results, and reiterate the value of the data analysis approach we used.
+
 ## Trends
 
-### Parameter by parameter discussion of trends
+- Parameter by parameter discussion of trends
+- How do trends break down by trophic state?
+- Local scale vs Regional/Global scale
+- RI is downstream - Bryan will write
 
-### How do trends break down by trophic state?
+## Broader Implications
 
-### Local scale vs Regional/Global scale
+- Provides feedback on long term changes for managers and stakeholders
+- Season Length Changes
+- How to stabilize trends?
+More reductions in nutrients?
+Engineered solutions to slow warming?
+Look to individual lakes that buck the trend for suggestions (e.g.
+local landcover may provide a buffer)
 
-### RI is downstream - Bryan will write
+## Data Analysis Approach
 
-## Management/Implications
-
-### Season Length Changes
-
-## Volunteer Monitoring Data
-
-### Why this is possible - Fantastic WW data
+- Why this is possible - Fantastic WW data
+- Use of site specific z-scores
+    - Other approaches available, but this one is useful becuase it provided readily interpretable results, robust to variations in sampling stations (i.e.
+added later in period, removed, etc.), and based off of approach with long history of use in exploring long term trends [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface]
 
@@
 
+- None yet...
+
 # Bibliography

### 2019-05-13 R/figures.R
updated figure caption to italics a

@@ -7,7 +7,8 @@ ww_data <- read_csv(here("data/ww_lake_trend_data.csv"))
 chla_gg <- ww_data %>%
   #filter(state == "RI") %>%
   wq_trend_gg( "chla", yvar = "measurement_scale", 
-               y = "Average Yearly Scaled Chlorophyll", x = "Year",
+               y = expression(paste("Average Yearly Scaled Chlorophyll  ", 
+                                    italic(" a"))), x = "Year",
                write = here("data/chla_data.csv"), error_bar = "sd",
                title = "A. URI Watershed Watch Chlorophyll")
 ggsave(here("figures/ww_chla_trends.jpg"), chla_gg[[1]], width = 7.5, height = 5.625, 
@@ -73,7 +74,8 @@ ww_np_rl <- np_gg[[4]]
 lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
 lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", 
-                             y = "Average Yearly Scaled Chlorophyll", x = "Year",
+                             expression(paste("Average Yearly Scaled Chlorophyll  ", 
+                                              italic(" a"))), x = "Year",
                              write = here("data/chla_data_lagos.csv"), 
                              error_bar = "sd",
                              title = "B. LAGOSNE Chlorophyll")
@@ -178,4 +180,20 @@ ww_map <- ggplot(st_geometry(ri)) +
 ww_map %>%  
   ggsave(here("figures/ww_map.jpg"), ., width = 7.5, height = 9.8,
          units = "in", dpi = 600)
+
+# Create supplemental dataset with data for all figures:
+figure_data <- chla_gg[[3]] %>%
+  mutate(source = "uriww", variable = "chla") %>%
+  rbind(mutate(temp_gg[[3]], source = "uriww", variable = "temp")) %>%
+  rbind(mutate(tn_gg[[3]], source = "uriww", variable = "total_n")) %>%
+  rbind(mutate(tp_gg[[3]], source = "uriww", variable = "total_p")) %>%
+  rbind(mutate(np_gg[[3]], source = "uriww", variable = "np_ratio")) %>%
+  rbind(mutate(lagos_chla_gg[[3]], source = "lagosne", variable = "chla")) %>%
[…]

### 2019-05-14 manuscript/manuscript.Rmd
updated figure caption to italics a
fixed schematic all the way down

@@
 
+As previously mentioned, both temperature and chlorophll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the nitrogen:phosphorus ratio are all relatively stable.
+These stable nutrient regimes may be partly explained by
+
 - Parameter by parameter discussion of trends
+- Local scale vs Regional/Global scale
+    - Why diffs?  
+        - LULC?
+        - Different trends
+- Other trends?
+    - monthly/seasonal
     - How do trends break down by trophic state?
-- Local scale vs Regional/Global scale
-- RI is downstream - Bryan will write
 
@@
 - How to stabilize trends?  
-More reductions in nutrients?
-Engineered solutions to slow warming?
-Look to individual lakes that buck the trend for suggestions (e.g.
-local landcover may provide a buffer)
+    - More reductions in nutrients?  
+    - Engineered solutions to slow warming? 
+    - Look to individual lakes that buck the trend for suggestions (e.g. local landcover may provide a buffer)
+    - stable nutrients doesn't mean they are low, just not getting worse
+    - interventions to reduce nutrients could offset the suggested increase caused by temperature
+        - e.g. freshwater mussels restoration
 
@@
 - Use of site specific z-scores
-    - Other approaches available, but this one is useful becuase it provided readily interpretable results, robust to variations in sampling stations (i.e.
+    - Other approaches available, but this one is useful because it provided readily interpretable results, robust to variations in sampling stations (i.e.
 added later in period, removed, etc.), and based off of approach with long history of use in exploring long term trends [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface]

### 2019-05-14 R/schematic.R

@@ -1,10 +1,13 @@
 library(readr)
 library(dplyr)
-schematic <- read_csv(here::here("data/schematic.csv"))
+library(here)
+
+schematic <- read_csv(here("data/schematic.csv"))
 
 station_year_mean <- schematic %>% 
   group_by(station, year) %>%
   summarize(station_year_mean = mean(meas)) %>%
+  round(2) %>%
   ungroup()
 
 station_year_z <- station_year_mean %>%
@@ -12,10 +15,13 @@ station_year_z <- station_year_mean %>%
   mutate(station_year_z = scale(station_year_mean),
          long_term_mean = mean(station_year_mean),
          long_term_sd = sd(station_year_mean)) %>%
+  round(2) %>%
   ungroup()
   
 mean_z <- station_year_z %>%
   group_by(year) %>%
   summarize(mean_z = mean(station_year_z),
             sd_z = sd(station_year_z),
-            n = n())
+            n = n()) %>%
+  round(2) %>%
+  ungroup()

### 2019-05-15 manuscript/manuscript.Rmd
sentence or two added to discussion

@@
 As previously mentioned, both temperature and chlorophll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the nitrogen:phosphorus ratio are all relatively stable.
-These stable nutrient regimes may be partly explained by
+These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [Need References Here]).
+Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
+For instance, controlling for nutrients is largely a local to watershed scale effort; whereas, warming lakes are driven by broader, climate patterns.
 
-- Parameter by parameter discussion of trends
-- Local scale vs Regional/Global scale
-    - Why diffs?  
-        - LULC?
-        - Different trends
 - Other trends?

### 2019-05-21 manuscript/manuscript.Rmd
one third of discussion!

@@
 
-As previously mentioned, both temperature and chlorophll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the nitrogen:phosphorus ratio are all relatively stable.
-These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [Need References Here]).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the nitrogen:phosphorus ratio are all relatively stable.
+These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [NEED REFERENCES HERE]).
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
-For instance, controlling for nutrients is largely a local to watershed scale effort; whereas, warming lakes are driven by broader, climate patterns.
+For instance, controlling for nutrients is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [NEED REFERENCES HERE].
+Similarly, warming lakes are driven by broader, climate patterns, yet waterbody specific factors such as percent impervious in the watershed and lake morphology can also impact temperature [NEED REFERENCES HERE].
+So, when examining differences in regional and state level trends those trends are the results of complex interacting drivers.
 
-- Other trends?
-    - monthly/seasonal
-    - How do trends break down by trophic state?
+In addition to the annualized trends we address with this study, there are other trends that may be of interest.
+First, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
+Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
+Second, are trends influenced by the current state of a given waterbody?
+For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trohpic status?
+These questions all warrant further investigation.
 
@@
 
-## Caveats
-
-- None yet...
-
 # Bibliography

### 2019-05-21 R/figures.R

@@ -73,8 +73,7 @@ ww_np_rl <- np_gg[[4]]
 
 lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
 
-lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", 
-                             expression(paste("Average Yearly Scaled Chlorophyll  ", 
+lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_scale", y = expression(paste("Average Yearly Scaled Chlorophyll  ", 
                                               italic(" a"))), x = "Year",
                              write = here("data/chla_data_lagos.csv"), 
                              error_bar = "sd",

### 2019-05-21 manuscript/manuscript.Rmd
bit more on discussion

@@
 Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
+Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
 Second, are trends influenced by the current state of a given waterbody?
 For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trohpic status?
-These questions all warrant further investigation.
+These questions all warrant further, careful investigation.
 
@@
     - Engineered solutions to slow warming? 
-    - Look to individual lakes that buck the trend for suggestions (e.g. local landcover may provide a buffer)
+    - Look to individual lakes that buck the trend for suggestions (e.g. local land cover may provide a buffer)
     - stable nutrients doesn't mean they are low, just not getting worse

### 2019-05-21 R/early_late_temp_compare.R

@@ -1,0 +1,18 @@
+source(here::here("R/functions.R"))
+
+ww_all <- read_csv(here("data/ww_all.csv"), 
+                   guess_max = 300000)
+ww_all_temp_month <- ww_all %>%
+  select(date = "Date", param = "Parameter", conc = "Concentration") %>%
+  mutate(month = month(date),
+         year = year(date),
+         early_late = case_when(year < 1996 & year >= 1993 ~
+                                  "early",
+                                year > 2013 ~
+                                  "late")) %>%
+  filter(param == "Temperature - 00011",
+         month == 5 | month == 10,
+         !is.na(early_late)) %>%
+  group_by(early_late, month) %>%
+  summarize(temp_mean = mean(conc, na.rm = TRUE)) %>%
+  arrange(month)

### 2019-05-24 manuscript/manuscript.Rmd
one more section till a draft ready to share with co-authors!

@@
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and reservoirs.
+Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and reservoirs.
 
-Examining long term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
+Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
 URI’s Watershed Watch is a citizen science project founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
@@
 Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-Another focus of this paper was to apply existing methodologies for examining long term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long term trends.
+Another focus of this paper was to apply existing methodologies for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
 Lastly, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOS project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
@@
 
-Average yearly scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
-Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`).
+Average yearly scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
 Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
@@
 shows yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
 
@@
 shows yearly averaged total nitrogen z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
 Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
@@
 shows yearly averaged total phosphorus z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
 The pattern is the same for the N:P ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of total nitrogen relative to the concentration of total phosphorus (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the total nitrogen:total phosphorus ratio (average z-score).
@@
 shows yearly averaged total nitrogen:total phosphorus ratio z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long term mean and red indicating an average above the long term mean.", display = "cite")`A.).
+Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
 Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].
@@
 One possible explanation for this is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
-Second,  scale does indeed matter when trying to identify long-term water quality trends as our analysis shows little increasing trend in broad scale (e.g.
-Northeastern and Mid-western United States) patterns of productivity, but do show clear increasing trends for just the state of Rhode Island.
-In the discussion that follows, we more closely address these trends, discuss broader implications of these results, and reiterate the value of the data analysis approach we used.
+Also, scale does indeed matter when trying to identify long-term water quality trends.
+Similar to the results of Oliver et al.
+[-@Oliver_2017, ]our analysis shows little increasing trend in broad scale (e.g.
+Northeastern and Mid-western United States) patterns of productivity.
+However, we do show clear increasing trends for just the state of Rhode Island.
+In the discussion that follows, we more closely address these trends, discuss management implications of these results, and reiterate the value of the data analysis approach we used.
 
@@
 For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trohpic status?
-These questions all warrant further, careful investigation.
+These questions are beyond the scope of this study, but all warrant further, careful investigation.
 
-## Broader Implications
+## Management Implications
 
-- Provides feedback on long term changes for managers and stakeholders
-- Season Length Changes
-- How to stabilize trends?  
-    - More reductions in nutrients?  
-    - Engineered solutions to slow warming? 
-    - Look to individual lakes that buck the trend for suggestions (e.g. local land cover may provide a buffer)
-    - stable nutrients doesn't mean they are low, just not getting worse
-    - interventions to reduce nutrients could offset the suggested increase caused by temperature
-        - e.g. freshwater mussels restoration
+There are several broader management implications from the results of our analysis and of examining long-term water quality trends in general.
+In particular, this analysis provides important feedback and identifies areas where additional information is required or a change in management approaches may be needed.
+First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term trends in the water quality.
+Specifically for our study, our results provide feedback to long time volunteer monitors and highlights the importance of volunteer monitoring programs.
+Second, with information on long-term trends it is possible to adapt management approaches to address undesirable trends.
+Our results show increasing productivity even though the general long-term nutrient trends have been stable.
+This result points to the need to reduce nutrients and attempt to counteract warming.
+
+There are several possible approaches that could be tried.
+First, nutrient load reductions may be possible through broader scale management of landscapes and watersheds.
+Green infrastructure approaches are one way to possibly achieve both goals [NEED REFERENCE HERE].
+Additionally, within-lake approaches may also be used.
+A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species.
+Some studies have shown reductions in both nutrients and algal biomass [NEED REFERENCES HERE].
 
@@
 
-- Why this is possible - Fantastic WW data
 - Use of site specific z-scores
     - Other approaches available, but this one is useful because it provided readily interpretable results, robust to variations in sampling stations (i.e.
-added later in period, removed, etc.), and based off of approach with long history of use in exploring long term trends [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface]
+added later in period, removed, etc.), and based off of approach with long history of use in exploring long-term trends [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface]
+
+- Why this is possible - Fantastic WW data
 

### 2019-05-24 manuscript/manuscript.Rmd
first round of writing done...  Lot's to fill in but my co-authors can do that...

@@
 Second, are trends influenced by the current state of a given waterbody?
-For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trohpic status?
+For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
 These questions are beyond the scope of this study, but all warrant further, careful investigation.
@@
 
-- Use of site specific z-scores
-    - Other approaches available, but this one is useful because it provided readily interpretable results, robust to variations in sampling stations (i.e.
-added later in period, removed, etc.), and based off of approach with long history of use in exploring long-term trends [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface]
-
-- Why this is possible - Fantastic WW data
+The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is novel as we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
+This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].
+We simply built on this methods and adapted it for use with long-term water quality trends.
+And while other methods are valid and robust (e.g.
+[@Oliver_2017]), we chose averaged site-specific z-scores as they, in our opinion, provide readily interpretable results, especially for communicating to less technical audiences.
+In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (i.e.
+added later in a time period or removed).
+Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.
+In other words, without the URIWW data and the commitment and participation of nearly XXX volunteers our analysis would have been impossible.
+Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in the analysis of long term environmental trends.
 

### 2019-05-24 manuscript/manuscript.Rmd
added URI WW numbers... Wow!

@@
 Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.
-In other words, without the URIWW data and the commitment and participation of nearly XXX volunteers our analysis would have been impossible.
+In other words, without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analysis would have been impossible.
 Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in the analysis of long term environmental trends.

### 2019-06-04 manuscript/manuscript.Rmd

--- 2019-05-24 manuscript/manuscript.Rmd
@@
 Temperature and chlorophyll *a* are increasing.
-Total nitrogen shows a weak increasing trend driven by low years in the early 1990s.
-Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
-Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.
+Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.
+Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.
+Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP all stable over time.
 In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
@@
 
-Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental scale water quality analysis.
+Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
 These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
@@
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality trends in Rhode Island lakes and reservoirs.
+Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.
 
 Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
-URI’s Watershed Watch is a citizen science project founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
+URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
 Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
@@
 Thus, the goals of this study were to look at approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
-First, what are the state-wide trends in total nitrogen, total phosphorus, nitrogen to phosphorus ratio (N:P), chlorophyll *a*, and lake temperature.
+First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
 Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
 Another focus of this paper was to apply existing methodologies for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
-Lastly, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOS project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
+Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
 
@@
 
-The study area for this analysis includes data from lakes and reservoirs in the state of Rhode Island that were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
+The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
 The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
 The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URI WW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
+URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
 Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
@@
 In particular we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, total nitrogen, total phosphorus, N:P, and chlorophyll *a*.
-For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for Total Nitrogen (TN), `r length(total_p_10yr_sites)` samples for Total Phosphorus (TP).
+For this analysis we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
+For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
@@
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
-Temperature data were not available, thus we only examine trends, using our analytical methods, for total nitrogen, total phosphorus, N:P, and chlorophyll *a* from the LAGOSNE database.
+Temperature data were not available, thus we only examine trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
 
@@
 There are many different methods for analyzing time series data for trends.
-One of the difficulties that is encountered when you have multiple sampling locations is how to identify a trend when you have variation at the scale of the sampling location as well as variation in when a sampling location might have been added to a dataset.
-For instance, if you have long-term data on water quality for several waterbodies, yet the cleaner waterbodies were sampled more frequently in early years then a simple comparison of raw-values over time might show a decrease in water quality.
-Thus, it is necessary to account for this type of site-specific variation.
-This is similar to how the long-term temperature trends are analyzed using temperature anomalies (e.g., [@jones1996calculating]).
-The general approach is to calculate site-specific deviations from a long-term mean over pre-determined reference period.
-This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change relative to the reference period.
+Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
+Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies (e.g., [@jones1996calculating]).
+The general approach is to calculate site-specific deviations from a long-term mean over a pre-determined reference period.
+This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change over the specified reference period.
 
@@
 
-Anomalies are very useful and informed the approach we chose to follow.
-However, using anomalies works best with a single measure (e.g temperature) or with multiple measurements that are on the same scale.
+Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature in Rhode Island lakes during the growing season will range from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
+For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
 To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
 The resultant value is commonly referred to as a z-score.
-We use these z-scores to examine each water quality variable for a trend over the time period of 1993 to 2016.
-Furthermore, since we are interested in water quality trends over time, we wanted to explore how each site was responding.
-Thus, our z-scores were calculated over the reference period, 1993-2016, for each site.
-We refer to this approach as the site-specific z-scores
+We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.
+Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.
+We refer to this approach as the site-specific z-scores.
 
@@
 
-Details for our approach of calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
+Methods for calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
 Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <fill in with zenodo link> [@zenodo_citation].
@@
 1.
-For each site, calculate the per site and per year mean values which results in a single mean value for each site and year.
-This step prevent bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].
+For each site, calculate the annual means, producing a single mean value for each site and year.
+This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].
 The per site means across years are assumed to be independent.
-2. Calculate the long-term reference mean and reference standard deviation for all per site and per year means.  This results in a single long-term mean and standard deviation for each of the sites.
-3. Calculate the z-score for each per site and per year mean by subtracting the reference mean and dividing by the reference standard deviation.
-4. Summarize the per site and per year z-scores by calculating the mean z-score per year.  The resultant values are analyzed for a trend over time.
+2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
+3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
+4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
 
-### Linear regression on yearly means
+### Linear regression on annual means
 
 Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
-We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, total nitrogen, and total phosphorus.
-For each parameter we fit a regression line to the z scores as a function of year and tested the null hypothesis that no trend existed (e.g.
+We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
+For each parameter we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
 $\beta$~1~ = 0).
@@
 Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local scale trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
+While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
 The analysis conducted by Oliver et al.
@@
 
-During the sampling period of 1990 to 2016 Rhode Island lakes and reservoirs average lake temperature was `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, average total nitrogen was `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, average N:P ratio was `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and average chlorophyll was `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the period of 1990 to 2016 Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 
-For lakes and reservoirs in the larger region represented by the LAGOSNE States,  average total nitrogen was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, average total phosphorus was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, average N:P ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and average chlorophyll was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
 
@@
 
-Average yearly scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
+Average annual scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
+Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
 Panel A.
-shows yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.
+Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.
 Panel B.
-shows yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
+Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
@@
 
-Average yearly trends for nutrients showed weaker or no trends over time.
-Total nitrogen did suggest a positive trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of total nitrogen (average z-score).
+Mean annual trends for nutrients were weaker or showed no trend over time.
+The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score).
 Panel A.
-shows yearly averaged total nitrogen z-scores from the URI Watershed Watch data.
+Yearly averaged TN z-scores from the URI Watershed Watch dataset.
 Panel B.
-shows yearly averaged total nitrogen z-scores from the LAGOSNE dataset.
+Yearly averaged TN z-scores from the LAGOSNE dataset.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
 Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
-Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of total phosphorus (average z-score).
+TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
 Panel A.
-shows yearly averaged total phosphorus z-scores from the URI Watershed Watch data.
+Yearly averaged TP z-scores from the URI Watershed Watch dataset.
 Panel B.
-shows yearly averaged total phosphorus z-scores from the LAGOSNE dataset.
+Yearly averaged TP z-scores from the LAGOSNE dataset.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-The pattern is the same for the N:P ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of total nitrogen relative to the concentration of total phosphorus (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the total nitrogen:total phosphorus ratio (average z-score).
+The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).
 Panel A.
-shows yearly averaged total nitrogen:total phosphorus ratio z-scores from the URI Watershed Watch data.
+Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.
 Panel B.
-shows yearly averaged total nitrogen:total phosphorus ratio z-scores from the LAGOSNE dataset.
+Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
@@
 
-In general, there was little to suggest broad regional trends.
-Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), total phosphorus shows a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), total nitrogen shows a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the total nitrogen:total phosphorus ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+In general, there was little to suggest broad regional changes.
+Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 Similar to the results of Oliver et al.
-[-@Oliver_2017, ]our analysis shows little increasing trend in broad scale (e.g.
-Northeastern and Mid-western United States) patterns of productivity.
-However, we do show clear increasing trends for just the state of Rhode Island.
-In the discussion that follows, we more closely address these trends, discuss management implications of these results, and reiterate the value of the data analysis approach we used.
+[-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g.
+Northeastern and Mid-western United States).
+However, at the local scale of the state of Rhode Island there is a clear increasing trend in productivity.
 
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the nitrogen:phosphorus ratio are all relatively stable.
-These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [NEED REFERENCES HERE]).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the TN:TP ratio are all relatively stable.
+These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
-For instance, controlling for nutrients is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [NEED REFERENCES HERE].
-Similarly, warming lakes are driven by broader, climate patterns, yet waterbody specific factors such as percent impervious in the watershed and lake morphology can also impact temperature [NEED REFERENCES HERE].
-So, when examining differences in regional and state level trends those trends are the results of complex interacting drivers.
+For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al.
+2002].
+Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007].
+Differences in regional and state level trends are driven by complex processes.
 
@@
 First, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
-Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
+Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
@@
 There are several broader management implications from the results of our analysis and of examining long-term water quality trends in general.
-In particular, this analysis provides important feedback and identifies areas where additional information is required or a change in management approaches may be needed.
-First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term trends in the water quality.
-Specifically for our study, our results provide feedback to long time volunteer monitors and highlights the importance of volunteer monitoring programs.
-Second, with information on long-term trends it is possible to adapt management approaches to address undesirable trends.
-Our results show increasing productivity even though the general long-term nutrient trends have been stable.
-This result points to the need to reduce nutrients and attempt to counteract warming.
+In particular, this analysis provides much needed information about the long-term effects of current nutrient control efforts and identifies areas where additional information is required or a change in management approaches may be needed.
+First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term water quality trends.
+Specifically for this study, the results provide feedback to long time volunteer monitors, highlighting the importance of volunteer monitoring programs.
+Second, with information on long-term trends it is possible to adapt management approaches to address areas of concern.
+Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a longer growing season and associated higher productivity.
 
-There are several possible approaches that could be tried.
-First, nutrient load reductions may be possible through broader scale management of landscapes and watersheds.
-Green infrastructure approaches are one way to possibly achieve both goals [NEED REFERENCE HERE].
+There are several possible approaches.
+First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
+Green infrastructure approaches are one way to possibly achieve both goals [see notes].
 Additionally, within-lake approaches may also be used.
 A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species.
-Some studies have shown reductions in both nutrients and algal biomass [NEED REFERENCES HERE].
+Some studies have shown reductions in both nutrients and algal biomass [see notes].
 
@@
 This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].
-We simply built on this methods and adapted it for use with long-term water quality trends.
+We simply built on these methods and adapted them for use with long-term water quality trends.
 And while other methods are valid and robust (e.g.
-[@Oliver_2017]), we chose averaged site-specific z-scores as they, in our opinion, provide readily interpretable results, especially for communicating to less technical audiences.
-In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (i.e.
-added later in a time period or removed).
+[@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
+In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).
 Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.
-In other words, without the URIWW data and the commitment and participation of nearly XXX volunteers our analysis would have been impossible.
-Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in the analysis of long term environmental trends.
+Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.
+Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
 

### 2019-06-10 manuscript/manuscript.Rmd
small edits to text

@@
 abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
-Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine its effectiveness requires long-term datasets.
-A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States, however, trends at smaller scales may be masked.
+Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine the effectiveness requires long-term datasets.
+A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at smaller scales may be masked.
 We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
@@
 Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.
-Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP all stable over time.
+Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP.
 In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
@@
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These contrasting studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales.
+These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.
+Let's discuss it).
+Maybe...
+These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.
 
+Attempts to explore these spatial dyanmics are available (my attempt at intro sentence for paragraph).
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
-These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
+These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.
+Again, let's chat).
 Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
@@
 Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
+(thought- Maybe this paragraph just needs a introduction sentence.
+)
 
-Previous studies using regional data from the northeastern and midwestern United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper) United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
 Macro-scale (i.e.
@@
 This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the Northeastern and Midwestern US [@filstrup2014regional; @filstrup2018evidence].
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
@@
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
-Additionally, the fact that volunteer monitors are used has allowed for this program to persist much longer than agency run programs that are often prone to uncertain funding.
 Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
@@
 
-Thus, the goals of this study were to look at approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
+Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
 First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
-Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-Another focus of this paper was to apply existing methodologies for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
+Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states.
+Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
 Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
@@
 
-For this study we combined a long-term dataset on water quality of lakes in Rhode Island, with a trend analysis based on centered and scaled water quality values (i.e.
+For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e.
 z-scores) to find increasing or decreasing annual water quality trends.
@@
 The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data, and is an integral part of the state's environmental data collection strategy.
-Data QA/QC is of paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data (^I don't know what this means- is a common knowledge I'm missing?
+If so, ignore), and is an integral part of the state's environmental data collection strategy.
+Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
@@
 
-In particular we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-For each of these parameters we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
+In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
+For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
@@
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
-Temperature data were not available, thus we only examine trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
+Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
 
@@
 For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
-To standardize these values we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
+To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
 The resultant value is commonly referred to as a z-score.
@@
 We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
-For each parameter we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
+For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
 $\beta$~1~ = 0).
@@
 
-During the period of 1990 to 2016 Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the period of 1990 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 ![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
+
+- Reminder- need to add "a" to figure titles
 
@@
 
-In general, there was little to suggest broad regional changes.
+In general, there was little evidence to suggest broad regional changes.
 Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
@@
 neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
-One possible explanation for this is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
+One possible explanation is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
 Also, scale does indeed matter when trying to identify long-term water quality trends.
@@
 Northeastern and Mid-western United States).
-However, at the local scale of the state of Rhode Island there is a clear increasing trend in productivity.
+However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.
 
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs, while total nutrients and the TN:TP ratio are all relatively stable.
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
 These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
@@
 In addition to the annualized trends we address with this study, there are other trends that may be of interest.
-First, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
+For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
 Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
-Second, are trends influenced by the current state of a given waterbody?
+Additionally, are trends influenced by the current state of a given waterbody?
 For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
@@
 Specifically for this study, the results provide feedback to long time volunteer monitors, highlighting the importance of volunteer monitoring programs.
-Second, with information on long-term trends it is possible to adapt management approaches to address areas of concern.
+Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
 Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a longer growing season and associated higher productivity.
@@
 We simply built on these methods and adapted them for use with long-term water quality trends.
-And while other methods are valid and robust (e.g.
+While other methods are valid and robust (e.g.
 [@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.

### 2019-07-23 manuscript/manuscript.Rmd
Bryan's edits

@@
 Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine the effectiveness requires long-term datasets.
-A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at smaller scales may be masked.
+A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at finer scales may be masked.
 We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
@@
 Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes].
-Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
+Increasing nutrient inputs, particularly of nitrogen (N) and phosphorus (P), derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
 This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.
-Let's discuss it).
-Maybe...
-These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.
+These studies indicate that complex relationships between N, P, and chlorophyll *a* exist and vary spatially; long-term studies are needed to identify trends at local, regional, and national scales.
 
-Attempts to explore these spatial dyanmics are available (my attempt at intro sentence for paragraph).
+Attempts to explore these spatial dynamics are available (my attempt at intro sentence for paragraph).
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
@@
 (thought- Maybe this paragraph just needs a introduction sentence.
-)
+) (B: why not just remove this paragraph completely)
 
@@
 
-Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
+Examining long-term trends in Rhode Island lakes is possible because of data gathered by University of Rhode Island’s Watershed Watch (URIWW).
 URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
-Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
+Volunteer monitors are trained to collect *in situ* data as well as whole water samples during the growing season.
 These efforts have been ongoing in some waterbodies since 1988.
@@
 
-Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
+The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.
 First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
@@
 # Methods
-
-For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e.
-z-scores) to find increasing or decreasing annual water quality trends.
-Details are outlined below.
 
@@
 
-In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
+We analyzed a subset of the URI data that were collected from 1990 to 2016 during the months of May to October at a depth of 2m or less.
+Only lakes with at least 10 years of data were included.
+For this study, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
+The final dataset included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
 
-Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+Prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
 The authors found little change in water quality trends across this region [@Oliver_2017] .
@@
 The resultant value is commonly referred to as a z-score.
-We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.
+We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016 (B: 1993 or 1990?).
 Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.
@@
 The per site means across years are assumed to be independent.
-2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
+2.
+Calculate the long-term reference mean and reference standard deviation for all sites.
+This results in a single long-term mean and standard deviation for each of the sites.
+(B: i'm confused-the first sentence suggests you get single values for the mean and SD across all sites but the second sentence says the a single mean and SD are calculated for each site.
+If it is the second then the index period will vary by site and this needs to be explained.)
 3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
@@
 
-Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
-The analysis conducted by Oliver et al.
-[-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al.
-[-@Oliver_2017] but using the trend analysis approach outlined above.
+Oliver et al (2017) have shown relatively stable water quality in the lakes of the Northeastern United States; the University of Rhode Island's Watershed Watch data were included in this regional study.
+However, we hypothesized in the case of Rhode Island regional trends were masking local trends.
+Therefore we decided to reanalyze the the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the approach outlined above.
 
@@
 neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
-One possible explanation is the long-term warming of Rhode Island lakes and reservoirs as also indicated by our analysis.
+Over the same period we also demonstrate long-term warming of RI lakes and reservoirs.
+This suggests that the observered increase in producticity, as measured by chlorophyll *a* is a direct result of warming waters and not a response to changes in nutrient condition.
 Also, scale does indeed matter when trying to identify long-term water quality trends.
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
-These stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are relatively stable.
+Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
@@
 Additionally, are trends influenced by the current state of a given waterbody?
-For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
+For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
 These questions are beyond the scope of this study, but all warrant further, careful investigation.
@@
 Some studies have shown reductions in both nutrients and algal biomass [see notes].
+(B: where did this come from?
+Are we sure mussels are this important?)
 
@@
 
-The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is novel as we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
+The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique; we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
 This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].

### 2019-07-25 manuscript/manuscript.Rmd
Mostly ready except for few spots in discussion.

--- 2019-06-10 manuscript/manuscript.Rmd
@@
 csl: ecology.csl
-abstract: 'Addressing anthropogenic impacts on aquatic ecosystems has long been the focus of lake management.
-Controlling phosphorus and nitrogen can mitigate the impacts of eutrophication, but to determine the effectiveness requires long-term datasets.
-A recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE), found stable water quality in the Northeast and Midwestern regions of the United States; however, trends at smaller scales may be masked.
-We address this by analyzing the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management.
+Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets.
+Recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE) found stable water quality in the Northeastern and Midwestern United States, however, sub-regional trends may be masked.
+We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
 URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
-These data, included in LAGOSNE, allow for comparison of water quality trends at regional and state extents.
-We assess for trends with z-scores (i.e.
-scaled anomalies) calculated on a per-station basis and examine yearly averages.
-Temperature and chlorophyll *a* are increasing.
-Total nitrogen (TN) shows a weak increasing trend driven by low years in the early 1990s.
-Total phosphorus (TP) and the total nitrogen:total phosphorus ratio (TN:TP) are stable.
-Applying the site-specific z-score approach to LAGOSNE found similar trends to prior studies with chlorophyll *a*, TN, TP, and TN:TP.
+The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and state extents, respectively.
+We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e.
+scaled anomalies) calculated on a per-station basis.
+In Rhode Island, temperature and chlorophyll a increased from 1993 to 2016.
+Total nitrogen shows a weak increase driven by low years in the early 1990s.
+Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
+At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time.
 In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
 Although not causal, this analysis suggests an association between lake temperature and productivity.
-Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.
-'
+Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
 ---
@@
 Because of their position in the landscape, lakes can function as integrators and sentinels for these anthropogenic effects [@williamson2008lakes; @schindler2009lakes].
-Increasing nutrient inputs, particularly of N and P, derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
+Increasing nutrient inputs, particularly of nitrogen (N) and phosphorus (P), derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
 This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These studies indicate that spatial differences and relationships between N, P, and chlorophyll *a* exist and that long-term studies are needed to identify trends at local, regional, and national scales (I'm confused about this sentence- not sure how to edit.
-Let's discuss it).
-Maybe...
-These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex; long-term studies are needed to identify trends at local, regional, and national scales.
+These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex.
+Thus, long-term data are needed to identify trends at local, regional, and national scales.
 
-Attempts to explore these spatial dyanmics are available (my attempt at intro sentence for paragraph).
+Lake datasets that cover longer time periods and broader spatial scales are now becoming available.
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
-These analyses can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.
+These data allow for analyses that can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.
 Again, let's chat).
-Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases[@stoddard2016continental].
+Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
 Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
@@
 Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
-(thought- Maybe this paragraph just needs a introduction sentence.
-)
 
-Previous studies using regional data from the Northeastern and Midwestern (I don't actually think these should be capitalized, but they are elsewhere in the paper) United States have investigated spatial and temporal water quality trends and have shown differences based on scale.
+Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale.
 Macro-scale (i.e.
@@
 This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the Northeastern and Midwestern US [@filstrup2014regional; @filstrup2018evidence].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
 Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
@@
 Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
-URI’s Watershed Watch is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of non-expert volunteer monitors.
+URIWW is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
 Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
-These efforts have been ongoing in some waterbodies since 1988.
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
 Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
-[@Oliver_2017], but also allowed us to examine the long-term trends in Rhode Island.
+[@Oliver_2017], but also allowed us to examine the long-term trends specifically in Rhode Island.
 
@@
 First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
-Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states.
+Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
 Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
@@
 
-The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch (URIWW) program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
+The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
 The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
 The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
-URIWW now provides more than 90% of Rhode Island's lake multi-year baseline data (^I don't know what this means- is a common knowledge I'm missing?
-If so, ignore), and is an integral part of the state's environmental data collection strategy.
+URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy.
 Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
-Water samples are collected three times per season (May through October) to be analyzed in the EPA-certified laboratory.
+Water samples are collected three times per season (May through October) to be analyzed for nutrients in the EPA-certified laboratory.
 
@@
 
-In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1990 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 timeframe.
+For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 time frame.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
@@
 
-Additionally, prior studies have looked at water quality trends across a larger region of the Northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+Additionally, prior studies have looked at water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
 The authors found little change in water quality trends across this region [@Oliver_2017] .
@@
 The general approach is to calculate site-specific deviations from a long-term mean over a pre-determined reference period.
-This allows all sites to be shifted to a common base line and the deviations, or anomalies, show change over the specified reference period.
+This allows all sites to be shifted to a common baseline and the deviations, or anomalies, show change over the specified reference period.
 
@@
 Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.
-We refer to this approach as the site-specific z-scores.
+We refer to this approach as "site-specific z-scores".
 
@@
 2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
-3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
+3. Calculate the z-score for each annual mean at each site by subtracting the annual and reference means and dividing by the reference standard deviation.
 4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
@@
 [-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al.
-[-@Oliver_2017] but using the trend analysis approach outlined above.
+[-@Oliver_2017] but using the site-specific z-score and trend analysis approach outlined above.
 
@@
 
-During the period of 1990 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
 
@@
 ![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
-
-- Reminder- need to add "a" to figure titles
 
@@
 [-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g.
-Northeastern and Mid-western United States).
+northeastern and midwestern United States).
 However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.
@@
 In addition to the annualized trends we address with this study, there are other trends that may be of interest.
-For example, what are the trends for water quality at finer temporal scales such as monthly or seasonal trends?
-Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to greater increases in temperature earlier and later in the growing season and suggests a lengthening of the growing season.
+For example, trends for water quality at finer temporal scales such as monthly or seasonal trends may be different than the annual trends we analyzed.
+Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season.
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
-Additionally, are trends influenced by the current state of a given waterbody?
+Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the productivity changes in that lake.
 For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
@@
 Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
-Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a longer growing season and associated higher productivity.
+Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a possibly longer growing season and associated higher productivity.
 
-There are several possible approaches.
+There are several potential approaches.
 First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.

### 2019-07-25 R/figures.R

@@ -49,7 +49,7 @@ temp_gg <- ww_data %>%
   wq_trend_gg("temp", yvar = "measurement_scale", 
               y = "Average Yearly Scaled Temperature", x = "Year",
               write = here("data/temp_data.csv"), error_bar = "sd",
-              title = "")
+              title = "URI Watershed Watch Temperature")
 ggsave(here("figures/ww_temp_trends.jpg"), temp_gg[[1]], width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
 

### 2019-07-29 manuscript/manuscript.Rmd

@@
 csl: ecology.csl
+
 abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management.
@@
 
-Thus, the goals of this study were to examine approximately 25 years of lake and reservoir data in Rhode Island and answer two questions.
+The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.
 First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
-Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
+Second, are water quality trends in Rhode Island similar to regional trends in the Northeastern United states.
 Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
@@
 The per site means across years are assumed to be independent.
-2. Calculate the long-term reference mean and reference standard deviation for all sites. This results in a single long-term mean and standard deviation for each of the sites.
-3. Calculate the z-score for each annual mean at each site by subtracting the annual and reference means and dividing by the reference standard deviation.
+2. Calculate the long-term reference mean and reference standard deviation for each site. This results in a single long-term mean and standard deviation for each of the sites.
+3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
 4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
@@
 
-Prior studies have shown relatively stable water quality in the lakes of the Northeastern United Sates [@Oliver_2017].
-While the University of Rhode Island's Watershed Watch data were included in that regional study,  we were curious if regional trends were masking local trends in Rhode Island and thus, wanted to compare the trends at the regional scale to the trends at the state scale.
-The analysis conducted by Oliver et al.
-[-@Oliver_2017] is a robust approach; however, to make direct comparisons between Rhode Island and the region,  we re-analyzed the same dataset used by Oliver et al.
-[-@Oliver_2017] but using the site-specific z-score and trend analysis approach outlined above.
+Oliver et al (2017) showed relatively stable water quality in the lakes of the Northeastern United States; the University of Rhode Island's Watershed Watch data were included in this regional study.
+However, we hypothesized in the case of Rhode Island regional trends were masking local trends.
+Therefore we decided to reanalyze the the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the approach outlined above.
 
@@
 [-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g.
-northeastern and midwestern United States).
+Northeastern and Mid-western United States).
 However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.
@@
 
-There are several potential approaches.
+There are several possible approaches.
 First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.

### 2019-07-29 manuscript/manuscript.Rmd
merged bryan and stephen, after much anguish (caused by me).

@@
 Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets.
-Recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE) found stable water quality in the Northeastern and Midwestern United States, however, sub-regional trends may be masked.
+Recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE) found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be masked.
 We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
@@
 # Methods
+
+For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e.
+z-scores) to find increasing or decreasing annual water quality trends.
+Details are outlined below.
 
@@
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
-Water samples are collected three times per season (May through October) to be analyzed for nutrients in the EPA-certified laboratory.
+Water samples are collected three times per season (May through October) to be analyzed for nutrients.
 
@@
 2. Calculate the long-term reference mean and reference standard deviation for each site. This results in a single long-term mean and standard deviation for each of the sites.
-3. Calculate the z-score for each annual mean at each site by subtracting the reference mean and dividing by the reference standard deviation.
+3. Calculate the z-score for each annual mean at each site by subtracting the annual and reference means and dividing by the reference standard deviation.
 4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
@@
 
-Oliver et al (2017) showed relatively stable water quality in the lakes of the Northeastern United States; the University of Rhode Island's Watershed Watch data were included in this regional study.
-However, we hypothesized in the case of Rhode Island regional trends were masking local trends.
-Therefore we decided to reanalyze the the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the approach outlined above.
+Oliver et al.
+[-@Oliver_2017] showed relatively stable water quality in the lakes of the northeastern United States.
+While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking local trends.
+Therefore, we decided to reanalyze the the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific z-score and trend analysis  approach outlined above.
 
@@
 
-Average annual scaled temperature in lakes and reservoirs appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
+Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
@@
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-Since 1995, the yearly trend is shows much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
-TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
+Since 1995, the yearly trend shows a much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
+TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tp_rl$slope,3)` , p-value: `r round(ww_tp_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
 Panel A.
@@
 In general, there was little evidence to suggest broad regional changes.
-Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP ratio was also flat (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 Over the same period we also demonstrate long-term warming of RI lakes and reservoirs.
-This suggests that the observered increase in producticity, as measured by chlorophyll *a* is a direct result of warming waters and not a response to changes in nutrient condition.
+This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.
 Also, scale does indeed matter when trying to identify long-term water quality trends.
 Similar to the results of Oliver et al.
-[-@Oliver_2017, ] our analysis shows little increasing trend in productivity at the regional scale (e.g.
-Northeastern and Mid-western United States).
+[-@Oliver_2017] our analysis shows little increasing trend in productivity at the regional scale (e.g.
+northeastern and mid-western United States).
 However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are relatively stable.
-Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide), ]).
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
+Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide)]).
+However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
@@
 Some studies have shown reductions in both nutrients and algal biomass [see notes].
-(B: where did this come from?
-Are we sure mussels are this important?)
 
@@
 
-The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique; we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
-This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global, hansen2010global, @jones1996calculating, @jones1999surface].
+The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
+This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].
 We simply built on these methods and adapted them for use with long-term water quality trends.
-While other methods are valid and robust (e.g.
-[@Oliver_2017]), we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
+While other methods are valid and robust [e.g.
+@Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
 In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).

### 2019-07-29 manuscript/manuscript.Rmd
manually merged qs_refs.  Getting closer to review ready!

@@
 As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
-Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings (e.g., [EQIP at USDA (QK will provide)]).
+Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings [e.g.
+@].
 However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.
 Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
-For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [Boyer et al.
-2002].
-Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [Nelson & Palmer, 2007].
+For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
+Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream].
 Differences in regional and state level trends are driven by complex processes.
@@
 First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
-Green infrastructure approaches are one way to possibly achieve both goals [see notes].
+Green infrastructure approaches are one way to possibly achieve both goals [e.g.
+@].
 Additionally, within-lake approaches may also be used.
 A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species.
-Some studies have shown reductions in both nutrients and algal biomass [see notes].
+Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
 

### 2019-07-30 manuscript/manuscript.Rmd
really close for STICS, need a few citations.  See issues on GitHub.

@@
 csl: ecology.csl
-
 abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management.
@@
 ```{r analysis, include=FALSE}
-#got to figure out path stuff before render.
 source(here::here("R/data_available.R"), encoding = "UTF-8") 
@@
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
-These data allow for analyses that can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing] (I'm not sure how this NLA fits into this.
-Again, let's chat).
+These data allow for analyses that can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
 Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
@@
 Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
-Despite these challenges, it is important to study at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
+Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
 
@@
 Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
-[@Oliver_2017], but also allowed us to examine the long-term trends specifically in Rhode Island.
+[-@Oliver_2017], but also allowed us to examine the long-term trends specifically in Rhode Island.
 
@@
 Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
-Another focus of this paper was to apply existing methods for examining long-term climate records (e.g., [@jones1996calculating]) to water quality data in order to examine long-term trends.
+Another focus of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
 Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
@@
 neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
-Over the same period we also demonstrate long-term warming of RI lakes and reservoirs.
+Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
 This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.
@@
 Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings [e.g.
-@].
+@NEED_CITATION].
 However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.
@@
 Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream].
-Differences in regional and state level trends are driven by complex processes.
+In short, differences in regional and state level trends are driven by complex and multi-scale processes.
 
@@
 For example, trends for water quality at finer temporal scales such as monthly or seasonal trends may be different than the annual trends we analyzed.
-Anecdotal evidence in Rhode Island [SHOULD REFERENCE THIS SOMEHOW - let's ask Elizabeth] points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season.
+Anecdotal evidence in Rhode Island points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season.
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
@@
 Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
-Our results show increasing productivity even though the general long-term nutrient trends have been stable, suggesting the need to further reduce nutrients to compensate for warmer water temperatures, a possibly longer growing season and associated higher productivity.
+Our results show increasing productivity even though the general long-term nutrient trends have been stable.
+This suggests the need to further reduce nutrients to compensate for warmer water temperatures and possible longer growing seasons.
 
-There are several possible approaches.
+There are several possible approaches to further reduce nutrient loads.
 First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
 Green infrastructure approaches are one way to possibly achieve both goals [e.g.
-@].
+@NEED_CITATION].
 Additionally, within-lake approaches may also be used.
@@
 
-The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find few examples of using it specifically for water quality trends [NEED TO DO THIS LIT SEARCH].
+The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find few examples of using it specifically for water quality trends [@NEED_CITATION].
 This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].
-We simply built on these methods and adapted them for use with long-term water quality trends.
+We built on these methods and adapted them for use with long-term water quality trends.
 While other methods are valid and robust [e.g.

### 2019-08-08 manuscript/manuscript.Rmd
Very close to internal review draft!

@@
 As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
-Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings [e.g.
-@NEED_CITATION].
+Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent.
 However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.
@@
 
-There are several possible approaches to further reduce nutrient loads.
+There are several possible approaches to further reduce nutrient loads [@yang2018nutrients].
 First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
-Green infrastructure approaches are one way to possibly achieve both goals [e.g.
-@NEED_CITATION].
+Green infrastructure approaches are one way to possibly achieve both goals [@pennino2016watershed; @reisinger2019changes].
 Additionally, within-lake approaches may also be used.
-A promising approach is the restoration of freshwater mussels to waterbodies that historically had those species.
+Another approach is the restoration of freshwater mussels to waterbodies that historically had those species.
 Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
@@
 
-The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find few examples of using it specifically for water quality trends [@NEED_CITATION].
+The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.
 This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].

### 2019-08-08 manuscript/manuscript.Rmd
ready for technical reviewers!

@@
 The authors found little change in water quality trends across this region [@Oliver_2017] .
-We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by [@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
+We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
+[-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
 Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
@@
 For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
-Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies (e.g., [@jones1996calculating]).
+Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
 The general approach is to calculate site-specific deviations from a long-term mean over a pre-determined reference period.
@@
 For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
-To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period (e.g., [@jones1996calculating]).
+To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].
 The resultant value is commonly referred to as a z-score.
@@
 
-![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
-
-![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends.jpg)
-
 ```{r no_early_tn, echo=FALSE}
@@
 Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].
+
+![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
+
+![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends.jpg)
 

### 2019-09-26 manuscript/manuscript.Rmd
started on response to internal reviews

@@
 ---
-title: "Increasing Productivity Amid Stable Nutrient Regimes in Rhode Island Lakes and Reservoirs"
+title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
 author:
@@
 At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time.
-In short, productivity in Rhode Island lakes and reservoirs is increasing, in spite of stable nutrient regimes.
-Although not causal, this analysis suggests an association between lake temperature and productivity.
+In short, algla biomass, as measured by cholorphyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
+Although not causal, this analysis suggests an association between lake temperature and primary production.
 Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
@@
 Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
-neither increasing nor decreasing over time), increases in productivity, as measured by chlorophyll *a*, occur.
+neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, occur.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
@@
 Similar to the results of Oliver et al.
-[-@Oliver_2017] our analysis shows little increasing trend in productivity at the regional scale (e.g.
+[-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g.
 northeastern and mid-western United States).
-However, at the local scale of the state of Rhode Island, there is a clear increasing trend in productivity.
+However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.
 
@@
 Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent.
-However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in productivity.
-Additionally, these results point to the fact that productivity is driven by processes operating at different scales.
+However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*  Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.
 For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
@@
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
-Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the productivity changes in that lake.
+Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.
 For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
@@
 Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
-Our results show increasing productivity even though the general long-term nutrient trends have been stable.
+Our results show increasing chlorophyll *a* even though the general long-term nutrient trends have been stable.
 This suggests the need to further reduce nutrients to compensate for warmer water temperatures and possible longer growing seasons.

### 2019-09-26 manuscript/manuscript.Rmd
more review edits

@@
 
-### Linear regression on annual means
+### Linear regression on annual mean z-scores
 
@@
 $\beta$~1~ = 0).
+Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].
+Our interpretation of the trends attempts to follow this advice and we assess trends with the information provided by the magnitude of the slopes, the p-values, and our understanding of the processes involved.
 

### 2019-10-01 manuscript/manuscript.Rmd
some progress on edits...

@@
 Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets.
-Recent analysis of the LAke multi-scaled GeOSpatial and temporal database (LAGOSNE) found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be masked.
+Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
 We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
@@
 Another focus of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAGOSNE project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
+Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
 
@@
 Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
-neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, occur.
+neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.

### 2019-10-04 manuscript/manuscript.Rmd
hagy response done, minus lab methods (still waiting on deets)

@@
 Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
-These data allow for analyses that can be used for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
+These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
 Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
 Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
-These trophic state models indicate that landscape variables (i.e.
+These trophic state models indicate that landscape variables (e.g.
 ecoregion, elevation, and latitude) are important and that regional trends exist.
 Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
-Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may not be evident when studying individual lakes [@cheruvelil2013multi; @lottig2014long].
+Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may or may not be present in individual lakes [@cheruvelil2013multi; @lottig2014long].
 
@@
 On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
-Thus, it was surprising when stasis was reported over a 25 year period for these regions [@Oliver_2017].
+Thus, it was surprising when little change was reported over a 25 year period for these regions [@Oliver_2017].
 Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.
@@
 Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
-URIWW is a citizen science program founded in the late 1980's that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
-Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season.
+URIWW is a citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
+Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g.
+May through October).
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
@@
 Additionally, prior studies have looked at water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
-The authors found little change in water quality trends across this region [@Oliver_2017] .
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
@@
 
-Anomaly analysis works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.
+Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
@@
 Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
-neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.
+neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
@@
 As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
-Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent.
+Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].
 However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*  Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.

### 2019-10-04 manuscript/manuscript.Rmd
working through nates comments.

@@
 Increasing nutrient inputs, particularly of nitrogen (N) and phosphorus (P), derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
-This eutrophication suggests an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+This eutrophication leads to an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
-These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially complex.
+These studies indicate that relationships between N, P, and chlorophyll *a* exist and these relationships are spatially and temporally complex.
 Thus, long-term data are needed to identify trends at local, regional, and national scales.
@@
 These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
-Studying trends across large spatial scales can evaluate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
+Studying trends across large spatial scales can illustrate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
 Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
@@
 subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
-This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the individual nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary among regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
-Thus, it was surprising when little change was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if those trends were also present in water quality in Rhode Island lakes and reservoirs.
+This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake].
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
+Thus, it was surprising when little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
+Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if the lack of trends was also present in water quality in Rhode Island lakes and reservoirs.
 
@@
 May through October).
-These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing].
-Here specifically, URIWW data contributed to not only the larger regional study by Oliver et al.
-[-@Oliver_2017], but also allowed us to examine the long-term trends specifically in Rhode Island.
+These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing, @Oliver_2017].
+URIWW data contributed to the larger regional study by Oliver et al.
+[-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.
 
@@
 Another focus of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-Finally, this analysis has also been done using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
+Finally, we conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
 
@@
 The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
-The URIWW is a volunteer monitoring program that has been collecting water quality data from Rhode Island lakes and reservoirs for over 25 years.
-The program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
+The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
 URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy.
@@
 
+For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
 In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
 For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 time frame.
-The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs which had approximately `r length(temp_10yr_sites)` samples for temperature, `r length(chla_10yr_sites)` samples for chlorophyll *a*, `r length(total_n_10yr_sites)` samples for TN, `r length(total_p_10yr_sites)` samples for TP.
+The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.
+Of these sites, our filtered dataset had approximately `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
 
-Additionally, prior studies have looked at water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+Prior studies have modeled water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
@@
 Oliver et al.
-[-@Oliver_2017] showed relatively stable water quality in the lakes of the northeastern United States.
+[-@Oliver_2017] used hierarchical linear models and showed relatively stable water quality in the lakes of the northeastern United States.
 While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking local trends.
@@
 As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
+While TN showed a weak positive trend, that trend was largely driven by the unusually low years for TN in 1993 and 1994.
+With those removed the positive trends weakens considerably.
+The general picture in Rhode Island appears to be one of little to no change in phosphorus, a very weak positive trend in nitrogen and little to no change in the TN:TP ratio.
+Furthermore, it has been shown that productivity in freshwater systems is likely a funciton of both phosphorus and nitrogen [@paerl2016takes].
+Thus, the increasing chlorophyll *a* in the face of stable TN:TP ratio suggests that the increase is being driven by something other than nutrients.
+We interpret these results as relative stability in nutrients in Rhode Island lakes and reservoirs.
+
 Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].
-However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*  Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.
+However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*.
+NEED TO ADD IN SENTENCE OR TWO ABOUT WHAT COULD BE DRIVING INCRESE (WARMING OR OTHER STUFF WE DIDN"T LOOK AT).
+Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.
 For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].

### 2019-10-17 manuscript/manuscript.Rmd
nates edits

@@
 Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets.
-Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
+Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
 We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
@@
 
-Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and reservoirs.
+Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs.
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
@@
 This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.
-Also, scale does indeed matter when trying to identify long-term water quality trends.
+Also, geographic extent does indeed matter when trying to identify long-term water quality trends.
 Similar to the results of Oliver et al.
@@
 However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*.
-NEED TO ADD IN SENTENCE OR TWO ABOUT WHAT COULD BE DRIVING INCRESE (WARMING OR OTHER STUFF WE DIDN"T LOOK AT).
 Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.
 For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
-Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as catchment percent impervious and lake morphology can also impact temperature [@nelson2007stream].
+Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream].
 In short, differences in regional and state level trends are driven by complex and multi-scale processes.
 
-In addition to the annualized trends we address with this study, there are other trends that may be of interest.
+In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.
 For example, trends for water quality at finer temporal scales such as monthly or seasonal trends may be different than the annual trends we analyzed.
@@
 Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.
-For instance, are oligotrophic lakes showing stronger trends than eutorphic lakes or are all lakes showing similar trends regardless of current trophic status?
+For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
+Lastly, changes in rainfall, extreme weather events, or other climate mediated factors can also be playing a role in increasing chlorophyll in Rhode Island lakes and reservoirs.
 These questions are beyond the scope of this study, but all warrant further, careful investigation.
@@
 Another approach is the restoration of freshwater mussels to waterbodies that historically had those species.
-Some studies have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
+Some studies using freshwater mussels have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
 
@@
 
-The analysis approach we used here, site-specific z-scores, is not a novel method, but using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.
-This approach, and in particular non-scaled anomalies, does have a long history in the analysis of trends in climate [@hansen2006global; hansen2010global; @jones1996calculating; @jones1999surface].
+The analysis approach we used here, site-specific z-scores, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface].
+However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.
 We built on these methods and adapted them for use with long-term water quality trends.

### 2019-10-17 manuscript/manuscript.Rmd
Anne's edits done

@@
 These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
-Studying trends across large spatial scales can illustrate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
-Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is predictive of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
+Studying temporal trends across large spatial scales can illustrate the effects of eutrophication such as the degradation of oligotrophic systems as P increases [@stoddard2016continental].
+Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is indicative of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
 These trophic state models indicate that landscape variables (e.g.
@@
 In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1990 to 2016 time frame.
+For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.
@@
 We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.
-Furthermore, since we are interested in water quality trends over time at individual sites, z-scores were calculated over the reference period, 1993-2016, for each site.
+Furthermore, since we are interested in water quality trends over time at individual sites (i.e.
+lakes), z-scores were calculated over the reference period, 1993-2016, for each site.
 We refer to this approach as "site-specific z-scores".
@@
 
-During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1990 to 2016.")`).
+During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
 
@@
 
-For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1990 to 2016.")`).
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
 
@@
 Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].
-However, in many lakes there are still likely sufficient nutrients present, as a result of eutrophication, to allow for increases in chlorophyll *a*.
+However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*.
 Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.

### 2019-10-17 manuscript/manuscript.Rmd
Elizabeth's comments plus sig digits

@@
 Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
-For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and dissolved oxygen.
+For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen.
 Water samples are collected three times per season (May through October) to be analyzed for nutrients.
@@
 ```{r ww_params, echo=FALSE, results="asis"}
-knitr::kable(ww_param_ranges, format = "markdown")
+pander::pander(ww_param_ranges[,c(-3,-6)], style = "rmarkdown")
 ```
@@
 ```{r lagos_params, echo=FALSE, results="asis"}
-knitr::kable(lagos_param_ranges, format = "markdown")
+pander::pander(lagos_param_ranges[,c(-3,-6)], style = "rmarkdown")
 ```

### 2019-10-17 R/data_clean.R

@@ -92,7 +92,22 @@ ww_lake_trend_data <- ww_all %>%
   summarize(mn_measurement = mean(measurement, na.rm = TRUE))%>%
   ungroup() 
 
-# Calculate and add n:p
+# Enforce sig digits
+
+ww_lake_trend_data <- ww_lake_trend_data %>%
+  mutate(mn_measurement = case_when(param == "temp"  ~ 
+                                      round(mn_measurement, 1),
+                                    param == "chla" ~
+                                      round(mn_measurement, 1),
+                                    param == "total_p" ~
+                                      round(mn_measurement, 0),
+                                    param == "total_n" ~
+                                      round(mn_measurement/5, 0)*5,
+                                    TRUE ~ mn_measurement))
+  
+
+# Calculate and add n:p 
+# Need to make molar conversions here...
 ww_np <- ww_lake_trend_data %>%
   filter(param == "total_n" | param == "total_p") %>% 
   unique() %>% #repeats from somewhere...
@@ -118,13 +133,53 @@ ww_lake_trend_data <- ww_lake_trend_data %>%
          lt_mean = mean(station_year_mean),
          lt_sd = sd(station_year_mean),
          lt_n = n()) %>%
+  ungroup() %>%
   left_join(ww_sites) %>%
   filter(WB_Type == "Lake or Pond" | WB_Type == "Reservoir") %>%
   select(station_name:lt_n, site_descr = Site_DESCR, 
          town = Town, county = COUNTY, state = State, lon_dd = LON_DD, 
          lat_dd = LAT_DD, huc_12 = HUC_12, huc_10_name = HUC_10_NAME, 
          huc_12_name = HUC_12_NAME)
-  
+
+# Enforcing sig digits again for station_year_mean, measurment_anmly, lt_mean, 
+# and lt_sd
+ww_lake_trend_data <- ww_lake_trend_data %>%
+  mutate(station_year_mean = case_when(param == "temp"  ~ 
+                                       round(station_year_mean, 1),
+                                     param == "chla" ~
+                                       round(station_year_mean, 1),
+                                     param == "total_p" ~
+                                       round(station_year_mean, 0),
+                                     param == "total_n" ~
+                                       round(station_year_mean/5, 0)*5,
+                                     TRUE ~ station_year_mean),
+         measurement_anmly = case_when(param == "temp"  ~ 
+                                         round(measurement_anmly, 1),
+                                       param == "chla" ~
+                                         round(measurement_anmly, 1),
[…]

### 2019-10-17 R/functions.R

@@ -1,7 +1,7 @@
 # Installations
 pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
           "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
-          "USAboundariesData","knitr","captioner")
+          "USAboundariesData","knitr","captioner", "pander")
 
 for(i in pkgs){
   if(!i %in% installed.packages()){

### 2019-10-17 R/summary_stats.R

@@ -24,6 +24,25 @@ ww_param_ranges <- ww_all %>%
          "75th Percentile" = percentile_75,
          Max = max, "Std. Dev" = sd )
 
+# All about the base...Enforce ww sig digits
+ww_param_ranges[ww_param_ranges$Parameter == "temp",][,3:8] <- 
+  round(ww_param_ranges[ww_param_ranges$Parameter == "temp",][,3:8], 1)
+
+ww_param_ranges[ww_param_ranges$Parameter == "total_n",][,3:8] <- 
+  round(ww_param_ranges[ww_param_ranges$Parameter == "total_n",][,3:8]/5)*5
+
+ww_param_ranges[ww_param_ranges$Parameter == "total_p",][,3:8] <- 
+  round(ww_param_ranges[ww_param_ranges$Parameter == "total_p",][,3:8], 0)
+
+ww_param_ranges[ww_param_ranges$Parameter == "np_ratio",][,3:8] <- 
+  round(ww_param_ranges[ww_param_ranges$Parameter == "np_ratio",][,3:8], 2)
+
+ww_param_ranges[ww_param_ranges$Parameter == "chla",][,3:8] <- 
+  round(ww_param_ranges[ww_param_ranges$Parameter == "chla",][,3:8], 1)
+
+ww_param_ranges <- ww_param_ranges %>%
+  mutate(Parameter = c("Temperature", "Total Nitrogen", "Total Phosphorus", "N:P", "Chlorophyll"))
+
 ww_param_anomaly_ranges <- ww_all %>%
   filter(param == "chla" |
            param == "temp" |
@@ -70,6 +89,25 @@ lagos_param_ranges <- lagos_all %>%
          Mean = mean, Median = median, 
          "75th Percentile" = percentile_75,
          Max = max, "Std. Dev" = sd )
+
+# All about the base...Enforce ww sig digits
+lagos_param_ranges[lagos_param_ranges$Parameter == "temp",][,3:8] <- 
+  round(lagos_param_ranges[lagos_param_ranges$Parameter == "temp",][,3:8], 1)
+
+lagos_param_ranges[lagos_param_ranges$Parameter == "total_n",][,3:8] <- 
+  round(lagos_param_ranges[lagos_param_ranges$Parameter == "total_n",][,3:8]/5)*5
+
+lagos_param_ranges[lagos_param_ranges$Parameter == "total_p",][,3:8] <- 
[…]

### 2019-10-17 manuscript/manuscript.Rmd
Most of arts edits done.  Need to add his comment about molar stuff from his email.

@@
 Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
-We analyze a sub-regional trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
 URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
-The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and state extents, respectively.
+The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively.
 We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e.
 scaled anomalies) calculated on a per-station basis.
-In Rhode Island, temperature and chlorophyll a increased from 1993 to 2016.
+Sub-regionally, temperature and chlorophyll a increased from 1993 to 2016.
 Total nitrogen shows a weak increase driven by low years in the early 1990s.
@@
 In short, algla biomass, as measured by cholorphyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
-Although not causal, this analysis suggests an association between lake temperature and primary production.
+This analysis suggests an association between lake temperature and primary production.
 Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
@@
 Increasing nutrient inputs, particularly of nitrogen (N) and phosphorus (P), derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
-This eutrophication leads to an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+This eutrophication often leads to an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
@@
 ecoregion, elevation, and latitude) are important and that regional trends exist.
-Lake-specific drivers are also important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
+Lake-specific drivers have also been shown to be important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
 Despite these challenges, it is important to study lakes at multiple spatial scales because emergent trends on regional or continental scales may or may not be present in individual lakes [@cheruvelil2013multi; @lottig2014long].
@@
 Thus, it was surprising when little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if the lack of trends was also present in water quality in Rhode Island lakes and reservoirs.
+Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
 
 Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
-URIWW is a citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
+URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
 Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g.
 May through October).
+The entire effort follow rigours quality control/quality assurance protocols.
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing, @Oliver_2017].
@@
 The goals of this study were to examine ~25 years of lake and reservoir data in Rhode Island and answer two questions.
-First, what are the state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature.
-Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states.
+First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature?
+Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states?
 Another focus of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
@@
 URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy.
-Data QA/QC is treated with paramount importance; volunteers are trained both in the classroom and the field, and are provided with all the necessary equipment and supplies, along with scheduled collection dates.
+Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates.
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen.
@@
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
+The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and the converting to a molar ratio by multiplying by 2.21 (e.g.
+atomic weight of P 30.974/atomic weight of 14.007).
 
@@
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
-The general approach is to calculate site-specific deviations from a long-term mean over a pre-determined reference period.
-This allows all sites to be shifted to a common baseline and the deviations, or anomalies, show change over the specified reference period.
+The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.
+This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, show change over the specified reference period.
 
@@
 [-@Oliver_2017] used hierarchical linear models and showed relatively stable water quality in the lakes of the northeastern United States.
-While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking local trends.
-Therefore, we decided to reanalyze the the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific z-score and trend analysis approach outlined above.
+While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking sub-regional trends.
+Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific z-score and trend analysis approach outlined above.
 
@@
 
-Our analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
+Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
 neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.
@@
 There are several broader management implications from the results of our analysis and of examining long-term water quality trends in general.
-In particular, this analysis provides much needed information about the long-term effects of current nutrient control efforts and identifies areas where additional information is required or a change in management approaches may be needed.
-First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term water quality trends.
+In particular, this analysis provides much needed information about the long-term effects of current nutrient control efforts at lake-specific and sub-regional scales and identifies areas where additional information is required or a change in management approaches may be needed.
+First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term water quality trends at multiple spatial scales.
 Specifically for this study, the results provide feedback to long time volunteer monitors, highlighting the importance of volunteer monitoring programs.
@@
 There are several possible approaches to further reduce nutrient loads [@yang2018nutrients].
-First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters.
+First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters [@kellogg2010geospatial].
 Green infrastructure approaches are one way to possibly achieve both goals [@pennino2016watershed; @reisinger2019changes].
-Additionally, within-lake approaches may also be used.
-Another approach is the restoration of freshwater mussels to waterbodies that historically had those species.
+Additionally, there is potential for within-lake approaches such as the restoration of freshwater mussels to waterbodies that historically had those species.
 Some studies using freshwater mussels have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].

### 2019-10-17 R/data_clean.R

@@ -112,7 +112,7 @@ ww_np <- ww_lake_trend_data %>%
   filter(param == "total_n" | param == "total_p") %>% 
   unique() %>% #repeats from somewhere...
   spread(key = param, value = mn_measurement) %>%
-  mutate(np_ratio = total_n/total_p) %>%
+  mutate(np_ratio = (total_n/total_p)*2.21) %>% #calc ratio and convert to molar
   filter(!is.na(np_ratio)) %>%
   filter(!is.infinite(np_ratio)) %>%
   gather(key = "param", value = "mn_measurement", np_ratio) %>%
@@ -191,7 +191,7 @@ lagos_data <- lagosne_select(table = "epi_nutr",
   mutate(year = as(year(ymd(sampledate)), "integer"),
          month = as(month(ymd(sampledate)), "integer"),
          day = as(day(ymd(sampledate)), "integer"),
-         `Station Name` = lagoslakeid, np_ratio = tn/tp) %>%
+         `Station Name` = lagoslakeid, np_ratio = (tn/tp)*2.21) %>%
   filter(year >= 1993) %>% #only get data for 1993 and beyond as that is earliest TN data for WW.  Make sure all ranges match
   #filter(year <= 2013) %>% #only get data from 2013 and prior to match Oliver
   select(`Station Name`, programname, year, month, day, chla, total_p = tp, total_n = tn, np_ratio) %>%

### 2019-10-17 R/summary_stats.R

@@ -17,7 +17,7 @@ ww_param_ranges <- ww_all %>%
             max = round(max(station_year_mean, na.rm = T), 2),
             sd = round(sd(station_year_mean, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l", "", "µg/l")) %>%
+  mutate(units = c("celsius", "µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 
@@ -59,7 +59,7 @@ ww_param_anomaly_ranges <- ww_all %>%
             max = round(max(measurement_anmly, na.rm = T), 2),
             sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l", "", "µg/l")) %>%
+  mutate(units = c("celsius", "µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 
@@ -83,7 +83,7 @@ lagos_param_ranges <- lagos_all %>%
             max = round(max(station_year_mean, na.rm = T), 2),
             sd = round(sd(station_year_mean, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("µg/l", "µg/l", "", "µg/l")) %>%
+  mutate(units = c("µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 
@@ -124,7 +124,7 @@ lagos_param_anomaly_ranges <- lagos_all %>%
             max = round(max(measurement_anmly, na.rm = T), 2),
             sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("µg/l", "µg/l", "", "µg/l")) %>%
+  mutate(units = c("µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 

### 2019-10-22 manuscript/manuscript.Rmd
technical reviews done - now to QA and BC

@@
 At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time.
-In short, algla biomass, as measured by cholorphyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
+In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
 This analysis suggests an association between lake temperature and primary production.
@@
 May through October).
-The entire effort follow rigours quality control/quality assurance protocols.
+The entire effort follow rigorous quality control/quality assurance protocols.
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing, @Oliver_2017].
@@
 
+Field and analytical methods are detailed on the URIWW website at <https://web.uri.edu/watershedwatch/uri-watershed-watch-monitoring-manuals/> and <https://web.uri.edu/watershedwatch/uri-watershed-watch-quality-assurance-project-plans-qapps/>, respectively.
+These methods, approved by both the state of Rhode Island and the US Environmental Protection Agency, have remained fairly consistent, although over the nearly 30 years changes did occur.
+When new methods were introduced, comparisons between old and new methods were conducted and in all cases no statistically significant differences were found with the new methods.
+Furthermore, the new methods did at times see an improvement in the limits of detection; however, this impacted a very small  number (less than 1%) of measurements in this study.
+We did run our analyses (see **Water Quality Trend Analysis** section) with all data and with only those data greater than the detection limit.
+There was no change in the trend analysis thus, the results we report are for all data as originally reported in the URIWW dataset.
+Given these facts, we assume the data to be consistent across the reported time period and appropriate for a long term assessment of trends.
+
 Prior studies have modeled water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
@@
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero ug/l to ~900 ug/l.
+For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero µg/l to ~900 µg/l.
 To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].
@@
 The general picture in Rhode Island appears to be one of little to no change in phosphorus, a very weak positive trend in nitrogen and little to no change in the TN:TP ratio.
-Furthermore, it has been shown that productivity in freshwater systems is likely a funciton of both phosphorus and nitrogen [@paerl2016takes].
+Furthermore, it has been shown that productivity in freshwater systems is likely a function of both phosphorus and nitrogen [@paerl2016takes].
 Thus, the increasing chlorophyll *a* in the face of stable TN:TP ratio suggests that the increase is being driven by something other than nutrients.
@@
 However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*.
-Additionally, these results point to the fact that chlorophyll *a* and aglal biomass is driven by processes operating at different scales.
+Additionally, these results point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.
 For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].

### 2019-10-22 R/data_clean.R

@@ -27,7 +27,7 @@ missing_nutrients_1 <- read_csv(here("data/WW_96_nutrients_nds.csv")) %>%
   rbind(read_csv(here("data/WW_98_nutrients.csv"))) %>%
   rbind(read_csv(here("data/WW_99_nutrients.csv"))) %>%
   rbind(read_csv(here("data/WW_00_nutrients.csv"))) %>%
-  rbind(read_csv(here("data/WW_00_nutrients.csv"))) %>%
+  rbind(read_csv(here("data/WW_01_nutrients.csv"))) %>%
   mutate(Day = 15) # Need this to conform to existing format - none of these had a day so assuming middle of month
 
 missing_nutrients <- missing_nutrients_1 %>%
@@ -35,7 +35,7 @@ missing_nutrients <- missing_nutrients_1 %>%
             Date = ymd(paste(Year, Month, Day, sep="-")),
             Time = NA,
             `Sample Type` = NA,
-            `Sample Media` = NA,
+            `Sample Media` = "Water",
             Depth = Depth,
             Parameter = Parameter,
             Concentration = Concentration,
@@ -69,28 +69,59 @@ ww_lake_trend_data <- ww_all %>%
   filter(year >= 1993) %>% #only get data for 1993 and beyond as that is earliest TN data for WW.  Make sure all ranges match
   #filter(year <= 2013) %>% #only get data from 2013 and prior to match Oliver
   filter(Parameter %in% params) %>% #Filter out subset of parameters
-  filter(Concentration != `Detection Limit`) %>% #filters out measurements at detect limit (temporary)
+  #filter(Concentration != `Detection Limit`) %>% #filters out measurements at detect limit (temporary)
   mutate(Parameter = case_when(Parameter == params[1] ~ "temp", #NAMES!!!!!
                                Parameter == params[2] ~ "secchi",
                                Parameter == params[3] ~ "total_p",
                                Parameter == params[4] ~ "total_n",
                                Parameter == params[5] ~ "enterococci",
                                Parameter == params[6] ~ "do",
-                               Parameter == params[7] ~ "chla",
                                Parameter == params[8] ~ "ph",
-                               Parameter == params[9] ~ "chloride")) %>% 
+                               Parameter == params[9] ~ "chloride",
+                               Parameter == params[7] ~ "chla",
+                               TRUE ~ Parameter)) %>%
[…]

### 2019-11-05 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
draft outline of NALMS talk

---
title: "Increasing productivity amid stable nutrient regmines in Rhode Island Lakes and Reservoirs"
subtitle: "NALMS 2019, Burlington VT"
author: "Jeff Hollister, Q Kellogg, Betty Kreakie, Stephen Shivers, Elizabeth Herron, Linda Green, Bryan Milstead, and Art Gold"
date: "2019-11-14"
output:  
  powerpoint_presentation:
    reference_doc: increasing_chla_stable_nutrients_template.pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
## Background
::::::::::::::{.columns}
:::{.column}
- Nutrient managment over the years
- One goal is to reduce increased phyto
- Long-term data needed
:::
:::{.column}
content
:::
::::::::::::::

## Stasis in Northeast

content

## But change at other scales

- VT example
- Other examples?
- What about RI?

## Watershed Watch

- Details on WW program and data collected

## Watershed Watch: filtered for trend analysis

- What we filtered
- resultant datasets details

## LAGOSNE

## Analysis: Site-specific z-scores

## Analysis: Yearly trends

## Results: Nutrient trends

## Results: Chlorophyll trends

## Results: Temperature trends

## Conclusions

### In Rhode Island
- More chlorophyll
- More heat
- Nutrients not changing much

### Within the Northeast
- Stasis (our analysis concurs)

## Thank you and Questions

### 2019-11-05 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
Getting started on NALMS talk

@@
 ```
-## Background
+
+## Nutrients and Productivity
 ::::::::::::::{.columns}
 :::{.column}
-- Nutrient managment over the years
+- Nutrients drive growth
 - One goal is to reduce increased phyto
@@
 :::{.column}
-content
+
+![](figures/es-2016-02575v_0006.jpeg)
+
+![](figures/tn_tp_chla_plot.jpg)
 :::
 ::::::::::::::
+
+## Nutrient Management
+
+- What has happened with P and N reductions?
+- What 
+- Did it work?
 

### 2019-11-05 nalms_2019_talk/tn_tp_chla_plot.R

@@ -1,0 +1,48 @@
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(hrbrthemes)
+
+nla_2007_url <- "https://www.epa.gov/sites/production/files/2014-10/nla2007_chemical_conditionestimates_20091123.csv"
+nla_2012_url <- "https://www.epa.gov/sites/production/files/2017-02/nla12_keyvariables_data.csv"
+
+nla_2007 <- read_csv(nla_2007_url)
+
+nla_2012 <- read_csv(nla_2012_url)
+
+nla_2007_nutr_chla <- nla_2007 %>%
+  mutate(year = 2007) %>%
+  rename_all(tolower) %>%
+  select(year, state = st, site_id, total_p = ptl, 
+         total_n = ntl, chla)
+
+nla_2012_nutr_chla <- nla_2012 %>%
+  mutate(year = 2012) %>%
+  rename_all(tolower) %>%
+  select(year, state, site_id, total_p = ptl_result, 
+         total_n = ntl_result, chla = chlx_result) %>%
+  mutate(total_n = total_n * 1000)
+
+nla <- rbind(nla_2007_nutr_chla, nla_2012_nutr_chla)  
+
+tn_tp_chla_gg <- nla %>%
+  group_by(year, state) %>%
+  summarize(mean_total_p = mean(total_p, na.rm = TRUE),
+            mean_total_n = mean(total_n, na.rm = TRUE),
+            mean_chla = mean(chla, na.rm = TRUE)) %>%
+  ungroup() %>%
+  ggplot(aes(x = log1p(mean_total_p), 
+             y = log1p(mean_total_n),
+             color = log1p(mean_chla), 
+             size = log1p(mean_chla))) +
+  geom_point() +
+  scale_color_continuous(low = "royalblue", high = "springgreen4", name = "Log10 Chlorophyll") +
+  scale_size_continuous(range = c(0.1, 10), guide = FALSE) +
+  labs(x = "Log10 Total Phosphorus", 
+       y = "Log10 Total Nitrogen",
+       title = "NLA 2007 and 2012 - State Averages") +
+  theme_ipsum_rc(base_size = 14, axis_title_size = 14,
+                 axis_text_size = 14)
+  
+ggsave("nalms_2019_talk/figures/tn_tp_chla_plot.jpg", tn_tp_chla_gg, width = 12, height = 9, dpi = 300)  
+  

### 2019-11-05 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
outline and draft of slide 1 done

@@
 
-## Nutrients and Productivity
+## What drives lake productivity
 ::::::::::::::{.columns}
 :::{.column}
-- Nutrients drive growth
-- One goal is to reduce increased phyto
-- Long-term data needed
+- Nutrients
+- Temperature
+- Other things
 :::
@@
 
+- One goal is to reduce increased phyto
 - What has happened with P and N reductions?
-- What 
 - Did it work?
+  - Long-term data needed
 

### 2019-11-07 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
progress on slides!

--- 2019-11-05 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
@@
 
-## Nutrients and Productivity
+## What drives changes in lake productivity?
 ::::::::::::::{.columns}
 :::{.column}
-- Nutrients drive growth
-- One goal is to reduce increased phyto
-- Long-term data needed
+- Nutrients
+- Temperature
+- Other things
 :::
@@
 ![](figures/tn_tp_chla_plot.jpg)
+
+![](figures/temp_chla_plot.jpg)
 :::
@@
 
-## Nutrient Management
+## The management problem
 
-- What has happened with P and N reductions?
-- What 
-- Did it work?
+- Desire to reduce and manage nutrient pollution
+- Phosphorus Bans
+- Nutrient Criteria
+- Enacted piecemeal over last ~20 years
+- Do the data show any change?
 
@@
 
-content
+- Oliver et al (2017) Unexpected stasis in a changing world: Lake nutrient and
+chlorophyll trends since 1990. Global Change Biology. DOI: 10.1111/gcb.13810
+
+![](figures/10.1111-gcb.13810Figure138100001.jpg)
 
@@
 
-- VT example
-- Other examples?
-- What about RI?
+- Mathews et al (2018) Is Vermont losing its oligotrophic lakes?. LakeLine. 38-2.
+- Rhode Island?
 
-## Watershed Watch
+![](figures/mathews_vt_olgio_lakeline_fig2.jpg)
 
-- Details on WW program and data collected
+## URI Watershed Watch
 
-## Watershed Watch: filtered for trend analysis
+- Volunteer monitoring program
+  - ~400 volunteers
+- Started in 1988 with 14 lakes
+- Now
+  - 250+ sites
+  - 120+ waterbodies
+- Rigorous QA/QC
+- Data used by RI DEM and US EPA
 
-- What we filtered
-- resultant datasets details
+## URI Watersehd Watch: Lake Sampling
+
+- May to October
+- Weekly
+  - Water Clarity
+  - Temperature
+- Every Other Week
+  - Chlorophyll
+  - Dissolved Oxygen
+- 3 Times per Season
+  - Nutrients
+  - Alakalinity
+  - pH
+  - Bacteria
+
+## URI Watershed Watch: Filtered for Trend Analysis
+
+- Parameters: Total Nitrogen, Total Phosphorus, N:P, Temperature, Chlorophyll
+- Years: 1993 to 2016
+- Months: May through October
+- Depths: <= 2 meters
+- Sites: At least 10 years of data
+- Total of 69 Sites
+
+![](figures/ww_map.jpg)
 

### 2019-11-07 nalms_2019_talk/tn_tp_chla_plot.R

@@ -10,18 +10,27 @@ nla_2007 <- read_csv(nla_2007_url)
 
 nla_2012 <- read_csv(nla_2012_url)
 
+nla_temp <- read_csv("../lake_photic_zone/data/nla_base.csv") %>%
+  select(nla_id, tmean_2m)
+
 nla_2007_nutr_chla <- nla_2007 %>%
   mutate(year = 2007) %>%
   rename_all(tolower) %>%
   select(year, state = st, site_id, total_p = ptl, 
-         total_n = ntl, chla)
+         total_n = ntl, chla) %>%
+  left_join(nla_temp, by = c("site_id" = "nla_id")) %>%
+  filter(tmean_2m < 75) %>% #some REALLY hot lakes - filtering out
+  filter(total_p < 4000) #same site in NV with REALLY high TP 
 
 nla_2012_nutr_chla <- nla_2012 %>%
   mutate(year = 2012) %>%
   rename_all(tolower) %>%
   select(year, state, site_id, total_p = ptl_result, 
          total_n = ntl_result, chla = chlx_result) %>%
-  mutate(total_n = total_n * 1000)
+  mutate(total_n = total_n * 1000) %>%
+  left_join(nla_temp, by = c("site_id" = "nla_id"))  %>%
+  filter(tmean_2m < 75) %>% #some REALLY hot lakes - filtering out
+  filter(total_p < 4000) #same site in NV with REALLY high TP 
 
 nla <- rbind(nla_2007_nutr_chla, nla_2012_nutr_chla)  
 
@@ -29,20 +38,35 @@ tn_tp_chla_gg <- nla %>%
   group_by(year, state) %>%
   summarize(mean_total_p = mean(total_p, na.rm = TRUE),
             mean_total_n = mean(total_n, na.rm = TRUE),
-            mean_chla = mean(chla, na.rm = TRUE)) %>%
+            mean_chla = mean(chla, na.rm = TRUE),
+            mean_temp = mean(tmean_2m, na.rm = TRUE)) %>%
   ungroup() %>%
-  ggplot(aes(x = log1p(mean_total_p), 
-             y = log1p(mean_total_n),
-             color = log1p(mean_chla), 
-             size = log1p(mean_chla))) +
+  ggplot(aes(x = mean_total_p, 
+             y = mean_total_n,
+             color = mean_chla, 
+             size = mean_chla)) +
   geom_point() +
[…]

### 2019-11-08 manuscript/manuscript.Rmd
more slides and lagosne available data stats added to script (not yet to ms)

@@
 The entire effort follow rigorous quality control/quality assurance protocols.
-These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing, @Oliver_2017].
+These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].
 URIWW data contributed to the larger regional study by Oliver et al.

### 2019-11-08 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
more stuff in slides
more slides and lagosne available data stats added to script (not yet to ms)

@@
 - Total of 69 Sites
+  - 66 with all parameters
+  - 1 without N:P
+  - 2 without chlorophyll and temperature
 
@@
 
-## LAGOSNE
+## LAke mult-scaled GeOSpatial and temoporal database for the Northeastern US (LAGOS-NE)
+
+- Soranno et al (2017). LAGOS-NE: a multi-scaled geospatial and temporal database of lake ecological context and water quality for thousands of US lakes. Gigascience, 6(12)
+- Stachelek and Oliver (2017). LAGOSNE: Interface to the lake multi-scaled geospatial and temporal database. R package version 1.1.0. https://cran.r-project.org/package=LAGOSNE
+- <https://lagoslakes.org/>
+- 17 States: CT, IL, IN, IA, ME, MA, MI, MN, MO, NH, NJ, NY, OH, PA, RI, VT, WI
+- ~ 50,000 Lakes
+
+![](figures/gix101fig1.jpeg)
+
+## LAke mult-scaled GeOSpatial and temoporal database for the Northeastern US (LAGOS-NE): Filtered for Trend Analysis
+
+- Parameters: Total Nitrogen, Total Phosphorus, N:P, Chlorophyll
+  - Temperature data not currently available
+- Years: 1993 to 2016
+- Months: May through October
+- Depths: <= 2 meters
+- Sites: At least 10 years of data
+- Total of 1482 Sites
+  - 637 with chlorophyll and phosphorus
+  - 377 with only chlorophyll
+  - 275 with only phosphorus
+  - 161 with all variables (except temperature)
+  - 25 with chlorophyll and nitrogen
+  - 7 with other combination of nutrients
 
@@
 
+![](figures/z-score_schematic.jpg)
+
 ## Analysis: Yearly trends
 
-## Results: Nutrient trends
+## Results: Nitrogen trends
+
+![](figures/ww_tn_trends.jpg)
+
+![](figures/lagos_tn_trends.jpg)
+
+## Results: Phosphorus trends
+
+![](figures/ww_tp_trends.jpg)
+
+![](figures/lagos_tp_trends.jpg)
+
+## Results: N:P trends
+
+![](figures/ww_np_trends.jpg)
+
+![](figures/lagos_np_trends.jpg)
 
@@
 
+![](figures/ww_chla_trends.jpg)
+
+![](figures/lagos_chla_trends.jpg)
+
 ## Results: Temperature trends
+
+![](figures/ww_temp_trends.jpg)
 

### 2019-11-08 R/data_available.R

@@ -2,6 +2,8 @@
 #' load up functions and packages
 source(here::here("R/functions.R"))
 
+# WW Available Data
+
 ww_data <- read_csv(here("data/ww_lake_trend_data.csv"))
 
 chla_10yr_sites <- ww_data %>%
@@ -60,3 +62,65 @@ ww_avail_data_locations <- ww_data %>%
   right_join(data_avail) %>%
   select(station_name, params_avail) %>%
   write_csv(here("data/ww_avail_data_stations.csv"))
+
+# LAGOSNE Available Data
+
+lagosne_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
+
+lagosne_chla_10yr_sites <- lagosne_data %>%
+  filter(param == "chla") %>%
+  filter_year(10)
+
+lagosne_total_n_10yr_sites <- lagosne_data %>%
+  filter(param == "total_n") %>%
+  filter_year(10)
+
+lagosne_total_p_10yr_sites <- lagosne_data %>%
+  filter(param == "total_p") %>%
+  filter_year(10)
+
+lagosne_np_10yr_sites <- lagosne_data %>%
+  filter(param == "np_ratio") %>%
+  filter_year(10)
+
+lagosne_possible_sites <- unique(c(lagosne_chla_10yr_sites,lagosne_total_p_10yr_sites, 
+                           lagosne_total_n_10yr_sites, lagosne_np_10yr_sites))
+lagosne_data_avail <- tibble(station_name = lagosne_possible_sites, 
+                     chla_avail = lagosne_possible_sites %in% lagosne_chla_10yr_sites,
+                     total_n_avail = lagosne_possible_sites %in% lagosne_total_n_10yr_sites, 
+                     total_p_avail = lagosne_possible_sites %in% lagosne_total_p_10yr_sites,
+                     np_avail = lagosne_possible_sites %in% lagosne_np_10yr_sites) %>%
+  mutate(params_avail = case_when(chla_avail == TRUE & total_n_avail == TRUE &
+                                    total_p_avail == TRUE & np_avail == TRUE ~ 
+                                    "chla, total_p, total_n, np",
+                                  chla_avail == TRUE & total_n_avail == TRUE &
+                                    total_p_avail == TRUE ~ 
+                                    "chla, total_p, total_n",
+                                  chla_avail == TRUE & total_n_avail == FALSE & 
+                                    total_p_avail == FALSE ~
+                                    "chla",
[…]

### 2019-11-08 nalms_2019_talk/increasing_chla_stable_nutrients_talk.Rmd
slides in Rmd done - future edits to pptx

@@
 
+![](figures/green_lake.png)
+
+![](figures/not_green_lake.png)
+
 ## Stasis in Northeast
@@
 
+![](figures/WW-Logo1.gif)
 ## URI Watersehd Watch: Lake Sampling
@@
   - Bacteria
+
+![](figures/boy-w-tube_EH.jpg)
+
+![](figures/14 V lowering deep sampler_EH.jpg)
 
@@
 
-## Analysis: Site-specific z-scores
+## Analysis: Site-specific z-scores and yearly trends
+
+- Calculate average yearly z-score
+- Look for monotonic yearly trend
+  - Slope of regression line
 
@@
 
-## Analysis: Yearly trends
+## Results: Nitrogen trends
 
-## Results: Nitrogen trends
+- RI: Increasing but driven by low years in 1993 and 1994
+  - Without those years slope = 0.01 and p-value = 0.06
+- LAGOSNE: Static
 
@@
 
+- RI: Static
+- LAGOSNE: Static
+
 ![](figures/ww_tp_trends.jpg)
@@
 ## Results: N:P trends
+
+- RI: Static
+- LAGOSNE: Static
 
@@
 
+- RI: Increasing
+- LAGOSNE: Static
+
 ![](figures/ww_chla_trends.jpg)
@@
 ## Results: Temperature trends
+
+- RI: Increasing
+- LAGOSNE: NA
 
@@
 ### Within the Northeast
-- Stasis (our analysis concurs)
+- Stasis
+  - Our results concur with Oliver et al. (2017)
 
 ## Thank you and Questions
+
+![](figures/yawgoo.jpg)

### 2019-11-18 manuscript/manuscript.Rmd
Ready for BC review

@@
 Another focus of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-Finally, we conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017].
+Finally, we conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation].
 
@@
 Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
+We used the same selction criteria on hte LAGOSNE dataset as was applied to the URIWW data.
 
@@
 Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
-For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
@@
 The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero µg/l to ~900 µg/l.
+For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero µg/L to ~900 µg/L.
 To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].
@@
 Methods for calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
-Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from <fill in with zenodo link> [@zenodo_citation].
+Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].
 
@@
 
-For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
 
@@
 Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from <link to zenodo> [@zenodo_citation].
+Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from "fill in with zenodo link" [@zenodo_citation].
 

### 2020-01-28 R/functions.R
munns edits

@@ -90,14 +90,15 @@ wq_trend_gg <- function(df, wqparam,
     slice(2) %>%
     select(slope = estimate, p.value) 
   
+  
   gg <- ggplot(df2,aes(x = year, y = mn_value)) + 
     geom_pointrange(aes(ymin=mn_value-!!error_bar, ymax=mn_value+!!error_bar, 
                         color = col_group), size = 1, fatten = 1.75) +
     #geom_point(aes(color = col_group), size=3.5) +
     geom_smooth(method = "lm", se=FALSE, color = "black") +
     theme_ipsum() +
-    labs(..., title = title, subtitle = paste0("slope: ", signif(regress$slope,2),
-                             " p-value: ", signif(regress$p.value, 2))) +
+    labs(..., title = title, subtitle = paste0("slope = ", signif(regress$slope,2),
+                             " p = ", signif(regress$p.value, 2))) +
     scale_color_manual(values = c("red3","darkblue")) + 
     theme(legend.position="none", 
           plot.title = element_text(size = 12, face = "bold"),

### 2020-01-31 manuscript/manuscript_anomaly_only.Rmd
one last read before submission!

--- 2019-11-18 manuscript/manuscript.Rmd
@@
 - Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
-  S. ^*3*^, Milstead, W. Bryan. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
+  S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
   ^*2*^
 - ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
-  Ecology Division, Narragansett, RI 02882
+  Coastal Environmental Sciences Division, Narragansett, RI 02882
 - ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
@@
 The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively.
-We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e.
-scaled anomalies) calculated on a per-station basis.
-Sub-regionally, temperature and chlorophyll a increased from 1993 to 2016.
+We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged anomalies calculated on a per-station basis.
+Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016.
 Total nitrogen shows a weak increase driven by low years in the early 1990s.
 Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
-At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time.
+At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.
 In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
 This analysis suggests an association between lake temperature and primary production.
-Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific z-scores for analyzing for long-term trends.'
+Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.'
 ---
@@
 source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
-source(here::here("R/figures.R"), encoding = "UTF-8")
+source(here::here("R/figures_anomaly_only.R"), encoding = "UTF-8")
 ```
@@
 Increasing nutrient inputs, particularly of nitrogen (N) and phosphorus (P), derived from intensive agriculture and densely populated urban areas have contributed to the eutrophication of many lakes [@carpenter1998nonpoint; @smith2003eutrophication].
-This eutrophication often leads to an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication ;@paerl2009climate].
+This eutrophication often leads to an increase in the frequency and severity of harmful algal blooms, greater risks for human and animal health, and potential economic costs associated with eutrophic waters [@michalak2013record; @kosten2012warmer; @taranu2015acceleration; @brooks2016harmful; @dodds2008eutrophication; @paerl2009climate].
 To address these problems, management strategies have historically focused on reducing P inputs to lakes, but research also suggests that reducing N inputs may be more effective in certain situations [@paerl2016takes; @schindler2008eutrophication].
@@
 Lake datasets that cover longer time periods and broader spatial scales are now becoming available.
-Programs such as USEPA's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
+Programs such as the US Environmental Protection Agency's National Lakes Assessment (NLA) provide data that allow for continental-scale water quality analysis.
 These data allow for analyses that can be useful for managing water resources by developing water quality criteria for N, P, and chlorophyll *a* [@herlihy2013using; @yuan2014managing].
@@
 Broad-scale data can also be used for water quality modeling across a range of spatial scales including for predicting lake trophic state, which is indicative of ecosystem condition [@hollister2016modeling; @nojavan2019rethinking].
-These trophic state models indicate that landscape variables (e.g.
-ecoregion, elevation, and latitude) are important and that regional trends exist.
+These trophic state models indicate that landscape variables (e.g., ecoregion, elevation, and latitude) are important and that regional trends exist.
 Lake-specific drivers have also been shown to be important for predicting continental-scale water quality which adds an additional layer of complexity [@read2015importance].
@@
 Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale.
-Macro-scale (i.e.
-subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
+Macro-scale (i.e., subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
 This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake].
@@
 Thus, it was surprising when little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long-term trends in water quality within the broader region of the northeastern United States, we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
+Given what is known about long-term trends in water quality within the broader region of the northeastern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
 
@@
 URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
-Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g.
-May through October).
+Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g., May through October).
 The entire effort follow rigorous quality control/quality assurance protocols.
@@
 Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states?
-Another focus of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-Finally, we conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation].
+Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation].
 
@@
 
-For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e.
-z-scores) to find increasing or decreasing annual water quality trends.
+For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on water quality anomalies (i.e., measured values with the long term mean subtracted) to find increasing or decreasing annual water quality trends.
 Details are outlined below.
@@
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
-The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and the converting to a molar ratio by multiplying by 2.21 (e.g.
-atomic weight of P 30.974/atomic weight of 14.007).
+The N:P ratio was calculated by dividing the mass concentrations of total nitrogen and total phosphorus and then converting to a molar ratio by multiplying by 2.21 (e.g., atomic weight of P 30.974/atomic weight of N 14.007).
 
@@
 When new methods were introduced, comparisons between old and new methods were conducted and in all cases no statistically significant differences were found with the new methods.
-Furthermore, the new methods did at times see an improvement in the limits of detection; however, this impacted a very small  number (less than 1%) of measurements in this study.
+Furthermore, the new methods did at times improve the limits of detection; however, this impacted a very small  number (less than 1%) of measurements in this study.
 We did run our analyses (see **Water Quality Trend Analysis** section) with all data and with only those data greater than the detection limit.
-There was no change in the trend analysis thus, the results we report are for all data as originally reported in the URIWW dataset.
-Given these facts, we assume the data to be consistent across the reported time period and appropriate for a long term assessment of trends.
+There was no change in the trend analysis and thus, the results we report are for all data as originally reported in the URIWW dataset.
+Given these results, we assume the data to be consistent across the reported time period and appropriate for a long term assessment of trends.
 
-Prior studies have modeled water quality trends across a larger region of the northeastern United states that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+Prior studies have modeled water quality trends across a larger region of the northeastern US that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
@@
 Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
-We used the same selction criteria on hte LAGOSNE dataset as was applied to the URIWW data.
+We used the same selction criteria on the LAGOSNE dataset as was applied to the URIWW data.
 
@@
 Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
-For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
 The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.
-This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, show change over the specified reference period.
+This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.
+We refer to this method as "site-specific anomalies".
 
-### Z-score
+### Summarizing site-specific anomalies
 
-Analysis that focuses on anomalies works best with a single measure (e.g., temperature) or with multiple measurements that are on the same scale.
-The water quality parameters that we explored have different scales and thus the anomaly alone is difficult to interpret across metrics.
-For instance, temperature in Rhode Island lakes during the growing season ranges from approximately 15 degrees Celsius to a high of 30 degrees Celsius, whereas phosphorus might range from near zero µg/L to ~900 µg/L.
-To standardize these values, we used the common approach of dividing each anomaly by the standard deviation for the reference period [e.g., @jones1996calculating].
-The resultant value is commonly referred to as a z-score.
-We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.
-Furthermore, since we are interested in water quality trends over time at individual sites (i.e.
-lakes), z-scores were calculated over the reference period, 1993-2016, for each site.
-We refer to this approach as "site-specific z-scores".
+Methods for calculating the site-specific anomalies and the yearly averages are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].
 
-### Summarizing site-specific z-scores
+![`r fig_num("anomaly_schematic")`](../figures/anomaly_schematic.jpg)
 
-Methods for calculating the site-specific z-scores and the yearly averages are as follows and are presented graphically in `r fig_num(name = "z_schematic", caption = "Example calculation of the site-specific z-scores and yearly mean z-scores.", display = "cite")`.
-Additionally, an example R script, `schematic.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("z_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].
-
-![`r fig_num("z_schematic")`](../figures/z-score_schematic.jpg)
-
-The general steps, outlined in `r fig_num("z_schematic", display = "cite")` and listed below, are repeated for each of the water quality parameters.
+The general steps, outlined in `r fig_num("anomaly_schematic", display = "cite")` and listed below, are repeated for each of the water quality parameters.
 
@@
 The per site means across years are assumed to be independent.
-2. Calculate the long-term reference mean and reference standard deviation for each site. This results in a single long-term mean and standard deviation for each of the sites.
-3. Calculate the z-score for each annual mean at each site by subtracting the annual and reference means and dividing by the reference standard deviation.
-4. Summarize by calculating the mean z-score per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
+2. Calculate the long-term reference mean for each site. This results in a single long-term mean for each of the sites.
+3. Calculate the anomaly for each annual mean at each site by subtracting the annual and reference means.
+4. Summarize by calculating the mean anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
 
-### Linear regression on annual mean z-scores
+### Linear regression on annual mean anomalies
 
@@
 We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
-For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
-$\beta$~1~ = 0).
+For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).
+The slope of this line provides information on the average yearly change of that paramter over the time period studied.
 Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].
@@
 
-### Comparison of Rhode Island to the Region
+### Comparison of Rhode Island to the region
 
@@
 While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking sub-regional trends.
-Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific z-score and trend analysis approach outlined above.
+Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific anomaly and trend analysis approach outlined above.
 
@@
 
-During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "temp")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_n")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "total_p")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "np_ratio")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "chla")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
+During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
 
@@
 
-Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
+Average annual temperature anomalies in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with average temperature greater than the long-term average occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average anomaly) in Rhode Island lakes and reservoirs.
+Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (average anomaly).
 Panel A.
-Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.
+Yearly averaged chlorophyll a anomalies from the URI Watershed Watch data.
 Panel B.
-Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
+Yearly averaged chlorophyll a anomalies from the LAGOSNE dataset.
+Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
 
@@
 Mean annual trends for nutrients were weaker or showed no trend over time.
-The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score).
+The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average anomaly).
 Panel A.
-Yearly averaged TN z-scores from the URI Watershed Watch dataset.
+Yearly averaged TN anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly averaged TN z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-Since 1995, the yearly trend shows a much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
-TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tp_rl$slope,3)` , p-value: `r round(ww_tp_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
+Yearly averaged TN anomalies from the LAGOSNE dataset.
+Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+Since 1995, the yearly trend shows a lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).
+TP does not show a trend over time in the yearly anomalies (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the average are more evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average anomaly).
 Panel A.
-Yearly averaged TP z-scores from the URI Watershed Watch dataset.
+Yearly averaged TP anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly averaged TP z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).
+Yearly averaged TP anomalies from the LAGOSNE dataset.
+Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average anomaly).
 Panel A.
-Yearly averaged TN:TP ratio z-scores from the URI Watershed Watch dataset.
+Yearly averaged TN:TP ratio anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from "fill in with zenodo link" [@zenodo_citation].
+Yearly averaged TN:TP ratio anomalies from the LAGOSNE dataset.
+Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from "fill in with zenodo link" [@zenodo_citation].
 
-![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends.jpg)
+![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends_anomaly.jpg)
 
-![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends.jpg)
+![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends_anomaly.jpg)
 
-![`r fig_num(name = "tn_trend")`](../figures/paired_tn_trends.jpg)
+![`r fig_num(name = "tn_trend")`](../figures/paired_tn_trends_anomaly.jpg)
 
-![`r fig_num(name = "tp_trend")`](../figures/paired_tp_trends.jpg)
+![`r fig_num(name = "tp_trend")`](../figures/paired_tp_trends_anomaly.jpg)
 
-![`r fig_num(name = "np_trend")`](../figures/paired_np_trends.jpg)
+![`r fig_num(name = "np_trend")`](../figures/paired_np_trends_anomaly.jpg)
 
@@
 In general, there was little evidence to suggest broad regional changes.
-Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+Chlorophyll *a* showed a very weak positive trend (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight decreasing trend (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight positive trend (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 
-Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
-neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.
+Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e., neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
+Chlorophyll has increased on average `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.
 This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.
@@
 Similar to the results of Oliver et al.
-[-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g.
-northeastern and mid-western United States).
+[-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).
 However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.
@@
 
-Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995, @litke1999review].
+Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].
 However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*.
@@
 
-## Management Implications
+## Management implications
 
@@
 
-## Data Analysis Approach
+## Data analysis approach
 
-The analysis approach we used here, site-specific z-scores, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface].
+The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface].
 However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.
 We built on these methods and adapted them for use with long-term water quality trends.
-While other methods are valid and robust [e.g.
-@Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
-In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).
+While other methods are valid and robust [e.g., @Oliver_2017], we chose averaged site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.
+For instance, reporting the changes in anomalies allows us to look at changes in the original units.
+With our analysis, the slope of the regression line for temperature suggests an average yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows and average yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
+Additionally, the site-specific anomalies are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).
 Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.

### 2020-01-31 R/data_clean.R

@@ -152,6 +152,12 @@ ww_np <- ww_lake_trend_data %>%
 ww_lake_trend_data <- ww_lake_trend_data %>%
   rbind(ww_np)
 
+# Almy Pond, 2016-07-23, Sample likely contaminated with bottom sediment per notes in data
+# Removed from analysis
+idx <- ww_lake_trend_data$year == "2016" & ww_lake_trend_data$month == 7 & ww_lake_trend_data$day == 23 & ww_lake_trend_data$station_name == "WW199"
+idx <- !idx
+ww_lake_trend_data <- ww_lake_trend_data %>% 
+  filter(idx)
   
 ww_lake_trend_data <- ww_lake_trend_data %>%
   filter(month >= 5 & month <= 10) %>%

### 2020-01-31 R/figures_anomaly_only.R

@@ -1,0 +1,203 @@
+#' script to generate figures
+#' load up functions and packages
+source(here::here("R/functions.R"))
+
+ww_data <- read_csv(here("data/ww_lake_trend_data.csv"))
+
+chla_gg <- ww_data %>%
+  #filter(state == "RI") %>%
+  wq_trend_gg( "chla", yvar = "measurement_anmly", 
+               y = expression(paste("Average Yearly Chlorophyll ", 
+                                    italic(" a"), " Anomaly (", mu, "g/L)")), 
+               x = "Year",
+               write = here("data/chla_data_anomaly.csv"), error_bar = "sd",
+               title = "A. URI Watershed Watch Chlorophyll")
+ggsave(here("figures/ww_chla_trends_anomaly.jpg"), chla_gg[[1]], width = 7.5, 
+       height = 5.625, 
+       units = "in", dpi = 600)
+
+ww_chla_kt <- chla_gg[[2]]
+ww_chla_df <- chla_gg[[3]]
+ww_chla_rl <- chla_gg[[4]]
+
+tn_gg <-  ww_data %>%
+  #filter(state == "RI") %>%
+  wq_trend_gg("total_n", yvar = "measurement_anmly", 
+              y = expression(paste("Average Yearly Total Nitrogen Anomaly (", mu, "g/L)")), 
+              x = "Year",
+              write = here("data/total_n_data_anomaly.csv"), error_bar = "sd",
+              title = "A. URI Watershed Watch Total Nitrogen")
+ggsave(here("figures/ww_tn_trends_anomaly.jpg"), tn_gg[[1]], width = 7.5, height = 5.625, 
+       units = "in", dpi = 600)
+
+ww_tn_kt <- tn_gg[[2]]
+ww_tn_df <- tn_gg[[3]]
+ww_tn_rl <- tn_gg[[4]]
+
+tp_gg <-  ww_data %>%
+  #filter(state == "RI") %>%
+  wq_trend_gg("total_p", yvar = "measurement_anmly", 
+              y = expression(paste("Average Yearly Total Phosphorus Anomaly (", mu, "g/L)")), 
+              x = "Year",
+              write = here("data/total_p_data_anomaly.csv"), error_bar = "sd",
+              title = "A. URI Watershed Watch Total Phosphorus")
+ggsave(here("figures/ww_tp_trends_anomaly.jpg"), tp_gg[[1]], width = 7.5, height = 5.625, 
[…]

### 2020-01-31 R/functions.R

@@ -59,6 +59,7 @@ wq_trend_gg <- function(df, wqparam,
                         yvar = c("measurement_anmly","measurement_scale"), 
                         num_yrs = 10, write = NULL, title = "",
                         start_yr = 1993, error_bar = c("se", "sd"), ...){
+  
   yvar <- rlang::sym(match.arg(yvar))
   error_bar <- rlang::sym(match.arg(error_bar))
   
@@ -90,13 +91,23 @@ wq_trend_gg <- function(df, wqparam,
     slice(2) %>%
     select(slope = estimate, p.value) 
   
+  #Calc y-axis ranges
+  if(yvar == "measurement_scale"){
+    label_break <- c(-2,-1,0,1,2)
+    limits <- c(-2.75, 2.75)
+  } else if(yvar == "measurement_anmly"){
+    limit<-max(abs(c(floor(min(df2$mn_value - df2[[error_bar]])), 
+                     ceiling(max(df2$mn_value + df2[[error_bar]])))))
+    label_break <- c(-limit, -limit/2, 0, limit/2, limit)
+    limits <- c(-limit,limit)
+  }
   
   gg <- ggplot(df2,aes(x = year, y = mn_value)) + 
     geom_pointrange(aes(ymin=mn_value-!!error_bar, ymax=mn_value+!!error_bar, 
                         color = col_group), size = 1, fatten = 1.75) +
     #geom_point(aes(color = col_group), size=3.5) +
     geom_smooth(method = "lm", se=FALSE, color = "black") +
-    theme_ipsum() +
+    theme_ipsum_rc() +
     labs(..., title = title, subtitle = paste0("slope = ", signif(regress$slope,2),
                              " p = ", signif(regress$p.value, 2))) +
     scale_color_manual(values = c("red3","darkblue")) + 
@@ -107,9 +118,9 @@ wq_trend_gg <- function(df, wqparam,
                        breaks = c(1990,1995,2000,2005,2010,2015),
                        minor_breaks = NULL,
                        limits = c(1993,2016)) +
-    scale_y_continuous(labels = c(-2, -1, 0, 1, 2),
-                       breaks = c(-2, -1, 0, 1, 2),
-                       limits = c(-2.75, 2.75))
+    scale_y_continuous(labels = label_break,
+                       breaks = label_break,
+                       limits = limits)
   
   list(gg, kt, df2, regress)
 }

### 2020-01-31 R/schematic_anomaly.R

@@ -1,0 +1,27 @@
+library(readr)
+library(dplyr)
+library(here)
+
+schematic <- read_csv(here("data/schematic.csv"))
+
+station_year_mean <- schematic %>% 
+  group_by(station, year) %>%
+  summarize(station_year_mean = mean(meas)) %>%
+  round(2) %>%
+  ungroup()
+
+station_year_anomaly <- station_year_mean %>%
+  group_by(station) %>%
+  mutate(long_term_mean = mean(station_year_mean),
+         long_term_sd = sd(station_year_mean)) %>%
+  mutate(station_year_anomaly = station_year_mean - long_term_mean)
+  round(2) %>%
+  ungroup()
+  
+mean_anomaly <- station_year_anomaly %>%
+  group_by(year) %>%
+  summarize(mean_anomaly = mean(station_year_anomaly),
+            sd_anomaly = sd(station_year_anomaly),
+            n = n()) %>%
+  round(2) %>%
+  ungroup()

### 2020-01-31 R/summary_stats.R

@@ -17,7 +17,7 @@ ww_param_ranges <- ww_all %>%
             max = round(max(station_year_mean, na.rm = T), 2),
             sd = round(sd(station_year_mean, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l", "molar", "µg/l")) %>%
+  mutate(units = c("°C", "µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 
@@ -59,7 +59,7 @@ ww_param_anomaly_ranges <- ww_all %>%
             max = round(max(measurement_anmly, na.rm = T), 2),
             sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
   ungroup() %>%
-  mutate(units = c("celsius", "µg/l", "µg/l", "molar", "µg/l")) %>%
+  mutate(units = c("°C", "µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
          "25th Percentile" = percentile_25,
          Mean = mean, Median = median, 

### 2020-02-11 manuscript/manuscript_zscore.Rmd
Ready to submit pre-print

--- 2019-11-18 manuscript/manuscript.Rmd
@@
 The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively.
-We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e.
-scaled anomalies) calculated on a per-station basis.
-Sub-regionally, temperature and chlorophyll a increased from 1993 to 2016.
+We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged z-scores (i.e., scaled anomalies) calculated on a per-station basis.
+Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016.
 Total nitrogen shows a weak increase driven by low years in the early 1990s.
 Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
-At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll a, total nitrogen, total phosphorus, and N:P all stable over time.
+At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.
 In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
@@
 
-For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e.
-z-scores) to find increasing or decreasing annual water quality trends.
+For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on centered and scaled water quality values (i.e., z-scores) to find increasing or decreasing annual water quality trends.
 Details are outlined below.
@@
 Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
-For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw-values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
 The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.
-This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, show change over the specified reference period.
+This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.
 
@@
 We used these z-scores to examine each water quality parameter for a trend over the time period of 1993 to 2016.
-Furthermore, since we are interested in water quality trends over time at individual sites (i.e.
-lakes), z-scores were calculated over the reference period, 1993-2016, for each site.
+Furthermore, since we are interested in water quality trends over time at individual sites (i.e., lakes), z-scores were calculated over the reference period, 1993-2016, for each site.
 We refer to this approach as "site-specific z-scores".
@@
 We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
-For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g.
-$\beta$~1~ = 0).
+For each parameter, we fit a regression line to the z-scores as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).
 Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].
@@
 
-Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll \\*a\\* (average z-score).
+Average annual scaled temperature in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with average temperature greater than the long-term average occurring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average z-score) in Rhode Island lakes and reservoirs.
+Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores.
+Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`).
+Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred since 2010 (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (average z-score).
 Panel A.
-Yearly averaged chlorophyll \\*a\\* z-scores from the URI Watershed Watch data.
+Yearly averaged chlorophyll a z-scores from the URI Watershed Watch data.
 Panel B.
-Yearly averaged chlorophyll \\*a\\* z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
+Yearly averaged chlorophyll a z-scores from the LAGOSNE dataset.
+Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores.
+Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).
 
@@
 Mean annual trends for nutrients were weaker or showed no trend over time.
-The data suggest a positive trend in TN (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score).
+The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average z-score).
 Panel A.
@@
 Yearly averaged TN z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-Since 1995, the yearly trend shows a much lower increase over time (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).
-TP does not show a trend over time in the yearly z-scores (slope: `r round(ww_tp_rl$slope,3)` , p-value: `r round(ww_tp_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
+Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores.
+Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+Since 1995, the yearly trend shows a much lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).
+TP does not show a trend over time in the yearly z-scores (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average z-score).
 Panel A.
@@
 Yearly averaged TP z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
-The pattern is the same for the TN:TP ratio (slope: `r round(ww_np_rl$slope,3)`, p-value: `r round(ww_np_rl$p.value,3)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).
+Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores.
+Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average z-score).
 Panel A.
@@
 Yearly averaged TN:TP ratio z-scores from the LAGOSNE dataset.
-Points are averages and ranges are standard deviations with blue indicating an average below the long-term mean and red indicating an average above the long-term mean.", display = "cite")`A.).
+Points are averages of site-specific z-scores and ranges are standard deviations of site-specific z-scores.
+Blue indicates yearly site-specific z-scores that were, on average, below the site-specific long-term means.
+Red indicates yearly site-specific z-scores that were, on average, above the site-specific long-term means.", display = "cite")`A.).
 Data for all figures are available as a comma-separated values file, `yearly_average_z.csv` from "fill in with zenodo link" [@zenodo_citation].
@@
 In general, there was little evidence to suggest broad regional changes.
-Chlorophyll *a* showed a very weak negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope: `r round(lagos_np_rl$slope, 3)`, p-value: `r round(lagos_np_rl$p.value, 5)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+Chlorophyll *a* showed a very weak negative trend (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight increasing trend (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight negative trend (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 
-Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e.
-neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can can occur.
+Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e., neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
@@
 
-## Data Analysis Approach
+## Data analysis approach
 
@@
 We built on these methods and adapted them for use with long-term water quality trends.
-While other methods are valid and robust [e.g.
-@Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
+While other methods are valid and robust [e.g., @Oliver_2017], we chose averaged site-specific z-scores as they can provide readily interpretable results, especially for communicating to general audiences.
 In addition, the site-specific z-scores are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).

### 2020-02-11 manuscript/peerj_preprint/manuscript.Rmd
Ready to submit pre-print

--- 2020-01-31 manuscript/manuscript_anomaly_only.Rmd
@@
 The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively.
-We assess regional (LAGOSNE) and state (URIWW) trends with yearly averaged anomalies calculated on a per-station basis.
+We assess regional (LAGOSNE) and state (URIWW) trends with yearly mean anomalies calculated on a per-station basis.
 Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016.
@@
 At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.
-In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, in spite of stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
+In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
 This analysis suggests an association between lake temperature and primary production.
@@
 Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g., May through October).
-The entire effort follow rigorous quality control/quality assurance protocols.
+The entire effort follows rigorous quality control/quality assurance protocols.
 These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].
@@
 For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen.
-Water samples are collected three times per season (May through October) to be analyzed for nutrients.
+Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.
 
@@
 In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
-For each of these parameters, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
+As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.
@@
 
-Methods for calculating the site-specific anomalies and the yearly averages are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
+Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
 Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].
@@
 For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).
-The slope of this line provides information on the average yearly change of that paramter over the time period studied.
+The slope of this line provides information on the mean yearly change of that paramter over the time period studied.
 Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].
@@
 
-Average annual temperature anomalies in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with average temperature greater than the long-term average occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (average anomaly) in Rhode Island lakes and reservoirs.
-Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Mean annual temperature anomalies in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with mean temperature greater than the long-term mean occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs.
+Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
 Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (average anomaly).
+Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).
 Panel A.
-Yearly averaged chlorophyll a anomalies from the URI Watershed Watch data.
+Yearly mean chlorophyll a anomalies from the URI Watershed Watch data.
 Panel B.
-Yearly averaged chlorophyll a anomalies from the LAGOSNE dataset.
-Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Yearly mean chlorophyll a anomalies from the LAGOSNE dataset.
+Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
@@
 Mean annual trends for nutrients were weaker or showed no trend over time.
-The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than average TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (average anomaly).
+The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than mean TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly).
 Panel A.
-Yearly averaged TN anomalies from the URI Watershed Watch dataset.
+Yearly mean TN anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly averaged TN anomalies from the LAGOSNE dataset.
-Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Yearly mean TN anomalies from the LAGOSNE dataset.
+Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
@@
 Since 1995, the yearly trend shows a lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).
-TP does not show a trend over time in the yearly anomalies (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the average are more evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (average anomaly).
+TP does not show a trend over time in the yearly anomalies (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the mean are more evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).
 Panel A.
-Yearly averaged TP anomalies from the URI Watershed Watch dataset.
+Yearly mean TP anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly averaged TP anomalies from the LAGOSNE dataset.
-Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Yearly mean TP anomalies from the LAGOSNE dataset.
+Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
 Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
-The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (average anomaly).
+The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).
 Panel A.
-Yearly averaged TN:TP ratio anomalies from the URI Watershed Watch dataset.
+Yearly mean TN:TP ratio anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly averaged TN:TP ratio anomalies from the LAGOSNE dataset.
-Points are averages of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
+Yearly mean TN:TP ratio anomalies from the LAGOSNE dataset.
+Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
@@
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
-Chlorophyll has increased on average `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.
+Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.
 This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.
@@
 We built on these methods and adapted them for use with long-term water quality trends.
-While other methods are valid and robust [e.g., @Oliver_2017], we chose averaged site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.
+While other methods are valid and robust [e.g., @Oliver_2017], we chose mean site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.
 For instance, reporting the changes in anomalies allows us to look at changes in the original units.
-With our analysis, the slope of the regression line for temperature suggests an average yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows and average yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
+With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
 Additionally, the site-specific anomalies are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).

### 2020-02-11 manuscript/peerj_preprint/manuscript.Rmd
closer for pre-print

@@
 
-![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../figures/ww_map.jpg)
+![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../../figures/ww_map.jpg)
 
@@
 
-![`r fig_num("anomaly_schematic")`](../figures/anomaly_schematic.jpg)
+![`r fig_num("anomaly_schematic")`](../../figures/anomaly_schematic.jpg)
 
@@
 
-![`r fig_num(name = "temp_ri")`](../figures/ww_temp_trends_anomaly.jpg)
+![`r fig_num(name = "temp_ri")`](../../figures/ww_temp_trends_anomaly.jpg)
 
-![`r fig_num(name = "chla_trend")`](../figures/paired_chla_trends_anomaly.jpg)
+![`r fig_num(name = "chla_trend")`](../../figures/paired_chla_trends_anomaly.jpg)
 
-![`r fig_num(name = "tn_trend")`](../figures/paired_tn_trends_anomaly.jpg)
+![`r fig_num(name = "tn_trend")`](../../figures/paired_tn_trends_anomaly.jpg)
 
-![`r fig_num(name = "tp_trend")`](../figures/paired_tp_trends_anomaly.jpg)
+![`r fig_num(name = "tp_trend")`](../../figures/paired_tp_trends_anomaly.jpg)
 
-![`r fig_num(name = "np_trend")`](../figures/paired_np_trends_anomaly.jpg)
+![`r fig_num(name = "np_trend")`](../../figures/paired_np_trends_anomaly.jpg)
 

### 2020-02-11 manuscript/preprint/manuscript.Rmd
preprint submission

--- 2020-02-11 manuscript/peerj_preprint/manuscript.Rmd
@@
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation].
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation].
 
@@
 Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
-Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from "fill in with zenodo link" [@zenodo_citation].
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
 
@@
 Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
-Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from "fill in with zenodo link" [@zenodo_citation].
+Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
 

### 2020-02-12 manuscript/ecosphere/manuscript.Rmd
working on ecosphere submission!

--- 2020-02-11 manuscript/preprint/manuscript.Rmd
@@
 ---
-title: "Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
+title: "**Title:** Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
 author:
-- Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers,
-  S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A.
-  ^*2*^
+- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
 - ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
@@
 - ^*3*^ ORISE, Narragansett, RI 02882
-- '^\*^  *corresponding author: hollister.jeff@epa.gov*'
+- '**Corresponding author: hollister.jeff@epa.gov**'
+keywords: '**keywords: ** This, is, a test'
 bibliography: manuscript.bib
@@
 csl: ecology.csl
-abstract: 'Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management.
+abstract: |
+  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management.
 Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets.
@@
 This analysis suggests an association between lake temperature and primary production.
-Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.'
+Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.
+  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
 ---
@@
 fig_num <- captioner(prefix = "Figure")
-options(scipen = 2, digits = 5)
+options(scipen=999)
 ```
@@
 
-![`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`](../../figures/ww_map.jpg)
-
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
@@
 
-### Summarizing site-specific anomalies
-
-Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
+*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
 Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
-
-![`r fig_num("anomaly_schematic")`](../../figures/anomaly_schematic.jpg)
 
@@
 
-### Linear regression on annual mean anomalies
-
-Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
+*Linear regression on annual mean anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
 We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
@@
 
-### Comparison of Rhode Island to the region
-
-Oliver et al.
+*Comparison of Rhode Island to the region:* Oliver et al.
 [-@Oliver_2017] used hierarchical linear models and showed relatively stable water quality in the lakes of the northeastern United States.
@@
 
-```{r ww_params, echo=FALSE, results="asis"}
-pander::pander(ww_param_ranges[,c(-3,-6)], style = "rmarkdown")
-```
-*`r tab_num("ww_params")`*
-
 For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
-
-```{r lagos_params, echo=FALSE, results="asis"}
-pander::pander(lagos_param_ranges[,c(-3,-6)], style = "rmarkdown")
-```
-*`r tab_num("lagos_params")`*
 
@@
 
-![`r fig_num(name = "temp_ri")`](../../figures/ww_temp_trends_anomaly.jpg)
-
-![`r fig_num(name = "chla_trend")`](../../figures/paired_chla_trends_anomaly.jpg)
-
-![`r fig_num(name = "tn_trend")`](../../figures/paired_tn_trends_anomaly.jpg)
-
-![`r fig_num(name = "tp_trend")`](../../figures/paired_tp_trends_anomaly.jpg)
-
-![`r fig_num(name = "np_trend")`](../../figures/paired_np_trends_anomaly.jpg)
-
 ## Regional trends in water quality
@@
 
-# Discussion and conclusions
+# Discussion and Conclusions
 
@@
 
+# Acknowledgments
+
+We would like to thank Anne Kuhn (US EPA, Atlantic Coastal Environmental Sciences Division), Nate Merrill (US EPA, Atlantic Coastal Environmental Sciences Division), Jim Hagy (US EPA, Gulf Ecosystem Measurement and Modeling Division), Joe LiVolsi (US EPA, Atlantic Coastal Environmental Sciences Division), Tim Gleason (US EPA, Atlantic Coastal Environmental Sciences Division), and Wayne Munns (US EPA, Atlantic Coastal Environmental Sciences Division) for constructive reviews of this paper.
+The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.
+Environmental Protection Agency.
+Any mention of trade names, products, or services does not imply an endorsement by the U.S.
+Government or the U.S.
+Environmental Protection Agency.
+The EPA does not endorse any commercial products, services, or enterprises.
+This contribution is identified by the tracking number ORD-032695 of the Atlantic Coastal Environmental Sciences Division, Office of Research and Development, Center for Environmental Measurement and Modeling, US Environmental Protection Agency.
+
 # Bibliography
+
+<div id="refs"></div>
+######
+# Tables
+
+```{r ww_params, echo=FALSE, results="asis"}
+pander::pander(ww_param_ranges[,c(-3,-6)], style = "rmarkdown")
+```
+`r tab_num("ww_params")`
+
+```{r lagos_params, echo=FALSE, results="asis"}
+pander::pander(lagos_param_ranges[,c(-3,-6)], style = "rmarkdown")
+```
+`r tab_num("lagos_params")`
+
+######
+
+# Figures
+
+![](../../figures/ww_map.jpg)
+`r fig_num("ww_map", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`
+
+######
+
+![](../../figures/anomaly_schematic.jpg)
+`r fig_num("anomaly_schematic")`
+
+######
+
+![](../../figures/ww_temp_trends_anomaly.jpg)
+`r fig_num(name = "temp_ri")`
+
+######
+
+![](../../figures/paired_chla_trends_anomaly.jpg)
+`r fig_num(name = "chla_trend")`
+
+######
+
+![](../../figures/paired_tn_trends_anomaly.jpg)
+`r fig_num(name = "tn_trend")`
+
+######
+
+![](../../figures/paired_tp_trends_anomaly.jpg)
+`r fig_num(name = "tp_trend")`
+
+######
+
+![](../../figures/paired_np_trends_anomaly.jpg)
+`r fig_num(name = "np_trend")`

### 2020-02-12 R/figures_anomaly_only.R

@@ -51,7 +51,7 @@ ww_tp_rl <- tp_gg[[4]]
 temp_gg <- ww_data %>%
   #filter(state == "RI") %>%
   wq_trend_gg("temp", yvar = "measurement_anmly", 
-              y = "Average Yearly Temperature Anomaly", x = "Year",
+              y = "Average Yearly Temperature Anomaly (°C)", x = "Year",
               write = here("data/temp_data_anomaly.csv"), error_bar = "sd",
               title = "URI Watershed Watch Temperature")
 ggsave(here("figures/ww_temp_trends_anomaly.jpg"), temp_gg[[1]], width = 7.5, height = 5.625, 
@@ -132,29 +132,25 @@ lagos_np_rl <- lagos_np_gg[[4]]
 chla_plot <- cowplot::plot_grid(chla_gg[[1]], lagos_chla_gg[[1]], 
                                 align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_chla_trends_anomaly.jpg"), chla_plot, width = 7.5, 
-       height = 9, 
+ggsave(here("figures/paired_chla_trends_anomaly.jpg"), chla_plot, width = 5.7, height = 7.45, 
        units = "in", dpi = 600)
 
 tn_plot <- cowplot::plot_grid(tn_gg[[1]], lagos_tn_gg[[1]], 
                               align = "h", ncol = 1)
 
-ggsave(here("figures/paired_tn_trends_anomaly.jpg"), tn_plot, width = 7.5, 
-       height = 9, 
+ggsave(here("figures/paired_tn_trends_anomaly.jpg"), tn_plot, width = 5.7, height = 7.45, 
        units = "in", dpi = 600)
 
 tp_plot <- cowplot::plot_grid(tp_gg[[1]], lagos_tp_gg[[1]], 
                               align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_tp_trends_anomaly.jpg"), tp_plot, width = 7.5, 
-       height = 9, 
+ggsave(here("figures/paired_tp_trends_anomaly.jpg"), tp_plot, width = 5.7, height = 7.45, 
        units = "in", dpi = 600)
 
 np_plot <- cowplot::plot_grid(np_gg[[1]], lagos_np_gg[[1]], 
                               align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_np_trends_anomaly.jpg"), np_plot, width = 7.5, 
-       height = 9, 
[…]

### 2020-02-14 manuscript/ecosphere/manuscript.Rmd
acknowledgements updated

@@
 We would like to thank Anne Kuhn (US EPA, Atlantic Coastal Environmental Sciences Division), Nate Merrill (US EPA, Atlantic Coastal Environmental Sciences Division), Jim Hagy (US EPA, Gulf Ecosystem Measurement and Modeling Division), Joe LiVolsi (US EPA, Atlantic Coastal Environmental Sciences Division), Tim Gleason (US EPA, Atlantic Coastal Environmental Sciences Division), and Wayne Munns (US EPA, Atlantic Coastal Environmental Sciences Division) for constructive reviews of this paper.
-The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.
+The URIWW program has received generous support from many different organization over its 25+ year history and we are grateful for this support.
+In particular, RI DEM and URI’s Cooperative Extension have provided sustained support for the program and deserve special mention.
+In part, this work was supported by the USDA National Institute of Food and Agriculture, McIntire Stennis project accession no.
+RI0020-MS984.
+Lastly, the views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.
 Environmental Protection Agency.

### 2020-07-21 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
only inlcude lakes with samples in may_jun, july_aug, sep_oct.  Need at least one in each.

---
output: html_document
editor_options: 
  chunk_output_type: console
---
```{r include = FALSE}
library(knitr)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
```
# Reviewer 1

## Major Comments

*1-1.* I would like more detail on how the averaging was done for each year. For each parameter, an annual average was calculated for each lake. In theory, Secchi depth and temperature were sampled every week and water samples were collected 3 times per season, May through October. If the monitoring design was followed consistently, then a simple average should be sufficient. However, I know from experience with other state databases that there are often gaps in sampling. If some years were represented with only samples late in the season and others only had samples early in the season, then seasonal trends could be misinterpreted as annual trends. How many samples were required per year to calculate an annual average and were there any additional measures taken to ensure that the data collected over the course of the season was weighted similarly between years (e.g., must have at least one sample in May/June, July/Aug, and Sept/Oct)? I think there should be some rules to this effect. If this is not a concern, please state some statistics showing that the majority (x%) of lake-years included in the analysis adhered to the full sampling schedule.

```{r include=FALSE}

ww_dat <- read_csv(here::here("data/ww_all_mostly_cleaned.csv"))

param_min_samp <- ww_dat %>%
  filter(month >= 5 & month <= 10) %>%
  select(station_name, year, param, month) %>%
  mutate(month = month(month, label = TRUE)) %>% 
  group_by(station_name, year, param, month) %>%
  summarize(num = n()) %>%
  ungroup() %>%
  pivot_wider(names_from = month, values_from = num) %>%
  group_by(station_name, year, param) %>%
  mutate(may_jun = sum(May, Jun, na.rm = TRUE),
         jul_aug = sum(Jul, Aug, na.rm = TRUE),
         sep_oct = sum(Sep, Oct, na.rm = TRUE),
         min_samp = sum(may_jun, jul_aug, sep_oct)>=3,
         least_two = sum(may_jun, jul_aug, sep_oct)>=2) %>%
  ungroup() %>%
  select(station_name, year, param, min_samp, least_two) %>%
  group_by(param) %>%
  summarize(percent_min_samp = sum(min_samp)/n()) 
```

- **Response:**  Good catch!  Fortunately, it has not greatly impacted the analysis.  The table below lists the percentage of lake-years for each parameter that had at least one sample in each of May/June, July/Aug, and Sept/Oct.

```{r echo = FALSE}
knitr::kable(param_min_samp)
```

While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.  This filtering step has been added to our `data_clean.R` script and we remove any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.  The changes to the analysis are minimal and do not alter any of our conclusions.

*1-2.* I also had questions regarding the way that interannual data were represented. At least 10 years of data from 1993 to 2016 had to be available for a lake to be included in the analysis (lines 121-122). This is reasonable. However, this is less than half of the years of the entire period of record. In theory, one lake could be analyzed from 1993 to 2002 and another lake from 2007 to 2016. These are not even overlapping periods of record. I think you need to take it one step further and ensure that there are years represented at the beginning and end of this 24-year period. Oliver et al. 2017 did include rules to this effect. I do not think your analytical method is immune to this possible difference in time frame. Again, please describe the time frames of your data set and adjust your rules if you find that you do have a substantial number of lakes with non-overlapping time frames.

- **Response:**

*1-3.* How did Secchi depth change? Why wasn’t this analyzed?

- **Response:**

*1-4.* It is unclear how the trends in temperature, TP, chl-a, TN, and TN:TP compare to water quality condition and any standards that might be set in Rhode Island. Tables 1 and 2 give the range of values across lakes and there is clearly a large gradient in trophic status. Following from the comment made in point 8, it would be helpful to better understand the trends observed here in context of water quality condition. If chl-a is rising, is it still relatively low, or is it reaching “polluted/degraded” levels? Do you observe a shift in trophic status in individual lakes? Somehow, try to put the trends or lack of trends observed in the context of how water quality condition is changing over time. Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.

- **Response:**

*1-5.* My biggest concern is possibly overstating the lack of trends in TP and TN and the positive trends in temperature and chl-a and then suggesting that temperature, not nutrients, are responsible for increasing chl-a. The authors are somewhat cautious when making this conclusion, but do not fully recognize the limitations of this interpretation. I will try to break my concerns into manageable pieces for the authors to respond to.

- **Response:**

*1-5a.* I think the title overstates this result. I agree with increasing Chl-a trends, but don’t agree with stable nutrient trends.

- **Response:**

*1-5b.* I have not read Wasserstein et al. 2016, but agree with the general concept that a specific p-value threshold is not the only weight of evidence for determining whether there is a trend. However, I do not think the authors conclusions about trends for each parameter in the Rhode Island vs. LAGOS datasets are consistent with one another need a more detailed framework laid out ahead of time to understand the more holistic way of interpreting whether there are trends. These are the trends that I interpret differently than do the authors:

- **Response:**

*1-5b-i.* TN in Rhode Island (Fig 5A) – The p-value is very small and the slope seems reasonably large. I also see more positive means later in the record. I do not think that 1993 and 1994 have an overly strong effect on the trend. Even when you take them out of the analysis, the slope is still 2.5 and p-value is small. Although the annual increase in TN is not large compared to the range of total TN anomalies, I interpret the stats as an increase in TN. In line 247, I disagree with the statement that removing data from 1993 and 1994 weakens the positive trend “considerably”.

- **Response:**

*1-5b-ii.* TN in Lagos data set (Fig 5B) – the text says that “TN showed a slight positive trend”. Here, the slope is smaller and the p-value is large. I would think that the positive slope is driven by the high TN values in 2013. I think there is more evidence for increasing TN in the Rhode Island data set than the LAGOS data set, but the text reads the opposite.

- **Response:**

*1-5b-iii.* TP in Rhode Island (Fig 6A) – The authors state that TP does not show a trend over time. I see a positive slope with a marginally significant p-value and more positive means later in the record. I wouldn’t oversell this as a major increasing trend in TP, but I see some evidence for a slight increase in TP.

- **Response:**

*1-5b-iv.* TP in LAGOS (Fig 6B) – “TP showed a slight decreasing trend”. I disagree. The slope was negative, but at -0.027, so a much smaller magnitude than the positive slope in the Rhode Island data set. The p-value was 0.32 and the graph itself looks like the slope is a horizontal line centered on 0.

- **Response:**

*1-5c.* To put the magnitude of each slope in context, you could report the slope as a percent of the full range of slopes observed in the data set. Even those trends that are “significant” have a relatively small slope compared to the variability between lakes.

- **Response:**

*1-6.* This analysis also seems to be saying something about coherence. In some years, there is a wide range of anomalies, meaning that some lakes are much higher than their long-term average at the same time as other lakes are much lower. This seems like an interesting result in and of itself that is not pointed out in the results or discussion. How does the variability in annual anomalies of water quality between lakes compare to that observed in the climate
literature, which traditionally uses this approach?

- **Response:**

*1-7.* In the discussion (lines 236-237) and in the title, the authors seem to want to conclude that because there were positive trends in temperature and chl-a but not in TP or TN (which I don’t agree as strongly with), that increasing chl-a is likely caused by increasing temperatures and not nutrients. This is certainly interesting if it were the case, but it is well beyond the scope of what can be concluded here.

- **Response:**

*1-7a.* This study simply lists the direction of trends of 5 parameters, but does not analyze correlations between them.

- **Response:**

*1-7b.* Even if it had tested correlations between parameters, it is still difficult to reach for causation.

- **Response:**

*1-7c.* The analysis of mean anomalies over time is interesting, but that simply means that the entire set of lakes in the state shows trends over time. What I’ve learned from reading some of the papers cited by this study and from my own investigations is that lakes near one another can have opposite trends. Processes at the scale of the individual lake and its watershed are important and can have opposite trends in neighboring watersheds. I think this study would benefit by adding an analysis at the individual lake scale to compliment the state scale. I would feel more comfortable with the statement that increasing chl-a may be a result of warming waters, not nutrients, if this was also tested at the lake scale.  One could build a mixed effects model:

- Chl-a anomaly ~ temperature anomaly + TP anomaly + TN anomaly + lake

where lake is a random effect that has a unique id for each lake in the study. Years would be replicates. I think it’s possible that you could see a nutrient effect if you examined the data this way. There is a lot of variability between lakes within a year, so while I find the statistical method used in this study interesting, I don’t think it tests of drivers of change.

- **Response:**

*1-8.* Line 274-275: the authors pose the interesting question about whether lakes of similar trophic state are showing similar trends. I disagree that this test is beyond the scope of the study. I think it is a timely question that would be good to test following the Stoddard et al. paper. It should be relatively easy to test – dividing the lakes into groups of hypereutrophic, eutrophic, mesotrophic, and oligotrophic categories based on their long-term mean TP or chl-a. If there are not enough lakes in each category, with would be interesting to compare a few.

- **Response:**

## Editorial Comments

*1-9.* Line 64: Better define the spatial scales of the studies referenced in this paragraph. Macro-scale, regional, subregional could be interpreted in different ways. Please be explicit about the geographic region that each study includes.

- **Response:**

*1-10.* Line 71: I’m not sure why it was surprising that little change in nutrients and chl-a was reported in Oliver et al. 2017? How is this study different from Lottig and Collins? All three use the LAGOS data set. I think it would be helpful to be more specific about both the spatial extent and the results of these studies – particularly when you talk about complexity in discussing the Lottig and Collins papers. Complex just doesn’t explain the patterns very well – please be more specific in describing what complex means in this case.

- **Response:**

*1-11.* Lines 99 – 102: please delete. Unnecessary.

- **Response:**

*1-12.* Line 113: capitalize Secchi

- **Response:**

*1-13.* Line 123: delete approximately. It seems that you are giving specific numbers here.

- **Response:**

*1-14.* Lines 202-207: The text simply repeats the numbers reported in tables 1 and 2. It would be more interesting to compare the populations at the two scales – in terms of the means and the ranges and the variety of trophic states represented and to also summarize the comparison of trends over time at the two scales.

- **Response:**

*1-15.* Line 285: “receive feedback” is a vague phrase here. Perhaps you mean that is important for managers and stakeholders to know what the long-term trends are? I think you could also add more references and specific language about why it is important to share results with volunteers.

- **Response:**

*1-16.* Line 292: I was a little bit confused by this paragraph. Why discuss ways of reducing nutrients when you have not identified increased nutrients as having an increasing trend? There are a lot of leaps made to get to this paragraph – temperature affects chl-a more than nutrients, resultant chl-a and nutrient values are too high, lower nutrients can offset warming temperatures, ways to lower nutrients. Some of the analyses I suggest adding above may better
motivate this paragraph being included in the discussion section.

- **Response:**

# Reviewer 2

## Major Comments

*2-1.* The paper needs some contextual description of the RI lakes that could be achieved from GIS analysis. For example, what is the distribution of size and depth of the lakes in the data set? What are the land-use area percentages (e.g. agriculture, residential, and forested areas) within the catchment areas of each lake in the data set? A contextual description will be of value to other researchers who may wish to compare these results to other areas. Smaller lakes potentially could show a different warming trend than larger lakes, for example, and therefore knowing a lake size distribution could help qualify the results. Some context on the land-use patterns is also important for understanding the nutrient-level data, for comparing results between study areas, and perhaps shedding some light on why the Rhode Island mean water quality values are less than mean values for the larger, LAGOSNE dataset (p. 11, lines 205-207). At a minimum, an overview description of the lakes that make up the dataset is needed.

- **Response:**

*2-2.* There should be no reason to rely on “anecdotal evidence” for explaining warming air temperatures over the timeframe of the investigation (p. 14, lines 269-272). There is no source/reference for this claim. Weather station data for RI should be available to assess whether regional climate is warming and to see if it is consistent with the observed increases in lake temperature.

- **Response:**

*2-3.* Some clarification on the methods is also needed. The authors say that they used the same analytical methods for the LAGOSNE data set as they did for the Rhode Island lakes. Please verify whether the LAGOSNE data represent the same seasonal timeframe (May – October) and lake sample depth (< or = 2m) as the Rhode Island data (as they should for a comparative analysis). Additionally, please state why the LAGOSNE data extend only until 2014 vs. the year 2016 for the RI dataset.

- **Response:**

## Editorial Comments

*2-4.* p. 6, line 90: Capitalize “States.”

- **Response:**

*2-5.* p. 10, line 188: Check spelling of “parameter.”

- **Response:**

### 2020-07-21 manuscript/ecosphere_resubmit/manuscript.Rmd
only inlcude lakes with samples in may_jun, july_aug, sep_oct.  Need at least one in each.

--- 2020-02-14 manuscript/ecosphere/manuscript.Rmd
@@
 author:
-- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
 - ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
@@
 - ^*3*^ ORISE, Narragansett, RI 02882
+- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
 - '**Corresponding author: hollister.jeff@epa.gov**'
@@
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation].
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
 
@@
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
-[-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017].
+[-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].
 Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
@@
 `r fig_num(name = "np_trend")`
+
+```{r session_info, echo=FALSE}
+#Send session info to a file: 
+capture.output(sessionInfo(),file=here::here("manuscript/ecosphere/sessioninfo.txt"))
+```

### 2020-07-21 R/data_clean.R

@@ -158,7 +158,32 @@ idx <- ww_lake_trend_data$year == "2016" & ww_lake_trend_data$month == 7 & ww_la
 idx <- !idx
 ww_lake_trend_data <- ww_lake_trend_data %>% 
   filter(idx)
-  
+
+write_csv(ww_lake_trend_data, here("data/ww_all_mostly_cleaned.csv"))
+
+# Reviewer 1 comment: To avoid possible seasonal effects due to missed samples, 
+# make sure to include only lake-years with a minimum of one sample in each of
+# May/June, Jul/Aug, Sep/Oct per parameter.
+idx_min_samp <- ww_lake_trend_data %>%
+  mutate(may_jun = case_when(month == 5 | month == 6 ~
+                             1, 
+                           TRUE ~ 0),
+       jul_aug = case_when(month == 7 | month == 8 ~
+                             1,
+                           TRUE ~ 0),
+       sep_oct = case_when(month == 9 | month == 10 ~
+                             1,
+                           TRUE ~ 0)) %>%
+  group_by(station_name, year, param) %>%
+  summarize(min_samp = sum(may_jun, jul_aug, sep_oct)>=3) %>% 
+  right_join(ww_lake_trend_data) %>%
+  pull(min_samp)
+
+ww_lake_trend_data <- ww_lake_trend_data %>%
+  filter(idx_min_samp)
+    
+
+# Calculates station/year/param stats
 ww_lake_trend_data <- ww_lake_trend_data %>%
   filter(month >= 5 & month <= 10) %>%
   group_by(station_name, year, param) %>%
@@ -241,7 +266,31 @@ lagos_data <- lagosne_select(table = "epi_nutr",
   #filter(year >= 1990) %>%
   #filter(`Station Name` %in% filter_year(., 20)) %>% #moved to plots and 10 years is min...
   filter(month >= 5 & month <= 10) %>%
-  select(station_name = `Station Name`,year, month, day, param, measurement) %>%
+  select(station_name = `Station Name`,year, month, day, param, measurement) 
+
+# Reviewer 1 comment: To avoid possible seasonal effects due to missed samples, 
+# make sure to include only lake-years with a minimum of one sample in each of
+# May/June, Jul/Aug, Sep/Oct per parameter.
+idx_min_samp <- lagos_data %>%
+  mutate(may_jun = case_when(month == 5 | month == 6 ~
+                               1, 
+                             TRUE ~ 0),
+         jul_aug = case_when(month == 7 | month == 8 ~
[…]

### 2020-07-21 R/functions.R

@@ -1,7 +1,7 @@
 # Installations
 pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
           "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
-          "USAboundariesData","knitr","captioner", "pander")
+          "USAboundariesData","knitr","captioner", "pander", "cowplot")
 
 for(i in pkgs){
   if(!i %in% installed.packages()){

### 2020-07-28 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
making progress on edits

@@
 While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.
-This filtering step has been added to our `data_clean.R` script and we remove any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.
+This filtering step has been added to our `data_clean.R` script at lines 163 and 259 and is in the `filter_months()` function in `functions.R`.
+This removes any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.
 The changes to the analysis are minimal and do not alter any of our conclusions.
+We added text in the second paragraph of the Methods section describing the selection criteria.
 
@@
 
-- **Response:**
+- **Response:**  Similar to the approach in Oliver et al.
+we have added an additional step to ensure that lakes are only included if they have data in the first and second half of the 1993 - 2016 time frame and now only lakes that have at least one measurement in 1993-2004 and 2005-2016, have a minimum of 10 years of data, and have at least 1 measurement in may/june, july/aug, sept/oct are included in the analysis.
+The filtering step has been added to the `data_clean.R` around line 233 and 316 and uses a new function, `filter_early_late()` which is in the `functions.R` file.
+The impact on the analysis was minimal.
+We added text in the second paragraph of the Methods section describing the selection criteria.
 
@@
 
-- **Response:**
+- **Response:** As we were comparing the results from our analysis in Rhode Island to the LAGOSNE data and specifically to the results found in the Oliver et al 2017 paper, we constrained the analysis to a similar set of parameters.
+The Oliver et al 2017 paper did not include secchi in their analysis and we decided to follow suit.
 
@@
 Somehow, try to put the trends or lack of trends observed in the context of how water quality condition is changing over time.
-Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.
 
-- **Response:**
+- **Response:** NEEDS A RESPONSE!
+
+*1-4a.* Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.
+
+- **Response:** NEEDS A RESPONSE!
 
@@
 
-- **Response:**
+- **Response:** We address the reviewers concern regarding the possible association between rising temperature and increasing chlorophyll in the individual pieces below.
+In short, we concur and have done all we can to avoid the appearance of suggesting causation.
 
@@
 
-- **Response:**
+- **Response:** NEEDS A RESPONSE!
 
@@
 
-- **Response:**
+- **Response:** We like the reviewers concept of a more holistic way of interpreting trends.
+We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.
+We now lay out a more explicit approach and describe how we use sign of the slope, magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.
+This is described in the paper in a new section in the methods, *Assessing regressions for trends*.
+One of the biggest changes is to suggest values for the slope that would indicate an ecologically meaningful trend.
+We describe this in the new section, but in general we are using a trophic state change from oligotrophic to eutrophic over one century to indicate ecologically meaningful change.
+For temperature we look at values associated with ecological change as well as policy values (e.g.
+the Paris Agreement).
+While these exact numbers are not universal, we use them as a starting point for assesing the trends.
+How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below.
 
@@
 
-- **Response:**
+- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*).
+This weakened the positive trend somewhat as the slope is smaller and the p-value larger.
+Additionally two years, 1995 and 1996, now have insufficient data to be included in the figures.
+Using the more explicit definition of a trend we not conclude that TN showed a weak statistically positive trend but not an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.
+This is reflected in the manuscript.
 
@@
 
-- **Response:**
+- **Response:** We applied the same within and across year filters to the LAGOSNE and these results have changed as well, largely becuase nutrient data across the full time frame was scarce in LAGOSNE in more recent years.
+Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing weak statistical support for a negative change in nitrogen, more below average years recently.
+The slope is less than the 3.5 we identified, albeit close.
+Thus we conclude a weak decrease in TN at the scale of the Region.
+This is indicated in the manuscript.
 
@@
 
-- **Response:**
+- **Response:** Again, with the changes to the filtering the trend lines have slightly changed.
+We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, a value greater than the threshold we identified and greater frequency of above average years in recent years.
+We have indicated as such in the text.
 
@@
 
-- **Response:**
+- **Response:** This is a good suggestion, but we have chosen not to implement it, primarily because of Reveiwer 1's suggestion to have a more holistic approach to addressing the trends.
+In developing our approach to assessing the trends we now have ecologically relevant thresholds we can compare to.
+Both approaches address the same issue of putting the magnitude of the slopes in context.
 
@@
 This seems like an interesting result in and of itself that is not pointed out in the results or discussion.
-How does the variability in annual anomalies of water quality between lakes compare to that observed in the climate
-literature, which traditionally uses this approach?
+How does the variability in annual anomalies of water quality between lakes compare to that observed in the climate literature, which traditionally uses this approach?
 
-- **Response:**
+- **Response:** NEEDS A RESPONSE!
 
@@
 
-- **Response:**
+- **Response:**  We have softened our language in the discussion on original lines 236-237 considerably.
+They now read, "While our analysis is not capable of detecting causation, this association between increasing chlorophyll *a* and warming waters is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature alone."
 
@@
 
-- **Response:**
+- **Response:** NEEDS A RESPONSE!
+
+```{r, echo=FALSE}
+x <- ww_data %>% 
+  select(station_name, site_descr, town, year, param, station_year_mean) %>%
+  spread(param, station_year_mean) %>%
+  filter(complete.cases(.))
+xyr <- group_by(x, station_name) %>% 
+  summarize(yrly_mean_chla = mean(chla),
+            yrly_mean_temp = mean(temp),
+            yrly_mean_np = mean(np_ratio),
+            yrl_mean_tn = mean(total_n),
+            yrly_mean_tp = mean(total_p))
+```
 
@@
 
-- **Response:**
+- **Response:** We have added a new analysis to look at the same plots across trophic states.
 
@@
 
-- **Response:**
+- **Response:** Fixed
 
@@
 
-- **Response:**
+- **Response:** Fixed.
 
@@
 
-- **Response:**
+- **Response:** Fixed.

### 2020-07-28 manuscript/ecosphere_resubmit/manuscript.Rmd
making progress on edits

@@
 - '**Corresponding author: hollister.jeff@epa.gov**'
-keywords: '**keywords: ** This, is, a test'
 bibliography: manuscript.bib
@@
 First, are there state-wide trends in total nitrogen (TN), total phosphorus (TP), total nitrogen to total phosphorus ratio (TN:TP), chlorophyll *a*, and lake temperature?
-Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United states?
+Second, are water quality trends in Rhode Island similar to regional trends in the northeastern United States?
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
@@
 # Methods
-
-For this study, we combined a long-term dataset on water quality of lakes in Rhode Island with a trend analysis based on water quality anomalies (i.e., measured values with the long term mean subtracted) to find increasing or decreasing annual water quality trends.
-Details are outlined below.
 
@@
 Data quality assurance and control is treated with paramount importance; volunteers are trained both in the classroom and the field, regular quality checks occur, and volunteers are provided with all the necessary equipment and supplies, along with scheduled collection dates.
-For freshwater lakes and reservoirs, weekly secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen.
+For freshwater lakes and reservoirs, weekly Secchi depth and water temperature are recorded, along with bi-weekly chlorophyll *a* and in deep lakes (greater than 5 meters) dissolved oxygen.
 Water samples are collected three times per season (May through October) to be analyzed for nutrients and bacteria.
@@
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-In particular, we selected URIWW data that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) and were sampled at a depth of 2 meters or less.
+In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event in each of May/June, July/August, and September/October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.
 As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
 The final dataset used in our analysis included `r nrow(ww_avail_data_locations)` lakes and reservoirs.
-Of these sites, our filtered dataset had approximately `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.
+Of these sites, our filtered dataset had `r length(temp_10yr_sites)` sites measured for temperature, `r length(chla_10yr_sites)` sites measured for chlorophyll *a*, `r length(total_n_10yr_sites)` sites measured for TN, and `r length(total_p_10yr_sites)` sites measured for TP.
 Of the `r nrow(ww_avail_data_locations)` sampling sites, `r sum(ww_avail_data_locations$params_avail == "chla, total_p, total_n, np, temp")` had data for all 5 parameters.
@@
 We used the same selction criteria on the LAGOSNE dataset as was applied to the URIWW data.
+The LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.
 
@@
 
+```{r get_year_n, echo=FALSE}
+year_num_ww <- figure_data %>%
+  filter(source == "uriww") %>%
+  select(year, variable, n) %>%
+  spread(variable, n)
+
+year_num_lagosne <- figure_data %>%
+  filter(source == "lagosne") %>%
+  select(year, variable, n) %>%
+  spread(variable, n)
+```
+
+After filtering and summarizing the data some years may not have sufficient number of sites to be included.
+We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data(`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).
+
 *Linear regression on annual mean anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
@@
 For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).
-The slope of this line provides information on the mean yearly change of that paramter over the time period studied.
-Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].
-Our interpretation of the trends attempts to follow this advice and we assess trends with the information provided by the magnitude of the slopes, the p-values, and our understanding of the processes involved.
+The slope of this line provides information on the mean yearly change of that parameter over the time period studied.
+
+*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].
+Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence.
+For this assessment, we are defining a trend as positive or negative with the following; 1) Direction of trend is determined by the sign of the slope, 2) p-value is used to inform interpretation and very small p-values suggest a statistically relevant trend, 3) relative frequency of high and low years in the beginning and end of the time frame, and 4) the magnitude of the slope suggests an ecologically relevant change.
+
+We recognize that ecologically relevant change is system dependent and no single value will be universally appropriate.
+We are using several values for this particular assessment, but other values could also be justified.
+There are various ways to identify important temperature changes in lakes and reservoirs.
+A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year.
+Alternatively, we could look at more ecologically focused temperature changes.
+For instance, Winder and Schindler [-@winder2004climate] saw disrupted trophic linkages between phytoplankton and zooplankton with 1.39°C over a 40 year data set, or 0.035°C per year.
+For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g.
+0.035°C per year for trophic linkages) and would also be indicative of meeting an important policy goal.
+For changes in chlorophyll, total phosphorus, and total nitrogen an ecologically meaningful trend is one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.
+We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].
+For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, the total phosphorus limit is 10 μg/L or 0.1 μg/L per year, and for total nitrogen the limit is 350 μg/L, or 3.5 μg/L.
+As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.
 
@@
 Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific anomaly and trend analysis approach outlined above.
+
+*Comparison of trends across trophic state:*  Just as
 
@@
 
-Mean annual temperature anomalies in lakes and reservoirs appears to be increasing (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`) with the majority of years with mean temperature greater than the long-term mean occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs.
+Mean annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests a statistical trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with mean temperature greater than the long-term mean are occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs.
 Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
@@
 Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).
-Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`) and with the exception of a slightly above-average year in 2003, the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).
+Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).
+The slope of 0.3 μg/L is much greater than our threshold of 0.035 μg/L and the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).
 Panel A.
@@
 ```{r no_early_tn, echo=FALSE}
-ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
+ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1),]) %>%
   tidy() %>%
@@
 
-Mean annual trends for nutrients were weaker or showed no trend over time.
-The data suggest a positive trend in TN (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`); however, that perceived trend is driven by the lower than mean TN values in 1993 and 1994 (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly).
+Mean annual trends for nutrients were not as clear.
+For total nitrogen, the slope of the line is positive with some statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) and the years greater than the mean are somewhat more frequent in more recent years.
+However, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly).
 Panel A.
@@
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
-Since 1995, the yearly trend shows a lower increase over time (slope = `r signif(ww_tn_rl_no9394$slope,2)`, p = `r signif(ww_tn_rl_no9394$p.value, 2)`).
-TP does not show a trend over time in the yearly anomalies (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over or under the mean are more evenly distributed over the years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
+Grey indicates yearly site specifica anomalies that were, on average equal to the long-term means.
+Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+We interpret this as a weak positive trend in TN, but with little ecological impact.
+Total phosphorus shows a positive trend over time in the yearly anomalies but with somewhat weak statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the mean are more common in recent years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).
 Panel A.
@@
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
-The pattern is the same for the TN:TP ratio (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) with little evidence suggesting a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
+Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is greater than our defined threshold of 0.1 μg/L and suggests a possible trophic state shift in less than a century.
+The pattern for the TN:TP ratio has a negative slope but no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).
 Panel A.
@@
 Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
+Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
 Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
@@
 
-In general, there was little evidence to suggest broad regional changes.
-Chlorophyll *a* showed a very weak positive trend (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B.), TP showed a slight decreasing trend (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.), TN showed a slight positive trend (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B.) and the TN:TP showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a weak positive trend, little statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B.).
+Total phosphorus showed a slight increasing trend as there is statistical support for a positive trend and the slope could result in a trophic state change over the course of a century (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.).
+Much of this trend is driven by the unusually high average anomaly in the last year.
+Additional years would indicate if this trend has continued at the regional scale.
+Total nitrogen showed a slight decreasing trend with weak statistical support and less than the 3.5 μg/L that would result in a trophic state change over a century.
+There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B.).
+Lastly, the TN:TP showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B.)
 
@@
 Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.
-This suggests that the observed increase in productivity, as measured by chlorophyll *a*, may be a result of warming waters and not a response to changes in nutrient condition.
+While our analysis is not capable of detecting causation, this association between increasing chlorophyll *a* and warming waters is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature alone.
 Also, geographic extent does indeed matter when trying to identify long-term water quality trends.
@@
 
+```{r year_num_table, echo=FALSE, results="asis"}
+pander::pander(year_num_ww, style = "rmarkdown")
+```
+`r tab_num("year_n_ww")`
+
+```{r year_num_table, echo=FALSE, results="asis"}
+pander::pander(year_num_lagosne, style = "rmarkdown")
+```
+`r tab_num("year_n_lagosne")`
+
 ```{r ww_params, echo=FALSE, results="asis"}
@@
 #Send session info to a file: 
-capture.output(sessionInfo(),file=here::here("manuscript/ecosphere/sessioninfo.txt"))
+capture.output(sessionInfo(),file=here::here("manuscript/ecosphere_resubmit/sessioninfo.txt"))
 ```

### 2020-07-28 R/data_available.R

@@ -123,4 +123,4 @@ lagosne_avail_data_locations <- lagosne_data %>%
   unique() %>%
   right_join(lagosne_data_avail) %>%
   select(station_name, params_avail) %>%
-  write_csv(here("data/lagosene_avail_data_stations.csv"))
+  write_csv(here("data/lagosne_avail_data_stations.csv"))

### 2020-07-28 R/data_clean.R

@@ -136,22 +136,6 @@ ww_lake_trend_data <- ww_lake_trend_data %>%
                                       round(mn_measurement/5, 0)*5,
                                     TRUE ~ mn_measurement))
   
-
-# Calculate and add n:p 
-# Need to make molar conversions here...
-ww_np <- ww_lake_trend_data %>%
-  filter(param == "total_n" | param == "total_p") %>% 
-  unique() %>% #repeats from somewhere...
-  spread(key = param, value = mn_measurement) %>%
-  mutate(np_ratio = (total_n/total_p)*2.21) %>% #calc ratio and convert to molar
-  filter(!is.na(np_ratio)) %>%
-  filter(!is.infinite(np_ratio)) %>%
-  gather(key = "param", value = "mn_measurement", np_ratio) %>%
-  select(station_name, year, month, day, param, location, mn_measurement)
-
-ww_lake_trend_data <- ww_lake_trend_data %>%
-  rbind(ww_np)
-
 # Almy Pond, 2016-07-23, Sample likely contaminated with bottom sediment per notes in data
 # Removed from analysis
 idx <- ww_lake_trend_data$year == "2016" & ww_lake_trend_data$month == 7 & ww_lake_trend_data$day == 23 & ww_lake_trend_data$station_name == "WW199"
@@ -159,29 +143,25 @@ idx <- !idx
 ww_lake_trend_data <- ww_lake_trend_data %>% 
   filter(idx)
 
-write_csv(ww_lake_trend_data, here("data/ww_all_mostly_cleaned.csv"))
-
-# Reviewer 1 comment: To avoid possible seasonal effects due to missed samples, 
-# make sure to include only lake-years with a minimum of one sample in each of
-# May/June, Jul/Aug, Sep/Oct per parameter.
-idx_min_samp <- ww_lake_trend_data %>%
-  mutate(may_jun = case_when(month == 5 | month == 6 ~
-                             1, 
-                           TRUE ~ 0),
-       jul_aug = case_when(month == 7 | month == 8 ~
-                             1,
-                           TRUE ~ 0),
-       sep_oct = case_when(month == 9 | month == 10 ~
-                             1,
-                           TRUE ~ 0)) %>%
-  group_by(station_name, year, param) %>%
-  summarize(min_samp = sum(may_jun, jul_aug, sep_oct)>=3) %>% 
-  right_join(ww_lake_trend_data) %>%
-  pull(min_samp)
-
[…]

### 2020-07-28 R/figures_anomaly_only.R

@@ -106,7 +106,7 @@ lagos_tn_rl <- lagos_tn_gg[[4]]
 lagos_tp_gg <- wq_trend_gg(lagos_data, "total_p", yvar = "measurement_anmly", 
                            y = expression(paste("Average Yearly Total Phosphorus Anomaly (", mu, "g/L)")), 
                            x = "Year",
-                           write = here("data/total_n_data_lagos_anomaly.csv"), 
+                           write = here("data/total_p_data_lagos_anomaly.csv"), 
                            error_bar = "sd",
                            title = "B. LAGOSNE Total Phosphorus")
 ggsave(here("figures/lagos_tp_trends_anomaly.jpg"), lagos_tp_gg[[1]], width = 7.5, height = 5.625, 

### 2020-07-28 R/functions.R

@@ -47,6 +47,52 @@ filter_year <- function(df, num_yrs){
     pull(station_name)
 }
 
+#' Function to filter on sample months
+#' 
+#' Reviewer comment 1-1: To avoid possible seasonal effects due to missed samples, 
+#' make sure to include only lake-years with a minimum of one sample in each of
+#' May/June, Jul/Aug, Sep/Oct per parameter.
+#' 
+#' @param df data frame to filter
+filter_months <- function(df){
+  
+  stations_min_samp <- df %>%
+    mutate(may_jun = case_when(month == 5 | month == 6 ~
+                                 1, 
+                               TRUE ~ 0),
+           jul_aug = case_when(month == 7 | month == 8 ~
+                                 1,
+                               TRUE ~ 0),
+           sep_oct = case_when(month == 9 | month == 10 ~
+                                 1,
+                               TRUE ~ 0)) %>%
+    group_by(station_name, year, param) %>%
+    filter(sum(any(may_jun>0), any(jul_aug>0), any(sep_oct>0))>=3) %>%
+    ungroup() 
+  
+  stations_min_samp
+}
+
+#' Function to filter on sample years
+#' 
+#' Reviewer comment 1-2: To avoid possible effects from having sites only 
+#' measured early or only measured late in the 1993-2016 time frame. Using 
+#' approach similar to Oliver et al. 
+#' 
+#' @param df data frame to filter
+#' @param cutoff The year to divide.  Included in the early period, not the late
+#'               period.
+filter_early_late <- function(df, cutoff){
+  
+  early_late <- df %>%
+    group_by(station_name, param) %>%
+    mutate(early = year <= cutoff,
+           late = year > cutoff) %>%
+    filter(any(early) & any(late)) %>%
+    ungroup() %>%
+    select(-early, -late)
+}
+
 #' Function to generate figures
 #' @param df data frame with data to plot
 #' @param wqparam the water quality parameter to plot
@@ -73,6 +119,8 @@ wq_trend_gg <- function(df, wqparam,
     group_by(year) %>%
     summarize(mn_value = mean(!!yvar),
               sd = sd(!!yvar),
+              min = min(!!yvar),
+              max = max(!!yvar),
               n = n(),
[…]

### 2020-07-28 R/summary_stats.R

@@ -15,7 +15,8 @@ ww_param_ranges <- ww_all %>%
             median = round(median(station_year_mean, na.rm = T), 2),
             percentile_75 = round(quantile(station_year_mean, probs = 0.75, na.rm = T), 2),
             max = round(max(station_year_mean, na.rm = T), 2),
-            sd = round(sd(station_year_mean, na.rm = T), 2)) %>%
+            sd = round(sd(station_year_mean, na.rm = T), 2),
+            n = n()) %>%
   ungroup() %>%
   mutate(units = c("°C", "µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
@@ -57,7 +58,8 @@ ww_param_anomaly_ranges <- ww_all %>%
             median = round(median(measurement_anmly, na.rm = T), 2),
             percentile_75 = round(quantile(measurement_anmly, probs = 0.75, na.rm = T), 2),
             max = round(max(measurement_anmly, na.rm = T), 2),
-            sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
+            sd = round(sd(measurement_anmly, na.rm = T), 2),
+            n = n()) %>%
   ungroup() %>%
   mutate(units = c("°C", "µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
@@ -81,7 +83,8 @@ lagos_param_ranges <- lagos_all %>%
             median = round(median(station_year_mean, na.rm = T), 2),
             percentile_75 = round(quantile(station_year_mean, probs = 0.75, na.rm = T), 2),
             max = round(max(station_year_mean, na.rm = T), 2),
-            sd = round(sd(station_year_mean, na.rm = T), 2)) %>%
+            sd = round(sd(station_year_mean, na.rm = T), 2),
+            n = n()) %>%
   ungroup() %>%
   mutate(units = c("µg/l", "µg/l", "molar", "µg/l")) %>%
   select(Parameter = param, Units = units, 
@@ -122,7 +125,8 @@ lagos_param_anomaly_ranges <- lagos_all %>%
             median = round(median(measurement_anmly, na.rm = T), 2),
             percentile_75 = round(quantile(measurement_anmly, probs = 0.75, na.rm = T), 2),
             max = round(max(measurement_anmly, na.rm = T), 2),
-            sd = round(sd(measurement_anmly, na.rm = T), 2)) %>%
[…]

### 2020-07-28 R/ww_all_sf.R

@@ -1,0 +1,113 @@
+library(readr)
+library(dplyr)
+library(readxl)
+library(lubridate)
+library(here)
+library(stringr)
+library(sf)
+library(tidyr)
+
+ww_all <- read_csv("data/ww_all.csv")
+# read in sites
+ww_sites <- read_excel(here("data/RIDOH Supporting Files.xlsx"), "Sites") %>%
+  mutate(station_name = case_when(str_length(`WW_Station`) == 4 ~ 
+                                    str_replace(`WW_Station`, "WW", "WW0"),
+                                  TRUE ~ `WW_Station`))
+
+
+params <- c("Temperature - 00011",
+            "Secchi Depth - 00069",
+            "Phosphorus, Total - 00665",
+            "Nitrogen, Total (unfiltered) - 00600",
+            "Enterococci - 31639",
+            "Dissolved Oxygen - 00300",
+            "Chlorophyll a, water, fluorometric method, corrected for pheophytin - 32209",
+            "pH - 00400",
+            "Chloride - 00940")
+
+# read in missing nutrient data
+
+missing_nutrients_1 <- read_csv(here("data/WW_96_nutrients_nds.csv")) %>%
+  rbind(read_csv(here("data/WW_97_nutrients.csv"))) %>%
+  rbind(read_csv(here("data/WW_98_nutrients.csv"))) %>%
+  rbind(read_csv(here("data/WW_99_nutrients.csv"))) %>%
+  rbind(read_csv(here("data/WW_00_nutrients.csv"))) %>%
+  rbind(read_csv(here("data/WW_01_nutrients.csv"))) %>%
+  mutate(Day = 15) # Need this to conform to existing format - none of these had a day so assuming middle of month
+
+missing_nutrients <- missing_nutrients_1 %>%
+  transmute(`Station Name` = `Station Name`,
+            Date = ymd(paste(Year, Month, Day, sep="-")),
+            Time = NA,
+            `Sample Type` = NA,
+            `Sample Media` = "Water",
+            Depth = Depth,
+            Parameter = Parameter,
+            Concentration = Concentration,
+            Unit = NA,
+            `Qualifier Code` = NA,
+            `Detection Limit` = 0, #zero becuase NA results getting filtered out
+            `Detection Limit Unit` = NA,
+            `Quantitation Level` = NA,
+            `Quantitation Level Unit` = NA,
+            `Lab Name` = NA,
+            `Analytical Method Number` = NA,
+            Comments = "Day assumed to be 15th, missing for original data",
[…]

### 2020-08-13 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
making slow progress on revisions...

@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** Another good suggestion.
+We have addressed in our response to comment *1-5b* and *1-8*.
+In particular, we have identified slope values that would indicate a meaningful ecological trend.
+We use literature values as well as state water quality criteria (when available) to inform those values.
+Additionally, we have added analysis to explore trends across trophic state.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** What we mean by this, is if sites with differing base lines (e.g.
+low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends becuase each of the sites will be re-scaled, relative to that sites long term mean.
+For instance, comparing measured values from a cool water site that shows no trend which is measured mostly early in a sampling period to a warm water site without a trend measured late in a sampling period would likely show an increasing trend when in fact none existed.
+Re-scaling these values relative to the mean sets both sites to a common baseline and would then not show a trend.
+To further illustrate this, look at the two figures below.
+The first figure shows simulated data from three sites.
+All sites have no trend with site one being measured across the time period with values mostly in the middle of a range.
+Site 2 has no trend with low values and is only in the first portion of the time period.
+Site 3 has no trend with high values and is only in the later portion of the time period.
+The second figure shows what the average yearly trend would look like if just looking at the values vs looking at the anomalies.
+The measured values show a trend, when none should exist.
+The anomalies correctly show no increasing trend.
+We have added in these figures and text to the Discussion
+
+```{r echo=TRUE}
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(stringr)
+library(ggplot2)
+library(hrbrthemes)
+library(broom)
+
+set.seed(0)
+options(scipen=10000)
+
+site1 <- tibble(site=rep("site 1",10),
+                year = 2011:2020,
+                rep_1 = rnorm(10,10,2.5),
+                rep_2 = rep_1*rnorm(10, 1, 0.05),
+                rep_3 = rep_1*rnorm(10, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+site2 <- tibble(site=rep("site 2",6),
+                year = 2011:2016,
+                rep_1 = rnorm(6,3,1),
+                rep_2 = rep_1 * rnorm(6, 1, 0.05),
+                rep_3 = rep_1 * rnorm(6, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+site3 <- tibble(site=rep("site 3",6),
+                year = 2015:2020,
+                rep_1 = rnorm(6, 17, 1),
+                rep_2 = rep_1*rnorm(6, 1, 0.05),
+                rep_3 = rep_1*rnorm(6, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+examp <- rbind(site1, site2, site3) %>%
+  mutate(year = year(ymd(paste0(year,"01-01"))))
+
+site_gg <- examp %>%
+  ggplot(aes(x = year, y = values)) +
+  geom_point() + 
+  facet_grid(. ~ site) +
+  theme_ipsum_rc()  +
+  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
+                       breaks = c(2012,2014,2016,2018,2020)) +
+  labs(x = "Year", y = "Simulated Measurements")
+site_gg
+
+examp_site_summ <- examp %>% 
+  group_by(site) %>%
+  mutate(lt_mean = mean(values)) %>%
+  ungroup() %>%
+  group_by(site, year) %>%
+  summarize(measured_value = mean(values),
+            anomaly = mean(values-lt_mean)) %>%
+  ungroup() %>%
+  pivot_longer(measured_value:anomaly, "Method")
+
+examp_yr_summ <- examp_site_summ %>%
+  group_by(year, Method) %>%
+  summarize(values = mean(value)) %>%
+  ungroup()
+
+meas_lm <- lm(values ~ year, 
+              data = examp_yr_summ[examp_yr_summ$Method == "measured_value",]) %>% 
+    tidy() %>%
+    slice(2) %>%
+    select(slope = estimate, p.value)
+anom_lm <- lm(values ~ year, 
+              data = examp_yr_summ[examp_yr_summ$Method == "anomaly",]) %>% 
+    tidy() %>%
+    slice(2) %>%
+    select(slope = estimate, p.value)
+
+simulated_trends_gg <- ggplot(examp_yr_summ, 
+                              aes(x = year, y = values, color = Method)) +
+  geom_point(size = 2) +
+  geom_smooth(method = "lm", se = FALSE) +
+  scale_color_manual(values = c("darkblue", "darkred"), 
+                     labels = c("Anomaly", "Measured Values")) +
+  labs(y = "Yearly Average Value", x = "Year", subtitle = paste0("Measured Values: slope = ", signif(meas_lm$slope, 2), 
+                                      ", p-value = ", signif(meas_lm$p.value, 2), "\nAnomaly: slope = ", signif(anom_lm$slope, 2), 
+                                      ", p-value = ", signif(anom_lm$p.value, 2))) +
+  theme_ipsum_rc() +
+  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
+                       breaks = c(2012,2014,2016,2018,2020)) +
+  theme(plot.subtitle = element_text(size=10, face="plain"))
+simulated_trends_gg
+```
 

### 2020-08-13 manuscript/ecosphere_resubmit/manuscript.Rmd
making slow progress on revisions...

@@
 There are various ways to identify important temperature changes in lakes and reservoirs.
-A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year.
+A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wg_regs].
 Alternatively, we could look at more ecologically focused temperature changes.
@@
 For this study, we will use the 0.02°C per year as it would provide protections against other ecologically relevant changes (e.g.
-0.035°C per year for trophic linkages) and would also be indicative of meeting an important policy goal.
+0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal.
 For changes in chlorophyll, total phosphorus, and total nitrogen an ecologically meaningful trend is one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.
 We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].
-For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, the total phosphorus limit is 10 μg/L or 0.1 μg/L per year, and for total nitrogen the limit is 350 μg/L, or 3.5 μg/L.
-As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.
+For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.
+This level for total phosphorus would also protect against As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.
 
@@
 With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
-Additionally, the site-specific anomalies are robust to variations in sampling effort and in the timing of inclusion of given sampling locations (e.g., added later in a time period or removed).
+
+The site-specific anomalies are also robust to variations in sampling effort and in the timing of inclusion of given sampling locations.
+For instance, if a site included early in a time period had low values and site included late in the time period, then analyzing the measured values over this time period would show an increase whether or not one existed.
+Using the site-specific anomalies rescales the values to a common framework and allows the real trend to be seen.
+We illustrate this with simulated data that have no trend present (`r fig_num(name = "simulated_data", caption = "Simulated data showing example of sites with no trend present and a low site early in a time preiod, a high site late in a time period, and on site measured across the time period.", display = "cite")`).
+The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated data with no trend present and a low site early in a time preiod, a high site late in a time period, and on site measured across the time period.
+Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).
+
 Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.
@@
 
-```{r year_num_table, echo=FALSE, results="asis"}
+```{r year_num_table_ww, echo=FALSE, results="asis"}
 pander::pander(year_num_ww, style = "rmarkdown")
@@
 
-```{r year_num_table, echo=FALSE, results="asis"}
+```{r year_num_table_lagos, echo=FALSE, results="asis"}
 pander::pander(year_num_lagosne, style = "rmarkdown")
@@
 
+######
+
+![](../../figures/simulated_data.jpg)
+`r fig_num("simulated_data")`
+
+######
+
+![](../../figures/simulated_analysis.jpg)
+`r fig_num("simulated_analysis")`
+
 ```{r session_info, echo=FALSE}

### 2020-08-13 R/figures.R

@@ -194,5 +194,92 @@ figure_data <- chla_gg[[3]] %>%
 
 write_csv(figure_data, here("data/yearly_average_z.csv"))
   
+# Create simulated data and analysis for figure in discussion.
+set.seed(0)
+options(scipen=10000)
+
+site1 <- tibble(site=rep("site 1",10),
+                year = 2011:2020,
+                rep_1 = rnorm(10,10,2.5),
+                rep_2 = rep_1*rnorm(10, 1, 0.05),
+                rep_3 = rep_1*rnorm(10, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+site2 <- tibble(site=rep("site 2",6),
+                year = 2011:2016,
+                rep_1 = rnorm(6,3,1),
+                rep_2 = rep_1 * rnorm(6, 1, 0.05),
+                rep_3 = rep_1 * rnorm(6, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+site3 <- tibble(site=rep("site 3",6),
+                year = 2015:2020,
+                rep_1 = rnorm(6, 17, 1),
+                rep_2 = rep_1*rnorm(6, 1, 0.05),
+                rep_3 = rep_1*rnorm(6, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))                
+
+examp <- rbind(site1, site2, site3) %>%
+  mutate(year = year(ymd(paste0(year,"01-01"))))
+
+site_gg <- examp %>%
+  ggplot(aes(x = year, y = values)) +
+  geom_point() + 
+  facet_grid(. ~ site) +
+  theme_ipsum_rc()  +
+  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
+                     breaks = c(2012,2014,2016,2018,2020)) +
+  labs(x = "Year", y = "Simulated Measurements")
+site_gg %>%  
+  ggsave(here("figures/simulated_data.jpg"), ., width = 7.5, height = 9.8,
+         units = "in", dpi = 600)
+
+
[…]

### 2020-08-13 R/figures_anomaly_only.R

@@ -197,3 +197,93 @@ figure_data <- chla_gg[[3]] %>%
   rbind(mutate(lagos_np_gg[[3]], source = "lagosne", variable = "np_ratio"))
 
 write_csv(figure_data, here("data/yearly_average_anomaly.csv"))
+
+# Create simulated data and analysis for figure in discussion.
+set.seed(0)
+options(scipen=10000)
+
+site1 <- tibble(site=rep("site 1",10),
+                year = 2011:2020,
+                rep_1 = rnorm(10,10,2.5),
+                rep_2 = rep_1*rnorm(10, 1, 0.05),
+                rep_3 = rep_1*rnorm(10, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+site2 <- tibble(site=rep("site 2",6),
+                year = 2011:2016,
+                rep_1 = rnorm(6,3,1),
+                rep_2 = rep_1 * rnorm(6, 1, 0.05),
+                rep_3 = rep_1 * rnorm(6, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))
+
+site3 <- tibble(site=rep("site 3",6),
+                year = 2015:2020,
+                rep_1 = rnorm(6, 17, 1),
+                rep_2 = rep_1*rnorm(6, 1, 0.05),
+                rep_3 = rep_1*rnorm(6, 1, 0.05)) %>%
+  pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
+  mutate(replicate = str_extract(replicate, "[0-9]"))                
+
+examp <- rbind(site1, site2, site3) %>%
+  mutate(year = year(ymd(paste0(year,"01-01"))))
+
+site_gg <- examp %>%
+  ggplot(aes(x = year, y = values)) +
+  geom_point() + 
+  facet_grid(. ~ site) +
+  theme_ipsum_rc()  +
+  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
+                     breaks = c(2012,2014,2016,2018,2020)) +
+  labs(x = "Year", y = "Simulated Measurements")
+site_gg %>%  
+  ggsave(here("figures/simulated_data.jpg"), ., width = 7.5, height = 5.625,
+         units = "in", dpi = 600)
+
+
[…]

### 2020-08-13 manuscript/ecosphere_resubmit/simulated_trends.R

@@ -1,0 +1,119 @@
+library(dplyr)
+library(tidyr)
+library(lubridate)
+library(stringr)
+library(ggplot2)
+library(hrbrthemes)
+
+set.seed(42)
+
+site1 <- tibble(site=rep("site 1",10),
+                year = 2011:2020,
+                rep_1 = rnorm(10,10,2.5),
+                rep_2 = rep_1*rnorm(10, 1, 0.05),
+                rep_3 = rep_1*rnorm(10, 1, 0.05),
+                inc_1 = seq(min(c(rep_1, rep_2, rep_3)),
+                            max(c(rep_1, rep_2, rep_3)),
+                            length.out = 10) *rnorm(10, 1, 0.05),
+                inc_2 = inc_1*rnorm(10, 1, 0.05),
+                inc_3 = inc_1*rnorm(10, 1, 0.05),
+                dec_1 = seq(max(c(rep_1, rep_2, rep_3)),
+                            min(c(rep_1, rep_2, rep_3)),
+                            length.out = 10) *rnorm(10, 1, 0.05),
+                dec_2 = dec_1*rnorm(10, 1, 0.05),
+                dec_3 = dec_1*rnorm(10, 1, 0.05)) %>%
+  pivot_longer(rep_1:dec_3, names_to = "replicate", values_to = "values") %>%
+  mutate(trend = case_when(grepl("rep", replicate) ~
+                             "flat",
+                           grepl("inc", replicate) ~
+                             "increase",
+                           grepl("dec", replicate) ~
+                             "decrease",
+                           TRUE ~ NA_character_),
+         replicate = str_extract(replicate, "[0-9]"))
+
+site2 <- tibble(site=rep("site 2",6),
+                year = 2011:2016,
+                rep_1 = rnorm(6,3,1),
+                rep_2 = rep_1 * rnorm(6, 1, 0.05),
+                rep_3 = rep_1 * rnorm(6, 1, 0.05),
+                inc_1 = seq(min(c(rep_1, rep_2, rep_3)),
+                            max(c(rep_1, rep_2, rep_3)),
+                            length.out = 6) * rnorm(6, 1, 0.05),
+                inc_2 = inc_1*rnorm(6, 1, 0.05),
+                inc_3 = inc_1*rnorm(6, 1, 0.05),
+                dec_1 = seq(max(c(rep_1, rep_2, rep_3)),
+                            min(c(rep_1, rep_2, rep_3)),
[…]

### 2020-08-13 manuscript/ecosphere_resubmit/manuscript.Rmd
some more revisions.

@@
 There are various ways to identify important temperature changes in lakes and reservoirs.
-A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wg_regs].
+A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs].
 Alternatively, we could look at more ecologically focused temperature changes.
@@
 Using the site-specific anomalies rescales the values to a common framework and allows the real trend to be seen.
-We illustrate this with simulated data that have no trend present (`r fig_num(name = "simulated_data", caption = "Simulated data showing example of sites with no trend present and a low site early in a time preiod, a high site late in a time period, and on site measured across the time period.", display = "cite")`).
-The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated data with no trend present and a low site early in a time preiod, a high site late in a time period, and on site measured across the time period.
+We illustrate this with simulated, random data (e.g.
+no trend) (`r fig_num(name = "simulated_data", caption = "Simulated, random data showing example of sites with a low site early in a time preiod, a high site late in a time period, and one site measured across the time period.", display = "cite")`).
+The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated, random data with a low site early in a time preiod, a high site late in a time period, and one site measured across the time period.
 Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).

### 2020-08-28 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
getting closer on revisions...

@@
 - **Response:** What we mean by this, is if sites with differing base lines (e.g.
-low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends becuase each of the sites will be re-scaled, relative to that sites long term mean.
+low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends because each of the sites will be re-scaled, relative to that sites long term mean.
 For instance, comparing measured values from a cool water site that shows no trend which is measured mostly early in a sampling period to a warm water site without a trend measured late in a sampling period would likely show an increasing trend when in fact none existed.
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** Changed title.
+It no longer focus on the trends.
+New title is: "Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
 
@@
 Additionally two years, 1995 and 1996, now have insufficient data to be included in the figures.
-Using the more explicit definition of a trend we not conclude that TN showed a weak statistically positive trend but not an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.
+Using the more explicit definition of a trend we now conclude that TN showed a weak statistically positive trend but not an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.
 This is reflected in the manuscript.
@@
 
-- **Response:** We applied the same within and across year filters to the LAGOSNE and these results have changed as well, largely becuase nutrient data across the full time frame was scarce in LAGOSNE in more recent years.
-Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing weak statistical support for a negative change in nitrogen, more below average years recently.
+- **Response:** We applied the same within and across year filters to the LAGOSNE and these results have changed as well.
+Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing weak statistical support for a negative change in nitrogen, with more below average years recently.
 The slope is less than the 3.5 we identified, albeit close.
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** Small changes due to filtering and new criteria for assessing the trend are also seen here.
+There is statistical support for a trend, that trend is positive and the magnitude of the slope suggests the possibility of ecological significant change in TP over a century.
+This has been added to the manuscript.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** Agreed.  Language softened and title changed to hopefully steer readers away from causation.  No changes made.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
-
-```{r, echo=FALSE}
-x <- ww_data %>% 
-  select(station_name, site_descr, town, year, param, station_year_mean) %>%
-  spread(param, station_year_mean) %>%
-  filter(complete.cases(.))
-xyr <- group_by(x, station_name) %>% 
-  summarize(yrly_mean_chla = mean(chla),
-            yrly_mean_temp = mean(temp),
-            yrly_mean_np = mean(np_ratio),
-            yrl_mean_tn = mean(total_n),
-            yrly_mean_tp = mean(total_p))
-```
+- **Response:** Agreed.  Language softened and title changed to hopefully steer readers away from causation.  No changes made.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** We agree with the reviewer that lake scale trends and the drivers behind those trends are both very interesting questions and something our approach could contribute to; however, we feel that it is beyond the scope of this particular paper.
+Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purposes, looking at long term trends in Rhode Island and the Northeast.
+We do feel this is a very good suggestion for next steps and my co-authors and I are in active discussion on how to move forward with such an analysis.
+In the Discussion under *Trends* we briefly mention this application and that the anomaly approach could also be used at the individual lake level to assess for trends.
 
@@
 
-- **Response:** We have added a new analysis to look at the same plots across trophic states.
+- **Response:** We agree that this is indeed a timely question and fairly easy to test.
+Initially we were going to include this analysis in our revisions and did some preliminary analysis.
+There were differences in trends across trophic state with many of the water quality variables and to fully address this would require significant additional analysis and discussion.
+After looking at these results and with quite a bit of discussion amongst the co-authors we do still feel that this type of analysis is 1) beyond the scope of this paper and 2) worthy of its own paper.
+We have added to the Discussion under *Trends* about this and briefly address preliminary results for differences in chlorophyll trends across just oligotrophic and hypereutrophic lakes.
+Our internal conversations are ongoing about completing this analysis for a subsequent paper.
 

### 2020-08-28 manuscript/ecosphere_resubmit/manuscript.Rmd
getting closer on revisions...

@@
 ---
-title: "**Title:** Increasing Chlorophyll *a* Amid Stable Nutrient Concentrations in Rhode Island Lakes and Reservoirs"
+title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
 author:
@@
 URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
-The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional extents, respectively.
+The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively.
 We assess regional (LAGOSNE) and state (URIWW) trends with yearly mean anomalies calculated on a per-station basis.
@@
 
-*Comparison of trends across trophic state:*  Just as
-
 # Results
@@
 ```{r no_early_tn, echo=FALSE}
-ww_tn_rl_no9394 <- lm(mn_value~year, data = ww_tn_df[c(-1),]) %>%
+ww_tn_rl_noearly <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
+  tidy() %>%
+  slice(2) %>%
+  select(slope = estimate, p.value)
+```
+
+```{r no_early_tp, echo=FALSE}
+ww_tp_rl_noearly <- lm(mn_value~year, data = ww_tp_df[c(-1,-2,-3),]) %>%
   tidy() %>%
@@
 Total phosphorus showed a slight increasing trend as there is statistical support for a positive trend and the slope could result in a trophic state change over the course of a century (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.).
-Much of this trend is driven by the unusually high average anomaly in the last year.
+Some of this trend is driven by the unusually high average anomaly in the last year.
 Additional years would indicate if this trend has continued at the regional scale.
@@
 
-Our sub-regional analysis indicates that even when nutrient regimes exhibit relative stability (i.e., neither increasing nor decreasing over time), increases in primary production, as measured by chlorophyll *a*, can occur.
+Our sub-regional analysis indicates that increases in primary production, as measured by chlorophyll *a*, occurred over that ~25 years of our study period.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
 Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.
-While our analysis is not capable of detecting causation, this association between increasing chlorophyll *a* and warming waters is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature alone.
+Trends in nutrients were less clear with nitrogen showing a slight statistical increase with little ecological importance and phosphorus showing a slight statistical and ecological increase.
+The nitrogen:phosphorus ratio showed no evidence of a trend.
+While our analysis is not capable of detecting causation, that both chlorophyll *a* and temperature is increasing with less obvious trends in nutrients is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature.
 Also, geographic extent does indeed matter when trying to identify long-term water quality trends.
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while total nutrients and the TN:TP ratio are all relatively stable.
-While TN showed a weak positive trend, that trend was largely driven by the unusually low years for TN in 1993 and 1994.
-With those removed the positive trends weakens considerably.
-The general picture in Rhode Island appears to be one of little to no change in phosphorus, a very weak positive trend in nitrogen and little to no change in the TN:TP ratio.
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while trends in total nutrients and the TN:TP ratio are less clear.
+While TN showed a weak statistically positive trend, data for the early years (1994-1998) were relatively sparse.
+There is no ecological or statistical trend in the years with consistently available data.
+The same pattern holds for total phosphorus.
+Some evidence for a weak increasing trend, but over the period of 199-2016, that trend disappears.
+The general picture in Rhode Island appears to be one of a slight increase in phosphorus, a very weak positive trend in nitrogen and little to no change in the TN:TP ratio.
 Furthermore, it has been shown that productivity in freshwater systems is likely a function of both phosphorus and nitrogen [@paerl2016takes].
@@
 In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.
-For example, trends for water quality at finer temporal scales such as monthly or seasonal trends may be different than the annual trends we analyzed.
+For example, changes in water quality at finer temporal scales, such as monthly or seasonal, may be different than the annual trends we analyzed.
 Anecdotal evidence in Rhode Island points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season.
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
-Additionally, it may be possible that the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.
-For instance, are oligotrophic lakes showing stronger trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
-Lastly, changes in rainfall, extreme weather events, or other climate mediated factors can also be playing a role in increasing chlorophyll in Rhode Island lakes and reservoirs.
-These questions are beyond the scope of this study, but all warrant further, careful investigation.
+Additionally, the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.
+For instance, are oligotrophic lakes showing different trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
+A full analysis and interpretation of trends by trophic state is beyond the scope of this study.
+However, we conducted a quick, preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes.
+This suggests that a full analysis of how lakes in different trophic states are changing over time is warrented.
+Lastly, there are other possible drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.
+The analysis we report in this paper focuses on long term trends at broader scales.
+It would be possible to use a similar approach to look at trends of individual lakes which could be a first step in trying to identify drivers of increasing chlorophyll.
+These questions are beyond the scope of this study, but all warrant further investigation.
 
@@
 Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
-Our results show increasing chlorophyll *a* even though the general long-term nutrient trends have been stable.
+Our results show increasing chlorophyll *a* even though the general long-term nutrient trends have been less clear.
 This suggests the need to further reduce nutrients to compensate for warmer water temperatures and possible longer growing seasons.
@@
 <div id="refs"></div>
+
 ######
+
 # Tables

### 2020-08-28 R/data_clean.R

@@ -163,8 +163,18 @@ ww_lake_trend_data <- select(ww_lake_trend_data, -may_jun, -jul_aug, -sep_oct) %
 
 write_csv(ww_lake_trend_data, here("data/ww_all_mostly_cleaned.csv"))
 
+#outliers <- ww_lake_trend_data %>% 
+#  group_by(station_name, param) %>% 
+#  mutate(station_mean = mean(mn_measurement), station_sd = sd(mn_measurement), 
+#         outlier_high = mn_measurement > station_mean + (3.291*station_sd), 
+#         outlier_low = mn_measurement < station_mean - (3.291*station_sd),
+#         outlier = outlier_high | outlier_low) %>%
+#  pull(outlier)
+
+
 # Calculates station/year/param stats
 ww_lake_trend_data <- ww_lake_trend_data %>%
+#  filter(!outliers) %>%
   filter(month >= 5 & month <= 10) %>%
   group_by(station_name, year, param) %>%
   summarize(station_year_mean = mean(mn_measurement)) %>%
@@ -239,15 +249,11 @@ ww_lake_trend_data <- ww_lake_trend_data %>%
 ww_lake_trend_data <- ww_lake_trend_data %>%
   filter_early_late(2004)
 
-# Add TSIs
-ww_lake_trend_data <- ww_lake_trend_data %>%
-  mutate(tsi = case_when(param == "chla" ~
-                           tsi(station_year_mean, "chla"),
-                         param == "total_p" ~
-                           tsi(station_year_mean, "tp"),
-                         param == "total_n" ~
-                           tsi(station_year_mean, "tn"),
-                         TRUE ~ NA_real_),
+# Add long term chla based TSIs
+station_tsi <- ww_lake_trend_data %>%
+  filter(param == "chla") %>%
+  select(station_name, lt_mean) %>%
+  mutate(tsi = tsi(lt_mean, "chla"),
          trophic_state = case_when(tsi <= 40 ~
                                      "oligotrophic",
                                    tsi > 40 & tsi <= 50 ~
@@ -257,7 +263,12 @@ ww_lake_trend_data <- ww_lake_trend_data %>%
                                    tsi > 60 ~
                                      "hypereutrophic",
                                    TRUE ~
-                                     NA_character_))
+                                     NA_character_)) %>%
+  unique()
+
+ww_lake_trend_data <- ww_lake_trend_data %>%
+  left_join(station_tsi, by = "station_name")
+
[…]

### 2020-08-28 R/figures_anomaly_only.R

@@ -2,7 +2,11 @@
 #' load up functions and packages
 source(here::here("R/functions.R"))
 
-ww_data <- read_csv(here("data/ww_lake_trend_data.csv"))
+ww_data <- read_csv(here("data/ww_lake_trend_data.csv")) %>%
+  mutate(trophic_state = factor(trophic_state, 
+                                levels = c("oligotrophic", "mesotrophic", 
+                                           "eutrophic", "hypereutrophic", 
+                                           ordered = TRUE)))
 
 chla_gg <- ww_data %>%
   #filter(state == "RI") %>%
@@ -75,7 +79,11 @@ ww_np_kt <- np_gg[[2]]
 ww_np_df <- np_gg[[3]]
 ww_np_rl <- np_gg[[4]]
 
-lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv"))
+lagos_data <- read_csv(here("data/lagos_lake_trend_data.csv")) %>%
+  mutate(trophic_state = factor(trophic_state, 
+                                levels = c("oligotrophic", "mesotrophic", 
+                                           "eutrophic", "hypereutrophic", 
+                                           ordered = TRUE)))
 
 lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_anmly", 
                              y = expression(paste("Average Yearly Chlorophyll  ", 
@@ -287,3 +295,150 @@ simulated_trends_gg %>%
          units = "in", dpi = 600)
 
 
+# Trophic State Plots
+chla_gg_ts <- ww_data %>%
+  #filter(state == "RI") %>%
+  wq_trophic_trend_gg("chla", yvar = "measurement_anmly", 
+               y = expression(paste("Average Yearly Chlorophyll ", 
+                                    italic(" a"), " Anomaly (", mu, "g/L)")), 
+               x = "Year",
+               write = here("data/chla_data_anomaly_ts.csv"), error_bar = "sd",
+               title = "A. URI Watershed Watch Chlorophyll")
+ggsave(here("figures/ww_chla_trends_anomaly_ts.jpg"), chla_gg_ts[[1]], width = 7.5, 
+       height = 5.625, 
[…]

### 2020-08-28 R/functions.R

@@ -134,7 +134,6 @@ wq_trend_gg <- function(df, wqparam,
   df2 <- df2 %>%
     filter(n >= 3)
   
-  
   if(!is.null(write)){
     write_csv(df2, write, append = FALSE)
   }
@@ -251,4 +250,109 @@ tsi_table <- function(){
   
 }
 
- 
+#' Function to generate trophic state figures
+#' @param df data frame with data to plot
+#' @param wqparam the water quality parameter to plot
+#' @param yvar the actual variable to plot, defaults to anomaly
+#' @param num_yrs minimum number of years needed to include a site
+#' @param write write out data used in the plot to a csv file
+#' @param title title for plot
+#' @param start_yr only plots years for which we have data for all params
+wq_trophic_trend_gg <- function(df, wqparam, 
+                        yvar = c("measurement_anmly","measurement_scale"), 
+                        num_yrs = 10, write = NULL, title = "",
+                        start_yr = 1993, error_bar = c("se", "sd"), ...){
+  
+  yvar <- rlang::sym(match.arg(yvar))
+  error_bar <- rlang::sym(match.arg(error_bar))
+  
+  df1 <- df %>%
+    #filter(param == wqparam) %>%
+    filter(year >= start_yr) %>%
+    filter(station_name %in% filter_year(., num_yrs))
+  
+  
+  df2 <- df1 %>%
+    filter(!is.na(trophic_state)) %>%
+    group_by(year, param, trophic_state) %>%
+    summarize(mn_value = mean(!!yvar),
+              sd = sd(!!yvar),
+              min = min(!!yvar),
+              max = max(!!yvar),
+              n = n(),
+              se = sd/sqrt(n())) %>%
+    mutate(col_group = case_when(mn_value < 0 ~ 
+                                   "Less than long-term site average",
+                                 mn_value > 0 ~ 
+                                   "Greater than long-term site average",
+                                 TRUE ~ "Equal to long-term site average"))
+  # Nutrient data were scarce in 1995-1998  To include data in plots, had to 
+  # have a minimum of three sites.  If only three sites, then data plotted are 
+  # mean and min/max. 
+  df2 <- df2 %>%
+    filter(n >= 3)
+  
+  browser()
+  if(!is.null(write)){
+    write_csv(df2, write, append = FALSE)
+  }
+  
+  kt <- with(df2, Kendall(year, mn_value))
[…]

### 2020-09-15 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
edits done

@@
 the Paris Agreement).
-While these exact numbers are not universal, we use them as a starting point for assesing the trends.
+While these exact numbers are not universal, we use them as a starting point for assessing the trends.
 How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below.
@@
 
-- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*).
-This weakened the positive trend somewhat as the slope is smaller and the p-value larger.
-Additionally two years, 1995 and 1996, now have insufficient data to be included in the figures.
-Using the more explicit definition of a trend we now conclude that TN showed a weak statistically positive trend but not an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.
+- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*) and made changes to use the median instead of the mean in our response to comment *1-6*.
+These changes have weakened the positive trend as the slope is smaller and the p-value larger.
+Additionally thre years, 1995, 1996, and 1997 now have insufficient data to be included in the figures and a third, 1998, has only three sites.
+Using the more explicit definition of a trend we now conclude that TN showed neither a statistically positive trend or an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.
 This is reflected in the manuscript.
@@
 
-- **Response:** We applied the same within and across year filters to the LAGOSNE and these results have changed as well.
-Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing weak statistical support for a negative change in nitrogen, with more below average years recently.
-The slope is less than the 3.5 we identified, albeit close.
-Thus we conclude a weak decrease in TN at the scale of the Region.
+- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instaed of the mean and these results have changed as well.
+Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a negative change in nitrogen, with more below average years recently.
+The slope is much less than the 3.5 we identified.
+Thus we conclude no trend in TN at the scale of the Region.
 This is indicated in the manuscript.
@@
 - **Response:** Again, with the changes to the filtering the trend lines have slightly changed.
-We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, a value greater than the threshold we identified and greater frequency of above average years in recent years.
+We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, athough the slope is not greater than the threshold we identified and greater frequency of above average years in recent years.
 We have indicated as such in the text.
@@
 
-- **Response:** This is a good suggestion, but we have chosen not to implement it, primarily because of Reveiwer 1's suggestion to have a more holistic approach to addressing the trends.
+- **Response:** This is a good suggestion, but we have chosen not to implement it, primarily because of Reviewer 1's suggestion to have a more holistic approach to addressing the trends.
 In developing our approach to assessing the trends we now have ecologically relevant thresholds we can compare to.
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** We were a little unclear about this comment.
+Is it referring to coherence over time within a single variable or coherence between the variables?
+We agree that for individual variables some years do show greater variability than others.
+These increases in the yearly standard deviations are largely the result of a few unusual measurements in those years for those variables.
+Essentially these are outliers, but ones that have no obvious reason to remove from the analysis so they are kept in.
+Further, there is no pattern to these outliers as they are from different ponds, aren't seen across variables in the same years, and in the case of the LAGOSNE data, they come from different states.
+There is only one exception to this and that is the high variability in 2011 for both TN and TP.
+That is driven by one sampling event where TN and TP were high for both.
+Given the lack of a pattern, there is little additional inference we can draw from this, other than the analysis might be susceptible to outliers.
+Given that our original analysis is based on the mean and std.
+dev.
+and that those statistics are sensitive to outliers we have decided, thanks to this comment, to change to median and the inter-quartile range, both of which are considerably more robust statistics in the presence of outliers.
+With this change the swings in variance are no longer seen as would be expected with statistics that are robust to outliers.
+Good catch, by the reviewer.
 
@@
 There were differences in trends across trophic state with many of the water quality variables and to fully address this would require significant additional analysis and discussion.
-After looking at these results and with quite a bit of discussion amongst the co-authors we do still feel that this type of analysis is 1) beyond the scope of this paper and 2) worthy of its own paper.
+After looking at these results and with quite a bit of discussion among the co-authors we do still feel that this type of analysis is 1) beyond the scope of this paper and 2) worthy of its own paper.
 We have added to the Discussion under *Trends* about this and briefly address preliminary results for differences in chlorophyll trends across just oligotrophic and hypereutrophic lakes.
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** NEEDS RESPONSE
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** Added an additional paragraph to the results that compares the two scales and includes a new table summarize the percentage of lakes at both scales in chlorophyll based trophic states.
+Another table was added to the trends sections of the results summarizing how we interepreted the trends at each scale.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** We changed the wording on this to reflect a focus on the need to know about long term trends.
+Also we added three citation that specifically address that sharing results back with volunteers meets their expectations and can motivate continued involvement.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** This section was re-written.  We mention nutrient reductions as one possible hypothesis that long-term trends like those seen in RI can inform.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** We added a summary table that summarizes average lake area, maximum depth, and average percent forest, ag, and developed for the LAGOSNE and URIWW.
+Some additional text is in the methods as well as in the results.
+Thanks for the suggestion.
+There are a few differences and are a useful addition to the paper.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** Thanks again.
+We have cleaned up this sentence to better reflect a focus on lengthening growing season and also refer to a source of data (NOAA Climate Time Series) that make this point as well.
 
@@
 
-- **Response:** NEEDS A RESPONSE!
+- **Response:** In the paragraph about LAGOSNE, we provided some additional clarification.
+We use the same selection criteria on both datasets to make sure the time frame is comparable.
+The data for LAGOSNE are reported to all be from the surface or epilimnion.
+While not a direct match to the depth filter we used for URIWW, they are roughly comparable.
+Data for LAGOSNE do not extend beyond 2013, hence the truncation.
+This is all reflected in the manuscript.
 

### 2020-09-15 manuscript/ecosphere_resubmit/manuscript.Rmd
edits done

@@
 The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively.
-We assess regional (LAGOSNE) and state (URIWW) trends with yearly mean anomalies calculated on a per-station basis.
+We assess regional (LAGOSNE) and state (URIWW) trends with yearly median anomalies calculated on a per-station basis.
 Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016.
@@
 Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale.
-Macro-scale (i.e., subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
+Macro-scale (i.e., ~ 1,800,000 km ^2^) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
 This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
-Thus, it was surprising when little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long-term trends in water quality within the broader region of the northeastern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
+On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern (i.e.
+~400,000 km^2^) and midwestern (i.e.
+~1,000,000 km^2^) US regions [@filstrup2014regional; @filstrup2018evidence].
+Furthermore, little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
+Given what is known about long-term trends in water quality within the broader regions of the northeastern and midwestern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
 
@@
 Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
-We used the same selction criteria on the LAGOSNE dataset as was applied to the URIWW data.
-The LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.
+We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data.
+This ensured that both datasets represented the same seasonal time frame.
+Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017].
+At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.
+
+Lastly, to facilitate a comparison of the context of each scale, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].
+We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculuture, Forest, and Developed.
+The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.
+Additional details are available in are detailed in Soranno et al.
+[-@sorannolagosne2017].
 
@@
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
-The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.
-This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.
+The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.
+This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.
+Many of the variable we looked at are non-normal and often have outliers (e.g.
+algal blooms).
+In this case the median is preferred.
+Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.
 We refer to this method as "site-specific anomalies".
 
-*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.
+*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.
 Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
@@
 1.
-For each site, calculate the annual means, producing a single mean value for each site and year.
+For each site, calculate the annual medians, producing a single median value for each site and year.
 This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].
-The per site means across years are assumed to be independent.
-2. Calculate the long-term reference mean for each site. This results in a single long-term mean for each of the sites.
-3. Calculate the anomaly for each annual mean at each site by subtracting the annual and reference means.
-4. Summarize by calculating the mean anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
+The per site medians across years are assumed to be independent.
+2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
+3. Calculate the anomaly for each annual median at each site by subtracting the annual and reference medians.
+4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
 
@@
 
-*Linear regression on annual mean anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
+*Linear regression on annual median anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
 We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
 For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).
-The slope of this line provides information on the mean yearly change of that parameter over the time period studied.
+The slope of this line provides information on the average yearly change of that parameter over the time period studied.
 
@@
 Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence.
-For this assessment, we are defining a trend as positive or negative with the following; 1) Direction of trend is determined by the sign of the slope, 2) p-value is used to inform interpretation and very small p-values suggest a statistically relevant trend, 3) relative frequency of high and low years in the beginning and end of the time frame, and 4) the magnitude of the slope suggests an ecologically relevant change.
+For this assessment, we are defining a trend as with the following: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame suggest an increasing or decreasing patter, and 4) the magnitude of the slope is used to infer an ecologically relevant change.
+Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.
 
@@
 
-During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
+Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
+Also, LAGOSNE lakes tended to occur in landscapes (i.e.
+a 500 meter buffer) with a higher average percent agriculture, less developed land, and comparable forested area than lakes in Rhode Island (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
 
-For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
+During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
+
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  median TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
+
+On average, the Rhode Island lakes and reservoirs show lower concentrations of nutrients with more nitrogen relative to phosphorus than in the larger LAGOSNE region (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).
+Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).
+Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic].
 
@@
 
-Mean annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests a statistical trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with mean temperature greater than the long-term mean are occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).
+Median annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (median anomaly) in Rhode Island lakes and reservoirs.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = "Summary of long term water quality trends in the URIWW and LAGOSENE data.", display = "cite")`).
 Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).
-The slope of 0.3 μg/L is much greater than our threshold of 0.035 μg/L and the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).
+The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (median anomaly).
 Panel A.
-Yearly mean chlorophyll a anomalies from the URI Watershed Watch data.
+Yearly median chlorophyll a anomalies from the URI Watershed Watch data.
 Panel B.
-Yearly mean chlorophyll a anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+Yearly median chlorophyll a anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 
 ```{r no_early_tn, echo=FALSE}
-ww_tn_rl_noearly <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
+ww_tn_rl_noearly <- lm(mdn_value~year, data = ww_tn_df[ww_tn_df$n > 3,]) %>%
   tidy() %>%
@@
 ```{r no_early_tp, echo=FALSE}
-ww_tp_rl_noearly <- lm(mn_value~year, data = ww_tp_df[c(-1,-2,-3),]) %>%
+ww_tp_rl_noearly <- lm(mdn_value~year, data = ww_tp_df[c(-1,-2,-3),]) %>%
   tidy() %>%
@@
 
-Mean annual trends for nutrients were not as clear.
-For total nitrogen, the slope of the line is positive with some statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) and the years greater than the mean are somewhat more frequent in more recent years.
-However, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly).
+Median annual trends for nutrients were not as clear.
+For total nitrogen, the slope of the line is positive with weak statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.
+Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (median anomaly).
 Panel A.
-Yearly mean TN anomalies from the URI Watershed Watch dataset.
+Yearly median TN anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly mean TN anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
-Grey indicates yearly site specifica anomalies that were, on average equal to the long-term means.
-Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
-We interpret this as a weak positive trend in TN, but with little ecological impact.
-Total phosphorus shows a positive trend over time in the yearly anomalies but with somewhat weak statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the mean are more common in recent years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).
+Yearly median TN anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
+Grey indicates yearly site specifica anomalies that were, on average equal to the long-term medians.
+Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+However, 1998 only had three sites with available total nitrogen data and may not be representative.
+This year also recorded the lowest median value.
+Combined, 1998 may be skewing these results.
+When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.
+Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`).
+Total phosphorus shows essentially no trend over time in the yearly anomalies with no statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (median anomaly).
 Panel A.
-Yearly mean TP anomalies from the URI Watershed Watch dataset.
+Yearly median TP anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly mean TP anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
-Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
-Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is greater than our defined threshold of 0.1 μg/L and suggests a possible trophic state shift in less than a century.
-The pattern for the TN:TP ratio has a negative slope but no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).
+Yearly median TP anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
+Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.
+The pattern for the TN:TP ratio has a small slope and no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (median anomaly).
 Panel A.
-Yearly mean TN:TP ratio anomalies from the URI Watershed Watch dataset.
+Yearly median TN:TP ratio anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly mean TN:TP ratio anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
-Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
-Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
+Yearly median TN:TP ratio anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
+Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 
@@
 
-In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a weak positive trend, little statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B.).
-Total phosphorus showed a slight increasing trend as there is statistical support for a positive trend and the slope could result in a trophic state change over the course of a century (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B.).
-Some of this trend is driven by the unusually high average anomaly in the last year.
-Additional years would indicate if this trend has continued at the regional scale.
+In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
 Total nitrogen showed a slight decreasing trend with weak statistical support and less than the 3.5 μg/L that would result in a trophic state change over a century.
-There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B.).
-Lastly, the TN:TP showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B.)
+The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`).
+There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Total phosphorus showed a very small increasing trend and there is statistical support for that trend.
+The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.
+This low median anomaly should be interpreted with this caveat in mind.
+Taken together we feel our results largely match the findings of Oliver et al.
+[-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.
+Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/usepa/ri_wq_trends> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
 
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs; while trends in total nutrients and the TN:TP ratio are less clear.
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs.
+However, 2006 and 2009 stand out as not following this trend.
+Average May-October air temperature for 2009 was third lowest on record for the 1993 to 2016 time period and also showed the lowest maximum temperatures [@noaatrends].
+On the other hand, 2006 was not unusually cool (11th lowest out of 24 years) but it was the wettest year on record for this period [@noaatrends].
+
+Trends in total nutrients and the TN:TP ratio are less clear.
 While TN showed a weak statistically positive trend, data for the early years (1994-1998) were relatively sparse.
 There is no ecological or statistical trend in the years with consistently available data.
-The same pattern holds for total phosphorus.
-Some evidence for a weak increasing trend, but over the period of 199-2016, that trend disappears.
-The general picture in Rhode Island appears to be one of a slight increase in phosphorus, a very weak positive trend in nitrogen and little to no change in the TN:TP ratio.
+The general picture in Rhode Island appears to be one of a very slight increase in nitrogen, no trend in phosphorus and little to no change in the TN:TP ratio.
 Furthermore, it has been shown that productivity in freshwater systems is likely a function of both phosphorus and nitrogen [@paerl2016takes].
@@
 For example, changes in water quality at finer temporal scales, such as monthly or seasonal, may be different than the annual trends we analyzed.
-Anecdotal evidence in Rhode Island points to warmer temperature earlier and later in the year and suggests a lengthening of the growing season.
+Evidence in Rhode Island points to possible lengthening of the growing season as both May and October show increasing temperature trends over time [@noaatrends].
 Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
@@
 This suggests that a full analysis of how lakes in different trophic states are changing over time is warrented.
-Lastly, there are other possible drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.
+Lastly, there are other drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.
 The analysis we report in this paper focuses on long term trends at broader scales.
@@
 
-## Management implications
+## Broader implications
 
-There are several broader management implications from the results of our analysis and of examining long-term water quality trends in general.
-In particular, this analysis provides much needed information about the long-term effects of current nutrient control efforts at lake-specific and sub-regional scales and identifies areas where additional information is required or a change in management approaches may be needed.
-First, as more long-term datasets become available, it is important for managers and stakeholders to receive feedback on long-term water quality trends at multiple spatial scales.
-Specifically for this study, the results provide feedback to long time volunteer monitors, highlighting the importance of volunteer monitoring programs.
+There are several broader implications from the results of our analysis and of examining long-term water quality trends in general.
+First, as more long-term datasets become available, it is important for managers, stakeholders, and researchers to work together to better understand long-term water quality trends at multiple spatial scales.
+Specifically for this study, the results provide feedback to long time volunteer monitors about the trends in long-term, volunteer collected data.
+This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some citizen science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].
+Were that long-term involvement not common, then consistent long-term data may be more difficult to obtain.
+
 Second, with information on long-term trends, it is possible to adapt management approaches to address areas of concern.
-Our results show increasing chlorophyll *a* even though the general long-term nutrient trends have been less clear.
-This suggests the need to further reduce nutrients to compensate for warmer water temperatures and possible longer growing seasons.
-
-There are several possible approaches to further reduce nutrient loads [@yang2018nutrients].
-First, nutrient load reductions may be possible through source controls and enhanced entrainment and treatment of ground and surface waters transporting nutrients to receiving waters [@kellogg2010geospatial].
-Green infrastructure approaches are one way to possibly achieve both goals [@pennino2016watershed; @reisinger2019changes].
-Additionally, there is potential for within-lake approaches such as the restoration of freshwater mussels to waterbodies that historically had those species.
-Some studies using freshwater mussels have shown reductions in both nutrients and algal biomass [@kreeger2018restoration].
+Our results show increasing chlorophyll *a* and increasing temperature even though the general long-term nutrient trends have been less clear.
+While the analysis described here can not be used to infer causation, it points to areas that may need to be addressed.
+For instance, warming waters are linked to increases in harmful algal blooms [@paerl2008blooms; @paerl2009climate; @paerl_climate_2012].
+An increase in blooms could be seen via increasing chlorophyll *a* levels.
+Broad scale warming would be difficult to manage on a sub-regional scale thus other mechanisms, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
+Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
+This is one possible hypothesis that is possible to form through information obtained by long-term trends.
+Additional work is certainly needed to examine the drivers of these trends in Rhode Island lakes and reservoirs.
 
@@
 We built on these methods and adapted them for use with long-term water quality trends.
-While other methods are valid and robust [e.g., @Oliver_2017], we chose mean site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.
+While other methods are valid and robust [e.g., @Oliver_2017], we chose median site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.
 For instance, reporting the changes in anomalies allows us to look at changes in the original units.
-With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
+With our analysis, the slope of the regression line for temperature suggests an average yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows an average yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
 
@@
 
+######
+
 ```{r year_num_table_lagos, echo=FALSE, results="asis"}
@@
 
+######
+
+```{r context, echo=FALSE, results="asis"}
+pander::pander(context_summ, style = "rmarkdown")
+```
+`r tab_num("context_summ")`
+
+######
+
 ```{r ww_params, echo=FALSE, results="asis"}
-pander::pander(ww_param_ranges[,c(-3,-6)], style = "rmarkdown")
+pander::pander(ww_param_ranges, style = "rmarkdown")
 ```
@@
 
+######
+
 ```{r lagos_params, echo=FALSE, results="asis"}
-pander::pander(lagos_param_ranges[,c(-3,-6)], style = "rmarkdown")
+pander::pander(lagos_param_ranges, style = "rmarkdown")
 ```
 `r tab_num("lagos_params")`
+
+######
+
+```{r ts_table, echo=FALSE, results="asis"}
+pander::pander(ts_table, style = "rmarkdown")
+```
+`r tab_num("ts_table")`
+
+######
+
+```{r trend_summ, echo=FALSE, results="asis"}
+
+trend_summ_df <- tibble(Source = c("URIWW", "LAGOSNE"), 
+                        Temperautre = c("++", ""), Chlorophyll= c("++", "0"), 
+                        `Total Nitrogen` = c("0", "0"), 
+                        `Total Phosphorus`= c("0", "+"), 
+                        `Nitrogen:Phosphorus` = c("0", "0"))
+
+pander::pander(trend_summ_df, style = "rmarkdown")
+```
+`r tab_num("trend_summ")`
 

### 2020-09-15 R/data_clean-2.R

@@ -198,31 +198,44 @@ write_csv(ww_lake_trend_data, here("data/ww_all_mostly_cleaned.csv"))
 # Calculates station/year/param stats
 ww_lake_trend_data <- ww_lake_trend_data %>%
 #  filter(!outliers) %>%
-  filter(month >= 5 & month <= 10) %>%
+  #filter(month >= 5 & month <= 10) %>%
+  #group_by(station_name, year, param) %>%
+  #summarize(station_year_mean = mean(mn_measurement),
+  #          station_year_median = median(mdn_measurement)) %>%
+  #ungroup() %>%
+  #group_by(station_name, param) %>%
+  #mutate(measurement_scale = scale(station_year_mean),
+  #       measurement_anmly = station_year_mean - mean(station_year_mean),
+  #       measurement_anmly_mdn = station_year_median - median(station_year_median),
+  #       lt_mean = mean(station_year_mean),
+  #       lt_sd = sd(station_year_mean),
+  #       lt_median = median(station_year_median),
+  #       lt_iqr_low = quantile(station_year_median, 0.25),
+  #       lt_iqr_high = quantile(station_year_median, 0.75),
+  #       lt_n = n()) %>%
+  group_by(station_name, param) %>%
+  mutate(measurement_scale = scale(mn_measurement),
+         measurement_anmly = mn_measurement - mean(mn_measurement),
+         measurement_anmly_mdn = mn_measurement - median(mn_measurement),
+         lt_mean = mean(mn_measurement),
+         lt_sd = sd(mn_measurement),
+         lt_median = median(mn_measurement),
+         lt_iqr_low = quantile(mn_measurement, 0.25),
+         lt_iqr_high = quantile(mn_measurement, 0.75),
+         lt_n = n()) %>%
+  ungroup() %>%
   group_by(station_name, year, param) %>%
   summarize(station_year_mean = mean(mn_measurement),
-            station_year_median = median(mn_measurement)) %>%
-  ungroup() %>%
-  group_by(station_name, param) %>%
-  mutate(measurement_scale = scale(station_year_mean),
-         measurement_anmly = station_year_mean - mean(station_year_mean),
-         measurement_anmly_mdn = station_year_median - median(station_year_median),
-         lt_mean = mean(station_year_mean),
-         lt_sd = sd(station_year_mean),
-         lt_median = median(station_year_median),
-         lt_iqr_low = quantile(station_year_median, 0.25),
[…]

### 2020-09-15 R/figures_anomaly_only.R

@@ -10,11 +10,11 @@ ww_data <- read_csv(here("data/ww_lake_trend_data.csv")) %>%
 
 chla_gg <- ww_data %>%
   #filter(state == "RI") %>%
-  wq_trend_gg( "chla", yvar = "measurement_anmly", 
-               y = expression(paste("Average Yearly Chlorophyll ", 
+  wq_trend_gg( "chla", yvar = "measurement_anmly_mdn", 
+               y = expression(paste("Median Yearly Chlorophyll ", 
                                     italic(" a"), " Anomaly (", mu, "g/L)")), 
                x = "Year",
-               write = here("data/chla_data_anomaly.csv"), error_bar = "sd",
+               write = here("data/chla_data_anomaly.csv"), error_bar = "iqr",
                title = "A. URI Watershed Watch Chlorophyll")
 ggsave(here("figures/ww_chla_trends_anomaly.jpg"), chla_gg[[1]], width = 7.5, 
        height = 5.625, 
@@ -26,10 +26,10 @@ ww_chla_rl <- chla_gg[[4]]
 
 tn_gg <-  ww_data %>%
   #filter(state == "RI") %>%
-  wq_trend_gg("total_n", yvar = "measurement_anmly", 
-              y = expression(paste("Average Yearly Total Nitrogen Anomaly (", mu, "g/L)")), 
+  wq_trend_gg("total_n", yvar = "measurement_anmly_mdn", 
+              y = expression(paste("Median Yearly Total Nitrogen Anomaly (", mu, "g/L)")), 
               x = "Year",
-              write = here("data/total_n_data_anomaly.csv"), error_bar = "sd",
+              write = here("data/total_n_data_anomaly.csv"), error_bar = "iqr",
               title = "A. URI Watershed Watch Total Nitrogen")
 ggsave(here("figures/ww_tn_trends_anomaly.jpg"), tn_gg[[1]], width = 7.5, height = 5.625, 
        units = "in", dpi = 600)
@@ -40,10 +40,10 @@ ww_tn_rl <- tn_gg[[4]]
 
 tp_gg <-  ww_data %>%
   #filter(state == "RI") %>%
-  wq_trend_gg("total_p", yvar = "measurement_anmly", 
-              y = expression(paste("Average Yearly Total Phosphorus Anomaly (", mu, "g/L)")), 
+  wq_trend_gg("total_p", yvar = "measurement_anmly_mdn", 
+              y = expression(paste("Median Yearly Total Phosphorus Anomaly (", mu, "g/L)")), 
[…]

### 2020-09-15 R/functions.R

@@ -102,9 +102,9 @@ filter_early_late <- function(df, cutoff){
 #' @param title title for plot
 #' @param start_yr only plots years for which we have data for all params
 wq_trend_gg <- function(df, wqparam, 
-                        yvar = c("measurement_anmly","measurement_scale"), 
+                        yvar = c("measurement_anmly","measurement_scale", "measurement_anmly_mdn"), 
                         num_yrs = 10, write = NULL, title = "",
-                        start_yr = 1993, error_bar = c("se", "sd"), ...){
+                        start_yr = 1993, error_bar = c("se", "sd", "iqr"), ...){
   
   yvar <- rlang::sym(match.arg(yvar))
   error_bar <- rlang::sym(match.arg(error_bar))
@@ -117,20 +117,23 @@ wq_trend_gg <- function(df, wqparam,
   
   df2 <- df1 %>%
     group_by(year) %>%
-    summarize(mn_value = mean(!!yvar),
+    summarize(mdn_value = median(!!yvar),
               sd = sd(!!yvar),
+              iqr = IQR(!!yvar),
+              iqr_low = quantile(!!yvar, 0.25),
+              iqr_high = quantile(!!yvar, 0.75),
               min = min(!!yvar),
               max = max(!!yvar),
               n = n(),
               se = sd/sqrt(n())) %>%
-    mutate(col_group = case_when(mn_value < 0 ~ 
-                                   "Less than long-term site average",
-                                 mn_value > 0 ~ 
-                                   "Greater than long-term site average",
-                                 TRUE ~ "Equal to long-term site average"))
+    mutate(col_group = case_when(mdn_value < 0 ~ 
+                                   "Less than long-term site median",
+                                 mdn_value > 0 ~ 
+                                   "Greater than long-term site median",
+                                 TRUE ~ "Equal to long-term site median"))
   # Nutrient data were scarce in 1995-1998  To include data in plots, had to 
   # have a minimum of three sites.  If only three sites, then data plotted are 
-  # mean and min/max. 
+  # median and min/max. 
   df2 <- df2 %>%
     filter(n >= 3)
   
[…]

### 2020-09-15 R/schematic_anomaly.R

@@ -4,24 +4,25 @@ library(here)
 
 schematic <- read_csv(here("data/schematic.csv"))
 
-station_year_mean <- schematic %>% 
+station_year_median <- schematic %>% 
   group_by(station, year) %>%
-  summarize(station_year_mean = mean(meas)) %>%
+  summarize(station_year_median = median(meas)) %>%
   round(2) %>%
   ungroup()
 
-station_year_anomaly <- station_year_mean %>%
+station_year_anomaly <- station_year_median %>%
   group_by(station) %>%
-  mutate(long_term_mean = mean(station_year_mean),
-         long_term_sd = sd(station_year_mean)) %>%
-  mutate(station_year_anomaly = station_year_mean - long_term_mean)
+  mutate(long_term_median = median(station_year_median),
+         long_term_iqr = IQR(station_year_median)) %>%
+  mutate(station_year_anomaly = station_year_median - long_term_median) %>%
   round(2) %>%
   ungroup()
   
-mean_anomaly <- station_year_anomaly %>%
+median_anomaly <- station_year_anomaly %>%
   group_by(year) %>%
-  summarize(mean_anomaly = mean(station_year_anomaly),
-            sd_anomaly = sd(station_year_anomaly),
+  summarize(median_anomaly = median(station_year_anomaly),
+            iqr_anomaly = IQR(station_year_anomaly),
             n = n()) %>%
   round(2) %>%
   ungroup()
+

### 2020-09-15 R/summary_stats.R

@@ -9,13 +9,13 @@ ww_param_ranges <- ww_all %>%
            param == "np_ratio") %>%
   mutate(param = factor(param,levels = c("temp", "total_n", "total_p", "np_ratio","chla"))) %>%
   group_by(param) %>%
-  summarize(min = round(min(station_year_mean, na.rm = T), 2),
-            percentile_25 = round(quantile(station_year_mean, probs = 0.25, na.rm = T), 2),
-            mean = round(mean(station_year_mean, na.rm = T), 2),
-            median = round(median(station_year_mean, na.rm = T), 2),
-            percentile_75 = round(quantile(station_year_mean, probs = 0.75, na.rm = T), 2),
-            max = round(max(station_year_mean, na.rm = T), 2),
-            sd = round(sd(station_year_mean, na.rm = T), 2),
+  summarize(min = round(min(station_year_median, na.rm = T), 2),
+            percentile_25 = round(quantile(station_year_median, probs = 0.25, na.rm = T), 2),
+            mean = round(mean(station_year_median, na.rm = T), 2),
+            median = round(median(station_year_median, na.rm = T), 2),
+            percentile_75 = round(quantile(station_year_median, probs = 0.75, na.rm = T), 2),
+            max = round(max(station_year_median, na.rm = T), 2),
+            sd = round(sd(station_year_median, na.rm = T), 2),
             n = n()) %>%
   ungroup() %>%
   mutate(units = c("°C", "µg/l", "µg/l", "molar", "µg/l")) %>%
@@ -77,13 +77,13 @@ lagos_param_ranges <- lagos_all %>%
            param == "np_ratio") %>%
   mutate(param = factor(param,levels = c("total_n", "total_p", "np_ratio","chla"))) %>%
   group_by(param) %>%
-  summarize(min = round(min(station_year_mean, na.rm = T), 2),
-            percentile_25 = round(quantile(station_year_mean, probs = 0.25, na.rm = T), 2),
-            mean = round(mean(station_year_mean, na.rm = T), 2),
-            median = round(median(station_year_mean, na.rm = T), 2),
-            percentile_75 = round(quantile(station_year_mean, probs = 0.75, na.rm = T), 2),
-            max = round(max(station_year_mean, na.rm = T), 2),
-            sd = round(sd(station_year_mean, na.rm = T), 2),
+  summarize(min = round(min(station_year_median, na.rm = T), 2),
[…]

### 2020-09-21 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
getting very close on revisions!

@@
 ---
-output: html_document
+output: 
+  word_document:
+    reference_docx: response_template.docx
 editor_options: 
@@
 library(lubridate)
+opts_chunk$set(message = FALSE, warning = FALSE)
 ```
@@
 
-*1-1.* I would like more detail on how the averaging was done for each year.
+***1-1.*** I would like more detail on how the averaging was done for each year.
 For each parameter, an annual average was calculated for each lake.
@@
 The table below lists the percentage of lake-years for each parameter that had at least one sample in each of May/June, July/Aug, and Sept/Oct.
+While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.
+This filtering step has been added to our `data_clean.R` script at lines 157 and 398 and is in the `filter_months()` function in `functions.R`.
+This removes any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.
+The changes to the analysis are minimal and do not significantly alter any of our conclusions.
+We added text in the second paragraph of the Methods section describing the selection criteria.
 
-```{r echo = FALSE}
+```{r echo = FALSE, results="asis"}
 knitr::kable(param_min_samp)
@@
 
-While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.
-This filtering step has been added to our `data_clean.R` script at lines 163 and 259 and is in the `filter_months()` function in `functions.R`.
-This removes any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.
-The changes to the analysis are minimal and do not alter any of our conclusions.
-We added text in the second paragraph of the Methods section describing the selection criteria.
-
-*1-2.* I also had questions regarding the way that interannual data were represented.
+***1-2.*** I also had questions regarding the way that interannual data were represented.
 At least 10 years of data from 1993 to 2016 had to be available for a lake to be included in the analysis (lines 121-122).
@@
 we have added an additional step to ensure that lakes are only included if they have data in the first and second half of the 1993 - 2016 time frame and now only lakes that have at least one measurement in 1993-2004 and 2005-2016, have a minimum of 10 years of data, and have at least 1 measurement in may/june, july/aug, sept/oct are included in the analysis.
-The filtering step has been added to the `data_clean.R` around line 233 and 316 and uses a new function, `filter_early_late()` which is in the `functions.R` file.
+The filtering step has been added to the `data_clean.R` around line 342 and 523 and uses a new function, `filter_early_late()` which is in the `functions.R` file.
 The impact on the analysis was minimal.
@@
 
-*1-3.* How did Secchi depth change? Why wasn’t this analyzed?
+***1-3.*** How did Secchi depth change? Why wasn’t this analyzed?
 
@@
 
-*1-4.* It is unclear how the trends in temperature, TP, chl-a, TN, and TN:TP compare to water quality condition and any standards that might be set in Rhode Island.
+***1-4.*** It is unclear how the trends in temperature, TP, chl-a, TN, and TN:TP compare to water quality condition and any standards that might be set in Rhode Island.
 Tables 1 and 2 give the range of values across lakes and there is clearly a large gradient in trophic status.
@@
 - **Response:** Another good suggestion.
-We have addressed in our response to comment *1-5b* and *1-8*.
+We have addressed this in our response to comment ***1-5b*** and ***1-8***.
 In particular, we have identified slope values that would indicate a meaningful ecological trend.
-We use literature values as well as state water quality criteria (when available) to inform those values.
-Additionally, we have added analysis to explore trends across trophic state.
+We use literature values to inform those values.
+Additionally, brief, preliminary results were added in the discussion looking at coarse trends across trophic state.
 
-*1-4a.* Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.
+***1-4a.*** Line 312: It is still unclear to me how this analysis of site-specific anomalies is robust to variation in timing of inclusion of given sampling locations.
 
@@
 
-```{r echo=TRUE}
+```{r include=FALSE, message=FALSE}
 library(dplyr)
@@
   mutate(year = year(ymd(paste0(year,"01-01"))))
+```
 
+```{r, echo=FALSE}
 site_gg <- examp %>%
@@
 site_gg
+```
 
+```{r, echo=FALSE}
 examp_site_summ <- examp %>% 
@@
 
-*1-5.* My biggest concern is possibly overstating the lack of trends in TP and TN and the positive trends in temperature and chl-a and then suggesting that temperature, not nutrients, are responsible for increasing chl-a.
+***1-5.*** My biggest concern is possibly overstating the lack of trends in TP and TN and the positive trends in temperature and chl-a and then suggesting that temperature, not nutrients, are responsible for increasing chl-a.
 The authors are somewhat cautious when making this conclusion, but do not fully recognize the limitations of this interpretation.
@@
 
-*1-5a.* I think the title overstates this result. I agree with increasing Chl-a trends, but don’t agree with stable nutrient trends.
+***1-5a.*** I think the title overstates this result. I agree with increasing Chl-a trends, but don’t agree with stable nutrient trends.
 
@@
 
-*1-5b.* I have not read Wasserstein et al.
+***1-5b.*** I have not read Wasserstein et al.
 2016, but agree with the general concept that a specific p-value threshold is not the only weight of evidence for determining whether there is a trend.
@@
 While these exact numbers are not universal, we use them as a starting point for assessing the trends.
-How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below.
+How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below and have changed somewhat from our initial submission, altough the broad patterns still hold.
 
-*1-5b-i.* TN in Rhode Island (Fig 5A) – The p-value is very small and the slope seems reasonably large.
+***1-5b-i.*** TN in Rhode Island (Fig 5A) – The p-value is very small and the slope seems reasonably large.
 I also see more positive means later in the record.
@@
 These changes have weakened the positive trend as the slope is smaller and the p-value larger.
-Additionally thre years, 1995, 1996, and 1997 now have insufficient data to be included in the figures and a third, 1998, has only three sites.
+Additionally three years, 1995, 1996, and 1997 now have insufficient data to be included in the figures and a third, 1998, has only three sites.
 Using the more explicit definition of a trend we now conclude that TN showed neither a statistically positive trend or an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.
@@
 
-*1-5b-ii.* TN in Lagos data set (Fig 5B) – the text says that “TN showed a slight positive trend”.
+***1-5b-ii.*** TN in Lagos data set (Fig 5B) – the text says that “TN showed a slight positive trend”.
 Here, the slope is smaller and the p-value is large.
@@
 
-- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instaed of the mean and these results have changed as well.
+- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instead of the mean and these results have changed as well.
 Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a negative change in nitrogen, with more below average years recently.
@@
 
-*1-5b-iii.* TP in Rhode Island (Fig 6A) – The authors state that TP does not show a trend over time.
+***1-5b-iii.*** TP in Rhode Island (Fig 6A) – The authors state that TP does not show a trend over time.
 I see a positive slope with a marginally significant p-value and more positive means later in the record.
@@
 - **Response:** Again, with the changes to the filtering the trend lines have slightly changed.
-We concur with Reviewer 1's take that TP is showing a slight increasing trend with some statistical support for a trend, athough the slope is not greater than the threshold we identified and greater frequency of above average years in recent years.
+TP is showing no trend with no statistical support for a trend, a very small slope and although above average years are evently distributed.
 We have indicated as such in the text.
 
-*1-5b-iv.* TP in LAGOS (Fig 6B) – “TP showed a slight decreasing trend”.
+***1-5b-iv.*** TP in LAGOS (Fig 6B) – “TP showed a slight decreasing trend”.
 I disagree.
@@
 - **Response:** Small changes due to filtering and new criteria for assessing the trend are also seen here.
-There is statistical support for a trend, that trend is positive and the magnitude of the slope suggests the possibility of ecological significant change in TP over a century.
+There is statistical support for a trend, that trend is positive, but the magnitude of the slope does not support the idea of an ecologically significant change in TP over a century.
 This has been added to the manuscript.
 
-*1-5c.* To put the magnitude of each slope in context, you could report the slope as a percent of the full range of slopes observed in the data set.
+***1-5c.*** To put the magnitude of each slope in context, you could report the slope as a percent of the full range of slopes observed in the data set.
 Even those trends that are “significant” have a relatively small slope compared to the variability between lakes.
@@
 In developing our approach to assessing the trends we now have ecologically relevant thresholds we can compare to.
-Both approaches address the same issue of putting the magnitude of the slopes in context.
+Both approaches address the same good suggestion of putting the magnitude of the slopes in context.
 
-*1-6.* This analysis also seems to be saying something about coherence.
+***1-6.*** This analysis also seems to be saying something about coherence.
 In some years, there is a wide range of anomalies, meaning that some lakes are much higher than their long-term average at the same time as other lakes are much lower.
@@
 
-*1-7.* In the discussion (lines 236-237) and in the title, the authors seem to want to conclude that because there were positive trends in temperature and chl-a but not in TP or TN (which I don’t agree as strongly with), that increasing chl-a is likely caused by increasing temperatures and not nutrients.
+***1-7.*** In the discussion (lines 236-237) and in the title, the authors seem to want to conclude that because there were positive trends in temperature and chl-a but not in TP or TN (which I don’t agree as strongly with), that increasing chl-a is likely caused by increasing temperatures and not nutrients.
 This is certainly interesting if it were the case, but it is well beyond the scope of what can be concluded here.
@@
 - **Response:**  We have softened our language in the discussion on original lines 236-237 considerably.
-They now read, "While our analysis is not capable of detecting causation, this association between increasing chlorophyll *a* and warming waters is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature alone."
+They now read, "While our analysis is not capable of detecting causation, that both chlorophyll *a* and temperature is increasing with less obvious trends in nutrients is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature."
 
-*1-7a.* This study simply lists the direction of trends of 5 parameters, but does not analyze correlations between them.
+***1-7a.*** This study simply lists the direction of trends of 5 parameters, but does not analyze correlations between them.
 
-- **Response:** Agreed.  Language softened and title changed to hopefully steer readers away from causation.  No changes made.
+- **Response:** Agreed.  Language softened and title changed to hopefully steer readers away from causation.
 
-*1-7b.* Even if it had tested correlations between parameters, it is still difficult to reach for causation.
+***1-7b.*** Even if it had tested correlations between parameters, it is still difficult to reach for causation.
 
-- **Response:** Agreed.  Language softened and title changed to hopefully steer readers away from causation.  No changes made.
+- **Response:** Agreed.  Language softened and title changed to hopefully steer readers away from causation.
 
-*1-7c.* The analysis of mean anomalies over time is interesting, but that simply means that the entire set of lakes in the state shows trends over time.
+***1-7c.*** The analysis of mean anomalies over time is interesting, but that simply means that the entire set of lakes in the state shows trends over time.
 What I’ve learned from reading some of the papers cited by this study and from my own investigations is that lakes near one another can have opposite trends.
@@
 - **Response:** We agree with the reviewer that lake scale trends and the drivers behind those trends are both very interesting questions and something our approach could contribute to; however, we feel that it is beyond the scope of this particular paper.
-Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purposes, looking at long term trends in Rhode Island and the Northeast.
+Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purposes, looking at long term trends in Rhode Island and the Northeast/Midwest.
 We do feel this is a very good suggestion for next steps and my co-authors and I are in active discussion on how to move forward with such an analysis.
-In the Discussion under *Trends* we briefly mention this application and that the anomaly approach could also be used at the individual lake level to assess for trends.
+In the Discussion at the the end of *Trends* we briefly mention this application and that the anomaly approach could also be used at the individual lake level to assess for trends.
 
-*1-8.* Line 274-275: the authors pose the interesting question about whether lakes of similar trophic state are showing similar trends.
+***1-8.*** Line 274-275: the authors pose the interesting question about whether lakes of similar trophic state are showing similar trends.
 I disagree that this test is beyond the scope of the study.
@@
 
-*1-9.* Line 64: Better define the spatial scales of the studies referenced in this paragraph.
+***1-9.*** Line 64: Better define the spatial scales of the studies referenced in this paragraph.
 Macro-scale, regional, subregional could be interpreted in different ways.
@@
 
-- **Response:** NEEDS RESPONSE
+- **Response:** Added information on extent, plus entire paragraph was re-written
 
-*1-10.* Line 71: I’m not sure why it was surprising that little change in nutrients and chl-a was reported in Oliver et al.
+***1-10.*** Line 71: I’m not sure why it was surprising that little change in nutrients and chl-a was reported in Oliver et al.
 2017?
@@
 
-- **Response:** NEEDS RESPONSE
+- **Response:** This paragraph was re-written and better describes the papers.
 
-*1-11.* Lines 99 – 102: please delete. Unnecessary.
+***1-11.*** Lines 99 – 102: please delete. Unnecessary.
 
@@
 
-*1-12.* Line 113: capitalize Secchi
+***1-12.*** Line 113: capitalize Secchi
 
@@
 
-*1-13.* Line 123: delete approximately. It seems that you are giving specific numbers here.
+***1-13.*** Line 123: delete approximately. It seems that you are giving specific numbers here.
 
@@
 
-*1-14.* Lines 202-207: The text simply repeats the numbers reported in tables 1 and 2.
+***1-14.*** Lines 202-207: The text simply repeats the numbers reported in tables 1 and 2.
 It would be more interesting to compare the populations at the two scales – in terms of the means and the ranges and the variety of trophic states represented and to also summarize the comparison of trends over time at the two scales.
@@
 - **Response:** Added an additional paragraph to the results that compares the two scales and includes a new table summarize the percentage of lakes at both scales in chlorophyll based trophic states.
-Another table was added to the trends sections of the results summarizing how we interepreted the trends at each scale.
+Another table was added to the trends sections of the results summarizing how we interpreted the trends at each scale.
 
-*1-15.* Line 285: “receive feedback” is a vague phrase here.
+***1-15.*** Line 285: “receive feedback” is a vague phrase here.
 Perhaps you mean that is important for managers and stakeholders to know what the long-term trends are?
@@
 - **Response:** We changed the wording on this to reflect a focus on the need to know about long term trends.
-Also we added three citation that specifically address that sharing results back with volunteers meets their expectations and can motivate continued involvement.
+Also we added three citations that specifically address that sharing results back with volunteers meets their expectations and can motivate continued involvement.
 
-*1-16.* Line 292: I was a little bit confused by this paragraph.
+***1-16.*** Line 292: I was a little bit confused by this paragraph.
 Why discuss ways of reducing nutrients when you have not identified increased nutrients as having an increasing trend?
@@
 
-- **Response:** This section was re-written.  We mention nutrient reductions as one possible hypothesis that long-term trends like those seen in RI can inform.
+- **Response:** This section was re-written.  We still mention nutrient reductions, but frame it as long-term trends, like those seen in RI, can help develop new hypotheses and inform existing ones.
 
@@
 
-*2-1.* The paper needs some contextual description of the RI lakes that could be achieved from GIS analysis.
+***2-1.*** The paper needs some contextual description of the RI lakes that could be achieved from GIS analysis.
 For example, what is the distribution of size and depth of the lakes in the data set?
@@
 Thanks for the suggestion.
-There are a few differences and are a useful addition to the paper.
+There are a few differences between the regions and this is a useful addition to the paper.
 
-*2-2.* There should be no reason to rely on “anecdotal evidence” for explaining warming air temperatures over the timeframe of the investigation (p.
+***2-2.*** There should be no reason to rely on “anecdotal evidence” for explaining warming air temperatures over the timeframe of the investigation (p.
 14, lines 269-272).
@@
 
-*2-3.* Some clarification on the methods is also needed.
+***2-3.*** Some clarification on the methods is also needed.
 The authors say that they used the same analytical methods for the LAGOSNE data set as they did for the Rhode Island lakes.
@@
 
-*2-4.* p. 6, line 90: Capitalize “States.”
+***2-4.*** p. 6, line 90: Capitalize “States.”
 
@@
 
-*2-5.* p. 10, line 188: Check spelling of “parameter.”
+***2-5.*** p. 10, line 188: Check spelling of “parameter.”
 

### 2020-09-21 manuscript/ecosphere_resubmit/manuscript.Rmd
getting very close on revisions!

@@
 - ^*3*^ ORISE, Narragansett, RI 02882
-- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
+- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
 - '**Corresponding author: hollister.jeff@epa.gov**'
@@
 Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
-We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+We analyze for a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
 URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
 The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively.
-We assess regional (LAGOSNE) and state (URIWW) trends with yearly median anomalies calculated on a per-station basis.
+We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis.
 Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016.
-Total nitrogen shows a weak increase driven by low years in the early 1990s.
-Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
-At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time.
+Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
+At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.
+Total phosphorus does show a very slight increase.
 In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
-This analysis suggests an association between lake temperature and primary production.
 Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.
@@
 
-Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale.
-Macro-scale (i.e., ~ 1,800,000 km ^2^) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
-This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern (i.e.
-~400,000 km^2^) and midwestern (i.e.
-~1,000,000 km^2^) US regions [@filstrup2014regional; @filstrup2018evidence].
+Previous studies using regional data from the northeastern and midwestern United States (US), approximately 400,000 km^2^ and 1,000,000 km^2^ respectively, have investigated spatial and temporal water quality trends and have demonstrated that drivers explaining changes spatially may not explain changes temporally.
+These differences may be due to complex interactions occurring at the different scales through time [@lottig2017macroscale].
+Additionally, differences between regions can cause nutrient (N and P) trends to have different drivers compared to ratios of the nutrients, and these may or may not align with sub-continental (i.e., ~ 1,800,000 km ^2^) trends [@collins2017lake].
+Similarly, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US regions [@filstrup2014regional; @filstrup2018evidence].
 Furthermore, little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
@@
 
-Prior studies have modeled water quality trends across a larger region of the northeastern US that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
@@
 Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017].
-At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.
+At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.
 
-Lastly, to facilitate a comparison of the context of each scale, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].
-We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculuture, Forest, and Developed.
+Lastly, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].
+We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.
 The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.
-Additional details are available in are detailed in Soranno et al.
+Additional details are available in Soranno et al.
 [-@sorannolagosne2017].
@@
 *Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.
-Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
 
@@
 2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
-3. Calculate the anomaly for each annual median at each site by subtracting the annual and reference medians.
+3. Calculate the anomaly for each annual median at each site by subtracting the sites reference long-term medians from the sites yearly median.
 4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.
@@
   select(year, variable, n) %>%
-  spread(variable, n)
+  spread(variable, n) %>%
+  rename("Year" = "year", "Chlorophyll" = "chla", 
+         "Nitrogen:Phosphorus" = "np_ratio", "Temperature" = "temp", 
+         "Total Nitrogen" = "total_n", "Total Phosphorus" = "total_p")
 
@@
   select(year, variable, n) %>%
-  spread(variable, n)
+  spread(variable, n)  %>%
+  rename("Year" = "year", "Chlorophyll" = "chla", 
+         "Nitrogen:Phosphorus" = "np_ratio", 
+         "Total Nitrogen" = "total_n", "Total Phosphorus" = "total_p")
 ```
@@
 *Linear regression on annual median anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].
-We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
+We used these standard procedures to look for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.
 For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0).
@@
 Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence.
-For this assessment, we are defining a trend as with the following: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame suggest an increasing or decreasing patter, and 4) the magnitude of the slope is used to infer an ecologically relevant change.
+For this assessment, we are defining a trend with the following criteria: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame are used to identify an increasing or decreasing pattern, and 3) the magnitude of the slope is used to infer an ecologically relevant change.
 Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.
@@
 0.035°C per year for trophic linkages), meet local criteria, and would also be indicative of meeting an important policy goal.
-For changes in chlorophyll, total phosphorus, and total nitrogen an ecologically meaningful trend is one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.
+
+For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.
 We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].
 For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.
-This level for total phosphorus would also protect against As we have defined it, slope magnitudes in excess of these per year values will be considered to have an ecologically meaningful trend.
+
+```{r np_thresh, include=FALSE, message=FALSE}
+d_and_m_1992 <- read_csv("data/downing_mccauley_1992_table_3.csv")
+np_thresh <- d_and_m_1992 %>%
+  mutate(molar_np = (total_n/total_p) * 2.21) %>%
+  group_by(n_lim) %>%
+  summarize(median_np = round(median(molar_np),0))
+```
+
+An ecologically meaningful trend for TN:TP could be one that would suggest a switch from nitrogen limitation to phosphorus limitation over a century.
+To identify this we use information from experimental manipulations of nitrogen in lakes that identify those lakes as being nitrogen limited [@downing1992nitrogen].
+In Downing and McCauley [-@downing1992nitrogen], they identified 34 studies and reported whether or not the study found the lake to be nitrogen limited and the ambient total nitrogen and total phosphorus concentration in μg/L.
+With this information we calculated a median molar TN:TP ratio for nitrogen limited lakes and for lakes that did not show nitrogen limitation.
+The difference between these two values would suggest a possible shift in the limiting nutrients.
+The median molar TN:TP for nitrogen limited lakes was `r np_thresh[np_thresh$n_lim == "Yes",]$median_np` and for lakes not limited by nitrogen was `r np_thresh[np_thresh$n_lim == "No",]$median_np`.
+The difference between these is `r abs(np_thresh[np_thresh$n_lim == "Yes",]$median_np - np_thresh[np_thresh$n_lim == "No",]$median_np)` and over a century a change of `r abs(np_thresh[np_thresh$n_lim == "Yes",]$median_np - np_thresh[np_thresh$n_lim == "No",]$median_np)/100` per year would be indicative of an ecologically meaningful change.
+As we have defined it, slope magnitudes in excess of these per year values (temperature: 0.02, chlorophyll: 0.035, total nitrogen: 3.5, total phosphorus: 0.1, TN:TP: 0.26) will be considered to have an ecologically meaningful trend.
 
@@
 [-@Oliver_2017] used hierarchical linear models and showed relatively stable water quality in the lakes of the northeastern United States.
-While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that in the case of Rhode Island regional trends were masking sub-regional trends.
+While the University of Rhode Island's Watershed Watch data were included in this regional study, we hypothesized that, in the case of Rhode Island, regional trends were masking sub-regional trends.
 Therefore, we decided to reanalyze the LAGOSNE data to compare the trends at the regional scale to the trends at the Rhode Island state scale using the site-specific anomaly and trend analysis approach outlined above.
@@
 Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
-Also, LAGOSNE lakes tended to occur in landscapes (i.e.
-a 500 meter buffer) with a higher average percent agriculture, less developed land, and comparable forested area than lakes in Rhode Island (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
+Also, Rhode Island lakes tended to occur in landscapes (i.e.
+a 500 meter buffer) with a lower average percent agriculture, more developed land, and comparable forested area than lakes in the LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
 
@@
 Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic].
+These results follow what would be expected given that, on average, LAGOSNE lakes are more dominated by agricultural lands than are Rhode Island lakes (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
 
@@
 
-Median annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (median anomaly) in Rhode Island lakes and reservoirs.
-Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = "Summary of long term water quality trends in the URIWW and LAGOSENE data.", display = "cite")`).
+Median annual temperature anomalies in lakes and reservoirs appear to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Twenty year trend for median lake temperature anomaly in Rhode Island.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles.
+Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSENE data.
+The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).
 Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).
-The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (median anomaly).
+The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorphyll anomaly.
 Panel A.
-Yearly median chlorophyll a anomalies from the URI Watershed Watch data.
+URI Watershed Watch yearly chlorophyll anomalies.
 Panel B.
-Yearly median chlorophyll a anomalies from the LAGOSNE dataset.
-Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+LAGOSNE yearly chlorophyll anomalies.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles.
+Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 
@@
 Median annual trends for nutrients were not as clear.
-For total nitrogen, the slope of the line is positive with weak statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.
-Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (median anomaly).
+For total nitrogen, the slope of the line is positive with some, albeit weak, statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.
+Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Twenty year trend for median TN anomaly.
 Panel A.
-Yearly median TN anomalies from the URI Watershed Watch dataset.
+URI Watershed Watch yearly TN anomalies.
 Panel B.
-Yearly median TN anomalies from the LAGOSNE dataset.
-Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
-Grey indicates yearly site specifica anomalies that were, on average equal to the long-term medians.
-Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
-However, 1998 only had three sites with available total nitrogen data and may not be representative.
+LAGOSNE yearly TN anomalies.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles.
+Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.
+Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+Also, 1998 only had three sites with available total nitrogen data and may not be representative.
 This year also recorded the lowest median value.
-Combined, 1998 may be skewing these results.
+Thus, 1998 may be skewing these results.
 When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.
 Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`).
-Total phosphorus shows essentially no trend over time in the yearly anomalies with no statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (median anomaly).
+Total phosphorus shows essentially no trend over time in the yearly anomalies with little statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Twenty year trend for median TP anomaly.
 Panel A.
-Yearly median TP anomalies from the URI Watershed Watch dataset.
+URI Watershed Watch yearly TP anomalies.
 Panel B.
-Yearly median TP anomalies from the LAGOSNE dataset.
-Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
-Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+LAGOSNE yearly TP anomalies.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles.
+Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.
+Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.
-The pattern for the TN:TP ratio has a small slope and no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (median anomaly).
+The TN:TP ratio has a small slope, very little statistical support, no pattern in the above and below years, and a slope magnitude less than 0.26 (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Twenty year trend for median TN:TP anomaly.
 Panel A.
-Yearly median TN:TP ratio anomalies from the URI Watershed Watch dataset.
+URI Watershed Watch yearly TN:TP anomalies.
 Panel B.
-Yearly median TN:TP ratio anomalies from the LAGOSNE dataset.
-Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
-Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+LAGOSNE yearly TN:TP anomalies.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles.
+Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.
+Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 
@@
 The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
-Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.
 Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.
@@
 
-Our sub-regional analysis indicates that increases in primary production, as measured by chlorophyll *a*, occurred over that ~25 years of our study period.
+Our sub-regional analysis indicates that increases in primary production, as measured by chlorophyll *a*, occurred over the ~25 years of our study period.
 Over the same period we also demonstrate long-term warming of Rhode Island lakes and reservoirs.
 Chlorophyll has increased, on average, `r signif(ww_chla_rl$slope, 2)` μg/L per year over the 23 years of our analysis, while temperature has increased `r signif(ww_temp_rl$slope, 2)` °C per year over the same period.
-Trends in nutrients were less clear with nitrogen showing a slight statistical increase with little ecological importance and phosphorus showing a slight statistical and ecological increase.
-The nitrogen:phosphorus ratio showed no evidence of a trend.
+Trends in nutrients suggested little change.
+Total nitrogen, total phosphorus, and the nitrogen:phosphorus ratio all showed no evidence of trends.
 While our analysis is not capable of detecting causation, that both chlorophyll *a* and temperature is increasing with less obvious trends in nutrients is interesting and warrants further exploration to see if increasing chlorophyll *a* can be described by temperature.
@@
 
-As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs.
+As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs (`r tab_num(name = "trend_summ", display = "cite")`).
 However, 2006 and 2009 stand out as not following this trend.
@@
 On the other hand, 2006 was not unusually cool (11th lowest out of 24 years) but it was the wettest year on record for this period [@noaatrends].
+These unusual weather patterns may help explain why 2006 and 2009 did not follow the trend.
 
 Trends in total nutrients and the TN:TP ratio are less clear.
-While TN showed a weak statistically positive trend, data for the early years (1994-1998) were relatively sparse.
+While TN showed a weak positive trend, data for the early years (1994-1998) were sparse.
 There is no ecological or statistical trend in the years with consistently available data.
-The general picture in Rhode Island appears to be one of a very slight increase in nitrogen, no trend in phosphorus and little to no change in the TN:TP ratio.
-Furthermore, it has been shown that productivity in freshwater systems is likely a function of both phosphorus and nitrogen [@paerl2016takes].
-Thus, the increasing chlorophyll *a* in the face of stable TN:TP ratio suggests that the increase is being driven by something other than nutrients.
+The general picture in Rhode Island appears to be no increase in nitrogen, no trend in phosphorus and little to no change in the TN:TP ratio (`r tab_num(name = "trend_summ", display = "cite")`).
 We interpret these results as relative stability in nutrients in Rhode Island lakes and reservoirs.
@@
 However, we conducted a quick, preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes.
-This suggests that a full analysis of how lakes in different trophic states are changing over time is warrented.
+This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.
 Lastly, there are other drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.
@@
 This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some citizen science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].
-Were that long-term involvement not common, then consistent long-term data may be more difficult to obtain.
+Were that long-term involvement not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.
 
@@
 For instance, warming waters are linked to increases in harmful algal blooms [@paerl2008blooms; @paerl2009climate; @paerl_climate_2012].
-An increase in blooms could be seen via increasing chlorophyll *a* levels.
+An increase in blooms could be assumed via increasing chlorophyll *a* levels as chlorophyll *a* has been linked to probability of the presence of bloom indicators such as microcystin [@Hollister_2016].
+Our results are consistent then with the prior research that temperature changes could be driving increased chlorophyll.
+
 Broad scale warming would be difficult to manage on a sub-regional scale thus other mechanisms, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
@@
 This is one possible hypothesis that is possible to form through information obtained by long-term trends.
-Additional work is certainly needed to examine the drivers of these trends in Rhode Island lakes and reservoirs.
+Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs.
 
@@
 The site-specific anomalies are also robust to variations in sampling effort and in the timing of inclusion of given sampling locations.
-For instance, if a site included early in a time period had low values and site included late in the time period, then analyzing the measured values over this time period would show an increase whether or not one existed.
-Using the site-specific anomalies rescales the values to a common framework and allows the real trend to be seen.
+For instance, if a site included early in a time period had low values and a site included late in the time period had high values, then analyzing the measured values over this time period would show an increase whether or not one existed.
+Using the site-specific anomalies rescales the values and allows the real trend to be seen.
 We illustrate this with simulated, random data (e.g.
-no trend) (`r fig_num(name = "simulated_data", caption = "Simulated, random data showing example of sites with a low site early in a time preiod, a high site late in a time period, and one site measured across the time period.", display = "cite")`).
-The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated, random data with a low site early in a time preiod, a high site late in a time period, and one site measured across the time period.
+no site-specific trend) (`r fig_num(name = "simulated_data", caption = "Simulated, random data showing example of sites with a low site early in a time period, a high site late in a time period, and one site measured across the time period.", display = "cite")`), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated_analysis", caption = "Analysis of simulated, random data with a low site early in a time period, a high site late in a time period, and one site measured across the time period.
+Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).
+The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated, random data (prior figure) with a low site early in a time period, a high site late in a time period, and one site measured across the time period.
 Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).
@@
 Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.
-Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long term environmental trends.
+Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long-term environmental trends.
 
@@
 RI0020-MS984.
-Lastly, the views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.
+The views expressed in this article are those of the authors and do not necessarily represent the views or policies of the U.S.
 Environmental Protection Agency.
@@
 ```{r year_num_table_ww, echo=FALSE, results="asis"}
-pander::pander(year_num_ww, style = "rmarkdown")
+#pander::pander(year_num_ww, format = "pipe")
+knitr::kable(year_num_ww, "pipe")
 ```
@@
 ```{r year_num_table_lagos, echo=FALSE, results="asis"}
-pander::pander(year_num_lagosne, style = "rmarkdown")
+knitr::kable(year_num_lagosne, format = "pipe")
 ```
@@
 ```{r context, echo=FALSE, results="asis"}
-pander::pander(context_summ, style = "rmarkdown")
+#knitr::kable(context_summ, format = "pipe")
+knitr::kable(context_summ)
 ```
@@
 ```{r ww_params, echo=FALSE, results="asis"}
-pander::pander(ww_param_ranges, style = "rmarkdown")
+knitr::kable(ww_param_ranges, format = "pipe")
 ```
@@
 ```{r lagos_params, echo=FALSE, results="asis"}
-pander::pander(lagos_param_ranges, style = "rmarkdown")
+knitr::kable(lagos_param_ranges, format = "pipe")
 ```
@@
 ```{r ts_table, echo=FALSE, results="asis"}
-pander::pander(ts_table, style = "rmarkdown")
+knitr::kable(ts_table, format = "pipe")
 ```
@@
 
-pander::pander(trend_summ_df, style = "rmarkdown")
+knitr::kable(trend_summ_df, format = "pipe")
 ```

### 2020-09-21 manuscript/ecosphere_resubmit/manuscript_sds.Rmd
getting very close on revisions!

--- 2020-08-28 manuscript/ecosphere_resubmit/manuscript.Rmd
@@
 
-Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale.
-Macro-scale (i.e., subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale].
-This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake].
-On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence].
-Thus, it was surprising when little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
-Given what is known about long-term trends in water quality within the broader region of the northeastern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
+Previous studies using regional data from the northeastern and midwestern United States (US), approximately 400,000 km^2^ and 1,000,000 km^2^ respectively, have investigated spatial and temporal water quality trends and have demonstrated that drivers explaining changes spatially may not explain changes temporally.
+These differences may be due to complex interactions occurring at the different scales through time [@lottig2017macroscale].
+Additionally, differences between regions can cause nutrient (N and P) trends to have different drivers compared to ratios of the nutrients, and these may or may not align with sub-continental (i.e., ~ 1,800,000 km ^2^) trends [@collins2017lake].
+Similarly, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US regions [@filstrup2014regional; @filstrup2018evidence].
+Furthermore, little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017].
+Given what is known about long-term trends in water quality within the broader regions of the northeastern and midwestern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.
 
@@
 
-During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
+During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
 
-For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
+For lakes and reservoirs in the larger region represented by the LAGOSNE States,  median TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).
 
@@
 
-Mean annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests a statistical trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with mean temperature greater than the long-term mean are occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).
+Median annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests a statistical trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (median anomaly) in Rhode Island lakes and reservoirs.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`).
 Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).
-The slope of 0.3 μg/L is much greater than our threshold of 0.035 μg/L and the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).
+The slope of 0.3 μg/L is much greater than our threshold of 0.035 μg/L and the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (median anomaly).
 Panel A.
-Yearly mean chlorophyll a anomalies from the URI Watershed Watch data.
+Yearly median chlorophyll a anomalies from the URI Watershed Watch data.
 Panel B.
-Yearly mean chlorophyll a anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).
+Yearly median chlorophyll a anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`A.).
 
@@
 
-Mean annual trends for nutrients were not as clear.
-For total nitrogen, the slope of the line is positive with some statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) and the years greater than the mean are somewhat more frequent in more recent years.
-However, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly).
+Median annual trends for nutrients were not as clear.
+For total nitrogen, the slope of the line is positive with some statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) and the years greater than the median are somewhat more frequent in more recent years.
+However, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (median anomaly).
 Panel A.
-Yearly mean TN anomalies from the URI Watershed Watch dataset.
+Yearly median TN anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly mean TN anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
-Grey indicates yearly site specifica anomalies that were, on average equal to the long-term means.
-Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+Yearly median TN anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
+Grey indicates yearly site specifica anomalies that were, on average equal to the long-term medians.
+Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
 We interpret this as a weak positive trend in TN, but with little ecological impact.
-Total phosphorus shows a positive trend over time in the yearly anomalies but with somewhat weak statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the mean are more common in recent years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).
+Total phosphorus shows a positive trend over time in the yearly anomalies but with somewhat weak statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median are more common in recent years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (median anomaly).
 Panel A.
-Yearly mean TP anomalies from the URI Watershed Watch dataset.
+Yearly median TP anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly mean TP anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
-Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+Yearly median TP anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
+Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
 Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is greater than our defined threshold of 0.1 μg/L and suggests a possible trophic state shift in less than a century.
-The pattern for the TN:TP ratio has a negative slope but no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).
+The pattern for the TN:TP ratio has a negative slope but no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (median anomaly).
 Panel A.
-Yearly mean TN:TP ratio anomalies from the URI Watershed Watch dataset.
+Yearly median TN:TP ratio anomalies from the URI Watershed Watch dataset.
 Panel B.
-Yearly mean TN:TP ratio anomalies from the LAGOSNE dataset.
-Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies.
-Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means.
-Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.
-Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+Yearly median TN:TP ratio anomalies from the LAGOSNE dataset.
+Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies.
+Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians.
+Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.
+Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
 Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
@@
 
+ADD PARAGRAPH ABOUT VARIANCE - Addresses Coherence comment
+
 Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].
@@
 
+######
+
 ```{r ww_params, echo=FALSE, results="asis"}
@@
 `r tab_num("ww_params")`
+
+######
 

### 2020-09-21 table.Rmd
getting very close on revisions!

---
title: "Untitled"
author: "Jeff Hollister"
date: "9/15/2020"
output:
  word_document:
    reference_docx: manuscript/manuscript_template.docx
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)
library(knitr)
library(kableExtra)
library(dplyr)
```

## Tables

```{r table, echo=FALSE}
mydf<-tibble(`Big Long Column Name Number1` = rnorm(10,1000,100),
       `Big Long Column Name Number2` = rnorm(10,1000,100),
       `Big Long Column Name Number3` = rnorm(10,1000,100),
       `Big Long Column Name Number4` = rnorm(10,1000,100),
       `Big Long Column Name Number5` = rnorm(10,1000,100),
       `Big Long Column Name Number6` = rnorm(10,1000,100)
       )
```

```{r kable, results="asis"}
kable(mydf,format = "markdown")
```

### 2020-09-21 R/data_clean.R

@@ -106,7 +106,7 @@ det_limits <- det_limit %>%
   select(year, param, det_limit)
   
 
-# Final Clean
+# First Clean
 ww_lake_trend_data <- ww_lake_trend_data %>%
   select(station_name = `Station Name`, year, month, day, depth = Depth, 
          location = Location, param = Parameter, 
@@ -119,7 +119,7 @@ ww_lake_trend_data <- ww_lake_trend_data %>%
                                  T ~ measurement)) %>% # Convert TN to ug/l
   #filter(year >= 1990) %>%
   group_by(station_name, year, month,day,param,location) %>%
-  summarize(mn_measurement = mean(measurement, na.rm = TRUE), #some sampling events had multiple measurements - standardizes to one pre site/date/param
+  summarize(mn_measurement = mean(measurement, na.rm = TRUE), #some sampling events had multiple measurements - standardizes to one per site/date/param
             mdn_measurement = median(measurement, na.rm = TRUE))%>%
   ungroup() 
 

### 2020-09-21 R/figures_anomaly_only.R

@@ -210,27 +210,27 @@ write_csv(figure_data, here("data/yearly_average_anomaly.csv"))
 set.seed(0)
 options(scipen=10000)
 
-site1 <- tibble(site=rep("site 1",10),
-                year = 2011:2020,
-                rep_1 = rnorm(10,10,2.5),
-                rep_2 = rep_1*rnorm(10, 1, 0.05),
-                rep_3 = rep_1*rnorm(10, 1, 0.05)) %>%
+site1 <- tibble(site=rep("site 1",24),
+                year = 1993:2016,
+                rep_1 = rnorm(24,10,2.5),
+                rep_2 = rep_1*rnorm(24, 1, 0.05),
+                rep_3 = rep_1*rnorm(24, 1, 0.05)) %>%
   pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
   mutate(replicate = str_extract(replicate, "[0-9]"))
 
-site2 <- tibble(site=rep("site 2",6),
-                year = 2011:2016,
-                rep_1 = rnorm(6,3,1),
-                rep_2 = rep_1 * rnorm(6, 1, 0.05),
-                rep_3 = rep_1 * rnorm(6, 1, 0.05)) %>%
+site2 <- tibble(site=rep("site 2",12),
+                year = 1993:2004,
+                rep_1 = rnorm(12,3,1),
+                rep_2 = rep_1 * rnorm(12, 1, 0.05),
+                rep_3 = rep_1 * rnorm(12, 1, 0.05)) %>%
   pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
   mutate(replicate = str_extract(replicate, "[0-9]"))
 
-site3 <- tibble(site=rep("site 3",6),
-                year = 2015:2020,
-                rep_1 = rnorm(6, 17, 1),
-                rep_2 = rep_1*rnorm(6, 1, 0.05),
-                rep_3 = rep_1*rnorm(6, 1, 0.05)) %>%
+site3 <- tibble(site=rep("site 3",13),
+                year = 2004:2016,
+                rep_1 = rnorm(13, 17, 1),
+                rep_2 = rep_1*rnorm(13, 1, 0.05),
+                rep_3 = rep_1*rnorm(13, 1, 0.05)) %>%
   pivot_longer(rep_1:rep_3, names_to = "replicate", values_to = "values") %>%
[…]

### 2020-09-21 R/schematic_anomaly.R

@@ -22,6 +22,8 @@ median_anomaly <- station_year_anomaly %>%
   group_by(year) %>%
   summarize(median_anomaly = median(station_year_anomaly),
             iqr_anomaly = IQR(station_year_anomaly),
+            percentile_25 = quantile(station_year_anomaly, 0.25),
+            percentile_75 = quantile(station_year_anomaly, 0.75),
             n = n()) %>%
   round(2) %>%
   ungroup()

### 2020-09-21 R/summary_stats.R

@@ -196,7 +196,7 @@ all_context_summ <- all_lakes_context %>%
   summarize(min = min(value),
             `25th percentile` = quantile(value, 0.25),
             med = median(value),
-            avg = mean(value),
+            avg = round(mean(value),1),
             `75th percentile` = quantile(value, 0.75),
             max = max(value)) %>%
   filter(param %in% c("Agriculture", "Forest", "Developed", "lake_area_ha", 
@@ -206,11 +206,11 @@ all_context_summ <- all_lakes_context %>%
                            param == "maxdepth" ~
                              "Maximum Depth (m)",
                            param == "Agriculture" ~
-                             "Agriculture (%)",
+                             "Percent Agriculture",
                            param == "Forest" ~
-                             "Forest (%)",
+                             "Percent Forest",
                            param == "Developed" ~
-                             "Developed (%)",
+                             "Percent Developed",
                            TRUE ~ param),
          units = c("%", "%", "%", "ha", "m")) %>%
   mutate(Source = "LAGOSNE") %>%
@@ -224,7 +224,7 @@ ww_context_summ <- all_lakes_context %>%
   summarize(min = min(value),
             `25th percentile` = quantile(value, 0.25),
             med = median(value),
-            avg = mean(value),
+            avg = round(mean(value), 1),
             `75th percentile` = quantile(value, 0.75),
             max = max(value)) %>%
   filter(param %in% c("Agriculture", "Forest", "Developed", "lake_area_ha", 
@@ -234,11 +234,11 @@ ww_context_summ <- all_lakes_context %>%
                            param == "maxdepth" ~
                              "Maximum Depth (m)",
                            param == "Agriculture" ~
-                             "Agriculture (%)",
+                             "Percent Agriculture",
                            param == "Forest" ~
-                             "Forest (%)",
+                             "Percent Forest",
                            param == "Developed" ~
-                             "Developed (%)",
+                             "Percent Developed",
                            TRUE ~ param),
          units = c("%", "%", "%", "ha", "m")) %>%
   mutate(Source = "URIWW") %>%

### 2020-09-25 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
push before resubmit to ecosphere

@@
 ```
+
+We would like to extend our thanks to the reviewers for two high quality reviews.
+We found the comments to be thoughtful and in all cases the final product, we feel, is much better that what we originally submitted.
+Below you will find our itemized response to each comment made by the reviewers.
+
 # Reviewer 1
@@
 - **Response:**  Good catch!
-Fortunately, it has not greatly impacted the analysis.
-The table below lists the percentage of lake-years for each parameter that had at least one sample in each of May/June, July/Aug, and Sept/Oct.
-While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.
+Fortunately, it has not greatly impacted the analysis as we have full seasonal coverage for the majority of lake-years.
+That being said, we still decided to follow the reviewer's suggestion and filter these out.
 This filtering step has been added to our `data_clean.R` script at lines 157 and 398 and is in the `filter_months()` function in `functions.R`.
@@
 We added text in the second paragraph of the Methods section describing the selection criteria.
-
-```{r echo = FALSE, results="asis"}
-knitr::kable(param_min_samp)
-```
 
@@
 - **Response:** What we mean by this, is if sites with differing base lines (e.g.
-low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends because each of the sites will be re-scaled, relative to that sites long term mean.
+low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends because each of the sites will be re-scaled, relative to that sites long term median.
 For instance, comparing measured values from a cool water site that shows no trend which is measured mostly early in a sampling period to a warm water site without a trend measured late in a sampling period would likely show an increasing trend when in fact none existed.
-Re-scaling these values relative to the mean sets both sites to a common baseline and would then not show a trend.
+Re-scaling these values relative to the median sets both sites to a common baseline and would then not show a trend.
 To further illustrate this, look at the two figures below.
@@
 The anomalies correctly show no increasing trend.
-We have added in these figures and text to the Discussion
+We have added in similar figures and text to the Discussion
 
@@
   group_by(site) %>%
-  mutate(lt_mean = mean(values)) %>%
+  mutate(lt_median = median(values)) %>%
   ungroup() %>%
   group_by(site, year) %>%
-  summarize(measured_value = mean(values),
-            anomaly = mean(values-lt_mean)) %>%
+  summarize(measured_value = median(values),
+            anomaly = median(values-lt_median)) %>%
   ungroup() %>%
@@
   group_by(year, Method) %>%
-  summarize(values = mean(value)) %>%
+  summarize(values = median(value)) %>%
   ungroup()
@@
 We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.
-We now lay out a more explicit approach and describe how we use sign of the slope, magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.
+We now lay out a more explicit approach and describe how we use magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.
 This is described in the paper in a new section in the methods, *Assessing regressions for trends*.
@@
 While these exact numbers are not universal, we use them as a starting point for assessing the trends.
-How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below and have changed somewhat from our initial submission, altough the broad patterns still hold.
+How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below and have changed somewhat from our initial submission, although the broad patterns still hold.
 
@@
 
-- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*) and made changes to use the median instead of the mean in our response to comment *1-6*.
+- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (**1-1** and **1-2**) and made changes to use the median instead of the mean in our response to comment **1-6**.
 These changes have weakened the positive trend as the slope is smaller and the p-value larger.
@@
 
-- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instead of the mean and these results have changed as well.
-Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a negative change in nitrogen, with more below average years recently.
+- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instead of the mean (see **1-6**) and these results have changed as well.
+Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a change in nitrogen with above and below average years showing little pattern.
 The slope is much less than the 3.5 we identified.
@@
 
-- **Response:** Again, with the changes to the filtering the trend lines have slightly changed.
-TP is showing no trend with no statistical support for a trend, a very small slope and although above average years are evently distributed.
+- **Response:** Again, with the changes to the filtering and the switch to the median the trend lines have slightly changed.
+TP is showing no trend with no statistical support, a very small slope and above/below average years are relatively evenly distributed.
 We have indicated as such in the text.
@@
 Further, there is no pattern to these outliers as they are from different ponds, aren't seen across variables in the same years, and in the case of the LAGOSNE data, they come from different states.
-There is only one exception to this and that is the high variability in 2011 for both TN and TP.
+There is only one exception to this and that is the high variability in 2011 for both TN and TP in Rhode Island.
 That is driven by one sampling event where TN and TP were high for both.
@@
 - **Response:** We agree with the reviewer that lake scale trends and the drivers behind those trends are both very interesting questions and something our approach could contribute to; however, we feel that it is beyond the scope of this particular paper.
-Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purposes, looking at long term trends in Rhode Island and the Northeast/Midwest.
+Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purpose of looking at long term trends in Rhode Island and the Northeast/Midwest.
 We do feel this is a very good suggestion for next steps and my co-authors and I are in active discussion on how to move forward with such an analysis.

### 2020-09-25 manuscript/ecosphere_resubmit/manuscript.Rmd
push before resubmit to ecosphere

@@
 
-```{r analysis, include=FALSE}
+```{r analysis, include=FALSE, cache=TRUE}
+source(here::here("R/data_clean.R"), encoding = "UTF-8")
 source(here::here("R/data_available.R"), encoding = "UTF-8") 
@@
 Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
-For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
@@
 *Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.
-Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
 
@@
 After filtering and summarizing the data some years may not have sufficient number of sites to be included.
-We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data(`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).
+We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data (`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).
 
@@
 ```{r np_thresh, include=FALSE, message=FALSE}
-d_and_m_1992 <- read_csv("data/downing_mccauley_1992_table_3.csv")
+d_and_m_1992 <- read_csv(here::here("data/downing_mccauley_1992_table_3.csv"))
 np_thresh <- d_and_m_1992 %>%
@@
 In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
-Total nitrogen showed a slight decreasing trend with weak statistical support and less than the 3.5 μg/L that would result in a trophic state change over a century.
-The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`).
-There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.
 Total phosphorus showed a very small increasing trend and there is statistical support for that trend.
@@
 [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.
-Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/usepa/ri_wq_trends> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
+Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
 
@@
 As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs (`r tab_num(name = "trend_summ", display = "cite")`).
-However, 2006 and 2009 stand out as not following this trend.
+However, 2006 and 2009 stand out as not following this trend for temperature.
 Average May-October air temperature for 2009 was third lowest on record for the 1993 to 2016 time period and also showed the lowest maximum temperatures [@noaatrends].
@@
 This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some citizen science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].
-Were that long-term involvement not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.
+If long-term involvement was not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.
 
@@
 
-Broad scale warming would be difficult to manage on a sub-regional scale thus other mechanisms, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
+One possible hypothesis is that broad scale warming is difficult to manage on a sub-regional scale thus other mechanisms and approaches, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
 Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
-This is one possible hypothesis that is possible to form through information obtained by long-term trends.
-Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs.
+There are of course many other hypotheses that we could form through information obtained by long-term trends such as those reported here.
+Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs and turn that information into management action.
 

### 2020-09-25 R/figures_anomaly_only.R

@@ -287,8 +287,8 @@ simulated_trends_gg <- ggplot(examp_yr_summ,
                                                                  ", p-value = ", signif(meas_lm$p.value, 2), "\nAnomaly: slope = ", signif(anom_lm$slope, 2), 
                                                                  ", p-value = ", signif(anom_lm$p.value, 2))) +
   theme_ipsum_rc() +
-  scale_x_continuous(labels = c(2012,2014,2016,2018,2020),
-                     breaks = c(2012,2014,2016,2018,2020)) +
+  scale_x_continuous(labels = c(1995,2000,2005,2010,2015),
+                     breaks = c(1995,2000,2005,2010,2015)) +
   theme(plot.subtitle = element_text(size=10, face="plain"))
 simulated_trends_gg %>%  
   ggsave(here("figures/simulated_analysis.jpg"), ., width = 7.5, height = 5.625,

### 2020-09-25 manuscript/ecosphere_resubmit/manuscript.Rmd
last resubmission commit

@@
 ---
-title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
+title: "**Title:** Analyzing long-term water quality of lakes in Rhode Island and the Northeastern United States with an anomaly approach"
 author:
@@
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
 
@@
 *Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.
-Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
 
@@
 [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.
-Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].
+Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
 

### 2020-09-30 manuscript/ecosphere_resubmit/manuscript.Rmd
few small changes - title was a bit off.

@@
 The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface].
-However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.
-We built on these methods and adapted them for use with long-term water quality trends.
+However, using it to examine water quality trends is a little-used application of the technique, as we only found a single study using anomalies in lakes and that study's focus was only on lake temperature [@oreilly2015warming].
+There is little evidence of using anomalies more broadly with water quality trends.
+Thus, we built on these methods and adapted them for use with long-term water quality trends.
 While other methods are valid and robust [e.g., @Oliver_2017], we chose median site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.

### 2020-11-25 manuscript/ecosphere_resubmit_2/ecosphere_response_2.Rmd
second resubmission revisions started

---
output: 
  word_document:
    reference_docx: response_template.docx
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE}
library(knitr)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
opts_chunk$set(message = FALSE, warning = FALSE)
```

General comments and thanks

**Subject-matter Editor Comments:**

1.) Line 15-16 - This sentence is a bit hard to read, consider rewriting and adding "water quality trends" somewhere in there.

*Response:* Changed to: "We use the University of Rhode Island's Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there are sub-regional (i.e. 3000km^2^) water quality trends."

2.) Line 75 - consider changing "we were curious" to "we investigated"

*Response:* Changed as suggested.

3.) Line 82 - not sure what in situ data as well as whole water samples means. It seems that whole water samples are in situ data, so I'm not sure what you're distinguishing here.

*Response:* Distinction was meant to be field data (e.g. sonde or secchi) vs samples for lab analysis. Changed to: "Volunteer monitors are trained and then collect field data (e.g., sonde measurements or secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab. "

4.) Line ~77 - One thing I think that is missing from this general area of the introduction is some context for RI lakes. You talk a lot about broad spatial scales and the types of landscape-scale changes that could influence water quality, but say little about RI specifically. I think a few sentences about what kinds of human pressures RI lakes are facing would be helpful context.

*Response:* NEEDS RESPONSE

5.) Line 78 - I appreciate the shout-out to to the watershed watch program. You could consider combining this info with the WW info that's in the first paragraph of the methods, and keeping all of that information in the methods. Especially if you decide to add more context about RI lakes in the introduction, this could be a way of keeping the introduction short and snappy.

*Response:* Moved this section from the Introduction to Methods.  Good suggestion.  Both Intro and Methods read better now, we think.

6.) Line 84 and throughout the manuscript - please consider a more inclusive term than "citizen science". One example I like is "community science" - here's an example of an org using it: <https://monarchjointventure.org/blog/were-changing-the-language-we-use-to-talk-about-public-participation-in-scientific-research>

*Response:* We like the new term!  Have replaced "citizen science" with "community science"

7.) Line 117-118 - do you mean had one sampling event in each month combination, in each year?

*Response:* Exactly.  Edited this to hopefully be more clear.  Now reads: "had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October"

8.) Line 175 - missing an s on "variable"

*Response:* Fixed

9.) Line 213-214 - consider changing "we are defining a trend with the following criteria" to "we evaluated trends based on the following criteria"

*Response:* Changed

10.) Line 225 - I think there should be an "or" after the comma

*Response:* Cleaned that sentence up.

11.) Line 237-239 - I'm slightly confused by these rate change calculations. Doesn't this rate of change assume you are starting from 0, which is not the case?

*Response:* NEEDS RESPONSE

12.) Line 257 - I think saying "we hypothesized regional trends were masking sub-regional trends" is good justification for including more RI-specific context in the introduction.

*Response:* Agreed.  Have included that in the Introduction.

13.) Line 262 - include a comma after "on average"

*Response:* Fixed.

14.) Tables 1 and 2 -- There is a lot of info here, and it's difficult to either look for trends through time or across sources in data availability. To be more concise plus enhance interpretability, I suggest a two panel plot, with panel A being URIWW and panel B being LAGOS, x axis is year and y axis is number of sites. Lines could be colored by constituent, and I think NA values should actually be zeros? If this turns out to be not as interpretable/useful than I thought, then I recommend putting these into a supplement.

*Response:* NEEDS RESPONSE

15.) Lines 266 -- might be a good place to remind readers they're dealing with surface/epi values here.

*Response:* Added in: "of surface and epilimnion measurements".

16.) Figure 2 -- really helpful figure!

*Response:* Thanks!  After about 100 failed attempts at trying to describe this approach, we finally decided to put this together.  Actually caught some initial errors in code with this!

17.) Trend plots -- describe what the line is in the caption, increase axis title text size

*Response:* NEEDS RESPONSE

18.) Multi-panel trend plots -- for publication, I suggest reducing the vertical space between panels.

*Response:* NEEDS RESPONSE

19.) Figure 8 and 9 -- Because these are based on hypothetical systems with generated data, and are a relatively small part of your overall story, I suggest trying to reduce to a single figure.

*Response:* NEEDS RESPONSE

20.) Figure 8 and elsewhere (like the paragraph that begins on line 432) -- be very careful with the language used here. I was confused at first with this caption, because you use site twice: "...example of sites with a low site early in a time...". Instead, I suggest using something like the language as follows: "...example of a hypothetical lake with data from three difference sites; one sampling location that was monitored throughout the record, one sampling location with low reported values early in the record, and another sampling location with high reported values late in the record."

*Response:* NEEDS RESPONSE

21.) Table 7 -- misspelled LAGOSNE

*Response:*  Oops!  Fixed.

22.) Lines 321-324 -- The data availability statement is a bit awkward in its current location. I suggest moving it to somewhere in the methods.

*Response:* Moved this from Results to Methods.

23.) Line 357-358 -- This is confusing -- enough nutrients to cause increasing chlorophyll? Do you mean, enough of an increase in nutrients?

*Response:* Was trying to make the case the nutrient reductions may not result in nutrient limitation.  Have changed this to state that more directly.  It now reads: "However, these nutrient reductions may not result in nutrient limitation and when when faced with other changes (e.g. temperature increases) increasing chlorophyll *a* may still occur."

24.) Line 377 -- suggest changing to "However, we conducted a preliminary"

*Response:* Changed this.

25.) Paragraph beginning on line 366 -- I find this paragraph to be a bit all over the place, and could use some direction/cleaning up. For example, I think you could provide more context: why do we care about temp trends in May, or chlorophyll changes in certain types of lakes? I think the Stoddard paper showing changes in oligotrophic systems is relevant here, and I think it's worth mentioning that a 3.5 ug/L per year change in an oligotrophic lake could be very different than the same change in a eutrophic lake. Also, the paragraph ends with a switch to suggesting individual lake trends, which seemingly comes out of nowhere -- please provide more context in what situations that would be useful.

*Response:* NEEDS RESPONSE

26.) Paragraph beginning on line 408 -- the language here confuses me; the hypothesis about what observation? I think you could drop that language all together, and start the paragraph with "Broad scale warming is difficult...". I think this paragraph should be cleaned up, and clearly convey 2-3 points: 1) it's difficult to manage for warming, which 2) emphasizes the need for nutrient control measures, and possibly 3) managing blooms will be an uphill battle against warming waters

*Response:* NEEDS RESPONSE

27.) Data analysis approach section -- are there any downsides to the anomaly approach?

*Response:* Added in a new paragraph before the last paragaph of the discussion mentioning a few caveats.

### 2020-11-25 manuscript/ecosphere_resubmit_2/manuscript.Rmd
second resubmission revisions started

--- 2020-09-30 manuscript/ecosphere_resubmit/manuscript.Rmd
@@
 Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
-We analyze for a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset.
+We use the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there are sub-regional (i.e.
+3000km^2^) water quality trends.
 URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
@@
 Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.
-  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
+  | **Keywords:** Volunteer Monitoring, community Science, Nutrients, Lakes, Trend Analysis, Long Term
 ---
@@
 
-Examining long-term trends in Rhode Island lakes is possible because of the data gathered by University of Rhode Island’s Watershed Watch (URIWW).
-URIWW is a scientist-led citizen science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
-Volunteer monitors are trained and then collect *in situ* data as well as whole water samples during the growing season (e.g., May through October).
-The entire effort follows rigorous quality control/quality assurance protocols.
-These types of citizen science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].
-URIWW data contributed to the larger regional study by Oliver et al.
-[-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.
+Add Paragraph about RI lakes: Context about human pressures on RI lakes + one sentence about adequate data being available
 
@@
 
-The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`).
+The study area for this analysis includes lakes and reservoirs in the state of Rhode Island where data were collected by the University of Rhode Island's Watershed Watch program (`r fig_num("ww_map", display = "cite", caption = "Map of URI Watershed Watch lake and reservoir sampling sites")`) which allows for the examination of long-term trends in Rhode Island lakes.
+The University of Rhode Island's Watershed Watch (URIWW) is a scientist-led community science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
+Volunteer monitors are trained and then collect field data (e.g., sonde measurements or secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab.
+The entire effort follows rigorous quality control/quality assurance protocols.
+These types of community science efforts allow for the collection of reliable data that in turn lead to crucial and frequently unexpected insights [@dickinson2012current; @kosmala2016assessing; @Oliver_2017].
+
+URIWW data contributed to the larger regional study by Oliver et al.
+[-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.
 The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
@@
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event in each of May/June, July/August, and September/October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.
+In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.
 As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
@@
 
-Lastly, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].
+We use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].
 We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.
@@
 [-@sorannolagosne2017].
+
+Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
+Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.
 
@@
 This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.
-Many of the variable we looked at are non-normal and often have outliers (e.g.
+Many of the variables we looked at are non-normal and often have outliers (e.g.
 algal blooms).
@@
 Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence.
-For this assessment, we are defining a trend with the following criteria: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame are used to identify an increasing or decreasing pattern, and 3) the magnitude of the slope is used to infer an ecologically relevant change.
+For this assessment, we evaluated trends based on the following criteria: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame are used to identify an increasing or decreasing pattern, and 3) the magnitude of the slope is used to infer an ecologically relevant change.
 Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.
@@
 There are various ways to identify important temperature changes in lakes and reservoirs.
-A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year and similarly, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs].
+A policy based approach could examine agreements that attempt to limit temperature increases, such as the Paris Agreement, which aims to limit increases over the next century to 2°C, which would be 0.02°C per year or, the state of Rhode Island criteria for temperature increase is 2.2°C or 0.022°C per year over a century [@paris; @ri_wq_regs].
 Alternatively, we could look at more ecologically focused temperature changes.
@@
 
-Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
+Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average, than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
 Also, Rhode Island lakes tended to occur in landscapes (i.e.
@@
 
-During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
+During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset of surface and epilimnion measurements had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).
 
@@
 Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSENE data.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data.
 The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).
@@
 [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region.
-Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
 
@@
 Stable nutrient regimes may be partly explained by efforts to curb nutrient loadings, for example through voluntary and state wide mandatory bans on phosphates in laundry detergent which were implemented in Rhode Island in 1995 [@riphosban1995; @litke1999review].
-However, in many lakes there are still likely sufficient nutrients present to allow for increases in chlorophyll *a*.
-Additionally, these results point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.
+However, these nutrient reductions may not result in nutrient limitation and when when faced with other changes (e.g.
+temperature increases) increasing chlorophyll *a* may still occur.
+Additionally, our analysis point to the fact that chlorophyll *a* and algal biomass is driven by processes operating at different scales.
 For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
@@
 A full analysis and interpretation of trends by trophic state is beyond the scope of this study.
-However, we conducted a quick, preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes.
+However, we conducted a preliminary analysis of chlorophyll trends in Rhode Island for oligotrophic and hypereutrophic lakes which showed increasing chlorophyll trends for both trophic states but the yearly increase was much greater in the hypereutrophic lakes.
 This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.
@@
 Specifically for this study, the results provide feedback to long time volunteer monitors about the trends in long-term, volunteer collected data.
-This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some citizen science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].
+This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some community science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].
 If long-term involvement was not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.
@@
 
-One possible hypothesis is that broad scale warming is difficult to manage on a sub-regional scale thus other mechanisms and approaches, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
+Broad scale warming is difficult to manage on a sub-regional scale thus other mechanisms and approaches, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
 Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
@@
 
+Site-specific anomalies do have broad utility for the analysis of water quality trends, however; there are a few caveats the must be considered.
+Our use of site-specific anomalies as outlined in this paper, requires aggregating over years.
+Any time period could be used for aggregating, but this is a decision that will need to be carefully addressed when conducting this type of analysis.
+Site-specific anomalies also require fairly large amounts of data.
+The required data will need to be consistently collected over a long period of time.
+Ideally, the data would cover a reference period (e.g.
+for our analysis that was 24 years) although other data could possibly be used to estimate the long-term medians.
+In short, site-specific anomalies do require relatively large datasets and decisions on appropriate levels of aggregation, as with any analysis using aggregation, must be made carefully.
+
 Lastly, this analysis is only possible because of the availability of sound, long-term data on water quality in Rhode Island.
 Without the URIWW data and the commitment and participation of more than 2500 volunteers over the years, our analyses would have been impossible.
-Going forward, it is important to appreciate the role that volunteer monitoring and citizen science programs can play in capturing and better understanding long-term environmental trends.
+Going forward, it is important to appreciate the role that volunteer monitoring and community science programs can play in capturing and better understanding long-term environmental trends.
 

### 2020-12-02 manuscript/ecosphere_resubmit_2/ecosphere_response_2.Rmd
Getting closer on edits.

@@
 
-*Response:* NEEDS RESPONSE
+*Response:* You are correct.
+This does assume that a change of the magnitude defined by the Nürnberg oligotrophic-mesotrophic cut-off would result in a change even for a lake starting with 0 chla, tn, or tp.
+To clarify, we made a slight change to the wording on this and it now reads "...
+an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century."   The main change being "any oligotrophic lake" and added the following at the end: "A slope at or above these values would result in a oligotrophic-mesotrophic transition for all oligotrophic lakes, even lakes with the impossible situation of chlorophyll, total nitrogen, or total phosphorus concentrations of zero."
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* We agree that this paragraph was not focused and was also quite long.
+We have split the paragraph apart.
+It currently now has two paragraphs, with the first paragraph focusing on lengthening of the growing season and the second paragraph focusing on trophic state.
+We also include your suggestions in this second paragraph.
+The last part of our original paragraph that included other drivers and individual lakes was an admittedly ham-handed effort at covering our bases as well as dealing with reviewer comments.
+Given that this last section of the paragraph does not add much to the discussion we have removed it.
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Re-wrote this paragraph with a focus on these three main points.
 
@@
 
-*Response:* Added in a new paragraph before the last paragaph of the discussion mentioning a few caveats.
+*Response:* Added in a new paragraph before the last paragraph of the discussion mentioning a few caveats.

### 2020-12-02 manuscript/ecosphere_resubmit_2/manuscript.Rmd
Getting closer on edits.

@@
 
+<!-- Jeff start here for change to comment #4 -->
 Add Paragraph about RI lakes: Context about human pressures on RI lakes + one sentence about adequate data being available
@@
 
-For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in a change from an oligotrophic state to a mesotrophic state over the course of a century.
+For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century.
 We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].
 For chlorophyll the oligotrophic-mesotrophic limit is 3.5 μg/L and over a century that is 0.035 μg/L per year, for total nitrogen the limit is 350 μg/L, or 3.5 μg/L per year, and the total phosphorus limit is 10 μg/L or 0.1 μg/L per year.
+A slope at or above these values would result in a oligotrophic-mesotrophic transition for all oligotrophic lakes, even lakes with the impossible situation of chlorophyll, total nitrogen, or total phosphorus concentrations of zero.
 
@@
 
-In addition to the annualized trends of the five variables we address with this study, there are other trends that may be of interest.
-For example, changes in water quality at finer temporal scales, such as monthly or seasonal, may be different than the annual trends we analyzed.
-Evidence in Rhode Island points to possible lengthening of the growing season as both May and October show increasing temperature trends over time [@noaatrends].
-Furthermore, preliminary analysis of the URIWW data back this up with mean temperature for May 1993 to May 1995 cooler by nearly a degree than mean temperature for May 2014 through May 2016.
-Additionally, the current trophic state of a given waterbody may partly explain the chlorophyll *a* changes in that lake.
-For instance, are oligotrophic lakes showing different trends than eutrophic lakes or are all lakes showing similar trends regardless of current trophic status?
+In addition to the sub-regional annualized trends of the five variables we address with this study, other trends may also be of interest.
+For example, a lengthening of the growing season could increase the number of days with conditions that favor the growth of harmful algal blooms.
+Many national and regional studies have documented longer growing seasons [e.g., @vega2019changes; @cooter1995recent; @kunkel2004temporal].
+Evidence in Rhode Island also points to possible lengthening of the growing season as both May and October, the beginning and end of the sampling dates in our dataset, show increasing temperature trends over time [@noaatrends].
+Furthermore, preliminary analysis of the URIWW data supports the idea that growing seasons may be getting longer in Rhode Island.
+Cyanobacteria grow better at temperatures greater than than 25°C [@paerl2008blooms; @johnk2008summer; @reynolds2006ecology].
+Early in our study period (1993-1995), the average first day of lake temperatures exceeding 25°C was June 11th, at the end of our study period (2014-2016) these warmer temperature were seen, on average, on June 1st.
+A site-specific anomaly analysis of growing season length could shed further light on potential changes to growing seasons in Rhode Island lakes.
+
+Trends, and their ecological importance, could also vary depending on trophic state of a given waterbody.
+We identified a minimum ecologically important change of 0.035 µg/l per year of chlorophyll.
+This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.
+Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only a single rate for all lakes.
+Furthermore, Stoddard et al.
+[-@stoddard2016continental] see an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.
+A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al.
+[-@stoddard2016continental], that might be occuring only on oligotrophic lakes.
 A full analysis and interpretation of trends by trophic state is beyond the scope of this study.
@@
 This suggests that a full analysis of how lakes in different trophic states are changing over time is warranted.
-Lastly, there are other drivers of these trends that could be analyzed including, changes in rainfall, extreme weather events, other climate mediated factors, or analyses at different spatial scales.
-The analysis we report in this paper focuses on long term trends at broader scales.
-It would be possible to use a similar approach to look at trends of individual lakes which could be a first step in trying to identify drivers of increasing chlorophyll.
-These questions are beyond the scope of this study, but all warrant further investigation.
 
@@
 
-Broad scale warming is difficult to manage on a sub-regional scale thus other mechanisms and approaches, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.
-Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
-There are of course many other hypotheses that we could form through information obtained by long-term trends such as those reported here.
-Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs and turn that information into management action.
+Our analysis has shown warming lakes in Rhode Island, but broad scale warming of lakes has also been documented and shown to be a result of both climatic and local drivers and can vary greatly within regions [e.g., @oreilly2015warming].
+Given that the drivers of warming are at both broad and local scales, suggests that managing warming lake temperatures will be a difficult task.
+To counteract the impacts of continued warming on increased chlorophyll in lakes will, at a minimum, require additional interventions.
+Reducing nutrient loads below current levels is one such intervention and could be acheieved via source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
+In any event, controlling future eutrophication and protecting against harmful aglal blooms will continue to be a challenge against the backdrop of warming lakes.
 

### 2020-12-02 R/preliminary_analyses.R

@@ -1,0 +1,23 @@
+source(here::here("R/functions.R"))
+library(magrittr)
+ww_all<-read.csv(here("data/ww_all_mostly_cleaned.csv"), stringsAsFactors = FALSE)
+
+ww_all_temp_gte25 <- ww_all %>%
+  filter(param == "temp") %>%
+  filter(mn_measurement > 25) %>%
+  mutate(day_of_year = yday(ymd(paste(year, month, day, "-")))) %>%
+  group_by(year) %>%
+  summarize(min_first_day = min(day_of_year))
+
+ww_all_temp_gte25 %>%
+  filter(year <= 1995) %>%
+  pull(min_first_day) %>%
+  mean() -> early
+as.Date("1995-01-01") -1 + early
+
+
+ww_all_temp_gte25 %>%
+  filter(year >= 2014) %>%
+  pull(min_first_day) %>%
+  mean() -> late
+as.Date("2014-01-01") -1 + late

### 2020-12-02 manuscript/ecosphere_resubmit_2/manuscript.Rmd
text revisions complete.  Now on to figure/tables.

@@
 source(here::here("R/figures_anomaly_only.R"), encoding = "UTF-8")
+source(here::here("R/ri_stats.R"), encoding = "UTF-8")
 ```
@@
 
-<!-- Jeff start here for change to comment #4 -->
-Add Paragraph about RI lakes: Context about human pressures on RI lakes + one sentence about adequate data being available
+In this study, we focus on the state of Rhode Island for several reasons.
+Rhode Island has the second highest population density in the country, yet still has a mix of land use/land cover with developed areas making up ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Development"]`% of the state, forested areas ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Forest"]`%, wetlands ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Wetlands"]`%, and agriculture ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Planted/Cultivated"]`%.
+Given these facts, Rhode Island lakes occur across a gradient of land use/land cover and face a suite of human pressures that most lakes in the nation are also facing with nutrients and pathogens of particular concern in the state [@ri_2016_waterquality].
+Lastly, Rhode Island provides an ideal study area as significant, long-term datasets are available for lakes making it possible to examine water quality trends.
 

### 2020-12-02 R/ri_stats.R

@@ -1,0 +1,25 @@
+library(dplyr)
+library(sf)
+library(tidycensus)
+library(FedData)
+library(nsink)
+
+us_pop <- state_laea %>%
+  left_join(get_acs("state", "B01003_001")) %>%
+  mutate(area = set_units(st_area(.), "mi^2"),
+         pop_density = estimate/area) %>%
+  arrange(desc(pop_density))
+us_pop
+
+ri_nlcd <- us_pop %>%
+  filter(NAME == "Rhode Island") %>%
+  get_nlcd(label = "ri", dataset = "Land_Cover")
+
+ri_nlcd_data <- data.frame(table(raster::values(ri_nlcd))) %>%
+  mutate(codes = as.numeric(as.character(Var1))) %>%
+  left_join(nsink:::n_load_idx_lookup) %>%
+  mutate(total_cells = sum(Freq)) %>%
+  group_by(nlcd_class_1) %>%
+  summarize(lulc_perc = round(sum(Freq)/max(total_cells)*100),0) %>%
+  ungroup()
+ri_nlcd_data

### 2020-12-02 manuscript/ecosphere_resubmit_2/manuscript.Rmd
saved by soft reset!

@@
 Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
+All analyses were conducted with R version 4.0.3 and details on R packages and operating system used for this analysis are included in a file, `sessioninfo.txt` at <>[].
 Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.
@@
 #Send session info to a file: 
-capture.output(sessionInfo(),file=here::here("manuscript/ecosphere_resubmit/sessioninfo.txt"))
+capture.output(sessionInfo(),file=here::here("sessioninfo.txt"))
 ```

### 2020-12-02 R/functions.R

@@ -1,17 +1,22 @@
 # Installations
 pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
           "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
-          "USAboundariesData","knitr","captioner", "pander", "cowplot")
+          "USAboundariesData","knitr","captioner", "pander", "cowplot", 
+          "tidycensus", "FedData", "nsink", "units")
+
+if(!requireNamespace("LAGOSNE")){
+  devtools::install_github("cont-limno/LAGOSNE")
+} else if(!requireNamespace("FedData")){
+  devtools::install_github("ropensci/FedData")
+} else if(!requireNamespace("nsink")){
+  devtools::install_github("jhollist/nsink")
+}
 
 for(i in pkgs){
   if(!i %in% installed.packages()){
     install.packages(i, repos = c("http://packages.ropensci.org", 
                                   "https://cran.rstudio.com"))
   }
-}
-
-if(!requireNamespace("LAGOSNE")){
-  devtools::install_github("cont-limno/LAGOSNE")
 }
 
 # Library
@@ -33,6 +38,10 @@ library(USAboundariesData)
 library(knitr)
 library(captioner)
 library(cowplot)
+library(tidycensus)
+library(FedData)
+library(nsink)
+library(units)
 
 #' Function to filter on number of years sampled per site
 #' @param df data frame to filter

### 2020-12-02 R/ri_stats.R

@@ -1,9 +1,3 @@
-library(dplyr)
-library(sf)
-library(tidycensus)
-library(FedData)
-library(nsink)
-
 us_pop <- state_laea %>%
   left_join(get_acs("state", "B01003_001")) %>%
   mutate(area = set_units(st_area(.), "mi^2"),

### 2020-12-04 manuscript/ecosphere_resubmit_2/ecosphere_response_2.Rmd
text revisions complete.  Now on to figure/tables.
draft of all SME revisions completed.

@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Good point.  We added an additional paragraph to the Introduction addressing RI LULC and pressures.
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Good suggestion.
+I never liked these tables, but figured readers would want details on variation (or lack thereof) in sampling effort.
+Created a faceted plot for each lagosne and uriww to replace the tables.
 
@@
 
-*Response:* Thanks!  After about 100 failed attempts at trying to describe this approach, we finally decided to put this together.  Actually caught some initial errors in code with this!
+*Response:* Thanks!  After about 100 failed attempts at trying to describe this approach to colleagues, we finally decided to put this together.  Actually caught some initial errors in code with this!
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Added description of regression line to caoptions and increased title text size.
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Have moved plots closer together
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Done.
 
@@
 
-*Response:* NEEDS RESPONSE
+*Response:* Good catch.
+The original language was confusing.
+Caption now reads: "Panel A.
+Simulated, random data showing example of three hypothetical lakes; one lake was monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.
+Panel B.
+Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.
+Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue."  Plus, have indicated in the methods (third paragaraph of Study Area and Data) that "lakes and reservoirs" are referred to as "sites" throughout the paper.
 

### 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
draft of all SME revisions completed.

@@
 Rhode Island has the second highest population density in the country, yet still has a mix of land use/land cover with developed areas making up ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Development"]`% of the state, forested areas ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Forest"]`%, wetlands ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Wetlands"]`%, and agriculture ~`r ri_nlcd_data$lulc_perc[ri_nlcd_data$nlcd_class_1 == "Planted/Cultivated"]`%.
-Given these facts, Rhode Island lakes occur across a gradient of land use/land cover and face a suite of human pressures that most lakes in the nation are also facing with nutrients and pathogens of particular concern in the state [@ri_2016_waterquality].
+The remainder of the land area is made up other land use/land cover classes.
+Given these facts, Rhode Island lakes occur across a gradient of land use/land cover and face a suite of human pressures that most lakes in the nation are also facing with nutrients and pathogens being of particular concern in the state [@ri_2016_waterquality].
 Lastly, Rhode Island provides an ideal study area as significant, long-term datasets are available for lakes making it possible to examine water quality trends.
@@
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
 
@@
 [-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.
-The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring sites on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
+The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
 URIWW now provides more than 90% of Rhode Island's lake baseline data and is an integral part of the state's environmental data collection strategy.
@@
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-In particular, we selected URIWW sites that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.
+In particular, we selected URIWW lakes and reservoirs (hereafter, referred to as sites) that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.
 As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
@@
 
-Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
-All analyses were conducted with R version 4.0.3 and details on R packages and operating system used for this analysis are included in a file, `sessioninfo.txt` at <>[].
+Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].
+All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].
 Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.
@@
 There are many different methods for analyzing time series data for trends.
-Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites.
+Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.
 For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.
@@
 *Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.
-Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275 > [@zenodo_citation].
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].
 
@@
 The per site medians across years are assumed to be independent.
-2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
+2. Calculate the long-term reference median for each site. This results in a single site-specific long-term median.
 3. Calculate the anomaly for each annual median at each site by subtracting the sites reference long-term medians from the sites yearly median.
@@
 After filtering and summarizing the data some years may not have sufficient number of sites to be included.
-We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data (`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).
+We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data and late in the time frame for the LAGOSNE data (`r fig_num("year_n", display = "cite", caption = "Number of sites available for trend analysis by parameter and year for both URIWW and LAGOSNE.")`).
 
@@
 
-For changes in chlorophyll, total phosphorus, and total nitrogen we define an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century.
+For changes in chlorophyll, total phosphorus, and total nitrogen, we define an ecologically meaningful trend as one that would result in any oligotrophic lake changing to a mesotrophic state over the course of a century.
 We use the values defined by the trophic state limits in Nürnberg [-@nurnberg1996trophic].
@@
 Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data.
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.
+Black line is the linear fit for year and median anomaly.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data.
 The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).
 Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).
-The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorphyll anomaly.
+The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Twenty year trend for median chlorophyll anomaly.
 Panel A.
@@
 Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.
+Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 
@@
 Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.
-Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A.).
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.
+Grey indicates yearly site-specific anomalies that were equal to the long-term medians.
+Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.
+Black line is the linear fit for year and median anomaly.", display = "cite")`A.).
 Also, 1998 only had three sites with available total nitrogen data and may not be representative.
@@
 Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.
-Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.
+Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.
+Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.
@@
 Blue indicates yearly site-specific anomalies that were below the site-specific long-term medians.
-Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site specific anomalies that were equal to the long-term medians.
-Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
+Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.Grey indicates yearly site-specific anomalies that were equal to the long-term medians.
+Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.
+Black line is the linear fit for year and median anomaly.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).
 
@@
 In addition to the sub-regional annualized trends of the five variables we address with this study, other trends may also be of interest.
-For example, a lengthening of the growing season could increase the number of days with conditions that favor the growth of harmful algal blooms.
-Many national and regional studies have documented longer growing seasons [e.g., @vega2019changes; @cooter1995recent; @kunkel2004temporal].
+For example, a lengthening of the growing season could increase the number of days with conditions that favor the growth of harmful algal blooms as cyanobacteria grow better at temperatures greater than than 25°C [@paerl2008blooms; @johnk2008summer; @reynolds2006ecology].
+Many national and regional studies have also documented longer growing seasons [e.g., @vega2019changes; @cooter1995recent; @kunkel2004temporal].
 Evidence in Rhode Island also points to possible lengthening of the growing season as both May and October, the beginning and end of the sampling dates in our dataset, show increasing temperature trends over time [@noaatrends].
 Furthermore, preliminary analysis of the URIWW data supports the idea that growing seasons may be getting longer in Rhode Island.
-Cyanobacteria grow better at temperatures greater than than 25°C [@paerl2008blooms; @johnk2008summer; @reynolds2006ecology].
 Early in our study period (1993-1995), the average first day of lake temperatures exceeding 25°C was June 11th, at the end of our study period (2014-2016) these warmer temperature were seen, on average, on June 1st.
@@
 This magnitude of change would result in a shift for any oligotrophic lake to a mesotrophic state; whereas, for any mesotrophic lake to become eutrophic over a century would require a change of 0.055 µg/l per year.
-Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only a single rate for all lakes.
+Examining these trends by trophic state provides the ability to look for ecologically important changes across the range of trophic state and not rely only on a single rate for all lakes.
 Furthermore, Stoddard et al.
-[-@stoddard2016continental] see an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.
+[-@stoddard2016continental] report an 18.2% reduction in the number of oligotrophic lakes in the United States from 2007 to 2012.
 A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al.
@@
 Our analysis has shown warming lakes in Rhode Island, but broad scale warming of lakes has also been documented and shown to be a result of both climatic and local drivers and can vary greatly within regions [e.g., @oreilly2015warming].
-Given that the drivers of warming are at both broad and local scales, suggests that managing warming lake temperatures will be a difficult task.
-To counteract the impacts of continued warming on increased chlorophyll in lakes will, at a minimum, require additional interventions.
-Reducing nutrient loads below current levels is one such intervention and could be acheieved via source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
+Given that the drivers of warming are at both broad and local scales, managing warming lake temperatures will be a difficult task.
+Counteracting the impacts of continued warming on increased chlorophyll in lakes will, at a minimum, require additional interventions.
+Reducing nutrient loads below current levels is one such intervention and could be achieved via source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].
 In any event, controlling future eutrophication and protecting against harmful aglal blooms will continue to be a challenge against the backdrop of warming lakes.
@@
 The site-specific anomalies are also robust to variations in sampling effort and in the timing of inclusion of given sampling locations.
-For instance, if a site included early in a time period had low values and a site included late in the time period had high values, then analyzing the measured values over this time period would show an increase whether or not one existed.
+For instance, if a site included only early in a time period had low values and a site included late in the time period had high values, then analyzing the measured values over this time period would show an increase whether or not one existed.
 Using the site-specific anomalies rescales the values and allows the real trend to be seen.
 We illustrate this with simulated, random data (e.g.
-no site-specific trend) (`r fig_num(name = "simulated_data", caption = "Simulated, random data showing example of sites with a low site early in a time period, a high site late in a time period, and one site measured across the time period.", display = "cite")`), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated_analysis", caption = "Analysis of simulated, random data with a low site early in a time period, a high site late in a time period, and one site measured across the time period.
-Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).
-The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated_analysis", caption = "Analysis of simulated, random data (prior figure) with a low site early in a time period, a high site late in a time period, and one site measured across the time period.
-Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`).
+no site-specific trend) (`r fig_num(name = "simulated", caption = "Panel A.
+Simulated, random data showing example of three hypothetical lakes; one lake was monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.
+Panel B.
+Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.
+Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`A.), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated", display = "cite")`B.).
+The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated", display = "cite")`) as each site did not change over time.
 
@@
 # Tables
-
-```{r year_num_table_ww, echo=FALSE, results="asis"}
-#pander::pander(year_num_ww, format = "pipe")
-knitr::kable(year_num_ww, "pipe")
-```
-`r tab_num("year_n_ww")`
-
-######
-
-```{r year_num_table_lagos, echo=FALSE, results="asis"}
-knitr::kable(year_num_lagosne, format = "pipe")
-```
-`r tab_num("year_n_lagosne")`
-
-######
 
@@
 
+![](../../figures/sample_sizes.jpg)
+
+`r fig_num("year_n")`
+
+######
+
 ![](../../figures/ww_temp_trends_anomaly.jpg)
@@
 
-![](../../figures/simulated_data.jpg)
-`r fig_num("simulated_data")`
-
-######
-
-![](../../figures/simulated_analysis.jpg)
-`r fig_num("simulated_analysis")`
+![](../../figures/simulated_data_analysis.jpg)
+`r fig_num("simulated")`
 

### 2020-12-04 R/figures.R

@@ -124,7 +124,7 @@ lagos_np_rl <- lagos_np_gg[[4]]
 chla_plot <- cowplot::plot_grid(chla_gg[[1]], lagos_chla_gg[[1]], 
                                 align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_chla_trends.jpg"), chla_plot, width = 7.5, 
+ggsave(here("figures/paired_chla_trendsx.jpg"), chla_plot, width = 7.5, 
        height = 9, 
        units = "in", dpi = 600)
 

### 2020-12-04 R/figures_anomaly_only.R

@@ -58,7 +58,7 @@ temp_gg <- ww_data %>%
               y = "Median Yearly Temperature Anomaly (°C)", x = "Year",
               write = here("data/temp_data_anomaly.csv"), error_bar = "iqr",
               title = "URI Watershed Watch Temperature")
-ggsave(here("figures/ww_temp_trends_anomaly.jpg"), temp_gg[[1]], width = 7.5, height = 5.625, 
+ggsave(here("figures/ww_temp_trends_anomaly.jpg"), temp_gg[[1]], width = 8.5, height = 5.5, 
        units = "in", dpi = 600)
 
 ww_temp_kt <- temp_gg[[2]]
@@ -140,25 +140,25 @@ lagos_np_rl <- lagos_np_gg[[4]]
 chla_plot <- cowplot::plot_grid(chla_gg[[1]], lagos_chla_gg[[1]], 
                                 align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_chla_trends_anomaly.jpg"), chla_plot, width = 5.7, height = 7.45, 
+ggsave(here("figures/paired_chla_trends_anomaly.jpg"), chla_plot, width = 8.5, height = 11, 
        units = "in", dpi = 600)
 
 tn_plot <- cowplot::plot_grid(tn_gg[[1]], lagos_tn_gg[[1]], 
                               align = "h", ncol = 1)
 
-ggsave(here("figures/paired_tn_trends_anomaly.jpg"), tn_plot, width = 5.7, height = 7.45, 
+ggsave(here("figures/paired_tn_trends_anomaly.jpg"), tn_plot, width = 8.5, height = 11, 
        units = "in", dpi = 600)
 
 tp_plot <- cowplot::plot_grid(tp_gg[[1]], lagos_tp_gg[[1]], 
                               align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_tp_trends_anomaly.jpg"), tp_plot, width = 5.7, height = 7.45, 
+ggsave(here("figures/paired_tp_trends_anomaly.jpg"), tp_plot, width = 8.5, height = 11, 
        units = "in", dpi = 600)
 
 np_plot <- cowplot::plot_grid(np_gg[[1]], lagos_np_gg[[1]], 
                               align = "h", ncol = 1) 
 
-ggsave(here("figures/paired_np_trends_anomaly.jpg"), np_plot, width = 5.7, height = 7.45, 
[…]

### 2020-12-04 R/functions.R

@@ -163,12 +163,12 @@ wq_trend_gg <- function(df, wqparam,
   } else if(yvar == "measurement_anmly"){
     limit<-max(abs(c(floor(1.1*min(df2$mdn_value - df2[[error_bar]])), 
                      ceiling(1.1*max(df2$mdn_value + df2[[error_bar]])))))
-    label_break <- c(-limit, floor(-limit/2), 0, ceiling(limit/2), limit)
+    label_break <- c(-limit, round((-limit/2)/5,1)*5, 0, round((limit/2)/5,1)*5, limit)
     limits <- c(-limit,limit)
   } else if(yvar == "measurement_anmly_mdn"){
     limit<-max(abs(c(floor(1.1*min(df2$iqr_low)), 
                      ceiling(1.1*max(df2$iqr_high)))))
-    label_break <- c(-limit, floor(-limit/2), 0, ceiling(limit/2), limit)
+    label_break <- c(-limit, round((-limit/2)/5,1)*5, 0, round((limit/2)/5,1)*5, limit)
     limits <- c(-limit,limit)
   }
   
@@ -204,6 +204,7 @@ wq_trend_gg <- function(df, wqparam,
                                              label = "*"), color = "darkblue")
     }
     #geom_point(aes(color = col_group), size=3.5) +
+  
   gg <- gg + 
     geom_smooth(method = "lm", se=FALSE, color = "black") +
     theme_ipsum_rc() +
@@ -211,8 +212,15 @@ wq_trend_gg <- function(df, wqparam,
                              " p = ", signif(regress$p.value, 2))) +
     scale_color_manual(values = my_colors) + 
     theme(legend.position="none", 
-          plot.title = element_text(size = 12, face = "bold"),
-          plot.subtitle = element_text(size=10, face="plain")) + 
+          plot.title = element_text(size = 14, face = "bold"),
+          plot.subtitle = element_text(size=14, face="plain"),
+          axis.title.x = element_text(size=14, face="plain",
+                                      margin = margin(t = 10, r = 0, b = 0, l = 0)),
+          axis.title.y = element_text(size=14, face="plain",
+                                      margin = margin(t = 0, r = 10, b = 0, l = 0)),
+          axis.text.x = element_text(size=14, face="plain"),
+          axis.text.y = element_text(size=14, face="plain"),
+          plot.margin = margin(t = 10, r = 0, b = 10, l = 0)) + 
[…]

### 2020-12-04 manuscript/ecosphere_resubmit_2/ecosphere_response_2.Rmd
edits on resubmit finished

@@
 
-General comments and thanks
+Thank you for the review comments.
+The end product is significantly better.
+We have itemized and responded to each of the comments below.
+If there are any questions about what we have done, please do not hesitate to contact us.
 
@@
 
-*Response:* Added description of regression line to caoptions and increased title text size.
+*Response:* Added description of regression line to captions and increased title text size.
 
@@
 Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.
-Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue."  Plus, have indicated in the methods (third paragaraph of Study Area and Data) that "lakes and reservoirs" are referred to as "sites" throughout the paper.
+Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue."  Plus, we have indicated in the methods (third paragaraph of Study Area and Data) that "lakes and reservoirs" are referred to as "sites" throughout the paper.
 

### 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
edits on resubmit finished

@@
 The University of Rhode Island's Watershed Watch (URIWW) is a scientist-led community science program founded in the late 1980s that has built a robust collaboration between URI scientists and a vast network of volunteer monitors.
-Volunteer monitors are trained and then collect field data (e.g., sonde measurements or secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab.
+Volunteer monitors are trained and then collect field data (e.g., sonde measurements or Secchi depth) as well as whole water samples during the growing season (e.g., May through October) for later analysis in the lab.
 The entire effort follows rigorous quality control/quality assurance protocols.
@@
 
-URIWW data contributed to the larger regional study by Oliver et al.
-[-@Oliver_2017], and, also allowed us to examine the long-term trends specifically in Rhode Island.
+URIWW data are included in the LAGOSNE and contributed to the larger regional study by Oliver et al.
+[-@Oliver_2017], which also allowed us to examine the long-term trends specifically in Rhode Island.
 The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
@@
 For this analysis, we were interested in trends in lake temperature, TN, TP, TN:TP, and chlorophyll *a*.
-In particular, we selected URIWW lakes and reservoirs (hereafter, referred to as sites) that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, 5) and were sampled at a depth of 2 meters or less.
+In particular, we selected URIWW lakes and reservoirs (hereafter, referred to as sites) that matched the following criteria: 1) were sampled between 1993 and 2016, 2) were sampled in May to October, 3) had at least one sampling event per year in May through June, at least one sampling event per year in July through August, and at least one sampling event per year in September through October, 4) had at least one sampling event from 1993-2004 and 2005-2016, and 5) were sampled at a depth of 2 meters or less.
 As not all sites have data for all selected years, we further filtered the data to select sites that had at least 10 years of data for a given parameter within the 1993 to 2016 time frame.
@@
 Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017].
-At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.
+At the time of our analysis, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013 to be included.
 
@@
 We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculture, Forest, and Developed.
-The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.
+The maximum lake depth is from various sources and lake area is calculated from the National Hydrography Dataset waterbody polygons.
 Additional details are available in Soranno et al.
@@
 All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].
-Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/raw/master/data/yearly_average_anomaly.csv>.
+Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.
 
@@
 Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling sites is how to identify a trend while there is variation within a sampling location as well as variation introduced by differing start years for sampling among the many sites.
-For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.
+For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degraded water bodies.
 Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
@@
 Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).
-Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic].
+Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region (`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; [@nurnberg1996trophic]).
 These results follow what would be expected given that, on average, LAGOSNE lakes are more dominated by agricultural lands than are Rhode Island lakes (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).
@@
 Missing years had insufficient data and an asterisk indicates years with only 3 sites and error bars are the range of the data.
-Black line is the linear fit for year and median anomaly.", display = "cite")`A.).
+Black line is the linear fit for year and median anomaly.", display = "cite")`A).
 Also, 1998 only had three sites with available total nitrogen data and may not be representative.
@@
 Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
-Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.
+Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively smaller number of sites (`r fig_num("year_n", display = "cite")`) and those values may not be representative.
 Total phosphorus showed a very small increasing trend and there is statistical support for that trend.
@@
 Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.
-Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.
+Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r fig_num("year_n", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.
 This low median anomaly should be interpreted with this caveat in mind.
@@
 For instance, nutrient management is largely a local to watershed scale effort, but may also be regional as atmospheric nitrogen deposition can be a significant source of nitrogen [@boyer2002anthropogenic].
-Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of a catchment that is impervious surface and lake morphology can also impact temperature [@nelson2007stream].
+Similarly, warming lakes are driven by broader climate patterns, yet waterbody-specific factors such as the percent of impervious surface in a catchment and lake morphology can also impact temperature [@nelson2007stream].
 In short, differences in regional and state level trends are driven by complex and multi-scale processes.
@@
 A site-specific anomaly analysis by trophic state would identify trends, such as those seen in Stoddard et al.
-[-@stoddard2016continental], that might be occuring only on oligotrophic lakes.
+[-@stoddard2016continental], that might be occurring only on oligotrophic lakes.
 A full analysis and interpretation of trends by trophic state is beyond the scope of this study.
@@
 Analysis of simulated, random data with one lake monitored throughout the record, one lake with low reported values monitored early in the record, and another lake with high reported values monitored late in the record.
-Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`A.), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated", display = "cite")`B.).
-The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly showed no trend (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated", display = "cite")`) as each site did not change over time.
+Yearly average of the actual values is shown in red and yearly average of the site-specific anomalies is shown in blue.", display = "cite")`A), and apply the site-specific analysis to those simulated data (`r fig_num(name = "simulated", display = "cite")`B).
+The average of the yearly measured values shows a trend (slope = `r signif(meas_lm$slope, 2)` and p-value = `r signif(meas_lm$p.value, 2)`), whereas the site-specific anomalies correctly shows no trend as each site did not change over time (slope = `r signif(anom_lm$slope, 2)` and p-value = `r signif(anom_lm$p.value, 2)`, `r fig_num(name = "simulated", display = "cite")`).
 

### 2020-12-04 R/figures_anomaly_only.R

@@ -521,7 +521,7 @@ ww_sample_gg <- year_num_ww %>%
         axis.text.y = element_text(size=14, face="plain"),
         plot.margin = margin(t = 10, r = 0, b = 10, l = 0),
         strip.text = element_text(size = 11)) +
-  labs(title = "A. Number of sampled lakes with measurments for each variable 
+  labs(title = "A. Number of sampled lakes for each variable 
        from URI Watershed Watch program", y = "Number of sampled lakes")
   
   lagos_sample_gg <- year_num_lagosne %>%
@@ -546,7 +546,7 @@ ww_sample_gg <- year_num_ww %>%
         axis.text.y = element_text(size=14, face="plain"),
         plot.margin = margin(t = 10, r = 0, b = 10, l = 0),
         strip.text = element_text(size = 11)) +
-    labs(title = "B. Number of sampled lakes with measurments for each variable 
+    labs(title = "B. Number of sampled lakes for each variable 
        from LAGOSNE", y = "Number of sampled lakes")
 
 sample_size_plot <- cowplot::plot_grid(ww_sample_gg, lagos_sample_gg, 

### 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
revisions submitted

@@
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
 
@@
 
-Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].
+Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].
 All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].
@@
 *Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.
-Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4050275> [@zenodo_citation].
+Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].
 

### 2021-03-19 manuscript/ecosphere_production/manuscript.Rmd
figures and tables and citations, oh my!

--- 2020-12-04 manuscript/ecosphere_resubmit_2/manuscript.Rmd
@@
 author:
+- '**List of Authors:** Hollister. J. W. ^†^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*1*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'  
 - ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
@@
   RI 02881
-- ^*3*^ ORISE, Narragansett, RI 02882
-- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. D. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
-- '**Corresponding author: hollister.jeff@epa.gov**'
+- '**†: hollister.jeff@epa.gov**'
+- "**Track:** Freshwater Ecology"
 bibliography: manuscript.bib
@@
 Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured.
-We use the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there are sub-regional (i.e.
+We used the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset to determine if there were sub-regional (i.e.
 3000km^2^) water quality trends.
 URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years.
-The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively.
-We assess regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis.
+The LAGOSNE and URIWW datasets allowed for comparison of water quality trends at regional and sub-regional scales, respectively.
+We assessed regional (LAGOSNE) and sub-regional (URIWW) trends with yearly median anomalies calculated on a per-station basis.
 Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016.
 Total nitrogen, total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable.
-At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.
-Total phosphorus does show a very slight increase.
-In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
-Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.
-  | **Keywords:** Volunteer Monitoring, community Science, Nutrients, Lakes, Trend Analysis, Long Term
+At the regional scale, the LAGOSNE dataset showed similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, and N:P all stable over time.
+Total phosphorus did show a very slight increase.
+In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs increased, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio.
+Additionally, we demonstrated both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends.
+  | **Keywords:** Anomaly; Chlorophyll; Community Science; Lakes; Lake Temperature; Long Term; Nutrients; Trend Analysis; Volunteer Monitoring.  
 ---
@@
 Another objective of this paper was to apply existing methods for examining long-term climate records [e.g., @jones1996calculating] to water quality data in order to examine long-term trends.
-We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@sorannolagosne2017; @stacheleklagosne2017; @zenodo_citation; @sorannolimno2019].
+We conducted this analysis using open data from the URI Watershed Watch program and the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) project and the analysis in its entirety is available for independent reproduction at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@sorannolagosne2017; @stacheleklagosne; @zenodo_citation; @sorannolimno2019].
 
@@
 
-## Study Area and Data
+## Study area and data
 
@@
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
-[-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].
+[-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne; @sorannolagosne2017; @sorannolimno2019].
 Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.
@@
 Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].
-All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2020; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc].
+All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2021; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc; @magick; @flextable].
 Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.
 
-## Water Quality Trend Analysis
+## Water quality trend analysis
 
@@
 Red indicates yearly site-specific anomalies that were above the site-specific long-term medians.
-Black line is the linear fit for year and median anomaly.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data.
+Black line is the linear fit for year and median anomaly.", display = "cite")`, `r tab_num(name = "trend_summ", caption = 'Summary of long term water quality trends in the URIWW and LAGOSNE data for termperature (Temp.) chlorophyll (Chl.), total nitrogen (TN), total phosphorus (TP), and the molar nitrogen to phosphorus ratio (N:P).
 The "++" indicate a positive trend, the "+" indicates a weak positive trend, and the "0" indicates no trend.', display = "cite")`).
@@
 Total phosphorus showed a very small increasing trend and there is statistical support for that trend.
-The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
-Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.
+The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "", display = "cite")`).
+Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "
+", display = "cite")`) as none of our criteria for a trend were met.
 Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r fig_num("year_n", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.
@@
 #knitr::kable(context_summ, format = "pipe")
-knitr::kable(context_summ)
+flextable(context_summ) %>%
+  width(width = 1) %>%
+  fontsize(size = 11, part = "all") %>%
+  bold(part = "header") %>%
+  font(fontname = "Times New Roman", part = "all") %>%
+  autofit()
 ```
@@
 ```{r ww_params, echo=FALSE, results="asis"}
-knitr::kable(ww_param_ranges, format = "pipe")
+#knitr::kable(ww_param_ranges, format = "pipe")
+flextable(ww_param_ranges) %>%
+  width(width = 1) %>%
+  fontsize(size = 11, part = "all") %>%
+  bold(part = "header") %>%
+  font(fontname = "Times New Roman", part = "all") %>%
+  autofit()
 ```
@@
 ```{r lagos_params, echo=FALSE, results="asis"}
-knitr::kable(lagos_param_ranges, format = "pipe")
+#knitr::kable(lagos_param_ranges, format = "pipe")
+flextable(lagos_param_ranges) %>%
+  width(width = 0.5) %>%
+  fontsize(size = 11, part = "all") %>%
+  bold(part = "header") %>%
+  font(fontname = "Times New Roman", part = "all") %>%
+  autofit()
 ```
@@
 ```{r ts_table, echo=FALSE, results="asis"}
-knitr::kable(ts_table, format = "pipe")
+#knitr::kable(ts_table, format = "pipe")
+flextable(ts_table) %>%
+  width(width = 1.2) %>%
+  fontsize(size = 11, part = "all") %>%
+  bold(part = "header") %>%
+  font(fontname = "Times New Roman", part = "all") %>%
+  autofit()
 ```
@@
 trend_summ_df <- tibble(Source = c("URIWW", "LAGOSNE"), 
-                        Temperautre = c("++", ""), Chlorophyll= c("++", "0"), 
-                        `Total Nitrogen` = c("0", "0"), 
-                        `Total Phosphorus`= c("0", "+"), 
-                        `Nitrogen:Phosphorus` = c("0", "0"))
+                        Temp. = c("++", ""), Chl. = c("++", "0"), 
+                        `TN` = c("0", "0"), 
+                        `TP`= c("0", "+"), 
+                        `N:P` = c("0", "0"))
 
-knitr::kable(trend_summ_df, format = "pipe")
+#knitr::kable(trend_summ_df, format = "pipe")
+flextable(trend_summ_df) %>%
+  width(width = 1) %>%
+  fontsize(size = 11, part = "all") %>%
+  bold(part = "header") %>%
+  font(fontname = "Times New Roman", part = "all") %>%
+  autofit()
 ```

### 2021-03-19 R/figures.R

@@ -11,8 +11,8 @@ chla_gg <- ww_data %>%
                                     italic(" a"))), x = "Year",
                write = here("data/chla_data.csv"), error_bar = "sd",
                title = "A. URI Watershed Watch Chlorophyll")
-ggsave(here("figures/ww_chla_trends.jpg"), chla_gg[[1]], width = 7.5, height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here("figures/ww_chla_trends.jpg"), chla_gg[[1]], width = 15.6, 
+       height = 11.7, units = "cm", dpi = 600)
 
 ww_chla_kt <- chla_gg[[2]]
 ww_chla_df <- chla_gg[[3]]
@@ -24,8 +24,8 @@ tn_gg <-  ww_data %>%
               y = "Average Yearly Scaled Total Nitrogen", x = "Year",
               write = here("data/total_n_data.csv"), error_bar = "sd",
               title = "A. URI Watershed Watch Total Nitrogen")
-ggsave(here("figures/ww_tn_trends.jpg"), tn_gg[[1]], width = 7.5, height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here("figures/ww_tn_trends.jpg"), tn_gg[[1]], width = 15.6, 
+       height = 11.7, units = "cm", dpi = 600)
 
 ww_tn_kt <- tn_gg[[2]]
 ww_tn_df <- tn_gg[[3]]
@@ -37,8 +37,8 @@ tp_gg <-  ww_data %>%
               y = "Average Yearly Scaled Total Phosphorus", x = "Year",
               write = here("data/total_p_data.csv"), error_bar = "sd",
               title = "A. URI Watershed Watch Total Phosphorus")
-ggsave(here("figures/ww_tp_trends.jpg"), tp_gg[[1]], width = 7.5, height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here("figures/ww_tp_trends.jpg"), tp_gg[[1]], width = 15.6, 
+       height = 11.7, units = "cm", dpi = 600)
 
 ww_tp_kt <- tp_gg[[2]]
 ww_tp_df <- tp_gg[[3]]
@@ -50,8 +50,8 @@ temp_gg <- ww_data %>%
               y = "Average Yearly Scaled Temperature", x = "Year",
               write = here("data/temp_data.csv"), error_bar = "sd",
[…]

### 2021-03-19 R/figures_anomaly_only.R

@@ -16,9 +16,8 @@ chla_gg <- ww_data %>%
                x = "Year",
                write = here("data/chla_data_anomaly.csv"), error_bar = "iqr",
                title = "A. URI Watershed Watch Chlorophyll")
-ggsave(here("figures/ww_chla_trends_anomaly.jpg"), chla_gg[[1]], width = 7.5, 
-       height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here("manuscript/ecosphere_production/figures/ww_chla_trends_anomaly.tiff"), chla_gg[[1]], width = 15.6, 
+       height = 11.7, units = "cm", dpi = 300, compress = "lzw")
 
 ww_chla_kt <- chla_gg[[2]]
 ww_chla_df <- chla_gg[[3]]
@@ -31,8 +30,8 @@ tn_gg <-  ww_data %>%
               x = "Year",
               write = here("data/total_n_data_anomaly.csv"), error_bar = "iqr",
               title = "A. URI Watershed Watch Total Nitrogen")
-ggsave(here("figures/ww_tn_trends_anomaly.jpg"), tn_gg[[1]], width = 7.5, height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here("manuscript/ecosphere_production/figures/ww_tn_trends_anomaly.tiff"), tn_gg[[1]], width = 15.6, 
+       height = 11.7, units = "cm", dpi = 300, compress = "lzw")
 
 ww_tn_kt <- tn_gg[[2]]
 ww_tn_df <- tn_gg[[3]]
@@ -45,8 +44,8 @@ tp_gg <-  ww_data %>%
               x = "Year",
               write = here("data/total_p_data_anomaly.csv"), error_bar = "iqr",
               title = "A. URI Watershed Watch Total Phosphorus")
-ggsave(here("figures/ww_tp_trends_anomaly.jpg"), tp_gg[[1]], width = 7.5, height = 5.625, 
-       units = "in", dpi = 600)
+ggsave(here("manuscript/ecosphere_production/figures/ww_tp_trends_anomaly.tiff"), tp_gg[[1]], width = 15.6, 
+       height = 11.7, units = "cm", dpi = 300, compress = "lzw")
 
 ww_tp_kt <- tp_gg[[2]]
 ww_tp_df <- tp_gg[[3]]
@@ -58,8 +57,10 @@ temp_gg <- ww_data %>%
               y = "Median Yearly Temperature Anomaly (°C)", x = "Year",
[…]

### 2021-03-19 R/functions.R

@@ -2,7 +2,7 @@
 pkgs <- c("lubridate","broom", "stringr", "dplyr", "readr", "readxl","tidyr", 
           "here","ggplot2", "hrbrthemes", "Kendall","sf", "USAboundaries",
           "USAboundariesData","knitr","captioner", "pander", "cowplot", 
-          "tidycensus", "FedData", "nsink", "units")
+          "tidycensus", "FedData", "nsink", "units", "magick", "flextable")
 
 if(!requireNamespace("LAGOSNE")){
   devtools::install_github("cont-limno/LAGOSNE")
@@ -42,6 +42,8 @@ library(tidycensus)
 library(FedData)
 library(nsink)
 library(units)
+library(magick)
+library(flextable)
 
 #' Function to filter on number of years sampled per site
 #' @param df data frame to filter
@@ -213,13 +215,13 @@ wq_trend_gg <- function(df, wqparam,
     scale_color_manual(values = my_colors) + 
     theme(legend.position="none", 
           plot.title = element_text(size = 14, face = "bold"),
-          plot.subtitle = element_text(size=14, face="plain"),
-          axis.title.x = element_text(size=14, face="plain",
+          plot.subtitle = element_text(size = 14, face="plain"),
+          axis.title.x = element_text(size = 14, face="plain",
                                       margin = margin(t = 10, r = 0, b = 0, l = 0)),
-          axis.title.y = element_text(size=14, face="plain",
+          axis.title.y = element_text(size = 14, face="plain",
                                       margin = margin(t = 0, r = 10, b = 0, l = 0)),
-          axis.text.x = element_text(size=14, face="plain"),
-          axis.text.y = element_text(size=14, face="plain"),
+          axis.text.x = element_text(size = 14, face="plain"),
+          axis.text.y = element_text(size = 14, face="plain"),
           plot.margin = margin(t = 10, r = 0, b = 10, l = 0)) + 
     scale_x_continuous(labels = c(1990,1995,2000,2005,2010,2015),
                        breaks = c(1990,1995,2000,2005,2010,2015),
@@ -227,7 +229,8 @@ wq_trend_gg <- function(df, wqparam,
                        limits = c(1993,2016)) +
     scale_y_continuous(labels = label_break,
                        breaks = label_break,
-                       limits = limits)
+                       limits = limits) +
+    theme(plot.margin=grid::unit(c(0,0,0,0), "mm"))
   
   list(gg, kt, df2, regress)
 }
[…]

### 2021-03-19 R/summary_stats.R

@@ -204,13 +204,13 @@ all_context_summ <- all_lakes_context %>%
   mutate(param = case_when(param == "lake_area_ha" ~
                              "Lake Area (ha)", 
                            param == "maxdepth" ~
-                             "Maximum Depth (m)",
+                             "Max. Depth (m)",
                            param == "Agriculture" ~
-                             "Percent Agriculture",
+                             "Agriculture (%)",
                            param == "Forest" ~
-                             "Percent Forest",
+                             "Forest (%)",
                            param == "Developed" ~
-                             "Percent Developed",
+                             "Developed (%)",
                            TRUE ~ param),
          units = c("%", "%", "%", "ha", "m")) %>%
   mutate(Source = "LAGOSNE") %>%
@@ -232,13 +232,13 @@ ww_context_summ <- all_lakes_context %>%
   mutate(param = case_when(param == "lake_area_ha" ~
                              "Lake Area (ha)", 
                            param == "maxdepth" ~
-                             "Maximum Depth (m)",
+                             "Max. Depth (m)",
                            param == "Agriculture" ~
-                             "Percent Agriculture",
+                             "Agriculture (%)",
                            param == "Forest" ~
-                             "Percent Forest",
+                             "Forest (%)",
                            param == "Developed" ~
-                             "Percent Developed",
+                             "Developed (%)",
                            TRUE ~ param),
          units = c("%", "%", "%", "ha", "m")) %>%
   mutate(Source = "URIWW") %>%

### 2021-03-19 manuscript/ecosphere_production/manuscript.Rmd
finally final?!?

@@
 
-URIWW data are included in the LAGOSNE and contributed to the larger regional study by Oliver et al.
-[-@Oliver_2017], which also allowed us to examine the long-term trends specifically in Rhode Island.
 The URIWW program began in 1988, monitoring 14 lakes and has now grown to include over 250 monitoring locations on over 120 waterbodies, including rivers/streams, and estuaries, with more than 400 trained volunteers.
@@
 
-Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
+Prior studies have modeled water quality trends across a larger region of the northeastern US within 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island (which includes the URIWW data), Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017].
 We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al.
@@
 Lastly, data and code for the entire analysis is available from <https://github.com/USEPA/ri_wq_trends/> and is archived at <https://doi.org/10.5281/zenodo.4306765> [@zenodo_citation].
-All analyses were conducted with R version 4.0.3 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2021; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc; @magick; @flextable].
+All analyses were conducted with R version 4.0.4 and details on R package versions and operating system used for this analysis are included in a file, `sessioninfo.txt` at <https://github.com/USEPA/ri_wq_trends/blob/master/sessioninfo.txt> [@r2021; @lubridate; @broom; @stringr; @dplyr; @readr; @readxl; @tidyr; @here; @ggplot2; @hrbrthemes; @kendall; @sf; @usaboundaries; @usaboundariesdata; @knitr; @captioner; @pander; @cowplot; @tidycensus; @feddata; @nsink; @units; @xie2015dynamicdoc; @magick; @flextable].
 Values included in each of the figures have also been separately saved as a comma-separated values file, `yearly_average_anomaly.csv`, and may be retrieved via <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv>.
@@
 For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degraded water bodies.
-Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
+Thus, it is necessary to account for this type of within-site and among-site variation.
+To do so, we used methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].
 The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.
@@
 [-@Oliver_2017] our analysis shows little increasing trend in chlorophyll *a* at the regional scale (e.g., northeastern and mid-western United States).
-However, at the local scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.
+However, at the more localized scale of the state of Rhode Island, there is a clear increasing trend in chlorophyll *a*.
 
@@
 This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some community science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].
-If long-term involvement was not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.
+If long-term involvement was not common, then consistent long-term data may be difficult to obtain and understanding of long-term trends would be challenging.
 

### 2021-04-15 manuscript/ecosphere_production/manuscript.Rmd
Updated table references

@@
 Total phosphorus showed a very small increasing trend and there is statistical support for that trend.
-The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "", display = "cite")`).
-Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "
-", display = "cite")`) as none of our criteria for a trend were met.
+The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).
+Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.
 Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r fig_num("year_n", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.

### 2021-05-12 R/figures_anomaly_only.R
updates to proofs

@@ -91,7 +91,7 @@ lagos_chla_gg <- wq_trend_gg(lagos_data, "chla", yvar = "measurement_anmly_mdn",
                              x = "Year",
                              write = here("data/chla_data_lagos_anomaly.csv"), 
                              error_bar = "iqr",
-                             title = "B. LAGOSNE Chlorophyll")
+                             title = "B. LAGOS-NE Chlorophyll")
 ggsave(here("manuscript/ecosphere_production/figures/lagos_chla_trends_anomaly.tiff"), lagos_chla_gg[[1]], 
        width = 15.6, 
        height = 11.7, units = "cm", dpi = 300, compress = "lzw")
@@ -104,7 +104,7 @@ lagos_tn_gg <- wq_trend_gg(lagos_data, "total_n", yvar = "measurement_anmly_mdn"
                            x = "Year",
                            write = here("data/total_n_data_lagos_anomaly.csv"), 
                            error_bar = "iqr",
-                           title = "B. LAGOSNE Total Nitrogen")
+                           title = "B. LAGOS-NE Total Nitrogen")
 ggsave(here("manuscript/ecosphere_production/figures/lagos_tn_trends_anomaly.tiff"), lagos_tn_gg[[1]], width = 15.6, 
        height = 11.7, units = "cm", dpi = 300, compress = "lzw")
 lagos_tn_kt <- lagos_tn_gg[[2]]
@@ -117,7 +117,7 @@ lagos_tp_gg <- wq_trend_gg(lagos_data, "total_p", yvar = "measurement_anmly_mdn"
                            x = "Year",
                            write = here("data/total_p_data_lagos_anomaly.csv"), 
                            error_bar = "iqr",
-                           title = "B. LAGOSNE Total Phosphorus")
+                           title = "B. LAGOS-NE Total Phosphorus")
 ggsave(here("manuscript/ecosphere_production/figures/lagos_tp_trends_anomaly.tiff"), lagos_tp_gg[[1]], width = 15.6, 
        height = 11.7, units = "cm", dpi = 300, compress = "lzw")
 
@@ -128,7 +128,7 @@ lagos_tp_rl <- lagos_tp_gg[[4]]
 lagos_np_gg <- wq_trend_gg(lagos_data, "np_ratio", yvar = "measurement_anmly_mdn", 
               y = "Median Yearly N:P Anomaly", x = "Year",
[…]
